---
title: Performance measures
filters:
  - line-highlight
date: "2025-11-05T11:21:31+00:00"
---

```{python}
#| echo: false
# pylint: disable=no-member
```

:::: {.pale-blue}

**Learning objectives:**

* Learn **how to record various performance measures** within the simulation model.
* Understand how **time-weighted calculations** work.

**Relevant reproducibility guidelines:**

* STARS Reproducibility Recommendations (⭐): Include code to calculate all required model outputs
* STARS Reproducibility Recommendations: Ensure clarity and consistency in the model results tables.
* NHS Levels of RAP (🥈): Data is handled and output in a Tidy data format.

**Pre-reading:**

This page continues on from: [Initialisation bias](warmup.qmd).

> *Entity generation → Entity processing → Initialisation bias → Performance measures*

::::

:::: {.very-pale-blue}

**Required packages:**

These should be available from environment setup on the [Structuring as a package](/pages/guide/setup/package.qmd) page.

::: {.python-content}

```{python}
from IPython.display import HTML
from itables import to_html_datatable
import numpy as np
import pandas as pd
import simpy
from sim_tools.distributions import Exponential
```

:::

::: {.r-content}

```{r}
#| output: false
library(dplyr)
library(knitr)
library(simmer)
library(tidyr)
```

```{r}
#| echo: false
#| output: false
# For illustrative plot only, code hidden
library(ggplot2)
library(plotly)
```

:::

::::

<br>

This page shows **how to calculate common performance measures** from your discrete‑event simulation model, such as total arrivals, mean waiting time and resource utilisation. These measures help you understand how busy your system is, how long entities wait, and how effectively resources are used during the data collection period.

Where relevant, we have included corresponding terms from **queueing theory** for each measure, so that readers familiar with analytic queueing models can see how the simulation outputs line up with classic results.

::: {.callout-note title="What is queueing theory?" collapse="true"}

Queueing theory is a field of mathematics that analyses queues under strict assumptions, and is suitable for simple well-characterised systems. Discrete event simulation (DES) can be applied in similar ways (e.g., replicating classic queueing models like M/M/s), but more often extends beyond simple theory to handle a wider range of arrival and service patterns and more complex resource constraints.

For a more comprehensive explanation of queueing theory, check out this video from 
[Bikey Bonn Kleiford Seranilla](https://www.youtube.com/@2334463) on YouTube:

{{< video https://youtu.be/SqSUJ0UYWMQ?si=Lgqws6sAhGLmRIcY >}}

:::

On this page, we introduce each performance measure in turn and demonstrate how to adapt the code from the [initialisation bias](warmup.qmd) page to include the new measure.

If you'd like to see code that incorporates **all** performance measures at once (rather than adding them individually), visit the [replications](replications.qmd) page next. There, we work with a model that includes all metrics, with utilisation implemented using the time-weighted approach.

## Preparation for results collection

::: {.python-content}

Before we start calculating any simulation performance measures, some set-up is required. This includes initial changes to `Parameters` and `Model`, and the creation of a new class: `Runner`.

### Parameter class

Set `verbose` to `False` in the `Parameters` class. This lets the workflow focus on the calculated results rather than individual patient activities.

*Bold lines show where the code has changed since the previous version. The highlight may correspond to a change anywhere on that line, not necessarily to the entire line.*

```{python}
class Parameters:
    """
    Parameter class.

    Attributes
    ----------
    interarrival_time : float
        Mean time between arrivals (minutes).
    consultation_time : float
        Mean length of doctor's consultation (minutes).
    number_of_doctors : int
        Number of doctors.
    warm_up_period : int
        Duration of the warm-up period (minutes).
    data_collection_period : int
        Duration of the data collection period (minutes).
    verbose : bool
        Whether to print messages as simulation runs.
    """
    def __init__(
        self, interarrival_time=5, consultation_time=10,
        number_of_doctors=3, warm_up_period=30, data_collection_period=40,
        verbose=False#<<
    ):
        """
        Initialise Parameters instance.

        Parameters
        ----------
        interarrival_time : float
            Time between arrivals (minutes).
        consultation_time : float
            Length of consultation (minutes).
        number_of_doctors : int
            Number of doctors.
        warm_up_period : int
            Duration of the warm-up period (minutes).
        data_collection_period : int
            Duration of the data collection period (minutes).
        verbose : bool
            Whether to print messages as simulation runs.
        """
        self.interarrival_time = interarrival_time
        self.consultation_time = consultation_time
        self.number_of_doctors = number_of_doctors
        self.warm_up_period = warm_up_period
        self.data_collection_period = data_collection_period
        self.verbose = verbose
```

### Patient class

No changes required for `Patient` at this stage.

```{python}
#| echo: false
{{< include outputs_resources/patient.py >}}
```

### Model class

We need to update the `Model` class so its `run` method returns a list of patient attributes. This makes patient-level data easy to analyse after simulation runs.

```{python}
{{< include outputs_resources/model.py >}}
```

### Runner class

We're also introducing a new `Runner` class, responsible for running simulations and performing results calculations. This keeps calculation logic separate from the core modelling code. Also, this class will be modified to perform multiple model runs on the [Replications](replications.qmd) page.

The basic structure of `Runner` is as follows. It initialises with simulation parameters, runs the model for a single replication, and organises results into separate patient-level and run-level outputs.

```{python}
{{< include outputs_resources/runner.py >}}
```

Running the model via `Runner` produces our new structured results.

```{python}
{{< include outputs_resources/run.py >}}
```

Now we're ready to record some performance measures. 😎

:::

::: {.r-content}

Before we start calculating any simulation performance measures, some set-up is required. This includes making a new function `get_run_results()`, and calling that and `filter_warmup()` from within `model()`.

### Imports

We need to add `dplyr::bind_cols()` and `tibble::tibble()`.

*Bold lines show where the code has changed since the previous version. The highlight may correspond to a change anywhere on that line, not necessarily to the entire line.*

```{.r}
#' @keywords internal
"_PACKAGE"

## usethis namespace: start
#' @importFrom dplyr arrange bind_cols filter group_by mutate n slice ungroup #<<
#' @importFrom rlang .data
#' @importFrom simmer add_generator add_resource get_mon_arrivals
#' @importFrom simmer get_mon_resources release run seize simmer
#' @importFrom simmer timeout trajectory
#' @importFrom tibble tibble #<<
## usethis namespace: end

NULL

```

### Parameter function

No changes required.

```{r}
#| echo: false
#' Generate parameter list.
#'
#' @param interarrival_time Numeric. Time between arrivals (minutes).
#' @param consultation_time Numeric. Mean length of doctor's
#'   consultation (minutes).
#' @param number_of_doctors Numeric. Number of doctors.
#' @param warm_up_period Numeric. Duration of the warm-up period (minutes).
#' @param data_collection_period Numeric. Duration of the data collection
#'   period (minutes).
#' @param verbose Boolean. Whether to print messages as simulation runs.
#'
#' @return A named list of parameters.
#' @export

create_params <- function(
  interarrival_time = 5L,
  consultation_time = 10L,
  number_of_doctors = 3L,
  warm_up_period = 30L,
  data_collection_period = 40L,
  verbose = FALSE
) {
  list(
    interarrival_time = interarrival_time,
    consultation_time = consultation_time,
    number_of_doctors = number_of_doctors,
    warm_up_period = warm_up_period,
    data_collection_period = data_collection_period,
    verbose = verbose
  )
}
```

### Warm-up function

No changes required.

```{r}
#| echo: false
#' Filters arrivals and resources to remove warm-up results.
#'
#' @param result Named list with two tables: monitored arrivals & resources.
#' @param warm_up_period Length of warm-up period.
#'
#' @return The named list `result`, but with the tables (`arrivals` and
#' `resources`) filtered to remove warm-up results.
#' @export

filter_warmup <- function(result, warm_up_period) {
  # Skip filtering if the warm-up period is zero
  if (warm_up_period == 0L) return(result)

  # Arrivals: Keep only patients who came during the data collection period
  result[["arrivals"]] <- result[["arrivals"]] |>
    group_by(.data[["name"]]) |>
    filter(all(.data[["start_time"]] >= warm_up_period)) |>
    ungroup()

  # Resources: Filter to resource events in the data collection period
  dc_resources <- filter(result[["resources"]],
                         .data[["time"]] >= warm_up_period)

  if (nrow(dc_resources) > 0L) {

    # For each resource, get the last even during warm-up, and replace time
    # for that event with the start time of the data collection period
    last_usage <- result[["resources"]] |>
      filter(.data[["time"]] < warm_up_period) |>
      arrange(.data[["time"]]) |>
      group_by(.data[["resource"]]) |>
      slice(n()) |>
      mutate(time = warm_up_period) |>
      ungroup()

    # Set that last event of the resource/s as the first row of the dataframe
    # else calculations would assume all resources were idle at the start of
    # the data collection period
    result[["resources"]] <- rbind(last_usage, dc_resources)

  } else {
    # No events after warm-up; use filtered resource table as is
    result[["resources"]] <- dc_resources
  }

  result
}
```

### Run results function

Let's create a new function `get_run_results()`. This will be used to combine the performance measures we calculate into a single table.

```{r}
#' Get average results for the provided single run.
#'
#' @param results Named list with `arrivals` from `get_mon_arrivals()` and
#'   `resources` from `get_mon_resources()`
#'   (`per_resource = TRUE` and `ongoing = TRUE`).
#' @param run_number Integer index of current simulation run.
#'
#' @return Tibble with processed results from replication.
#' @export

get_run_results <- function(results, run_number) {
  metrics <- list()
  dplyr::bind_cols(tibble(replication = run_number), metrics)
}
```

### Model function

Having introduced the `filter_warmup()` function on the [Initialisation bias](warmup.qmd) page and `get_run_results()` above, we'll now call these from within `model()`.

We also need to correct the "replication" column. This is generated by `get_mon_arrivals()` and `get_mon_resources()`, and will have been set to 1. This is because these functions assume, when they haven't been supplied with a list of environments, that there was one replication. We need to set it to the correct `run_number`.

```{r}
#| file: outputs_resources/model.R
```

### Run the model

```{r}
param <- create_params()
result <- model(param = param, run_number = 1L)
kable(arrange(result[["arrivals"]], start_time))
kable(result[["resources"]])
kable(result[["run_results"]])
```

At the moment, `result[["run_results"]]` is pretty empty (just has the replication number). Let's add some performance measures!

:::

## Overview of performance measures

On this page, we breakdown how to add several different performance measures to your model. These include:

* Total arrivals
* Mean wait time
* Mean time in consultation
* Mean doctor utilisation
* Mean queue length
* Mean time in system
* Mean number of patients in the system
* Backlogged patient count
* Backlogged patient mean wait time

We introduce these on-by-one below, with each measure added and the required code changes clearly highlighted so it is easy to see what is needed for each statistic.

::: {.pale-blue}

**If you would prefer to work from a complete example with all of the measures implemented together, you can find that in the following callout:**

:::

:::: {.callout title="▶&nbsp;&nbsp;View how to record all performance measures" collapse="true"}

::: {.python-content}

Here, we just provide a brief description of the changes. For a full explanation of how these measures are implementation, please refer to the information on individual measures below.

**Note:** There are two approaches to recording utilisation - here, we have used the time-weighted approach (approach 2).

### Patient class

There are three new attributes:

* `wait_time` - used for mean wait time.
* `time_with_doctor` - used for mean time in consultation, backlogged patient count and backlogged patient mean wait time.
* `end_time` - used for mean time in system.

```{python}
{{< include outputs_resources/all_measures/patient.py >}}
```

### Monitored resource class

A new class `MonitoredResource` is used for mean doctor utilisation and mean queue length.

```{python}
{{< include outputs_resources/all_measures/monitored_resource.py >}}
```

### Model class

There are several new attributes and a method used for mean number of patients in the system (`area_n_in_system`, `time_last_n_in_system`, `n_in_system`, `update_n_in_system()`).

The doctor is now a `MonitoredResource`, as required for mean doctor utilisation and mean queue length.

During `consultation()`, the patient's `wait_time` is recorded for mean wait time. Time with the doctor is recorded for mean time in consultation, backlogged patient count and backlogged patient mean wait time. End time is recorded for mean time in system. 

```{python}
{{< include outputs_resources/all_measures/model.py >}}
```

### Runner class

In the Runner class, we calculate each of the metrics...

```{python}
{{< include outputs_resources/all_measures/runner.py >}}
```

### Run the model

```{python}
runner = Runner(param=Parameters())
result = runner.run_single(run=0)
HTML(to_html_datatable(result["patient"]))
print(result["run"])
```

:::

::: {.r-content}

### Imports

### Run results function

### Model function

:::

::::

## Total arrivals

:::: {.callout title="▶&nbsp;&nbsp;View how to record the total arrivals" collapse="true"}

::: {.python-content}

Recording the total number of arrivals is super simple! We just need to find the **length of the patient list**.

### Patient class

No changes required.

```{python}
#| echo: false
{{< include outputs_resources/patient.py >}}
```

### Model class

No changes required.

```{python}
#| echo: false
{{< include outputs_resources/model.py >}}
```

### Runner class

```{python}
class Runner:
    """
    Run the simulation.

    Attributes
    ----------
    param : Parameters
        Simulation parameters.
    """
    def __init__(self, param):
        """
        Initialise a new instance of the Runner class.

        Parameters
        ----------
        param : Parameters
            Simulation parameters.
        """
        self.param = param

    def run_single(self, run):
        """
        Runs the simulation once and performs results calculations.

        Parameters
        ----------
        run : int
            Run number for the simulation.

        Returns
        -------
        dict
            Contains patient-level results and results from each run.
        """
        model = Model(param=self.param, run_number=run)
        model.run()

        # Patient results
        patient_results = pd.DataFrame(model.results_list)
        patient_results["run"] = run

        # Run results
        run_results = {
            "run_number": run,
            "arrivals": len(patient_results)#<<
        }

        return {
            "patient": patient_results,
            "run": run_results
        }
```

### Run the model

```{python}
{{< include outputs_resources/run.py >}}
```

:::

::: {.r-content}

<div class="h3-tight"></div>

### Imports

We add `dplyr::n_distinct()` and `dplyr::summarise()`.

```{.r}
#' @keywords internal
"_PACKAGE"

## usethis namespace: start
#' @importFrom dplyr arrange bind_cols filter group_by mutate n n_distinct #<<
#' @importFrom dplyr slice summarise ungroup #<<
#' @importFrom rlang .data
#' @importFrom simmer add_generator add_resource get_mon_arrivals
#' @importFrom simmer get_mon_resources release run seize simmer
#' @importFrom simmer timeout trajectory
#' @importFrom tibble tibble
## usethis namespace: end

NULL

```

### Arrivals function

We introduce a new function `calc_arrivals()`. This determines the number of arrivals by counting the number of unique patient names in the arrival records.

```{r}
#' Calculate the number of arrivals
#'
#' @param arrivals Dataframe with times for each patient with each resource.
#'
#' @return Tibble with column containing total number of arrivals.
#' @export

calc_arrivals <- function(arrivals) {
  arrivals |>
    summarise(arrivals = n_distinct(.data[["name"]])) |>
    ungroup()
}
```

### Run results function

In `get_run_results()`, the table `results[["arrivals"]]` is passed to `calc_arrivals()`.

```{r}
#' Get average results for the provided single run.
#'
#' @param results Named list with `arrivals` from `get_mon_arrivals()` and
#'   `resources` from `get_mon_resources()`
#'   (`per_resource = TRUE` and `ongoing = TRUE`).
#' @param run_number Integer index of current simulation run.
#'
#' @return Tibble with processed results from replication.
#' @export

get_run_results <- function(results, run_number) {
  metrics <- list(
    calc_arrivals(results[["arrivals"]])#<<
  )
  dplyr::bind_cols(tibble(replication = run_number), metrics)
}
```

### Model function

No changes required.

```{r}
#| echo: false
#| file: outputs_resources/model.R
```

### Run the model

In `arrivals` we can see six patients listed, which aligns with our results table listing `arrivals = 6`.

```{r}
param <- create_params()
result <- model(param = param, run_number = 1L)
kable(arrange(result[["arrivals"]], start_time))
kable(result[["run_results"]])
```

:::

::::

## Mean wait time

> **Corresponding term in queueing theory:** Average wait time in queue, $\\W_q$

:::: {.callout title="▶&nbsp;&nbsp;View how to record the mean wait time" collapse="true"}

::: {.python-content}

For this measure, we need to record the **time each patient spent waiting** for the doctor.

### Patient class

A new attribute `wait_time` is added to the `Patient` class.

As it is an attribute, we have to give `wait_time` an initial value. A value like 0 would wrongly look like "no wait", so `np.nan` is used instead to clearly mean "we don't know this patient's wait time yet". Patients who still have `wait_time = np.nan` at the end of the simulation are those who were never seen, and this is used below when calculating metrics for unseen (backlogged) patients. Patients who have `wait_time = 0` at the end of the simulation did not wait for the resource (doctor) as the resource was free.

```{python}
class Patient:
    """
    Represents a patient.

    Attributes
    ----------
    patient_id : int
        Unique patient identifier.
    period : str
        Arrival period (warm up or data collection) with emoji.
    arrival_time : float
        Time patient entered the system (minutes).
    wait_time : float#<<
        Time spent waiting for the doctor (minutes).#<<
    """
    def __init__(self, patient_id, period, arrival_time):
        """
        Initialises a new patient.

        Parameters
        ----------
        patient_id : int
            Unique patient identifier.
        period : str
            Arrival period (warm up or data collection) with emoji.
        arrival_time : float
            Time patient entered the system (minutes).
        """
        self.patient_id = patient_id
        self.period = period
        self.arrival_time = arrival_time
        self.wait_time = np.nan#<<
```

### Model class

Within `Model.consultation()`, the start time of the consultation is recorded (`start_wait`). When the resource request has been granted, the time elapsed is saved under `patient.wait_time`.

```{python}
class Model:
    """
    Simulation model.

    Attributes
    ----------
    param : Parameters
        Simulation parameters.
    run_number : int
        Run number for random seed generation.
    env : simpy.Environment
        The SimPy environment for the simulation.
    doctor : simpy.Resource
        SimPy resource representing doctors.
    patients : list
        List of Patient objects.
    results_list : list
        List of dictionaries with the attributes of each patient.
    arrival_dist : Exponential
        Distribution used to generate random patient inter-arrival times.
    consult_dist : Exponential
        Distribution used to generate length of a doctor's consultation.
    """
    def __init__(self, param, run_number):
        """
        Create a new Model instance.

        Parameters
        ----------
        param : Parameters
            Simulation parameters.
        run_number : int
            Run number for random seed generation.
        """
        self.param = param
        self.run_number = run_number

        # Create SimPy environment
        self.env = simpy.Environment()

        # Create resource
        self.doctor = simpy.Resource(
            self.env, capacity=self.param.number_of_doctors
        )

        # Create a random seed sequence based on the run number
        ss = np.random.SeedSequence(self.run_number)
        seeds = ss.spawn(2)

        # Set up attributes to store results
        self.patients = []
        self.results_list = []

        # Initialise distributions
        self.arrival_dist = Exponential(mean=self.param.interarrival_time,
                                        random_seed=seeds[0])
        self.consult_dist = Exponential(mean=self.param.consultation_time,
                                        random_seed=seeds[1])

    def generate_arrivals(self):
        """
        Process that generates patient arrivals.
        """
        while True:
            # Sample and pass time to next arrival
            sampled_iat = self.arrival_dist.sample()
            yield self.env.timeout(sampled_iat)

            # Check whether arrived during warm-up or data collection
            if self.env.now < self.param.warm_up_period:
                period = "\U0001F538 WU"
            else:
                period = "\U0001F539 DC"

            # Create a new patient
            patient = Patient(patient_id=len(self.patients)+1,
                              period=period,
                              arrival_time=self.env.now)
            self.patients.append(patient)

            # Print arrival time
            if self.param.verbose:
                print(f"{patient.period} Patient {patient.patient_id} " +
                      f"arrives at time: {patient.arrival_time:.3f}")

            # Start process of consultation
            self.env.process(self.consultation(patient))

    def consultation(self, patient):
        """
        Process that simulates a consultation.

        Parameters
        ----------
        patient :
            Instance of the Patient() class representing a single patient.
        """
        start_wait = self.env.now#<<
        # Patient requests access to a doctor (resource)
        with self.doctor.request() as req:
            yield req

            # Record how long patient waited before consultation#<<
            patient.wait_time = self.env.now - start_wait#<<

            if self.param.verbose:
                print(f"{patient.period} Patient {patient.patient_id} starts" +
                      f" consultation at: {self.env.now:.3f}")

            # Sample consultation duration and pass time spent with doctor
            time_with_doctor = self.consult_dist.sample()
            yield self.env.timeout(time_with_doctor)

            # Record end time
            if self.param.verbose:
                print(f"{patient.period} Patient {patient.patient_id} " +
                      f"leaves at: {self.env.now:.3f}")

    def reset_results(self):
        """
        Reset results.
        """
        self.patients = []

    def warmup(self):
        """
        Reset result collection after the warm-up period.
        """
        if self.param.warm_up_period > 0:
            # Delay process until warm-up period has completed
            yield self.env.timeout(self.param.warm_up_period)
            # Reset results variables
            self.reset_results()
            if self.param.verbose:
                print(f"Warm up period ended at time: {self.env.now}")

    def run(self):
        """
        Run the simulation.
        """
        # Schedule arrival generator and warm-up
        self.env.process(self.generate_arrivals())
        self.env.process(self.warmup())

        # Run the simulation
        self.env.run(until=(self.param.warm_up_period +
                            self.param.data_collection_period))

        # Create list of dictionaries containing each patient's attributes
        self.results_list = [x.__dict__ for x in self.patients]
```

### Runner class

The mean `wait_time` from the patients is calculated.

```{python}
class Runner:
    """
    Run the simulation.

    Attributes
    ----------
    param : Parameters
        Simulation parameters.
    """
    def __init__(self, param):
        """
        Initialise a new instance of the Runner class.

        Parameters
        ----------
        param : Parameters
            Simulation parameters.
        """
        self.param = param

    def run_single(self, run):
        """
        Runs the simulation once and performs results calculations.

        Parameters
        ----------
        run : int
            Run number for the simulation.

        Returns
        -------
        dict
            Contains patient-level results and results from each run.
        """
        model = Model(param=self.param, run_number=run)
        model.run()

        # Patient results
        patient_results = pd.DataFrame(model.results_list)
        patient_results["run"] = run

        # Run results
        run_results = {
            "run_number": run,
            "mean_wait_time": patient_results["wait_time"].mean()#<<
        }

        return {
            "patient": patient_results,
            "run": run_results
        }
```

### Run the model

We can see each patient's wait time in `result["patient"]`, and the mean wait time in `result["run"]`.

```{python}
{{< include outputs_resources/run.py >}}
```

*The mean wait time is returned as `np.float64()` because NumPy's `mean()` function returns a NumPy float type instead of a built-in Python `float`. You can just treat it like a regular `float`, or convert it with `float()` if you want it to print without `np.float64()`.*

:::

::: {.r-content}

<div class="h3-tight"></div>

### Imports

We add:

* Functions from `dplyr`: `left_join` and `summarise`.
* Functions from `simmer`: `now` and `set_attribute`.
* Imports from two new packages: `tidyr` and `tidyselect`.

```{.r}
#' @keywords internal
"_PACKAGE"

## usethis namespace: start
#' @importFrom dplyr arrange bind_cols filter group_by left_join mutate n #<<
#' @importFrom dplyr slice summarise ungroup #<<
#' @importFrom rlang .data
#' @importFrom simmer add_generator add_resource get_mon_arrivals
#' @importFrom simmer get_mon_resources now release run seize set_attribute #<<
#' @importFrom simmer simmer timeout trajectory
#' @importFrom tibble tibble
#' @importFrom tidyr drop_na pivot_wider #<<
#' @importFrom tidyselect any_of #<<
## usethis namespace: end

NULL
```

### Wait time function

A new `calc_mean_wait()` function removes patients still waiting (since their wait times are censored and including them would underestimate the mean), then calculates the mean wait time for each resource.

```{r}
#' Calculate the mean wait time for each resource
#'
#' @param arrivals Dataframe with times for each patient with each resource.
#'
#' @return Tibble with columns containing result for each resource.
#' @export

calc_mean_wait <- function(arrivals) {

  # Create subset of data that removes patients who were still waiting
  complete_arrivals <- drop_na(arrivals, any_of("wait_time"))

  # Calculate mean wait time for each resource
  complete_arrivals |>
    group_by(.data[["resource"]]) |>
    summarise(mean_wait_time = mean(.data[["wait_time"]])) |>
    pivot_wider(names_from = "resource",
                values_from = "mean_wait_time",
                names_glue = "mean_wait_time_{resource}") |>
    ungroup()
}
```

### Run results function

In `get_run_results()`, the table `results[["arrivals"]]` is passed to `calc_mean_wait()`.

```{r}
#' Get average results for the provided single run.
#'
#' @param results Named list with `arrivals` from `get_mon_arrivals()` and
#'   `resources` from `get_mon_resources()`
#'   (`per_resource = TRUE` and `ongoing = TRUE`).
#' @param run_number Integer index of current simulation run.
#'
#' @return Tibble with processed results from replication.
#' @export

get_run_results <- function(results, run_number) {
  metrics <- list(
    calc_mean_wait(results[["arrivals"]])#<<
  )
  dplyr::bind_cols(tibble(replication = run_number), metrics)
}
```

### Model function

In `model()`, we save the time the patient started their consultation (`doctor_serve_start`) using `set_attribute()`.

To retrieve this later using simmer's `get_mon_attributes()` function, we have to set `mon = 2L` within the patient generator, as this enables attribute monitoring beyond basic arrival tracking.

When retrieving the attribute, we restructure the returned dataframe using a generic approach that handles any attributes following the pattern `resourcename_attribute` - for example, transforming `doctor_serve_start` into `serve_start` (aligned with the corresponding doctor rows).

We can calculate the wait time for each patient by subtracting their arrival time (`start_time`) from when service began (`serve_start`).

```{r}
#' Run simulation.
#'
#' @param param List. Model parameters.
#' @param run_number Numeric. Run number for random seed generation.
#'
#' @return Named list with tables `arrivals`, `resources` and `run_results`.
#' @export

model <- function(param, run_number) {

  # Set random seed based on run number
  set.seed(run_number)

  # Create simmer environment
  env <- simmer("simulation", verbose = param[["verbose"]])

  # Define the patient trajectory
  patient <- trajectory("consultation") |>
    seize("doctor", 1L) |>
    # Record time resource is obtained #<<
    set_attribute("doctor_serve_start", function() now(env)) |> #<<
    timeout(function() {
      rexp(n = 1L, rate = 1L / param[["consultation_time"]])
    }) |>
    release("doctor", 1L)

  env <- env |>
    # Add doctor resource
    add_resource("doctor", param[["number_of_doctors"]]) |>
    # Add patient generator (mon=2 so can get manual attributes) #<<
    add_generator("patient", patient, function() {
      rexp(n = 1L, rate = 1L / param[["interarrival_time"]])
    }, mon = 2L) |> #<<
    # Run the simulation
    simmer::run(until = (param[["warm_up_period"]] +
                           param[["data_collection_period"]]))

  # Extract information on arrivals and resources from simmer environment
  result <- list(
    arrivals = get_mon_arrivals(env, per_resource = TRUE, ongoing = TRUE),
    resources = get_mon_resources(env)
  )

  # Get the extra arrivals attributes #<<
  extra_attributes <- get_mon_attributes(env) |> #<<
    dplyr::select("name", "key", "value") |> #<<
    # Add column with resource name, and remove that from key #<<
    mutate(resource = gsub("_.+", "", .data[["key"]]), #<<
           key = gsub("^[^_]+_", "", .data[["key"]])) |> #<<
    pivot_wider(names_from = "key", values_from = "value") #<<

  # Merge extra attributes with the arrival data #<<
  result[["arrivals"]] <- left_join( #<<
    result[["arrivals"]], extra_attributes, by = c("name", "resource") #<<
  ) #<<

  # Filter to remove results from the warm-up period
  result <- filter_warmup(result, param[["warm_up_period"]])

  # Replace "replication" value with appropriate run number
  result[["arrivals"]] <- mutate(result[["arrivals"]],
                                 replication = run_number)
  result[["resources"]] <- mutate(result[["resources"]],
                                  replication = run_number)

  # Calculate wait time for each patient #<<
  result[["arrivals"]] <- mutate( #<<
    result[["arrivals"]], #<<
    wait_time = .data[["serve_start"]] - .data[["start_time"]]#<<
  )#<<

  # Calculate the average results for that run
  result[["run_results"]] <- get_run_results(result, run_number)

  result
}
```

### Run the model

```{r}
param <- create_params()
result <- model(param = param, run_number = 1L)
kable(arrange(result[["arrivals"]], start_time))
kable(result[["run_results"]])
```

:::

::::

## Mean time in consultation

> **Corresponding term in queueing theory:** Service time, 1/$\mu$

:::: {.callout title="▶&nbsp;&nbsp;View how to record the mean time spent in the consultation" collapse="true"}

::: {.python-content}

<div class="h3-tight"></div>

### Patient class

A new attribute `time_with_doctor` is add to the `Patient` class.

```{python}
{{< include outputs_resources/patient_timewithdoctor.py >}}
```

### Model class

Within `Model.consultation()`, the sampled time with the doctor is saved under `patient.time_with_doctor`.

```{python}
{{< include outputs_resources/model_timewithdoctor.py >}}
```

### Runner class

The mean `time_with_doctor` from the patients is calculated.

```{python}
class Runner:
    """
    Run the simulation.

    Attributes
    ----------
    param : Parameters
        Simulation parameters.
    """
    def __init__(self, param):
        """
        Initialise a new instance of the Runner class.

        Parameters
        ----------
        param : Parameters
            Simulation parameters.
        """
        self.param = param

    def run_single(self, run):
        """
        Runs the simulation once and performs results calculations.

        Parameters
        ----------
        run : int
            Run number for the simulation.

        Returns
        -------
        dict
            Contains patient-level results and results from each run.
        """
        model = Model(param=self.param, run_number=run)
        model.run()

        # Patient results
        patient_results = pd.DataFrame(model.results_list)
        patient_results["run"] = run

        # Run results
        run_results = {
            "run_number": run,
            "mean_time_with_doctor": (#<<
                patient_results["time_with_doctor"].mean()#<<
            )#<<
        }

        return {
            "patient": patient_results,
            "run": run_results
        }
```

### Run the model

We can see each patient's consultation length in `result["patient"]`, and the mean time in `result["run"]`.

```{python}
{{< include outputs_resources/run.py >}}
```

*The mean consultation length is returned as `np.float64()` because NumPy's `mean()` function returns a NumPy float type instead of a built-in Python `float`. You can just treat it like a regular `float`, or convert it with `float()` if you want it to print without `np.float64()`.*

:::

::: {.r-content}

<div class="h3-tight"></div>

### Imports

We add:

* Functions from `dplyr`: `left_join` and `summarise`.
* Functions from `simmer`: `get_attribute` and `set_attribute`.
* Imports from two new packages: `tidyr` and `tidyselect`.

```{.r}
#' @keywords internal
"_PACKAGE"

## usethis namespace: start
#' @importFrom dplyr arrange bind_cols filter group_by left_join mutate n slice #<<
#' @importFrom dplyr summarise ungroup #<<
#' @importFrom rlang .data
#' @importFrom simmer add_generator add_resource get_attribute get_mon_arrivals #<<
#' @importFrom simmer get_mon_resources release run seize set_attribute simmer #<<
#' @importFrom simmer timeout trajectory
#' @importFrom tibble tibble
#' @importFrom tidyr drop_na pivot_wider #<<
#' @importFrom tidyselect any_of #<<
## usethis namespace: end

NULL
```

### Service length function

A new function `calc_mean_serve_length()` finds the mean length of time patients spent with each resource using a new column we'll introduce in `model()` called `serve_length`.

```{r}
#' Calculate the mean length of time patients spent with each resource
#'
#' @param arrivals Dataframe with times for each patient with each resource.
#' @param resources Dataframe with times patients use or queue for resources.
#' @param groups Optional list of columns to group by for the calculation.
#'
#' @return Tibble with columns containing result for each resource.
#' @export

calc_mean_serve_length <- function(arrivals, resources, groups = NULL) {

  # Create subset of data that removes patients who were still waiting
  complete_arrivals <- drop_na(arrivals, any_of("serve_length"))

  # Calculate mean serve time for each resource
  complete_arrivals |>
    group_by(.data[["resource"]]) |>
    summarise(mean_serve_time = mean(.data[["serve_length"]])) |>
    pivot_wider(names_from = "resource",
                values_from = "mean_serve_time",
                names_glue = "mean_time_with_{resource}") |>
    ungroup()
}
```

### Run results function

In `get_run_results()`, the table `results[["arrivals"]]` is passed to `calc_mean_serve_length()`.

```{r}
#' Get average results for the provided single run.
#'
#' @param results Named list with `arrivals` from `get_mon_arrivals()` and
#'   `resources` from `get_mon_resources()`
#'   (`per_resource = TRUE` and `ongoing = TRUE`).
#' @param run_number Integer index of current simulation run.
#'
#' @return Tibble with processed results from replication.
#' @export

get_run_results <- function(results, run_number) {
  metrics <- list(
    calc_mean_serve_length(results[["arrivals"]])#<<
  )
  dplyr::bind_cols(tibble(replication = run_number), metrics)
}
```

### Model function

In `model()`, we save the sampled length of the consultation (`doctor_serve_length`) using `set_attribute()`. This is then used for `timeout()`, retrieving it again with `get_attribute()`.

Next, we repeat the steps we did for calculating mean wait time: setting `mon = 2L` in the patient generator, retrieving results using `get_mon_attributes()`, and restructuring the returned dataframe to transform `doctor_serve_length` into `serve_length` (aligned with the corresponding doctor rows).

```{r}
#' Run simulation.
#'
#' @param param List. Model parameters.
#' @param run_number Numeric. Run number for random seed generation.
#'
#' @return Named list with tables `arrivals`, `resources` and `run_results`.
#' @export

model <- function(param, run_number) {

  # Set random seed based on run number
  set.seed(run_number)

  # Create simmer environment
  env <- simmer("simulation", verbose = param[["verbose"]])

  # Define the patient trajectory
  patient <- trajectory("consultation") |>
    seize("doctor", 1L) |>
    # Record sampled length of consultation #<<
    set_attribute("doctor_serve_length", function() { #<<
      rexp(n = 1L, rate = 1L / param[["consultation_time"]]) #<<
    }) |> #<<
    timeout(function() get_attribute(env, "doctor_serve_length")) |> #<<
    release("doctor", 1L)

  env <- env |>
    # Add doctor resource
    add_resource("doctor", param[["number_of_doctors"]]) |>
    # Add patient generator (mon=2 so can get manual attributes) #<<
    add_generator("patient", patient, function() {
      rexp(n = 1L, rate = 1L / param[["interarrival_time"]])
    }, mon = 2L) |> #<<
    # Run the simulation
    simmer::run(until = (param[["warm_up_period"]] +
                           param[["data_collection_period"]]))

  # Extract information on arrivals and resources from simmer environment
  result <- list(
    arrivals = get_mon_arrivals(env, per_resource = TRUE, ongoing = TRUE),
    resources = get_mon_resources(env)
  )

  # Get the extra arrivals attributes#<<
  extra_attributes <- get_mon_attributes(env) |> #<<
    dplyr::select("name", "key", "value") |> #<<
    # Add column with resource name, and remove that from key #<<
    mutate(resource = gsub("_.+", "", .data[["key"]]), #<<
           key = gsub("^[^_]+_", "", .data[["key"]])) |> #<<
    pivot_wider(names_from = "key", values_from = "value") #<<

  # Merge extra attributes with the arrival data#<<
  result[["arrivals"]] <- left_join( #<<
    result[["arrivals"]], extra_attributes, by = c("name", "resource") #<<
  ) #<<

  # Filter to remove results from the warm-up period
  result <- filter_warmup(result, param[["warm_up_period"]])

  # Replace "replication" value with appropriate run number
  result[["arrivals"]] <- mutate(result[["arrivals"]],
                                 replication = run_number)
  result[["resources"]] <- mutate(result[["resources"]],
                                  replication = run_number)

  # Calculate the average results for that run
  result[["run_results"]] <- get_run_results(result, run_number)

  result
}
```

### Run the model

**Why do the patient results look different?** This measure requires saving sampled times using `set_attribute()` ahead of resource use. This changes the order in which random numbers are drawn from the simulation's global random number generator. As a result, even with the same seed and parameters, the results will differ from previous runs.

```{r}
param <- create_params()
result <- model(param = param, run_number = 1L)
kable(arrange(result[["arrivals"]], start_time))
kable(result[["run_results"]])
```

:::

::::

## Mean doctor utilisation

> **Corresponding term in queueing theory:** Server utilisation, $\rho$

:::: {.python-content}

Mean utilisation measures the proportion of available doctor time that is actually used during the data collection period. We have suggested two different approaches for measuring utilisation - both return the same result.

### Approach 1

The first approach is **easier to understand**, as it just sums consultation times for each patient. However, it relies on making **corrections** when consultations overlap the end of the simulation or span the warm-up and data collection periods.

::: {.callout title="▶&nbsp;&nbsp;View how to record utilisation" collapse="true"}

<div class="h3-tight"></div>

### Patient class

No changes required.

```{python}
#| echo: false
{{< include outputs_resources/patient.py >}}
```

### Model class

A new attribute `doctor_time_used` is used to record the total time patients spend with doctor. This value is reset at the end of any warm-up period, so that only data from the main collection phase is analysed.

Recording the time used requires two corrections:

**End of simulation overrun**. If a consultation begins during the data collection period and extends past the simulation end, only the time up to simulation end is included. This prevents overestimating usage by excluding extra time beyond the simulation window.

**Warm-up overrun**. For models with a warm-up, consultations that begin during warm-up but finish during data collection would be missed after the reset, leading to underestimated usage. To correct for this, any time overlapping data collection (for these patients) is recorded in the `doctor_time_used_correction` attribute and added back in after the warm-up reset.

```{python}
class Model:
    """
    Simulation model.

    Attributes
    ----------
    param : Parameters
        Simulation parameters.
    run_number : int
        Run number for random seed generation.
    env : simpy.Environment
        The SimPy environment for the simulation.
    doctor : simpy.Resource
        SimPy resource representing doctors.
    patients : list
        List of Patient objects.
    doctor_time_used : float#<<
        Total time that doctor resources were used for (minutes).#<<
    doctor_time_used_correction : float#<<
        Adjustment for doctor time. Without this, usage is underestimated#<<
        since patients whose consultations began in warm-up but ended#<<
        during data collection are excluded.#<<
    results_list : list
        List of dictionaries with the attributes of each patient.
    arrival_dist : Exponential
        Distribution used to generate random patient inter-arrival times.
    consult_dist : Exponential
        Distribution used to generate length of a doctor's consultation.
    """
    def __init__(self, param, run_number):
        """
        Create a new Model instance.

        Parameters
        ----------
        param : Parameters
            Simulation parameters.
        run_number : int
            Run number for random seed generation.
        """
        self.param = param
        self.run_number = run_number

        # Create SimPy environment
        self.env = simpy.Environment()

        # Create resource
        self.doctor = simpy.Resource(
            self.env, capacity=self.param.number_of_doctors
        )

        # Create a random seed sequence based on the run number
        ss = np.random.SeedSequence(self.run_number)
        seeds = ss.spawn(2)

        # Set up attributes to store results
        self.patients = []
        self.results_list = []
        self.doctor_time_used = 0#<<
        self.doctor_time_used_correction = 0#<<

        # Initialise distributions
        self.arrival_dist = Exponential(mean=self.param.interarrival_time,
                                        random_seed=seeds[0])
        self.consult_dist = Exponential(mean=self.param.consultation_time,
                                        random_seed=seeds[1])

    def generate_arrivals(self):
        """
        Process that generates patient arrivals.
        """
        while True:
            # Sample and pass time to next arrival
            sampled_iat = self.arrival_dist.sample()
            yield self.env.timeout(sampled_iat)

            # Check whether arrived during warm-up or data collection
            if self.env.now < self.param.warm_up_period:
                period = "\U0001F538 WU"
            else:
                period = "\U0001F539 DC"

            # Create a new patient
            patient = Patient(patient_id=len(self.patients)+1,
                              period=period,
                              arrival_time=self.env.now)
            self.patients.append(patient)

            # Print arrival time
            if self.param.verbose:
                print(f"{patient.period} Patient {patient.patient_id} " +
                      f"arrives at time: {patient.arrival_time:.3f}")

            # Start process of consultation
            self.env.process(self.consultation(patient))

    def consultation(self, patient):
        """
        Process that simulates a consultation.

        Parameters
        ----------
        patient :
            Instance of the Patient() class representing a single patient.
        """
        # Patient requests access to a doctor (resource)
        with self.doctor.request() as req:
            yield req

            if self.param.verbose:
                print(f"{patient.period} Patient {patient.patient_id} " +
                      f"starts consultation at: {self.env.now:.3f}")

            # Sample consultation duration
            time_with_doctor = self.consult_dist.sample()

            # Add to total doctor time used#<<
            # If it runs past simulation end, only count the time until end#<<
            remaining_time = (#<<
                self.param.warm_up_period +#<<
                self.param.data_collection_period) - self.env.now#<<
            self.doctor_time_used += min(#<<
                time_with_doctor, remaining_time)#<<

            # During warm-up: check if consultation continues past warm-up.#<<
            # If so, record the portion overlapping with data collection in#<<
            # doctor_time_used_correction (capped at the simulation end).#<<
            remaining_warmup = self.param.warm_up_period - self.env.now#<<
            if remaining_warmup > 0:#<<
                time_exceeding_warmup = time_with_doctor - remaining_warmup#<<
                if time_exceeding_warmup > 0:#<<
                    self.doctor_time_used_correction += min(#<<
                        time_exceeding_warmup,#<<
                        self.param.data_collection_period)#<<

            # Pass time spent with the doctor
            yield self.env.timeout(time_with_doctor)

            # Record end time
            if self.param.verbose:
                print(f"{patient.period} Patient {patient.patient_id} " +
                      f"leaves at: {self.env.now:.3f}")

    def reset_results(self):
        """
        Reset results.
        """
        self.patients = []
        self.doctor_time_used = 0#<<

    def warmup(self):
        """
        Reset result collection after the warm-up period.
        """
        if self.param.warm_up_period > 0:
            # Delay process until warm-up period has completed
            yield self.env.timeout(self.param.warm_up_period)

            # Reset results variables
            self.reset_results()
            if self.param.verbose:
                print(f"Warm up period ended at time: {self.env.now}")

            # Add correction for patients whose consultations began in#<<
            # warm-up but continued into data collection.#<<
            self.doctor_time_used += self.doctor_time_used_correction#<<

    def run(self):
        """
        Run the simulation.
        """
        # Schedule arrival generator and warm-up
        self.env.process(self.generate_arrivals())
        self.env.process(self.warmup())

        # Run the simulation
        self.env.run(until=(self.param.warm_up_period +
                            self.param.data_collection_period))

        # Create list of dictionaries containing each patient's attributes
        self.results_list = [x.__dict__ for x in self.patients]
```

### Runner class

Mean utilisation can now be calculated as:

$$
\text{mean utilisation} = \frac{\text{doctor\_time\_used}}{\text{number of doctors} \times \text{data collection period}}
$$

```{python}
class Runner:
    """
    Run the simulation.

    Attributes
    ----------
    param : Parameters
        Simulation parameters.
    """
    def __init__(self, param):
        """
        Initialise a new instance of the Runner class.

        Parameters
        ----------
        param : Parameters
            Simulation parameters.
        """
        self.param = param

    def run_single(self, run):
        """
        Runs the simulation once and performs results calculations.

        Parameters
        ----------
        run : int
            Run number for the simulation.

        Returns
        -------
        dict
            Contains patient-level results and results from each run.
        """
        model = Model(param=self.param, run_number=run)
        model.run()

        # Patient results
        patient_results = pd.DataFrame(model.results_list)
        patient_results["run"] = run

        # Run results
        run_results = {
            "run_number": run,
            "mean_utilisation": model.doctor_time_used / (#<<
                self.param.number_of_doctors *#<<
                self.param.data_collection_period#<<
            )#<<
        }

        return {
            "patient": patient_results,
            "run": run_results
        }
```

### Run the model

```{python}
{{< include outputs_resources/run.py >}}
```

:::

### Approach 2

This second approach introduces a **new class** and may initially seem **harder to grasp**, but it has the advantage of requiring **fewer changes** to the rest of the simulation code. It is also handy for [choosing the length of the warm-up period](length_warmup.qmd), as it provides you with utilisation at each timepoint (rather than a final overall utilisation).

::: {.callout title="▶&nbsp;&nbsp;View how to record utilisation (time-weighted approach)" collapse="true"}

<div class="h3-tight"></div>

### Patient class

No changes required.

```{python}
#| echo: false
{{< include outputs_resources/patient.py >}}
```

### Monitored resource class

For this approach, we create a new class called `MonitoredResource`. This is explained line-by-line below.

```{python}
class MonitoredResource(simpy.Resource):
    """
    SimPy Resource subclass that tracks resource utilisation.

    Attributes
    ----------
    time_last_event : list
        Time of the most recent resource request or release event.
    area_resource_busy : list
        Cumulative time resources were busy, for computing utilisation.

    Notes
    -----
    Class adapted from Monks, Harper and Heather 2025.
    """

    def __init__(self, *args, **kwargs):
        """
        Initialise the MonitoredResource, resetting monitoring statistics.

        Parameters
        ----------
        *args :
            Positional arguments to be passed to the parent class constructor.
        **kwargs :
            Keyword arguments to be passed to the parent class constructor.
        """
        # Initialise a SimPy Resource
        super().__init__(*args, **kwargs)
        # Run the init_results() method
        self.init_results()

    def init_results(self):
        """
        Reset monitoring attributes.
        """
        self.time_last_event = [self._env.now]
        self.area_resource_busy = [0.0]

    def request(self, *args, **kwargs):
        """
        Update time-weighted statistics before requesting a resource.

        Parameters
        ----------
        *args :
            Positional arguments to be passed to the parent class.
        **kwargs :
            Keyword arguments to be passed to the parent class.

        Returns
        -------
        simpy.events.Event
            Event for the resource request.
        """
        # Update time-weighted statistics
        self.update_time_weighted_stats()
        # Request a resource
        return super().request(*args, **kwargs)

    def release(self, *args, **kwargs):
        """
        Update time-weighted statistics before releasing a resource.

        Parameters
        ----------
        *args :
            Positional arguments to be passed to the parent class.
        **kwargs :
            Keyword arguments to be passed to the parent class.

        Returns
        -------
        simpy.events.Event
            Event for the resource release.
        """
        # Update time-weighted statistics
        self.update_time_weighted_stats()
        # Release a resource
        return super().release(*args, **kwargs)

    def update_time_weighted_stats(self):
        """
        Update the time-weighted statistics for resource usage.

        Notes
        -----
        - These sums can be referred to as "the area under the curve".
        - They are called "time-weighted" statistics as they account for how
          long certain events or states persist over time.
        """
        # Calculate time since last event
        time_since_last_event = self._env.now - self.time_last_event[-1]

        # Add record of current time
        self.time_last_event.append(self._env.now)

        # Add "area under curve" of resources in use
        # self.count is the number of resources in use
        self.area_resource_busy.append(self.count * time_since_last_event)
```

#### Explaining `MonitoredResource`...

```{.python}
class MonitoredResource(simpy.Resource):
    """
    SimPy Resource subclass that tracks resource utilisation.

    Attributes
    ----------
    time_last_event : list
        Time of the most recent resource request or release event.
    area_resource_busy : list
        Cumulative time resources were busy, for computing utilisation.

    Notes
    -----
    Class adapted from Monks, Harper and Heather 2025.
    """

    def __init__(self, *args, **kwargs):
        """
        Initialise the MonitoredResource, resetting monitoring statistics.

        Parameters
        ----------
        *args :
            Positional arguments to be passed to the parent class constructor.
        **kwargs :
            Keyword arguments to be passed to the parent class constructor.
        """
        # Initialise a SimPy Resource
        super().__init__(*args, **kwargs)
        # Run the init_results() method
        self.init_results()
```

This class inherits SimPy's `Resource` class (if new to inheritance, see the [Code organisation](../setup/code_structure.qmd) page for an introduction). This means it has same functionality, but we can modify and add to it.

In the `__init__` method, we initialise the object by calling the constructor of the parent `simpy.Resource` class via `super().__init__(*args, **kwargs)`, passing through all positional and keyword arguments unchanged. We then call the new `init_results` method to set up the monitoring attributes.

> Those with a C/C++ background might see `*args` and `**kwargs` and think of variable argument lists. In Python, `*` and `**` are used to collect (and later unpack) positional and keyword arguments, which we forward directly to the parent class constructor and methods using `super(...)(*args, **kwargs)`.

<br>

```{.python}
def init_results(self):
    """
    Reset monitoring attributes.
    """
    self.time_last_event = [self._env.now]
    self.area_resource_busy = [0.0]
```

This method resets all tracking attributes. This is called when the object is created or when usage statistics need to be cleared, such as after a warm-up period.

<br>

```{.python}
def request(self, *args, **kwargs):
    """
    Update time-weighted statistics before requesting a resource.

    Parameters
    ----------
    *args :
        Positional arguments to be passed to the parent class.
    **kwargs :
        Keyword arguments to be passed to the parent class.

    Returns
    -------
    simpy.events.Event
        Event for the resource request.
    """
    # Update time-weighted statistics
    self.update_time_weighted_stats()
    # Request a resource
    return super().request(*args, **kwargs)
```

Whenever a resource is requested, this method first updates time-weighted statistics based on the current system state. It then makes the actual resource request via the parent class, ensuring all tracking is up to date.

<br>

```{.python}
def release(self, *args, **kwargs):
    """
    Update time-weighted statistics before releasing a resource.

    Parameters
    ----------
    *args :
        Positional arguments to be passed to the parent class.
    **kwargs :
        Keyword arguments to be passed to the parent class.

    Returns
    -------
    simpy.events.Event
        Event for the resource release.
    """
    # Update time-weighted statistics
    self.update_time_weighted_stats()
    # Release a resource
    return super().release(*args, **kwargs)
```

Just as we did with the `request()` method, the `release()` method updates usage statistics before a resource if released.

<br>

```{.python}
def update_time_weighted_stats(self):
    """
    Update the time-weighted statistics for resource usage.

    Notes
    -----
    - These sums can be referred to as "the area under the curve".
    - They are called "time-weighted" statistics as they account for how
        long certain events or states persist over time.
    """
    # Calculate time since last event
    time_since_last_event = self._env.now - self.time_last_event[-1]

    # Add record of current time
    self.time_last_event.append(self._env.now)

    # Add "area under curve" of resources in use
    # self.count is the number of resources in use
    self.area_resource_busy.append(self.count * time_since_last_event)
```

Calculates and records usage statistics since the last resource event (request or release). It determines how long the resource usage lasted, then adds its "area under the curve" to the running total.

### Model class

In `Model`, we make three changes:

1. The `doctor` attribute is now set up using `MonitoredResource` (instead of `simpy.Resource`).

2. In `reset_results()`, we call the `MonitoredResource` method `init_results()` to reset their monitoring at the end of a warm-up period.

3. In `run()`, once the simulation ends, the `MonitoredResource` method `update_time_weighted_stats()` is called one final time, to update with the usage between the last record resource event (request or release) and the simulation end.

```{python}
{{< include outputs_resources/model_monitored.py >}}
```

### Runner class

We've tracked the number of busy doctors over time by summing `busy doctors x duration` across all intervals - referred to as the "area under the curve". We now divide this by `number of doctors x data collection length` to give the average fraction of doctor time that was used.

For a clear visual explanation of what we mean by "area under the curve" and how this calculation works, see the [Time-weighted averages](#time-weighted-averages) section at the end of this page.

```{python}
class Runner:
    """
    Run the simulation.

    Attributes
    ----------
    param : Parameters
        Simulation parameters.
    """
    def __init__(self, param):
        """
        Initialise a new instance of the Runner class.

        Parameters
        ----------
        param : Parameters
            Simulation parameters.
        """
        self.param = param

    def run_single(self, run):
        """
        Runs the simulation once and performs results calculations.

        Parameters
        ----------
        run : int
            Run number for the simulation.

        Returns
        -------
        dict
            Contains patient-level results and results from each run.
        """
        model = Model(param=self.param, run_number=run)
        model.run()

        # Patient results
        patient_results = pd.DataFrame(model.results_list)
        patient_results["run"] = run

        # Run results
        run_results = {
            "run_number": run,
            "mean_utilisation_tw": (#<<
                sum(model.doctor.area_resource_busy) / (#<<
                    self.param.number_of_doctors *#<<
                    self.param.data_collection_period#<<
                )#<<
            )#<<
        }

        return {
            "patient": patient_results,
            "run": run_results
        }
```

### Run the model

```{python}
{{< include outputs_resources/run.py >}}
```

:::

::::

:::: {.r-content}

Mean utilisation measures the proportion of available doctor time that is actually used during the data collection period.

::: {.callout title="▶&nbsp;&nbsp;View how to record utilisation" collapse="true"}

<div class="h3-tight"></div>

### Imports

We add `dplyr::row_number()`, `dplyr::summarise()` and `tidyr::pivot_wider()`.

```{.r}
#' @keywords internal
"_PACKAGE"

## usethis namespace: start
#' @importFrom dplyr arrange bind_cols filter group_by mutate n row_number #<<
#' @importFrom dplyr slice summarise ungroup #<<
#' @importFrom rlang .data
#' @importFrom simmer add_generator add_resource get_mon_arrivals
#' @importFrom simmer get_mon_resources release run seize simmer
#' @importFrom simmer timeout trajectory
#' @importFrom tibble tibble
#' @importFrom tidyr pivot_wider #<<
## usethis namespace: end

NULL
```

### Utilisation function

Our function for utilisation is a **time-weighted calculation**: it computes a weighted mean, multiplying each interval's utilisation by its interval duration.

For a clear visual explanation of how these calculations work, refer to the [Time-weighted averages](#time-weighted-averages) section below.

As described in the docstring, the calculation is adapted from the `plot.resources.utilization()` function in simmer.plot 0.1.18, which is shared under an MIT Licence (Ucar I, Smeets B (2023). simmer.plot: Plotting Methods for 'simmer'. <https://r-simmer.org> <https://github.com/r-simmer/simmer.plot>).

The function is explained line-by-line below...

```{r}
#' Calculate the resource utilisation
#'
#' Utilisation is given by the total effective usage time (`in_use`) over
#' the total time intervals considered (`dt`).
#'
#' Credit: The utilisation calculation is adapted from the
#' `plot.resources.utilization()` function in simmer.plot 0.1.18, which is
#' shared under an MIT Licence (Ucar I, Smeets B (2023). simmer.plot: Plotting
#' Methods for 'simmer'. https://r-simmer.org
#' https://github.com/r-simmer/simmer.plot.).
#'
#' @param resources Dataframe with times patients use or queue for resources.
#' @param simulation_end_time Time at end of simulation run.
#' @param summarise If TRUE, return overall utilisation. If FALSE, just return
#' the resource dataframe with the additional columns interval_duration,
#' effective_capacity and utilisation.
#'
#' @return Tibble with columns containing result for each resource.
#' @export

calc_utilisation <- function(
  resources, simulation_end_time, summarise = TRUE
) {

  # Calculate utilisation
  util_df <- resources |>
    group_by(.data[["resource"]]) |>
    mutate(
      # Calculate time between this row and the next. For final row,
      # lead(time) returns NA, so use simulation_end_time - time
      interval_duration = ifelse(
        row_number() == n(),
        simulation_end_time - .data[["time"]],
        lead(.data[["time"]]) - .data[["time"]]
      ),
      # Ensures effective capacity is never less than number of servers in
      # use (in case of situations where servers may exceed "capacity").
      effective_capacity = pmax(.data[["capacity"]], .data[["server"]]),
      # Divide number of servers in use by effective capacity
      # Set to NA if effective capacity is 0
      utilisation = ifelse(.data[["effective_capacity"]] > 0L,
                           .data[["server"]] / .data[["effective_capacity"]],
                           NA_real_)
    )

  # If summarise = TRUE, find total utilisation
  if (summarise) {
    util_df |>
      summarise(
        # Multiply each utilisation by its own unique duration. The total of
        # those is then divided by the total duration of all intervals.
        # Hence, we are calculated a time-weighted average utilisation.
        utilisation = (sum(.data[["utilisation"]] *
                             .data[["interval_duration"]], na.rm = TRUE) /
                         sum(.data[["interval_duration"]], na.rm = TRUE))
      ) |>
      pivot_wider(names_from = "resource",
                  values_from = "utilisation",
                  names_glue = "utilisation_{resource}") |>
      ungroup()
  } else {
    # If summarise = FALSE, just return the util_df with no further processing
    ungroup(util_df)
  }
}
```

#### Explaining the utilisation function

```{.r}
#' Calculate the resource utilisation
#'
#' Utilisation is given by the total effective usage time (`in_use`) over
#' the total time intervals considered (`dt`).
#'
#' Credit: The utilisation calculation is adapted from the
#' `plot.resources.utilization()` function in simmer.plot 0.1.18, which is
#' shared under an MIT Licence (Ucar I, Smeets B (2023). simmer.plot: Plotting
#' Methods for 'simmer'. https://r-simmer.org
#' https://github.com/r-simmer/simmer.plot.).
#'
#' @param resources Dataframe with times patients use or queue for resources.
#' @param simulation_end_time Time at end of simulation run.
#' @param summarise If TRUE, return overall utilisation. If FALSE, just return
#' the resource dataframe with the additional columns interval_duration,
#' effective_capacity and utilisation.
#'
#' @return Tibble with columns containing result for each resource.
#' @export

calc_utilisation <- function(
  resources, simulation_end_time, summarise = TRUE
) {
```

The function takes the resources dataframe from `get_mon_resources()`.

The `simulation_end_time` parameter will be used in the time-weighted calculations.

The `summarise` parameter controls whether to return the overall time-weighted utilisation (`TRUE`), or a detailed dataframe with per-interval metrics (`FALSE`).

<br>

```{.r}
# Calculate utilisation
util_df <- resources |>
  group_by(resource) |>
  mutate(
    ...
```

The code groups by resource, so we can calculate utilisation of each resource separately. The `mutate()` call will then add new columns to the dataframe...

<br>

```{.r}
    ...
    # Calculate time between this row and the next. For final row,
    # lead(time) returns NA, so use simulation_end_time - time
    interval_duration = ifelse(
      row_number() == n(),
      simulation_end_time - .data[["time"]],
      lead(.data[["time"]]) - .data[["time"]]
    ),
    ...
```

The `interval_duration` column computes how long each state lasts by taking the time difference between consecutive events. This is the basis for time-weighting.

<br>

```{.r}
    ...
    # Ensures effective capacity is never less than number of servers in
    # use (in case of situations where servers may exceed "capacity").
    effective_capacity = pmax(.data[["capacity"]], .data[["server"]]),
    ...
```

The `effective_capacity` ensures the denominator in utilisation calculations is always at least as large as the number of servers in use.

<br>

```{.r}
    ...
    # Divide number of servers in use by effective capacity
    # Set to NA if effective capacity is 0
    utilisation = ifelse(.data[["effective_capacity"]] > 0L,
                         .data[["server"]] / .data[["effective_capacity"]],
                         NA_real_)
  )
```

The `utilisation` column calculates, for each interval, the proportion of servers in use relative to available capacity. If capacity is zero, it returns NA.

<br>

```{.r}
# If summarise = TRUE, find total utilisation
if (summarise) {
  util_df |>
    summarise(
      # Multiply each utilisation by its own unique duration. The total of
      # those is then divided by the total duration of all intervals.
      # Hence, we are calculated a time-weighted average utilisation.
      utilisation = (sum(.data[["utilisation"]] *
                           .data[["interval_duration"]], na.rm = TRUE) /
                       sum(.data[["interval_duration"]], na.rm = TRUE))
    ) |>
    pivot_wider(names_from = "resource",
                values_from = "utilisation",
                names_glue = "utilisation_{resource}") |>
    ungroup()
```

If `summarise` is true, the final result is a time-weighted average: per-interval utilisations are weighted by their durations and averaged. Output is one value per resource.

<br>

```{.r}
} else {
  # If summarise = FALSE, just return the util_df with no further processing
  ungroup(util_df)
}
```

When `summarise` is false, detailed interval-level results are returned for further analysis or plotting.

### Run results function

In `get_run_results()`, the table `results[["resources"]]` is passed to `calc_utilisation()`.

We also add a new parameter `simulation_end_time`.

```{r}
#' Get average results for the provided single run.
#'
#' @param results Named list with `arrivals` from `get_mon_arrivals()` and
#'   `resources` from `get_mon_resources()`
#'   (`per_resource = TRUE` and `ongoing = TRUE`).
#' @param run_number Integer index of current simulation run.
#' @param simulation_end_time Time at end of simulation run.#<<
#'
#' @return Tibble with processed results from replication.
#' @export

get_run_results <- function(results, run_number, simulation_end_time) {#<<
  metrics <- list(
    calc_utilisation(results[["resources"]], simulation_end_time)#<<
  )
  dplyr::bind_cols(tibble(replication = run_number), metrics)
}
```

### Model function

We need to pass the full simulation run time (i.e. warm-up period + data collection period) to `get_run_results()` in `model()`. To avoid calculating this twice (for `simmer::run()` and `get_run_results()`), we create a new variable `full_run_length` and use that for both.

```{r}
#' Run simulation.
#'
#' @param param List. Model parameters.
#' @param run_number Numeric. Run number for random seed generation.
#'
#' @return Named list with tables `arrivals`, `resources` and `run_results`.
#' @export

model <- function(param, run_number) {

  # Set random seed based on run number
  set.seed(run_number)

  # Create simmer environment
  env <- simmer("simulation", verbose = param[["verbose"]])

  # Calculate full run length#<<
  full_run_length <- (param[["warm_up_period"]] + #<<
                        param[["data_collection_period"]])#<<

  # Define the patient trajectory
  patient <- trajectory("consultation") |>
    seize("doctor", 1L) |>
    timeout(function() {
      rexp(n = 1L, rate = 1L / param[["consultation_time"]])
    }) |>
    release("doctor", 1L)

  env <- env |>
    # Add doctor resource
    add_resource("doctor", param[["number_of_doctors"]]) |>
    # Add patient generator
    add_generator("patient", patient, function() {
      rexp(n = 1L, rate = 1L / param[["interarrival_time"]])
    }) |>
    # Run the simulation
    simmer::run(until = full_run_length)#<<

  # Extract information on arrivals and resources from simmer environment
  result <- list(
    arrivals = get_mon_arrivals(env, per_resource = TRUE, ongoing = TRUE),
    resources = get_mon_resources(env)
  )

  # Filter to remove results from the warm-up period
  result <- filter_warmup(result, param[["warm_up_period"]])

  # Replace "replication" value with appropriate run number
  result[["arrivals"]] <- mutate(result[["arrivals"]],
                                 replication = run_number)
  result[["resources"]] <- mutate(result[["resources"]],
                                  replication = run_number)

  # Calculate the average results for that run
  result[["run_results"]] <- get_run_results(
    result, run_number, simulation_end_time = full_run_length#<<
  )#<<

  result
}
```

### Run the model

```{r}
param <- create_params()
result <- model(param = param, run_number = 1L)
kable(result[["run_results"]])
```

<br>

We can also view the result with `summarise = FALSE`:

```{r}
util_df <- calc_utilisation(
  resources = result[["resources"]],
  summarise = FALSE,
  simulation_end_time = (param[["warm_up_period"]] +
                           param[["data_collection_period"]])
)
kable(util_df)
```

:::

::::

## Mean queue length

> **Corresponding term in queueing theory:** Average number in queue, $L_q$

:::: {.callout title="▶&nbsp;&nbsp;View how to record mean queue length" collapse="true"}

::: {.python-content}

<div class="h3-tight"></div>

### Patient class

No changes required.

```{python}
#| echo: false
{{< include outputs_resources/patient.py >}}
```

### Monitored resource class

We need the new `MonitoredResource` class (introduced and explained under time-weighted utilisation).

For queue length, the class is similar, except that we are monitoring `area_n_in_queue` (`len(self.queue) * time_since_last_event`) instead of `area_resource_busy`. These changes are highlighted.

```{python}
class MonitoredResource(simpy.Resource):
    """
    SimPy Resource subclass that tracks queue length.#<<

    Attributes
    ----------
    time_last_event : list
        Time of the most recent resource request or release event.
    area_n_in_queue : list#<<
        Cumulative time patients spent queueing for a resource, for#<<
        computing average queue length.#<<

    Notes
    -----
    Class adapted from Monks, Harper and Heather 2025.
    """

    def __init__(self, *args, **kwargs):
        """
        Initialise the MonitoredResource, resetting monitoring statistics.

        Parameters
        ----------
        *args :
            Positional arguments to be passed to the parent class.
        **kwargs :
            Keyword arguments to be passed to the parent class.
        """
        # Initialise a SimPy Resource
        super().__init__(*args, **kwargs)
        # Run the init_results() method
        self.init_results()

    def init_results(self):
        """
        Reset monitoring attributes.
        """
        self.time_last_event = [self._env.now]
        self.area_n_in_queue = [0.0]#<<

    def request(self, *args, **kwargs):
        """
        Update time-weighted statistics before requesting a resource.

        Parameters
        ----------
        *args :
            Positional arguments to be passed to the parent class.
        **kwargs :
            Keyword arguments to be passed to the parent class.

        Returns
        -------
        simpy.events.Event
            Event for the resource request.
        """
        # Update time-weighted statistics
        self.update_time_weighted_stats()
        # Request a resource
        return super().request(*args, **kwargs)

    def release(self, *args, **kwargs):
        """
        Update time-weighted statistics before releasing a resource.

        Parameters
        ----------
        *args :
            Positional arguments to be passed to the parent class.
        **kwargs :
            Keyword arguments to be passed to the parent class.

        Returns
        -------
        simpy.events.Event
            Event for the resource release.
        """
        # Update time-weighted statistics
        self.update_time_weighted_stats()
        # Release a resource
        return super().release(*args, **kwargs)

    def update_time_weighted_stats(self):
        """
        Update the time-weighted statistics for resource usage.

        Notes
        -----
        - These sums can be referred to as "the area under the curve".
        - They are called "time-weighted" statistics as they account for how
          long certain events or states persist over time.
        """
        # Calculate time since last event
        time_since_last_event = self._env.now - self.time_last_event[-1]

        # Add record of current time
        self.time_last_event.append(self._env.now)

        # Add "area under curve" of people in queue#<<
        # len(self.queue) is the number of requests queued#<<
        self.area_n_in_queue.append(len(self.queue) * time_since_last_event)#<<
```

### Model class

Same changes as made for time-weighted utilisation.

```{python}
#| echo: false
{{< include outputs_resources/model_monitored.py >}}
```

### Runner class

We've tracked the number of patients in the queue over time by summing `queue length x duration` across all intervals - this is the total "area under the curve" for the queue. We now divide this by the `data collection length` to obtain the average queue length experienced over the period.

For a clear visual explanation of what we mean by "area under the curve" and how this calculation works, see the [Time-weighted averages](#time-weighted-averages) section at the end of this page.

```{python}
class Runner:
    """
    Run the simulation.

    Attributes
    ----------
    param : Parameters
        Simulation parameters.
    """
    def __init__(self, param):
        """
        Initialise a new instance of the Runner class.

        Parameters
        ----------
        param : Parameters
            Simulation parameters.
        """
        self.param = param

    def run_single(self, run):
        """
        Runs the simulation once and performs results calculations.

        Parameters
        ----------
        run : int
            Run number for the simulation.

        Returns
        -------
        dict
            Contains patient-level results and results from each run.
        """
        model = Model(param=self.param, run_number=run)
        model.run()

        # Patient results
        patient_results = pd.DataFrame(model.results_list)
        patient_results["run"] = run

        # Run results
        run_results = {
            "run_number": run,
            "mean_queue_length": (#<<
                sum(model.doctor.area_n_in_queue) /#<<
                self.param.data_collection_period#<<
            )#<<
        }

        return {
            "patient": patient_results,
            "run": run_results
        }
```

### Run the model

```{python}
{{< include outputs_resources/run.py >}}
```

:::

::: {.r-content}

<div class="h3-tight"></div>

### Imports

We add:

* Functions from `dplyr`: `lead`, `left_join`, `row_number` and `summarise`.
* Functions from `simmer`: `get_queue_count` and `set_attribute`.
* Functions from `tidyr`: `pivot_wider`.

```{.r}
#' @keywords internal
"_PACKAGE"

## usethis namespace: start
#' @importFrom dplyr arrange bind_cols filter group_by lead left_join mutate n #<<
#' @importFrom dplyr row_number slice summarise ungroup #<<
#' @importFrom rlang .data
#' @importFrom simmer add_generator add_resource get_mon_arrivals
#' @importFrom simmer get_mon_resources get_queue_count release run seize #<<
#' @importFrom simmer set_attribute simmer timeout trajectory #<<
#' @importFrom tibble tibble
#' @importFrom tidyr pivot_wider #<<
## usethis namespace: end

NULL
```

### Queue length function

A new function `calc_mean_queue_length()` finds the **time-weighted** mean queue length for each resource. It does this by:

* Sorting the arrivals by start time.
* Calculating the duration until the next arrival for each patient.
* Multiplying each observed queue length by its corresponding interval duration ("time weighting").

The sum of these products is then divided by the total observed time to find the average queue length. This approach ensures that, for example, intervals where the queue was longer and persisted for more time have a greater effect on the mean than short-lived fluctuations.

For a clear visual explanation of how these calculations work, refer to the [Time-weighted averages](#time-weighted-averages) section below.

```{r}
#' Calculate the time-weighted mean queue length.
#'
#' @param arrivals Dataframe with times for each patient with each resource.
#' @param simulation_end_time Time at end of simulation run.
#'
#' @return Tibble with column containing mean queue length.
#' @export

calc_mean_queue_length <- function(arrivals, simulation_end_time) {
  arrivals |>
    group_by(.data[["resource"]]) |>
    # Sort by arrival time
    arrange(.data[["start_time"]]) |>
    # Calculate time between this row and the next. For final row,
    # lead(start_time) returns NA, so use simulation_end_time - start_time
    mutate(
      interval_duration = ifelse(
        row_number() == n(),
        simulation_end_time - .data[["start_time"]],
        lead(.data[["start_time"]]) - .data[["start_time"]]
      )
    ) |>
    # Multiply each queue length by its own unique duration. The total of
    # those is then divided by the total duration of all intervals.
    # Hence, we are calculated a time-weighted average queue length.
    summarise(mean_queue_length = (
      sum(.data[["queue_on_arrival"]] *
            .data[["interval_duration"]], na.rm = TRUE) /
        sum(.data[["interval_duration"]], na.rm = TRUE)
    )
    ) |>
    # Reshape dataframe
    pivot_wider(names_from = "resource",
                values_from = "mean_queue_length",
                names_glue = "mean_queue_length_{resource}") |>
    ungroup()
}
```

### Run results function

In `get_run_results()`, the table `results[["arrivals"]]` is passed to `calc_mean_queue_length()`. A new parameter `simulation_end_time` is add to `get_run_results()` (as likewise done for utilisation).

```{r}
#' Get average results for the provided single run.
#'
#' @param results Named list with `arrivals` from `get_mon_arrivals()` and
#'   `resources` from `get_mon_resources()`
#'   (`per_resource = TRUE` and `ongoing = TRUE`).
#' @param run_number Integer index of current simulation run.
#' @param simulation_end_time Time at end of simulation run.#<<
#'
#' @return Tibble with processed results from replication.
#' @export

get_run_results <- function(results, run_number, simulation_end_time) {#<<
  metrics <- list(
    calc_mean_queue_length(results[["arrivals"]], simulation_end_time)#<<
  )
  dplyr::bind_cols(tibble(replication = run_number), metrics)
}
```

### Model function

We need to record the length of the queue when each patient arrives (`doctor_queue_on_arrival`).

We can achieve this using `set_attribute()` and `get_queue_count()`.

However, add this attribute to our patient results, we need to make the same changes described above for mean wait time (and mean time in consultation): setting `mon = 2L` in the patient generator, retrieving results using `get_mon_attributes()`, and restructuring the returned dataframe to transform `doctor_serve_length` into `serve_length` (aligned with the corresponding doctor rows).

We also need to pass the full run time to `get_run_results()`, so calculate `full_run_length` (as likewise done for utilisation).

```{r}
#' Run simulation.
#'
#' @param param List. Model parameters.
#' @param run_number Numeric. Run number for random seed generation.
#'
#' @return Named list with tables `arrivals`, `resources` and `run_results`.
#' @export

model <- function(param, run_number) {

  # Set random seed based on run number
  set.seed(run_number)

  # Create simmer environment
  env <- simmer("simulation", verbose = param[["verbose"]])

  # Calculate full run length#<<
  full_run_length <- (param[["warm_up_period"]] + #<<
                        param[["data_collection_period"]])#<<

  # Define the patient trajectory
  patient <- trajectory("consultation") |>
    # Record queue length on arrival #<<
    set_attribute("doctor_queue_on_arrival", #<<
                  function() get_queue_count(env, "doctor")) |> #<<
    seize("doctor", 1L) |>
    timeout(function() {
      rexp(n = 1L, rate = 1L / param[["consultation_time"]])
    }) |>
    release("doctor", 1L)

  env <- env |>
    # Add doctor resource
    add_resource("doctor", param[["number_of_doctors"]]) |>
    # Add patient generator (mon=2 so can get manual attributes) #<<
    add_generator("patient", patient, function() {
      rexp(n = 1L, rate = 1L / param[["interarrival_time"]])
    }, mon = 2L) |> #<<
    # Run the simulation
    simmer::run(until = full_run_length)#<<

  # Extract information on arrivals and resources from simmer environment
  result <- list(
    arrivals = get_mon_arrivals(env, per_resource = TRUE, ongoing = TRUE),
    resources = get_mon_resources(env)
  )

  # Get the extra arrivals attributes#<<
  extra_attributes <- get_mon_attributes(env) |> #<<
    dplyr::select("name", "key", "value") |> #<<
    # Add column with resource name, and remove that from key #<<
    mutate(resource = gsub("_.+", "", .data[["key"]]), #<<
           key = gsub("^[^_]+_", "", .data[["key"]])) |> #<<
    pivot_wider(names_from = "key", values_from = "value") #<<

  # Merge extra attributes with the arrival data#<<
  result[["arrivals"]] <- left_join( #<<
    result[["arrivals"]], extra_attributes, by = c("name", "resource") #<<
  ) #<<

  # Filter to remove results from the warm-up period
  result <- filter_warmup(result, param[["warm_up_period"]])

  # Replace "replication" value with appropriate run number
  result[["arrivals"]] <- mutate(result[["arrivals"]],
                                 replication = run_number)
  result[["resources"]] <- mutate(result[["resources"]],
                                  replication = run_number)

  # Calculate the average results for that run
  result[["run_results"]] <- get_run_results(
    result, run_number, simulation_end_time = full_run_length#<<
  )#<<

  result
}
```

### Run the model

```{r}
param <- create_params()
result <- model(param = param, run_number = 1L)
kable(arrange(result[["arrivals"]], start_time))
kable(result[["run_results"]])
```

:::

::::

## Mean time in system

> **Corresponding term in queueing theory:** Average time in system, $\\W$

:::: {.callout title="▶&nbsp;&nbsp;View how to record mean time in system" collapse="true"}

::: {.python-content}

<div class="h3-tight"></div>

### Patient class

We create a new attribute for recording the `end_time` of each patient.

As it is an attribute, we have to give `end_time` an initial value. A value like 0 would wrongly look like "departure at time 0", so `np.nan` is used instead to clearly mean "we don't yet know when this patient will depart".

```{python}
class Patient:
    """
    Represents a patient.

    Attributes
    ----------
    patient_id : int
        Unique patient identifier.
    period : str
        Arrival period (warm up or data collection) with emoji.
    arrival_time : float
        Time patient entered the system (minutes).
    end_time : float#<<
        Time that patient leaves (minutes), or NaN if remain in system.#<<
    """
    def __init__(self, patient_id, period, arrival_time):
        """
        Initialises a new patient.

        Parameters
        ----------
        patient_id : int
            Unique patient identifier.
        period : str
            Arrival period (warm up or data collection) with emoji.
        arrival_time : float
            Time patient entered the system (minutes).
        """
        self.patient_id = patient_id
        self.period = period
        self.arrival_time = arrival_time
        self.end_time = np.nan#<<
```

### Model class

For each patient, the time at the end of their consultation is recorded as their `end_time`.

```{python}
class Model:
    """
    Simulation model.

    Attributes
    ----------
    param : Parameters
        Simulation parameters.
    run_number : int
        Run number for random seed generation.
    env : simpy.Environment
        The SimPy environment for the simulation.
    doctor : simpy.Resource
        SimPy resource representing doctors.
    patients : list
        List of Patient objects.
    results_list : list
        List of dictionaries with the attributes of each patient.
    arrival_dist : Exponential
        Distribution used to generate random patient inter-arrival times.
    consult_dist : Exponential
        Distribution used to generate length of a doctor's consultation.
    """
    def __init__(self, param, run_number):
        """
        Create a new Model instance.

        Parameters
        ----------
        param : Parameters
            Simulation parameters.
        run_number : int
            Run number for random seed generation.
        """
        self.param = param
        self.run_number = run_number

        # Create SimPy environment
        self.env = simpy.Environment()

        # Create resource
        self.doctor = simpy.Resource(
            self.env, capacity=self.param.number_of_doctors
        )

        # Create a random seed sequence based on the run number
        ss = np.random.SeedSequence(self.run_number)
        seeds = ss.spawn(2)

        # Set up attributes to store results
        self.patients = []
        self.results_list = []

        # Initialise distributions
        self.arrival_dist = Exponential(mean=self.param.interarrival_time,
                                        random_seed=seeds[0])
        self.consult_dist = Exponential(mean=self.param.consultation_time,
                                        random_seed=seeds[1])

    def generate_arrivals(self):
        """
        Process that generates patient arrivals.
        """
        while True:
            # Sample and pass time to next arrival
            sampled_iat = self.arrival_dist.sample()
            yield self.env.timeout(sampled_iat)

            # Check whether arrived during warm-up or data collection
            if self.env.now < self.param.warm_up_period:
                period = "\U0001F538 WU"
            else:
                period = "\U0001F539 DC"

            # Create a new patient
            patient = Patient(patient_id=len(self.patients)+1,
                              period=period,
                              arrival_time=self.env.now)
            self.patients.append(patient)

            # Print the arrival time
            if self.param.verbose:
                print(f"{patient.period} Patient {patient.patient_id} " +
                      f"arrives at time: {patient.arrival_time:.3f}")

            # Start process of consultation
            self.env.process(self.consultation(patient))

    def consultation(self, patient):
        """
        Process that simulates a consultation.

        Parameters
        ----------
        patient :
            Instance of the Patient() class representing a single patient.
        """
        # Patient requests access to a doctor (resource)
        with self.doctor.request() as req:
            yield req

            if self.param.verbose:
                print(f"{patient.period} Patient {patient.patient_id} " +
                      f"starts consultation at: {self.env.now:.3f}")

            # Sample consultation duration and pass time spent with doctor
            time_with_doctor = self.consult_dist.sample()
            yield self.env.timeout(time_with_doctor)

            # Record end time#<<
            patient.end_time = self.env.now#<<
            if self.param.verbose:
                print(f"{patient.period} Patient {patient.patient_id} " +
                      f"leaves at: {patient.end_time:.3f}")#<<

    def reset_results(self):
        """
        Reset results.
        """
        self.patients = []

    def warmup(self):
        """
        Reset result collection after the warm-up period.
        """
        if self.param.warm_up_period > 0:
            # Delay process until warm-up period has completed
            yield self.env.timeout(self.param.warm_up_period)
            # Reset results variables
            self.reset_results()
            if self.param.verbose:
                print(f"Warm up period ended at time: {self.env.now}")

    def run(self):
        """
        Run the simulation.
        """
        # Schedule arrival generator and warm-up
        self.env.process(self.generate_arrivals())
        self.env.process(self.warmup())

        # Run the simulation
        self.env.run(until=(self.param.warm_up_period +
                            self.param.data_collection_period))

        # Create list of dictionaries containing each patient's attributes
        self.results_list = [x.__dict__ for x in self.patients]
```

### Runner class

The time each patient spent in the system is calculated from `patient_results["end_time"] - patient_results["arrival_time"]`. Backlogged patients will not have an end time so will automatically set to NaN.

```{python}
class Runner:
    """
    Run the simulation.

    Attributes
    ----------
    param : Parameters
        Simulation parameters.
    """
    def __init__(self, param):
        """
        Initialise a new instance of the Runner class.

        Parameters
        ----------
        param : Parameters
            Simulation parameters.
        """
        self.param = param

    def run_single(self, run):
        """
        Runs the simulation once and performs results calculations.

        Parameters
        ----------
        run : int
            Run number for the simulation.

        Returns
        -------
        dict
            Contains patient-level results and results from each run.
        """
        model = Model(param=self.param, run_number=run)
        model.run()

        # Patient results
        patient_results = pd.DataFrame(model.results_list)
        patient_results["run"] = run
        patient_results["time_in_system"] = (#<<
            patient_results["end_time"] - patient_results["arrival_time"]#<<
        )#<<

        # Run results
        run_results = {
            "run_number": run,
            "mean_time_in_system": patient_results["time_in_system"].mean()#<<
        }

        return {
            "patient": patient_results,
            "run": run_results
        }
```

### Run the model

```{python}
{{< include outputs_resources/run.py >}}
```

*The mean time in system is returned as `np.float64()` because NumPy's `mean()` function returns a NumPy float type instead of a built-in Python `float`. You can just treat it like a regular `float`, or convert it with `float()` if you want it to print without `np.float64()`.*

:::

::: {.r-content}

<div class="h3-tight"></div>

### Imports

We only need to add `dplyr::summarise()`.

```{.r}
#' @keywords internal
"_PACKAGE"

## usethis namespace: start
#' @importFrom dplyr arrange bind_cols filter group_by mutate n slice summarise #<<
#' @importFrom dplyr ungroup
#' @importFrom rlang .data
#' @importFrom simmer add_generator add_resource get_mon_arrivals
#' @importFrom simmer get_mon_resources release run seize simmer
#' @importFrom simmer timeout trajectory
#' @importFrom tibble tibble
## usethis namespace: end

NULL
```

### Time in system function

The `calc_mean_time_in_system()` function finds the mean of a new column `time_in_system`.

```{r}
#' Calculate the mean time in the system for finished patients.
#'
#' @param arrivals Dataframe with times for each patient with each resource.
#'
#' @return Tibble with column containing the mean time in the system.
#' @export

calc_mean_time_in_system <- function(arrivals) {
  arrivals |>
    summarise(
      mean_time_in_system = mean(.data[["time_in_system"]], na.rm = TRUE)
    ) |>
    ungroup()
}
```

### Run results function

In `get_run_results()`, the table `results[["arrivals"]]` is passed to `calc_mean_time_in_system()`.

```{r}
#' Get average results for the provided single run.
#'
#' @param results Named list with `arrivals` from `get_mon_arrivals()` and
#'   `resources` from `get_mon_resources()`
#'   (`per_resource = TRUE` and `ongoing = TRUE`).
#' @param run_number Integer index of current simulation run.
#'
#' @return Tibble with processed results from replication.
#' @export

get_run_results <- function(results, run_number) {
  metrics <- list(
    calc_mean_time_in_system(results[["arrivals"]])#<<
  )
  dplyr::bind_cols(tibble(replication = run_number), metrics)
}
```

### Model function

In `model()` we create the new `time_in_system` column by finding the difference between each patient's start and end time. This metric excludes unfinished patients - their `time_in_system` will set to `NaN` by default as they will have no recorded end time.

```{r}
#' Run simulation.
#'
#' @param param List. Model parameters.
#' @param run_number Numeric. Run number for random seed generation.
#'
#' @return Named list with tables `arrivals`, `resources` and `run_results`.
#' @export

model <- function(param, run_number) {

  # Set random seed based on run number
  set.seed(run_number)

  # Create simmer environment
  env <- simmer("simulation", verbose = param[["verbose"]])

  # Define the patient trajectory
  patient <- trajectory("consultation") |>
    seize("doctor", 1L) |>
    timeout(function() {
      rexp(n = 1L, rate = 1L / param[["consultation_time"]])
    }) |>
    release("doctor", 1L)

  env <- env |>
    # Add doctor resource
    add_resource("doctor", param[["number_of_doctors"]]) |>
    # Add patient generator
    add_generator("patient", patient, function() {
      rexp(n = 1L, rate = 1L / param[["interarrival_time"]])
    }) |>
    # Run the simulation
    simmer::run(until = (param[["warm_up_period"]] +
                           param[["data_collection_period"]]))

  # Extract information on arrivals and resources from simmer environment
  result <- list(
    arrivals = get_mon_arrivals(env, per_resource = TRUE, ongoing = TRUE),
    resources = get_mon_resources(env)
  )

  # Add time in system (unfinished patients will set to NaN) #<<
  result[["arrivals"]][["time_in_system"]] <- ( #<<
    result[["arrivals"]][["end_time"]] - #<<
      result[["arrivals"]][["start_time"]] #<<
  ) #<<

  # Filter to remove results from the warm-up period
  result <- filter_warmup(result, param[["warm_up_period"]])

  # Replace "replication" value with appropriate run number
  result[["arrivals"]] <- mutate(result[["arrivals"]],
                                 replication = run_number)
  result[["resources"]] <- mutate(result[["resources"]],
                                  replication = run_number)

  # Calculate the average results for that run
  result[["run_results"]] <- get_run_results(result, run_number)

  result
}
```

### Run the model

```{r}
param <- create_params()
result <- model(param = param, run_number = 1L)
kable(arrange(result[["arrivals"]], start_time))
kable(result[["run_results"]])
```

:::

::::

## Mean number of patients in the system

> **Corresponding term in queueing theory:** Mean system size or average number in the system, $L$

:::: {.callout title="▶&nbsp;&nbsp;View how to record mean number of patients in the system" collapse="true"}

::: {.python-content}

This is a time-weighted calculation, but unlike utilisation and queue length, we record this within `Model` itself (and not with an additional `MonitoredResource` class).

### Patient class

No changes required.

```{python}
#| echo: false
{{< include outputs_resources/patient.py >}}
```

### Model class

We introduce three new attributes:

* `area_n_in_system` - a list storing the area under the curve for the number of patients in the system over time.
* `time_last_n_in_system` - records last time the number-in-system statistic was updated.
* `n_in_system` - maintains current count of patients (whether waiting or being served)

When a patient arrives or leaves, we call a new `update_n_in_system(inc)` method. This:

1. Calculates the time since the last update, then appends `n_in_system x duration` to `area_n_in_system`.
2. Then updates `time_last_n_insystem` and `n_in_system` accordingly.

Warmup logic resets only the necessary area and timing components, preserving the correct patient count for accurate post-warmup statistics.

At the simulation end, a final update with `inc=0` ensures statistics reflect the last interval up to the end of the run.

```{python}
class Model:
    """
    Simulation model.

    Attributes
    ----------
    param : Parameters
        Simulation parameters.
    run_number : int
        Run number for random seed generation.
    env : simpy.Environment
        The SimPy environment for the simulation.
    doctor : simpy.Resource
        SimPy resource representing doctors.
    patients : list
        List of Patient objects.
    results_list : list
        List of dictionaries with the attributes of each patient.
    area_n_in_system : list of float#<<
        List containing incremental area contributions used for#<<
        time-weighted statistics of the number of patients in the system.#<<
    time_last_n_in_system : float#<<
        Simulation time at last update of the number-in-system statistic.#<<
    n_in_system: int#<<
        Current number of patients present in the system, including#<<
        waiting and being served.#<<
    arrival_dist : Exponential
        Distribution used to generate random patient inter-arrival times.
    consult_dist : Exponential
        Distribution used to generate length of a doctor's consultation.
    """
    def __init__(self, param, run_number):
        """
        Create a new Model instance.

        Parameters
        ----------
        param : Parameters
            Simulation parameters.
        run_number : int
            Run number for random seed generation.
        """
        self.param = param
        self.run_number = run_number

        # Create SimPy environment
        self.env = simpy.Environment()

        # Create resource
        self.doctor = simpy.Resource(
            self.env, capacity=self.param.number_of_doctors
        )

        # Create a random seed sequence based on the run number
        ss = np.random.SeedSequence(self.run_number)
        seeds = ss.spawn(2)

        # Set up attributes to store results
        self.patients = []
        self.results_list = []
        self.area_n_in_system = [0]#<<
        self.time_last_n_in_system = self.env.now#<<
        self.n_in_system = 0#<<

        # Initialise distributions
        self.arrival_dist = Exponential(mean=self.param.interarrival_time,
                                        random_seed=seeds[0])
        self.consult_dist = Exponential(mean=self.param.consultation_time,
                                        random_seed=seeds[1])

    def update_n_in_system(self, inc):#<<
        """#<<
        Update the time-weighted statistics for number of patients in system.#<<

        Parameters#<<
        ----------#<<
        inc : int#<<
            Change in the number of patients (+1, 0, -1).#<<
        """#<<
        # Compute time since last event and calculate area under curve for that#<<
        duration = self.env.now - self.time_last_n_in_system#<<
        self.area_n_in_system.append(self.n_in_system * duration)#<<
        # Update time and n in system#<<
        self.time_last_n_in_system = self.env.now#<<
        self.n_in_system += inc#<<

    def generate_arrivals(self):
        """
        Process that generates patient arrivals.
        """
        while True:
            # Sample and pass time to next arrival
            sampled_iat = self.arrival_dist.sample()
            yield self.env.timeout(sampled_iat)

            # Check whether arrived during warm-up or data collection
            if self.env.now < self.param.warm_up_period:
                period = "\U0001F538 WU"
            else:
                period = "\U0001F539 DC"

            # Create a new patient
            patient = Patient(patient_id=len(self.patients)+1,
                              period=period,
                              arrival_time=self.env.now)
            self.patients.append(patient)

            # Update the number in the system#<<
            self.update_n_in_system(inc=1)#<<

            # Print arrival time
            if self.param.verbose:
                print(f"{patient.period} Patient {patient.patient_id} " +
                      f"arrives at time: {patient.arrival_time:.3f}")

            # Start process of consultation
            self.env.process(self.consultation(patient))

    def consultation(self, patient):
        """
        Process that simulates a consultation.

        Parameters
        ----------
        patient :
            Instance of the Patient() class representing a single patient.
        """
        # Patient requests access to a doctor (resource)
        with self.doctor.request() as req:
            yield req

            if self.param.verbose:
                print(f"{patient.period} Patient {patient.patient_id} " +
                      f"starts consultation at: {self.env.now:.3f}")

            # Sample consultation duration and pass time spent with doctor
            time_with_doctor = self.consult_dist.sample()
            yield self.env.timeout(time_with_doctor)

            # Update number in system#<<
            self.update_n_in_system(inc=-1)#<<

            # Record end time
            if self.param.verbose:
                print(f"{patient.period} Patient {patient.patient_id} " +
                      f"leaves at: {self.env.now:.3f}")

    def reset_results(self):
        """
        Reset results.
        """
        self.patients = []
        # For number in system, we reset area and time but not the count, as#<<
        # it should include any remaining warm-up patients in the count#<<
        self.area_n_in_system = [0]#<<
        self.time_last_n_in_system = self.env.now#<<

    def warmup(self):
        """
        Reset result collection after the warm-up period.
        """
        if self.param.warm_up_period > 0:
            # Delay process until warm-up period has completed
            yield self.env.timeout(self.param.warm_up_period)
            # Reset results variables
            self.reset_results()
            if self.param.verbose:
                print(f"Warm up period ended at time: {self.env.now}")

    def run(self):
        """
        Run the simulation.
        """
        # Schedule arrival generator and warm-up
        self.env.process(self.generate_arrivals())
        self.env.process(self.warmup())

        # Run the simulation
        self.env.run(until=(self.param.warm_up_period +
                            self.param.data_collection_period))

        # Run final calculation of number in system#<<
        self.update_n_in_system(inc=0)#<<

        # Create list of dictionaries containing each patient's attributes
        self.results_list = [x.__dict__ for x in self.patients]
```

### Runner class

We've tracked how many patients are in the system (waiting or in consultation) by summing `patients in system x duration` across all intervals - the "area under the curve" for system size. Dividing this by the `data collection length` gives the average number of patients in the system over time.

For a clear visual explanation of what we mean by "area under the curve" and how this calculation works, see the [Time-weighted averages](#time-weighted-averages) section at the end of this page.

```{python}
class Runner:
    """
    Run the simulation.

    Attributes
    ----------
    param : Parameters
        Simulation parameters.
    """
    def __init__(self, param):
        """
        Initialise a new instance of the Runner class.

        Parameters
        ----------
        param : Parameters
            Simulation parameters.
        """
        self.param = param

    def run_single(self, run):
        """
        Runs the simulation once and performs results calculations.

        Parameters
        ----------
        run : int
            Run number for the simulation.

        Returns
        -------
        dict
            Contains patient-level results and results from each run.
        """
        model = Model(param=self.param, run_number=run)
        model.run()

        # Patient results
        patient_results = pd.DataFrame(model.results_list)
        patient_results["run"] = run

        # Run results
        run_results = {
            "run_number": run,
            "mean_patients_in_system": (#<<
                sum(model.area_n_in_system) /#<<
                self.param.data_collection_period#<<
            )#<<
        }

        return {
            "patient": patient_results,
            "run": run_results
        }
```

### Run the model

```{python}
{{< include outputs_resources/run.py >}}
```

:::

::: {.r-content}

<div class="h3-tight"></div>

### Imports

We add:

* Functions from `dplyr`: `bind_rows`, `desc`, `lead`, `row_number`, `summarise` and `transmute`.
* Imports from new packages: `tidyr` and `tidyselect`.

```{.r}
#' @keywords internal
"_PACKAGE"

## usethis namespace: start
#' @importFrom dplyr arrange bind_cols bind_rows desc filter group_by lead #<<
#' @importFrom dplyr mutate n row_number slice summarise transmute ungroup #<<
#' @importFrom rlang .data
#' @importFrom simmer add_generator add_resource get_mon_arrivals
#' @importFrom simmer get_mon_resources release run seize simmer
#' @importFrom simmer timeout trajectory
#' @importFrom tibble tibble
#' @importFrom tidyr drop_na #<<
#' @importFrom tidyselect all_of #<<
## usethis namespace: end

NULL
```

### Patients in system function

The function `calc_mean_patients_in_system()` does not use the arrivals or resources dataframes - instead, it uses a new dataframe which contains patient counts over time.

It is a **time-weighted calculation** which multiplies the patient count in each time interval by the interval duration. The sum of these is then divided by the total simulation time to find the average system occupancy.

For a clear visual explanation of how these calculations work, refer to the [Time-weighted averages](#time-weighted-averages) section below.

```{r}
#' Calculate the time-weighted mean number of patients in the system.
#'
#' @param patient_count Dataframe with patient counts over time.
#' @param simulation_end_time Time at end of simulation run.
#'
#' @return Tibble with column containing mean number of patients in the system.
#' @export

calc_mean_patients_in_system <- function(patient_count, simulation_end_time) {
  patient_count |>
    # Sort by time
    arrange(.data[["time"]]) |>
    # Calculate time between this row and the next. For final row,
    # lead(time) returns NA, so use simulation_end_time - time
    mutate(
      interval_duration = ifelse(
        row_number() == n(),
        simulation_end_time - .data[["time"]],
        lead(.data[["time"]]) - .data[["time"]]
      )
    ) |>
    # Multiply each patient count by its own unique duration. The total of
    # those is then divided by the total duration of all intervals.
    # Hence, we are calculated a time-weighted average patient count.
    summarise(
      mean_patients_in_system = (
        sum(.data[["count"]] * .data[["interval_duration"]], na.rm = TRUE) /
          sum(.data[["interval_duration"]], na.rm = TRUE)
      )
    ) |>
    ungroup()
}
```

### Run results function

In `get_run_results()`, the table `results[["patients_in_system"]]` is passed to `calc_mean_patients_in_system()`. We also add a `simulation_end_time` parameter (as likewise done for utilisation and queue length).

```{r}
#' Get average results for the provided single run.
#'
#' @param results Named list with `arrivals` from `get_mon_arrivals()` and
#'   `resources` from `get_mon_resources()`
#'   (`per_resource = TRUE` and `ongoing = TRUE`).
#' @param run_number Integer index of current simulation run.
#' @param simulation_end_time Time at end of simulation run.#<<
#'
#' @return Tibble with processed results from replication.
#' @export

get_run_results <- function(results, run_number, simulation_end_time) {#<<
  metrics <- list(
    calc_mean_patients_in_system(results[["patients_in_system"]], #<<
                                 simulation_end_time)#<<
  )
  dplyr::bind_cols(tibble(replication = run_number), metrics)
}
```

### Model function

In `model()`, we create a new dataframe that gives the count of patients in the system at every time there was an arrival or exit.

To do this, we first make two dataframes: one recording arrival times (`arrivals_start`) and another recording exit times (`arrivals_end`). Each event is tagged as either a +1 or -1 change.

These event records are then combined into a single dataframe and sorted, allowing calculation of the cumulative sum of the +1 and -1 values to give us the patient count over time.

We also add a column with the `run_number` (so - if running multiple replications later - the results can be linked to the relevant run).

We also need to pass the full run time to `get_run_results()`, so calculate `full_run_length` (as likewise done for utilisation and queue length).

```{r}
#' Run simulation.
#'
#' @param param List. Model parameters.
#' @param run_number Numeric. Run number for random seed generation.
#'
#' @return Named list with tables `arrivals`, `resources` and `run_results`.
#' @export

model <- function(param, run_number) {

  # Set random seed based on run number
  set.seed(run_number)

  # Create simmer environment
  env <- simmer("simulation", verbose = param[["verbose"]])

  # Calculate full run length#<<
  full_run_length <- (param[["warm_up_period"]] + #<<
                        param[["data_collection_period"]])#<<

  # Define the patient trajectory
  patient <- trajectory("consultation") |>
    seize("doctor", 1L) |>
    timeout(function() {
      rexp(n = 1L, rate = 1L / param[["consultation_time"]])
    }) |>
    release("doctor", 1L)

  env <- env |>
    # Add doctor resource
    add_resource("doctor", param[["number_of_doctors"]]) |>
    # Add patient generator
    add_generator("patient", patient, function() {
      rexp(n = 1L, rate = 1L / param[["interarrival_time"]])
    }) |>
    # Run the simulation
    simmer::run(until = full_run_length)#<<

  # Extract information on arrivals and resources from simmer environment
  result <- list(
    arrivals = get_mon_arrivals(env, per_resource = TRUE, ongoing = TRUE),
    resources = get_mon_resources(env)
  )

  # Filter to remove results from the warm-up period
  result <- filter_warmup(result, param[["warm_up_period"]])

  # Gather all start and end times, with a row for each, marked +1 or -1 #<<
  # Drop NA for end time, as those are patients who haven't left system #<<
  # at the end of the simulation #<<
  arrivals_start <- transmute( #<<
    result[["arrivals"]], time = .data[["start_time"]], change = 1L #<<
  ) #<<
  arrivals_end <- result[["arrivals"]] |> #<<
    drop_na(all_of("end_time")) |> #<<
    transmute(time = .data[["end_time"]], change = -1L) #<<
  events <- bind_rows(arrivals_start, arrivals_end) #<<

  # Determine the count of patients in the service with each entry/exit #<<
  result[["patients_in_system"]] <- events |> #<<
    # Sort events by time #<<
    arrange(.data[["time"]], desc(.data[["change"]])) |> #<<
    # Use cumulative sum to find number of patients in system at each time #<<
    mutate(count = cumsum(.data[["change"]])) |> #<<
    dplyr::select(c("time", "count")) #<<

  # Replace "replication" value with appropriate run number
  result[["arrivals"]] <- mutate(result[["arrivals"]],
                                 replication = run_number)
  result[["resources"]] <- mutate(result[["resources"]],
                                  replication = run_number)
  result[["patients_in_system"]] <- mutate(result[["patients_in_system"]], #<<
                                           replication = run_number) #<<

  # Calculate the average results for that run
  result[["run_results"]] <- get_run_results(
    result, run_number, simulation_end_time = full_run_length#<<
  )#<<

  result
}
```

### Run the model

Here, we see the new `result[["patient_in_system"]]` dataframe, as well as the result for `mean_patients_in_system` in `result[["run_results"]]`.

```{r}
param <- create_params()
result <- model(param = param, run_number = 1L)
kable(result[["patients_in_system"]])
kable(result[["run_results"]])
```

:::

::::

## Unseen (backlogged) patients

Unseen patients are those **still waiting in the queue**, having not started their consultation when the simulation ends.

Measuring them is important, otherwise **system congestion** and long wait times may be missed. Their wait time is not included with completed consultations because these waits are **censored** - i.e., the full duration is unknown until consultation begins. Tracking them separately avoids distorting the calculated averages, whilst accurately reflecting **system backlog** at simulation end.

:::: {.python-content}

### Backlogged patient count

:::: {.callout title="▶&nbsp;&nbsp;View how to record the backlogged patient count" collapse="true"}

#### Patient class

This requires the same change as for mean time in consultation - adding an attribute `time_with_doctor`.

```{python}
{{< include outputs_resources/patient_timewithdoctor.py >}}
```

#### Model class

Again, we make the same change as for mean time in consultation - recording the sampled time with the doctor as `patient.time_with_doctor` during `Model.consultation()`.

```{python}
{{< include outputs_resources/model_timewithdoctor.py >}}
```

#### Runner class

We can count the number of unseen patients by finding how many have `np.nan` for `time_with_doctor`.

```{python}
class Runner:
    """
    Run the simulation.

    Attributes
    ----------
    param : Parameters
        Simulation parameters.
    """
    def __init__(self, param):
        """
        Initialise a new instance of the Runner class.

        Parameters
        ----------
        param : Parameters
            Simulation parameters.
        """
        self.param = param

    def run_single(self, run):
        """
        Runs the simulation once and performs results calculations.

        Parameters
        ----------
        run : int
            Run number for the simulation.

        Returns
        -------
        dict
            Contains patient-level results and results from each run.
        """
        model = Model(param=self.param, run_number=run)
        model.run()

        # Patient results
        patient_results = pd.DataFrame(model.results_list)
        patient_results["run"] = run

        # Run results
        run_results = {
            "run_number": run,
            "unseen_count": patient_results["time_with_doctor"].isna().sum()#<<
        }

        return {
            "patient": patient_results,
            "run": run_results
        }
```

#### Run the model

In `result["patient"]` we can see one person had NaN for `time_with_doctor`. Correspondingly, the `unseen_count` in `result["run"]` is 1.

```{python}
{{< include outputs_resources/run.py >}}
```

*The count is returned as `np.int64()` because NumPy's `.isna().sum()` functions return a NumPy int type instead of a built-in Python `int`. You can just treat it like a regular `int`, or convert it with `int()` if you want it to print without `np.int64()`.*

:::

### Backlogged patient mean wait time

::: {.callout title="▶&nbsp;&nbsp;View how to record the backlogged patient mean wait time" collapse="true"}

#### Patient class

As done in mean time in consultation and unseen patient count, we add a new attribute `time_with_doctor`.

As it is an attribute, we have to give `time_with_doctor` an initial value. A value like 0 would wrongly look like "patient spent 0 minutes with the doctor", so `np.nan` is used instead to clearly mean "we don't yet know how long this patient will spend with the doctor".

```{python}
class Patient:
    """
    Represents a patient.

    Attributes
    ----------
    patient_id : int
        Unique patient identifier.
    period : str
        Arrival period (warm up or data collection) with emoji.
    arrival_time : float
        Time patient entered the system (minutes).
    time_with_doctor : float#<<
        Time spent in consultation with a doctor (minutes).#<<
    """
    def __init__(self, patient_id, period, arrival_time):
        """
        Initialises a new patient.

        Parameters
        ----------
        patient_id : int
            Unique patient identifier.
        period : str
            Arrival period (warm up or data collection) with emoji.
        arrival_time : float
            Time patient entered the system (minutes).
        """
        self.patient_id = patient_id
        self.period = period
        self.arrival_time = arrival_time
        self.time_with_doctor = np.nan#<<
```

#### Model class

We also repeat the change to `Model` - recording the sampled time with doctor as `patient.time_with_doctor` during `Model.consultation()`.

```{python}
class Model:
    """
    Simulation model.

    Attributes
    ----------
    param : Parameters
        Simulation parameters.
    run_number : int
        Run number for random seed generation.
    env : simpy.Environment
        The SimPy environment for the simulation.
    doctor : simpy.Resource
        SimPy resource representing doctors.
    patients : list
        List of Patient objects.
    results_list : list
        List of dictionaries with the attributes of each patient.
    arrival_dist : Exponential
        Distribution used to generate random patient inter-arrival times.
    consult_dist : Exponential
        Distribution used to generate length of a doctor's consultation.
    """
    def __init__(self, param, run_number):
        """
        Create a new Model instance.

        Parameters
        ----------
        param : Parameters
            Simulation parameters.
        run_number : int
            Run number for random seed generation.
        """
        self.param = param
        self.run_number = run_number

        # Create SimPy environment
        self.env = simpy.Environment()

        # Create resource
        self.doctor = simpy.Resource(
            self.env, capacity=self.param.number_of_doctors
        )

        # Create a random seed sequence based on the run number
        ss = np.random.SeedSequence(self.run_number)
        seeds = ss.spawn(2)

        # Set up attributes to store results
        self.patients = []
        self.results_list = []

        # Initialise distributions
        self.arrival_dist = Exponential(mean=self.param.interarrival_time,
                                        random_seed=seeds[0])
        self.consult_dist = Exponential(mean=self.param.consultation_time,
                                        random_seed=seeds[1])

    def generate_arrivals(self):
        """
        Process that generates patient arrivals.
        """
        while True:
            # Sample and pass time to next arrival
            sampled_iat = self.arrival_dist.sample()
            yield self.env.timeout(sampled_iat)

            # Check whether arrived during warm-up or data collection
            if self.env.now < self.param.warm_up_period:
                period = "\U0001F538 WU"
            else:
                period = "\U0001F539 DC"

            # Create a new patient
            patient = Patient(patient_id=len(self.patients)+1,
                              period=period,
                              arrival_time=self.env.now)
            self.patients.append(patient)

            # Record and print the arrival time
            patient.arrival_time = self.env.now
            if self.param.verbose:
                print(f"{patient.period} Patient {patient.patient_id} " +
                      f"arrives at time: {patient.arrival_time:.3f}")

            # Start process of consultation
            self.env.process(self.consultation(patient))

    def consultation(self, patient):
        """
        Process that simulates a consultation.

        Parameters
        ----------
        patient :
            Instance of the Patient() class representing a single patient.
        """
        # Patient requests access to a doctor (resource)
        with self.doctor.request() as req:
            yield req

            if self.param.verbose:
                print(f"{patient.period} Patient {patient.patient_id} starts" +
                      f" consultation at: {self.env.now:.3f}")

            # Sample consultation duration and pass time spent with doctor
            patient.time_with_doctor = self.consult_dist.sample()#<<
            yield self.env.timeout(patient.time_with_doctor)#<<

            # Record end time
            if self.param.verbose:
                print(f"{patient.period} Patient {patient.patient_id} " +
                      f"leaves at: {self.env.now:.3f}")

    def reset_results(self):
        """
        Reset results.
        """
        self.patients = []

    def warmup(self):
        """
        Reset result collection after the warm-up period.
        """
        if self.param.warm_up_period > 0:
            # Delay process until warm-up period has completed
            yield self.env.timeout(self.param.warm_up_period)
            # Reset results variables
            self.reset_results()
            if self.param.verbose:
                print(f"Warm up period ended at time: {self.env.now}")

    def run(self):
        """
        Run the simulation.
        """
        # Schedule arrival generator and warm-up
        self.env.process(self.generate_arrivals())
        self.env.process(self.warmup())

        # Run the simulation
        self.env.run(until=(self.param.warm_up_period +
                            self.param.data_collection_period))

        # Create list of dictionaries containing each patient's attributes
        self.results_list = [x.__dict__ for x in self.patients]
```

#### Runner class

After the simulation finishes, we check whether patients have been seen yet by a doctor by looking for missing values in the `time_with_doctor` column.

For those patients, their waiting time (`unseen_wait_time`) is calculated as the difference between the current simulation time and their recorded arrival time. We can then find the mean of this column.

Patients who have already seen a doctor have their `unseen_wait_time` set to `NaN`, indicating that it does not apply to them.

```{python}
class Runner:
    """
    Run the simulation.

    Attributes
    ----------
    param : Parameters
        Simulation parameters.
    """
    def __init__(self, param):
        """
        Initialise a new instance of the Runner class.

        Parameters
        ----------
        param : Parameters
            Simulation parameters.
        """
        self.param = param

    def run_single(self, run):
        """
        Runs the simulation once and performs results calculations.

        Parameters
        ----------
        run : int
            Run number for the simulation.

        Returns
        -------
        dict
            Contains patient-level results and results from each run.
        """
        model = Model(param=self.param, run_number=run)
        model.run()

        # Patient results
        patient_results = pd.DataFrame(model.results_list)
        patient_results["run"] = run

        # For each patient, if they haven't seen a doctor yet, calculate#<<
        # their wait as current time minus arrival, else set as missing#<<
        patient_results["unseen_wait_time"] = np.where(#<<
            patient_results["time_with_doctor"].isna(),#<<
            model.env.now - patient_results["arrival_time"], np.nan#<<
        )#<<

        # Run results
        run_results = {
            "run_number": run,
            "unseen_wait_time": patient_results["unseen_wait_time"].mean()#<<
        }

        return {
            "patient": patient_results,
            "run": run_results
        }
```

#### Run the model

```{python}
{{< include outputs_resources/run.py >}}
```

*The unseen mean wait time is returned as `np.float64()` because NumPy's `mean()` function returns a NumPy float type instead of a built-in Python `float`. You can just treat it like a regular `float`, or convert it with `float()` if you want it to print without `np.float64()`.*

:::

::::

:::: {.r-content}

::: {.callout title="▶&nbsp;&nbsp;View how to record the backlogged patient count and mean wait time" collapse="true"}

We will record two performance measures here: the **count** of unseen patients and their **mean wait time** (ie., time between arrival and end of simulation).

### Imports

We add:

* Functions from `dplyr`: `lead`, `left_join` and `summarise`.
* Functions from `simmer`: `now` and `set_attribute`.
* New import `tidyr::pivot_wider()`.

```{.r}
#' @keywords internal
"_PACKAGE"

## usethis namespace: start
#' @importFrom dplyr arrange bind_cols filter group_by lead left_join mutate n #<<
#' @importFrom dplyr slice summarise ungroup #<<
#' @importFrom rlang .data
#' @importFrom simmer add_generator add_resource get_mon_arrivals
#' @importFrom simmer get_mon_resources now release run seize set_attribute #<<
#' @importFrom simmer simmer timeout trajectory
#' @importFrom tibble tibble
#' @importFrom tidyr pivot_wider #<<
## usethis namespace: end

NULL
```

### Unseen patient count function

The `calc_unseen_n` function finds the count of patients who have a result in the new `wait_time_unseen` column.

```{r}
#' Calculate the number of patients still waiting for resource at end
#'
#' @param arrivals Dataframe with times for each patient with each resource.
#'
#' @return Tibble with columns containing result for each resource.
#' @export

calc_unseen_n <- function(arrivals, groups = NULL) {
  arrivals |>
    group_by(.data[["resource"]]) |>
    summarise(value = sum(!is.na(.data[["wait_time_unseen"]]))) |>
    pivot_wider(names_from = "resource",
                values_from = "value",
                names_glue = "count_unseen_{resource}") |>
    ungroup()
}
```

### Unseen patient mean wait time function

The `calc_unseen_mean` function finds the mean of a new column `wait_time_unseen`.

```{r}
#' Calculate the mean wait time of patients who are still waiting for a
#' resource at the end of the simulation
#'
#' @param arrivals Dataframe with times for each patient with each resource.
#'
#' @return Tibble with columns containing result for each resource.
#' @export

calc_unseen_mean <- function(arrivals) {
  arrivals |>
    group_by(.data[["resource"]]) |>
    summarise(value = mean(.data[["wait_time_unseen"]], na.rm = TRUE)) |>
    pivot_wider(names_from = "resource",
                values_from = "value",
                names_glue = "mean_waiting_time_unseen_{resource}") |>
    ungroup()
}
```

### Run results function

In `get_run_results()`, the table `results[["arrivals"]]` is passed to `calc_unseen_n()` and `calc_unseen_mean`.

```{r}
#' Get average results for the provided single run.
#'
#' @param results Named list with `arrivals` from `get_mon_arrivals()` and
#'   `resources` from `get_mon_resources()`
#'   (`per_resource = TRUE` and `ongoing = TRUE`).
#' @param run_number Integer index of current simulation run.
#'
#' @return Tibble with processed results from replication.
#' @export

get_run_results <- function(results, run_number) {
  metrics <- list(
    calc_unseen_n(results[["arrivals"]]), #<<
    calc_unseen_mean(results[["arrivals"]]) #<<
  )
  dplyr::bind_cols(tibble(replication = run_number), metrics)
}
```

### Model function

The modifications in `model()` are very similar to those described for calculating mean wait time of finished patients.

The only difference here is that we make a column called `wait_time_unseen`, which records the time elapsed between their arrival and the end of the simulation, for patients who are unseen at the end of the simulation (i.e., their `serve_start` is NA).

The earlier changes to add `serve_start` to the arrivals dataframe are the same as before though - saving with `set_attribute`, setting `mon = 2L` in the patient generator, retrieving results using `get_mon_attributes()`, and restructuring the returned dataframe to transform `doctor_serve_length` into `serve_length` (aligned with the corresponding doctor rows).

```{r}
#' Run simulation.
#'
#' @param param List. Model parameters.
#' @param run_number Numeric. Run number for random seed generation.
#'
#' @return Named list with tables `arrivals`, `resources` and `run_results`.
#' @export

model <- function(param, run_number) {

  # Set random seed based on run number
  set.seed(run_number)

  # Create simmer environment
  env <- simmer("simulation", verbose = param[["verbose"]])

  # Define the patient trajectory
  patient <- trajectory("consultation") |>
    seize("doctor", 1L) |>
    # Record time resource is obtained #<<
    set_attribute("doctor_serve_start", function() now(env)) |> #<<
    timeout(function() {
      rexp(n = 1L, rate = 1L / param[["consultation_time"]])
    }) |>
    release("doctor", 1L)

  env <- env |>
    # Add doctor resource
    add_resource("doctor", param[["number_of_doctors"]]) |>
    # Add patient generator (mon=2 so can get manual attributes) #<<
    add_generator("patient", patient, function() {
      rexp(n = 1L, rate = 1L / param[["interarrival_time"]])
    }, mon = 2L) |> #<<
    # Run the simulation
    simmer::run(until = (param[["warm_up_period"]] +
                           param[["data_collection_period"]]))

  # Extract information on arrivals and resources from simmer environment
  result <- list(
    arrivals = get_mon_arrivals(env, per_resource = TRUE, ongoing = TRUE),
    resources = get_mon_resources(env)
  )

  # Get the extra arrivals attributes #<<
  extra_attributes <- get_mon_attributes(env) |> #<<
    dplyr::select("name", "key", "value") |> #<<
    # Add column with resource name, and remove that from key #<<
    mutate(resource = gsub("_.+", "", .data[["key"]]), #<<
           key = gsub("^[^_]+_", "", .data[["key"]])) |> #<<
    pivot_wider(names_from = "key", values_from = "value") #<<

  # Merge extra attributes with the arrival data #<<
  result[["arrivals"]] <- left_join( #<<
    result[["arrivals"]], extra_attributes, by = c("name", "resource") #<<
  ) #<<

  # Filter to remove results from the warm-up period
  result <- filter_warmup(result, param[["warm_up_period"]])

  # Replace "replication" value with appropriate run number
  result[["arrivals"]] <- mutate(result[["arrivals"]],
                                 replication = run_number)
  result[["resources"]] <- mutate(result[["resources"]],
                                  replication = run_number)

  # Calculate wait time for each patients who remain unseen at the end of #<<
  # the simulation #<<
  result[["arrivals"]] <- result[["arrivals"]] |> #<<
    mutate( #<<
      wait_time_unseen = ifelse( #<<
        is.na(.data[["serve_start"]]), now(env) - .data[["start_time"]], NA #<<
      ) #<<
    ) #<<

  # Calculate the average results for that run
  result[["run_results"]] <- get_run_results(result, run_number)

  result
}
```

### Run the model

```{r}
param <- create_params()
result <- model(param = param, run_number = 1L)
kable(result[["arrivals"]])
kable(result[["run_results"]])
```

:::

::::

## Time-weighted averages

A **time-weighted average** calculates the mean value of a performance measure by weighting each observed value by the length of time it persisted, then dividing by the total time observed.

This is used by three of the performance measures above: **utilisation, queue length and patients in system**. But why do we use it? Let's use **queue length** as an example...

Consider a queue with the following time periods and sizes:

| Interval    | Queue Size | Duration |
|-------------|------------|----------|
| 0 - 14.4    | 0          | 14.4     |
| 14.4 - 15.2 | 1          | 0.8      |
| 15.2 - 16.1 | 2          | 0.9      |
| 16.1 - 17.0 | 3          | 0.9      |


::: {.pale-red}

**Simple (unweighted) average**

This method ignores duration and only averages the observed queue sizes:

$$
\text{Simple average} = \frac{0 + 1 + 2 + 3}{4} = 1.5
$$

It overestimates the typical queue size because most of the time, the queue was empty.

:::

::: {.pale-green}

**Time-weighted average**

This method weights each value by how long it lasted:

$$
\text{Time-weighted average} = \frac{(0 \times 14.4) + (1 \times 0.8) + (2 \times 0.9) + (3 \times 0.9)}{17.0} \approx 0.312
$$

The value is much closer to the true "average" experienced in time, because the queue was empty for most of the observation period.

:::

The same logic applies for the other performance measures. In the code, we track the product of value and duration ("area under the curve") for each metric, then divide the total by the observation time.

It's called "**area under the curve**" because, when you plot the measure (e.g., queue length) over time, the total area of all the rectangles you get under the curve (height=queue x width=time) returns your average.

```{r}
#| echo: false
# Create a data frame for queue size and time
queue_df <- data.frame(
  time = c(0.0, 14.4, 15.2, 16.1, 17.0),
  queue = c(0L, 1L, 2L, 3L, 3L)
)

# Use geom_step for step plot, horizontal-then-vertical (hv) style
p <- ggplot(queue_df, aes(x = time, y = queue)) +
  geom_step(direction = "hv", color = "blue", linewidth = 1.2) +
  scale_x_continuous(breaks = seq(0L, 18L, by = 2L)) +
  scale_y_continuous(breaks = seq(0L, 3L, by = 1L)) +
  labs(x = "Time", y = "Queue size") +
  theme_minimal()

ggplotly(p)
```

## Explore the example models

<div class="h3-tight"></div>

### Nurse visit simulation

:::: {.python-content}

{{< include /html/pydesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `simulation/patient.py`<br>`simulation/monitoredresource.py`<br>`simulation/model.py`<br>`simulation/runner.py` |
| **What to look for?** | All of the metrics above can be seen in this example repository. |
| **Why it matters?** | Seeing all these metrics together in the example repository can feel overwhelming! So this page breaks them down individually and shows how each is captured in the code. In the nurse visit repository, there is some additional logic in places to allow the model to handle a scenario with zero arrivals. |
: {tbl-colwidths="[20,80]"}
:::

::::

:::: {.r-content}

{{< include /html/rdesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `R/get_run_results.R`<br>`R/model.R` |
| **What to look for?** | All of the metrics above can be seen in this example repository. |
| **Why it matters?** | Seeing all these metrics together in the example repository can feel overwhelming! So this page breaks them down individually and shows how each is captured in the code. In the nurse visit repository, there is some additional logic in places to allow the model to handle a scenario with zero arrivals, and also to allow calculations to be performed by group. |
: {tbl-colwidths="[20,80]"}
:::

::::

### Stroke pathway simulation

:::: {.python-content}

{{< include /html/pydesrapstroke.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `simulation/patient.py`<br>`simulation/model.py`<br>`simulation/runner.py` |
| **What to look for?** | This model computes less standard metrics: occupancy frequency, percentages, cumulative percentages, probability of delay, and "1 in n" delays. |
| **Why it matters?** | While this page focuses on some common DES metrics, the stroke simulation shows examples of some more unusual metrics. Indeed, depending on your model structure, there is a huge array of metrics you could calculate, well beyond those described on this page. |
: {tbl-colwidths="[20,80]"}
:::

::::

:::: {.r-content}

{{< include /html/rdesrapstroke.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `R/get_occupancy_stats.R`<br>`R/model.R` |
| **What to look for?** | This model computes less standard metrics: occupancy frequency, percentages, cumulative percentages, probability of delay, and "1 in n" delays. |
| **Why it matters?** | While this page focuses on some common DES metrics, the stroke simulation shows examples of some more unusual metrics. Indeed, depending on your model structure, there is a huge array of metrics you could calculate, well beyond those described on this page. |
: {tbl-colwidths="[20,80]"}
:::

::::

## Test yourself

Try adding one or more of the measures described above to your own DES model.

See how the results change as you adjust the model (e.g. arrival rate, number of resources).

## Acknowledgements

The `MonitoredResource` class is based on Tom Monks, Alison Harper and Amy Heather (2025) An introduction to Discrete-Event Simulation (DES) using Free and Open Source Software (<https://github.com/pythonhealthdatascience/intro-open-sim/tree/main>) (MIT Licence). They based it on the method described in Law. Simulation Modeling and Analysis 4th Ed. Pages 14 - 17.

<br><br>
