---
title: Replications
filters:
  - line-highlight
date: "2025-11-05T11:21:31+00:00"
---

{{< include /scripts/_reticulate-setup.md >}}

<!-- Using source-line-numbers rather than in-line #<< at some points on this page as I want to display this same code on a later page without the line highlights -->

```{python}
#| echo: false
# pylint: disable=undefined-variable, unused-import
```

:::: {.pale-blue}

**Learning objectives:**

* **Adapt** our model to run multiple replications.

**Pre-reading:**

This page continues on from: [Performance measures](outputs.qmd).

> *Entity generation → Entity processing → Initialisation bias → Performance measures → Replications*

::::

:::: {.very-pale-blue}

**Required packages:**

These should be available from environment setup on the [Structuring as a package](/pages/guide/setup/package.qmd) page.

::: {.python-content}

```{python}
from IPython.display import HTML
from itables import to_html_datatable
import numpy as np
import pandas as pd
import scipy.stats as st
import simpy
from sim_tools.distributions import Exponential
```

:::

::: {.r-content}

```{r}
#| output: false
library(dplyr)
library(kableExtra)
library(knitr)
library(simmer)
library(tidyr)
```

```{r}
#| echo: false
#| output: false
# For illustrative plot only, code hidden
library(ggplot2)
```

:::

::::

<br>

Replications in discrete event simulation are necessary because each run includes randomness, leading to different results every time.

Running multiple replications reveals the **range and variability** of possible system outcomes, allowing the calculation of **averages and confidence intervals** for results.

For example, in the chart below each point represents a single replication, showing the values of performance measure 1 (PM1) and performance measure 2 (PM2) for that run. Across runs, PM1 varies only slightly, whereas PM2 shows much larger fluctuations, illustrating how replications reveal the differing levels of variability in each performance measure and why averages and confidence intervals are needed rather than relying on a single run.

```{r}
#| echo: false
set.seed(9L)

df <- data.frame(
  Category = rep(c("Performance measure 1", "Performance measure 2"),
                 each = 7L),
  Value = runif(14L, min = 0L, max = 1L),
  Run = rep(paste("Run", 1L:7L), 2L),
  stringsAsFactors = FALSE
)

ggplot(df, aes(x = Category, y = Value)) +
  geom_point(size = 3L) +
  geom_text(aes(label = Run), vjust = -1L, size = 3L) +
  labs(x = NULL, y = "Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 12L))
```

## How to run replications

We will continue working with the model developed on the previous pages. This model contains all the metrics from the [Performance measures](outputs.qmd) page (with utilisation implemented via the time-weighted approach).

::: {.r-content}

> **Note:** The results will look different to the previous page for several of the metrics. This is due to mean time in consultation. As mentioned on the [Performance measures](outputs.qmd) page, mean time in consultation requires saving sampled times using `set_attribute()` ahead of resource use. This changes the order in which random numbers are drawn from the simulation’s global random number generator. As a result, even with the same seed and parameters, the results will differ from previous runs before we added the metric.

:::

:::: {.python-content}

### Parameter class

Let's add a new attribute `number_of_runs`. We've arbitrarily set this to five, but the [Number of replications](n_reps.qmd) page shows how to choose a suitable number to run.

```{python}
#| source-line-numbers: "17-18,25,42-43,52"
{{< include replications_resources/parameters.py >}}
```

### Patient class

No changes required.

::: {.callout title="▶&nbsp;&nbsp; Patient class" collapse="true" style="minimal"}

```{python}
{{< include replications_resources/patient.py >}}
```

:::

### Monitored resource class

No changes required.

::: {.callout title="▶&nbsp;&nbsp; Monitored resource class" collapse="true" style="minimal"}

```{python}
{{< include replications_resources/monitored_resource.py >}}
```

:::

### Model class

No changes required.

::: {.callout title="▶&nbsp;&nbsp;Model class" collapse="true" style="minimal"}

```{python}
{{< include replications_resources/model.py >}}
```

:::

### Summary statistics function

This function computes descriptive statistics for a dataset, returning the mean, standard deviation, and 95% confidence interval.

* If the input has **no valid values**, all outputs are `NaN`.

* If the input has **fewer than 3 values**, it returns the mean only, with the rest as `NaN`.

* If the input has **3 or more values**, it calculates the mean, standard deviation, and a 95% confidence interval using the t-distribution (`scipy.stats`).

* If all values are identical (zero variance), the confidence interval equals the mean.

```{python}
{{< include replications_resources/summary_stats.py >}}
```

### Runner class

We run replications with a new method `run_reps()`, explained line-by-line below.

```{python}
#| source-line-numbers: "83-125"
{{< include replications_resources/runner.py >}}
```

#### Explaining `Runner.run_reps()`

```{.python}
def run_reps(self):
    """
    Execute a single model configuration for multiple runs.

    Returns
    -------
    dict
        Contains patient-level results, results from each run and
        overall results.
    """
    # Run replications
    all_results = [self.run_single(run)
                    for run in range(self.param.number_of_runs)]
```

It starts by calling `run_single()` repeatedly for the requested number of simulation runs, passing the run number to set the random seed (ensuring each run uses a different seed!).

<br>

```{.python}
# Separate results from each run into appropriate lists
patient_results_list = [result["patient"] for result in all_results]
run_results_list = [result["run"] for result in all_results]

# Convert lists into dataframes
patient_results_df = pd.concat(
    patient_results_list, ignore_index=True
)
run_results_df = pd.DataFrame(run_results_list)
```

The results then get stacked on top of each other, so you end up with one big dataframe for all patients, and another for all runs. The `replication` column shows which run each row came from.

<br>

```{.python}
# Calculate average results and uncertainty from across all runs
uncertainty_metrics = {}
run_col = run_results_df.columns

# Loop through the run performance measure columns
# Calculate mean, standard deviation and 95% confidence interval
for col in run_col[~run_col.isin(["run_number", "scenario"])]:
    uncertainty_metrics[col] = dict(zip(
        ["mean", "std_dev", "lower_95_ci", "upper_95_ci"],
        summary_stats(run_results_df[col])
    ))
# Convert to dataframe
overall_results_df = pd.DataFrame(uncertainty_metrics)
```

We then calculate summary statistics (mean, standard deviation, and 95% confidence interval) for each run in `run_results_df` using `summary_stats()`.

<br>

```{.python}
return {
    "patient": patient_results_df,
    "run": run_results_df,
    "overall": overall_results_df
}
```

At the end, the function returns a dictionary with all patient results (`patient`), all run results (`run`), and a summary table with statistics calculated across all runs (`overall`).

### Run the model

```{python}
runner = Runner(param=Parameters())
result = runner.run_reps()
HTML(to_html_datatable(result["patient"]))
HTML(to_html_datatable(result["run"]))
HTML(to_html_datatable(result["overall"]))
```

::::

:::: {.r-content}

### Parameter function

Let's add a new attribute `number_of_runs`. We've arbitrarily set this to five, but the [Number of replications](n_reps.qmd) page shows how to choose a suitable number to run.

```{r}
#| source-line-numbers: "10,22,31"
#| file: replications_resources/create_params.R
```

### Warm-up function

No changes required.

::: {.callout title="▶&nbsp;&nbsp;Warm-up function" collapse="true" style="minimal"}

```{r}
#| file: replications_resources/filter_warmup.R
```

:::

### Functions to calculate different metrics

No changes required.

::: {.callout title="▶&nbsp;&nbsp;View metrics functions" collapse="true" style="minimal"}

```{r}
#| file: replications_resources/metrics.R
```

:::

### Run results function

No changes required.

::: {.callout title="▶&nbsp;&nbsp;Run results function" collapse="true" style="minimal"}

```{r}
#| file: replications_resources/get_run_results.R
```

:::

### Model function

No changes required.

::: {.callout title="▶&nbsp;&nbsp;Model function" collapse="true" style="minimal"}

```{r}
#| file: replications_resources/model.R
```

:::

### Runner function

This new function `runner()` runs the simulation model multiple times based on the specified `param[["number_of_runs"]]`. It passes the run number to set the random seed (ensuring each run uses a different seed!).

It combines the results into unified output tables for arrivals, resources and run-level results. If only one replication is requested, it simply returns that single result without combining.

```{r}
#| file: replications_resources/runner.R
```

### Run the model

```{r}
param <- create_params()
result <- runner(param = param)
result[["arrivals"]] |>
  arrange(replication, start_time) |>
  kable() |>
  scroll_box(height = "400px")
kable(result[["resources"]]) |> scroll_box(height = "400px")
kable(result[["run_results"]])
result[["patients_in_system"]] |>
  kable() |>
  scroll_box(height = "400px")
```

::::

## Explore the example models

<div class="h3-tight"></div>

### Nurse visit simulation

:::: {.python-content}

{{< include /html/pydesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `simulation/param.py`<br>`simulation/summary_stats.py`<br>`simulation/runner.py` |
| **What to look for?** | This is set up to run replications in parallel by default, as introduced on the [Parallel processing](parallel.qmd) page. |
: {tbl-colwidths="[20,80]"}
:::

::::

:::: {.r-content}

{{< include /html/rdesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `R/parameters.R`<br>`R/runner.R` |
| **What to look for?** | This is set up to run replications sequentially by default, but has functionality for running in parallel as introduced on the [Parallel processing](parallel.qmd) page. This means some differences in set up (with sequential also using `future_lapply` in that case - see parallel page for more details). |
: {tbl-colwidths="[20,80]"}
:::

::::

### Stroke pathway simulation

:::: {.python-content}

{{< include /html/pydesrapstroke.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `simulation/parameters.py`<br>`simulation/runner.py` |
| **What to look for?** | Like nurse visit simulation, this is also set up to support parallel processing. |
: {tbl-colwidths="[20,80]"}
:::

::::

:::: {.r-content}

{{< include /html/rdesrapstroke.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `R/parameters.R`<br>`R/runner.R` |
| **What to look for?** | Like nurse visit simulation, this is also set up to support parallel processing. |
: {tbl-colwidths="[20,80]"}
:::

::::

## Test yourself

Extend your model to support multiple replications, as shown above.

<br><br>
