---
title: ❌ Entity processing
filters:
  - line-highlight
---

{{< include ../../scripts/_reticulate-setup.md >}}

:::: {.pale-blue}

🎯 **Objectives**

::: {.python-content}
This page adds a process to the model (which uses a resource), with inspiration from Rosser et al. ([2025](https://des.hsma.co.uk/)).
:::
::: {.r-content}
This page adds a process to the model (which uses a resource).
:::

* **🥼 Adding a consultation with a doctor to the model**.

[🔗](../intro/guidelines.qmd) **Reproducibility guidelines**

Entity processing is not directly addressed by the guidelines, but is an essential step in building DES.

::::

<br>

On the last page, we created a simple model with patient arrivals. As we continue building our model iteratively, we now add add a feature typical fo DES models: a process using a resource. In this case, a consultation with a doctor.

> ⚠️ This page continues on from the [Entity generation](patients.qmd) page - work through that first!

<br>

Required packages:

::: {.python-content}

```{python}
import numpy as np
import simpy
from sim_tools.distributions import Exponential
```

:::

::: {.r-content}

```{r}
library(simmer)
```

:::

## 🥼 Adding a consultation with a doctor to the model

::: {.python-content}

<div class="h3-tight"></div>

### Parameter class

There are two new parameters: `consultation_time` and `number_of_doctors`.

```{python}
class Parameters:
    """
    Parameter class.

    Attributes
    ----------
    interarrival_time : float
        Mean time between arrivals (minutes).
    consultation_time : float#<<
        Mean length of doctor's consultation (minutes).#<<
    number_of_doctors : int#<<
        Number of doctors.#<<
    run_length : int
        Total duration of simulation (minutes).
    """
    def __init__(
        self, interarrival_time=5, consultation_time=10,#<<
        number_of_doctors=3, run_length=50#<<
    ):
        """
        Initialise Parameters instance.

        Parameters
        ----------
        interarrival_time : float
            Time between arrivals (minutes).
        consultation_time : float#<<
            Length of consultation (minutes).#<<
        number_of_doctors : int#<<
            Number of doctors.#<<
        run_length : int
            Total duration of simulation (minutes).
        """
        self.interarrival_time = interarrival_time
        self.consultation_time = consultation_time#<<
        self.number_of_doctors = number_of_doctors#<<
        self.run_length = run_length
```

### Patient class

No changes needed!

```{python}
class Patient:
    """
    Represents a patient.

    Attributes
    ----------
    patient_id : int
        Unique patient identifier.
    """
    def __init__(self, patient_id):
        """
        Initialises a new patient.

        Parameters
        ----------
        patient_id : int
            Unique patient identifier.
        """
        self.patient_id = patient_id
```

### Model class

These changes are explained below...

```{python}
class Model:
    """
    Simulation model.

    Attributes
    ----------
    param : Parameters
        Simulation parameters.
    run_number : int
        Run number for random seed generation.
    env : simpy.Environment
        The SimPy environment for the simulation.
    doctor : simpy.Resource#<<
        SimPy resource representing doctors.#<<
    patients : list
        List of Patient objects.
    arrival_dist : Exponential
        Distribution used to generate random patient inter-arrival times.
    consult_dist : Exponential#<<
        Distribution used to generate length of a doctor's consultation.#<<
    """
    def __init__(self, param, run_number):
        """
        Create a new Model instance.

        Parameters
        ----------
        param : Parameters
            Simulation parameters.
        run_number : int
            Run number for random seed generation.
        """
        self.param = param
        self.run_number = run_number

        # Create SimPy environment
        self.env = simpy.Environment()

        # Create resource
        self.doctor = simpy.Resource(#<<
            self.env, capacity=self.param.number_of_doctors#<<
        )#<<

        # Create a random seed sequence based on the run number
        ss = np.random.SeedSequence(self.run_number)
        seeds = ss.spawn(2)#<<

        # Create patient list
        self.patients = []

        # Initialise distributions
        self.arrival_dist = Exponential(mean=self.param.interarrival_time,
                                        random_seed=seeds[0])
        self.consult_dist = Exponential(mean=self.param.consultation_time,#<<
                                        random_seed=seeds[1])#<<

    def generate_arrivals(self):
        """
        Process that generates patient arrivals.
        """
        while True:
            # Sample and pass time to next arrival
            sampled_iat = self.arrival_dist.sample()
            yield self.env.timeout(sampled_iat)

            # Create a new patient
            patient = Patient(patient_id=len(self.patients)+1)
            self.patients.append(patient)

            # Print arrival time
            print(f"Patient {patient.patient_id} arrives " +
                  f"at time: {self.env.now:.3f}")

            # Start process of consultation#<<
            self.env.process(self.consultation(patient))#<<

    def consultation(self, patient):#<<
        """#<<
        Process that simulates a consultation.#<<

        Parameters#<<
        ----------#<<
        patient :#<<
            Instance of the Patient() class representing a single patient.#<<
        """#<<
        with self.doctor.request() as req:#<<
            yield req#<<
            print(f"Patient {patient.patient_id} starts consultation " +#<<
                  f"at: {self.env.now:.3f}")#<<
            time_with_doctor = self.consult_dist.sample()#<<
            yield self.env.timeout(time_with_doctor)#<<

    def run(self):
        """
        Run the simulation.
        """
        # Schedule arrival generator
        self.env.process(self.generate_arrivals())
        # Run the simulation
        self.env.run(until=self.param.run_length)

```

<br>

```{.python}
# Create resource
self.doctor = simpy.Resource(                                       
    self.env, capacity=self.param.number_of_doctors                 
)
```

This creates a resource called doctor using SimPy. It represents the group of doctors, with the number available set up `number_of_doctors`.

In simulations, a resource is something a process (here, a patient) needs to use, but only a limited number can access it at once - patients will wait if all doctors are busy.

<br>

```{.python}
seeds = ss.spawn(2)
```

We now create two seeds: one for patient arrivals and one for consultations.

<br>

```{.python}
self.consult_dist = Exponential(mean=self.param.consultation_time,  
                                random_seed=seeds[1])
```

Alike arrivals, the length of each consultation is sampled from an exponential distribution using the sim-tools `Exponential` class.

<br>

```{.python}
# Start process of consultation                                 
self.env.process(self.consultation(patient))
```

As soon as a patient arrives, we tell SimPy to begin handling this patient's consultation using `self.env.process()`. SimPy will run this in parallel with other events like arrivals, so multiple patients and activities can progress at the same time.

<br>

```{.python}
def consultation(self, patient):
    """
    Process that simulates a consultation.

    Parameters
    ----------
    patient :
        Instance of the Patient() class representing a single patient.
    """  
    with self.doctor.request() as req:
        yield req
        print(f"Patient {patient.patient_id} starts consultation " +
                f"at: {self.env.now:.3f}")

```

Each patient asks for a doctor using `self.doctor.request()`. If they're all busy, the patient waits in a queue until one is free.

Once a doctor is available, SimPy assigns the resource to that patient, and the consultation start time is printed.

```{.python}
time_with_doctor = self.consult_dist.sample()
yield self.env.timeout(time_with_doctor)
```

The duration of the consultation is then randomly sampled from the distribution set up earlier. The simulation then waits for this amount of time using `timeout()`, after which the doctor is released and free for the next patient.

### Run the model

```{python}
param = Parameters()
model = Model(param=param, run_number=0)
model.run()
```

:::

::: {.r-content}

### Parameter function

There are two new parameters: `consultation_time` and `number_of_doctors`.

```{r}
#' Generate parameter list.
#'
#' @param interarrival_time Numeric. Time between arrivals (minutes).
#' @param consultation_time Numeric. Mean length of doctor's#<<
#'   consultation (minutes).#<<
#' @param number_of_doctors Numeric. Number of doctors.#<<
#' @param run_length Numeric. Total duration of simulation (minutes).
#' @param verbose Boolean. Whether to print messages as simulation runs.
#'
#' @return A named list of parameters.

create_params <- function(
    interarrival_time = 5L,
    consultation_time = 10L,#<<
    number_of_doctors=3L,#<<
    run_length = 50L,
    verbose = TRUE
) {
  list(
    interarrival_time = interarrival_time,
    consultation_time = consultation_time,#<<
    number_of_doctors = number_of_doctors,#<<
    run_length = run_length,
    verbose = verbose
  )
}
```

### Model function

We make two main changes to the `model()` function.

* **Patient trajectory.** Patients must now wait for a doctor, spend a random amount of time in consultation (`timeout()` with `rexp()` function), and then release a doctor.

* **Doctor resource setup.** The simulation environment adds a "doctor" resource with a specified capacity (`number_of_doctors`) using the `add_resource()` function.

```{r}
#' Run simulation.
#'
#' @param param List. Model parameters.
#' @param run_number Int. Run number for random seed generation.
#'
#' @importFrom simmer add_generator run simmer timeout trajectory

model <- function(param, run_number) {

  # Set random seed based on run number
  set.seed(run_number)

  # Create simmer environment
  env <- simmer("simulation", verbose = param[["verbose"]])

  # Define the patient trajectory
  patient <- trajectory("consultation") |>
    seize("doctor", 1L) |>#<<
    timeout(function() {#<<
      rexp(n = 1L, rate = 1L / param[["consultation_time"]])#<<
    }) |>#<<
    release("doctor", 1L)#<<

  env <- env |>
    # Add doctor resource#<<
    add_resource("doctor", param[["number_of_doctors"]]) |>#<<
    # Add patient generator
    add_generator("patient", patient, function() {
      rexp(n = 1L, rate = 1L / param[["interarrival_time"]])
    }) |>
    # Run the simulation
    simmer::run(until = param[["run_length"]])

}
```

### Run the model

```{r}
param <- create_params()
model(param = param, run_number = 1L)
```

:::

## 🔗 Further information

::: {.python-content}

* "[HSMA - little book of DES](https://des.hsma.co.uk/)" from Sammi Rosser, Dan Chalk and Amy Heather 2025

    *Introduction to writing discrete event simulation models for healthcare in SimPy.*

:::

::: {.r-content}

N/A

:::

<br><br>
