---
title: Logging
filters:
  - line-highlight
date: "2025-11-03T10:48:30+00:00"
---

```{python}
#| echo: false
# pylint: disable=unexpected-keyword-arg,no-value-for-parameter
```

:::: {.pale-blue}

**Learning objectives:**

* Learn how to set up alternative logs for your model.
* Add parameter list to your log.
* Understand how to save the log to a file.

**Relevant reproducibility guidelines:**

* NHS Levels of RAP (🥈): Logs are automatically recorded by the pipeline to ensure outputs are as expected.

**Pre-reading:**

This page continues on from: [Entity processing](../model/process.qmd).

> *Entity generation → Entity processing → Logging*

::::

:::: {.very-pale-blue}

**Required packages:**

These should be available from environment setup on the [Structuring as a package](/pages/guide/setup/package.qmd) page.

::: {.python-content}

```{python}
import logging
import os
from pprint import pformat

import numpy as np
from rich.logging import RichHandler
from rich.console import Console
import simpy
from sim_tools.distributions import Exponential
```

:::

::: {.r-content}

```{r}
library(simmer)
```

:::

::::

<br>

::: {.python-content}

In the basic model we built on the previous pages, we printed out the arrival and consultation times as the simulation ran.

:::

::: {.r-content}

In the basic simmer model we built on the previous pages, we ran the simulation with `verbose = TRUE` and looked at the logs output by simmer.

:::

This allowed us to **check that the model is working as expected**. It is important to check this when **building a simulation**, so you can have confidence that events are coded correctly - e.g., that arrivals happen at the right time, that patients get seen in the right order by the right people. Logs may help you spot bugs, so are handy for debugging as part of [model verification](../verification_validation/verification_validation.qmd#debugging).

::: {.python-content}

Instead of print statements, this page shows you how to make more detailed visual logs, with functionality including saving them to a file.

:::

::: {.r-content}

The default simmer logs are useful but can be quite confusing to use. Instead, we can make custom logs that are easier to read and interpret.

:::

Logging should be **optional**. It's best to use them while developing and testing the model with short runs and no replications. They will be excessive and slow down the simulation if used for long runs or multiple replications.

## Setting up alternative logs for the model

<div class="h3-tight"></div>

::: {.python-content}

### Parameter class

No changes required.

::: {.callout title="▶&nbsp;&nbsp;Parameter class" collapse="true" style="minimal"}

```{python}
class Parameters:
    """
    Parameter class.

    Attributes
    ----------
    interarrival_time : float
        Mean time between arrivals (minutes).
    consultation_time : float
        Mean length of doctor's consultation (minutes).
    number_of_doctors : int
        Number of doctors.
    run_length : int
        Total duration of simulation (minutes).
    verbose : bool
        Whether to print messages as simulation runs.
    """
    def __init__(
        self, interarrival_time=5, consultation_time=10,
        number_of_doctors=3, run_length=50, verbose=True
    ):
        """
        Initialise Parameters instance.

        Parameters
        ----------
        interarrival_time : float
            Time between arrivals (minutes).
        consultation_time : float
            Length of consultation (minutes).
        number_of_doctors : int
            Number of doctors.
        run_length : int
            Total duration of simulation (minutes).
        verbose : bool
            Whether to print messages as simulation runs.
        """
        self.interarrival_time = interarrival_time
        self.consultation_time = consultation_time
        self.number_of_doctors = number_of_doctors
        self.run_length = run_length
        self.verbose = verbose
```

:::

### Patient class

No changes required.

::: {.callout title="▶&nbsp;&nbsp;Patient class" collapse="true" style="minimal"}

```{python}
class Patient:
    """
    Represents a patient.

    Attributes
    ----------
    patient_id : int
        Unique patient identifier.
    arrival_time : float
            Time patient entered the system (minutes).
    """
    def __init__(self, patient_id, arrival_time):
        """
        Initialises a new patient.

        Parameters
        ----------
        patient_id : int
            Unique patient identifier.
        arrival_time : float
            Time patient entered the system (minutes).
        """
        self.patient_id = patient_id
        self.arrival_time = arrival_time
```

:::

### Logging class

We create a new class `SimLogger` for logging, which is explained line-by-line below.

```{python}
class SimLogger:
    """
    Provides log of events as the simulation runs.

    Attributes
    ----------
    verbose : bool
        Whether to print messages as simulation runs.
    """
    def __init__(self, verbose):
        """
        Initialise logging class.

        Parameters
        ----------
        verbose : bool
            Whether to print messages as simulation runs.
        """
        self.verbose = verbose
        if self.verbose:
            self.logger = logging.getLogger(__name__)
            self._configure_logging()

    def _configure_logging(self):
        """
        Configure the logger.
        """
        # Ensure any existing handlers are removed to avoid duplication
        for handler in self.logger.handlers[:]:
            self.logger.removeHandler(handler)

        # Configure RichHandler without INFO/ERROR labels, times or paths
        # to log message. Set up console with jupyter-specific behaviour
        # disabled to prevent large gaps between each log message on ipynb.
        console = Console()
        console.is_jupyter = False
        rich_handler = RichHandler(
            console=console, show_time=False,
            show_level=False, show_path=False
        )

        # Add handler to the logger
        self.logger.addHandler(rich_handler)

        # Set logging level and format. If don't set level info, it would
        # only show log messages which are warning, error or critical.
        self.logger.setLevel(logging.INFO)
        formatter = logging.Formatter("%(message)s")
        rich_handler.setFormatter(formatter)

    def log(self, msg, sim_time=None):
        """
        Log a message if logging is enabled.

        Parameters
        ----------
        msg : str
            Message to log.
        sim_time : float or None, optional
            Current simulation time. If provided, prints before message.
        """
        if self.verbose:
            if sim_time is not None:
                self.logger.info("%0.3f: %s", sim_time, msg)
            else:
                self.logger.info(msg)
```

#### Explaining `SimLogger`

```{.python}
class SimLogger:
    """
    Provides log of events as the simulation runs.

    Attributes
    ----------
    verbose : bool
        Whether to print messages as simulation runs.
    """
    def __init__(self, verbose):
        """
        Initialise logging class.

        Parameters
        ----------
        verbose : bool
            Whether to print messages as simulation runs.
        """
        self.verbose = verbose
        if self.verbose:
            self.logger = logging.getLogger(__name__)
            self._configure_logging()
```

If `verbose` is `True`, the class configures a logger using Python's built-in `logging` module, then calls our `_configure_logging` method.

If `verbose` is `False`, the logger is not set up and later calls to the `log()` method will not produce an output.

<br>

```{.python}
def _configure_logging(self):
    """
    Configure the logger.
    """
    # Ensure any existing handlers are removed to avoid duplication
    for handler in self.logger.handlers[:]:
        self.logger.removeHandler(handler)
```

The `_configure_logging` method begins `_`, a Python convention which indicates that it is a method for internal use within the class (i.e., users wouldn't call the method directly).

This method starts by removing any existing logging handlers. This is because Python loggers are persistent and global, so if you create multiple `SimLogger` instances or re-run code in an interactive environment like Jupyter, each new handler would be add to the logger, causing duplicate logging messages.

<br>

```{.python}
    # Configure RichHandler without INFO/ERROR labels, times or paths
    # to log message. Set up console with jupyter-specific behaviour
    # disabled to prevent large gaps between each log message on ipynb.
    console = Console()
    console.is_jupyter = False
    rich_handler = RichHandler(
        console=console, show_time=False,
        show_level=False, show_path=False
    )
```

A `RichHandler` is created for styled, readable messages in the console. The handler's settings are adjusted so that log messages are clear and compact: labels for log level, timestamps, and file paths are hidden to keep each line focused on the actual message.

Setting `console.is_jupyter = False` ensures that when working in a Jupyter notebook, the log lines are formatted without large gaps between them.

<br>

```{.python}
    # Add handler to the logger
    self.logger.addHandler(rich_handler)
```

The prepared handler is then added to the logger, which means all log output from this logger will be processed and shown using the `RichHandler`'s formatting rules.

<br>

```{.python}
    # Set logging level and format. If don't set level info, it would
    # only show log messages which are warning, error or critical.
    self.logger.setLevel(logging.INFO)
    formatter = logging.Formatter("%(message)s")
    rich_handler.setFormatter(formatter)
```

The logger's level is set to `INFO`, ensuring that our simulation logging messages (which use `logger.info`) are shown. If you do not set the level, the default is `WARNING`, which means only warnings and severe messages woudl appear.

<br>

```{.python}
def log(self, msg, sim_time=None):
    """
    Log a message if logging is enabled.

    Parameters
    ----------
    msg : str
        Message to log.
    sim_time : float or None, optional
        Current simulation time. If provided, prints before message.
    """
    if self.verbose:
        if sim_time is not None:
            self.logger.info("%0.3f: %s", sim_time, msg)
        else:
            self.logger.info(msg)
```

The `log()` method is used to record messages as the simulation runs. If `verbose` is `True`, this method will feed the provided message to the `logger`.

### Model class

In `Model`, a new attribute `logger` stores an instance of the `SimLogger` class.

Where previously we had statements "if verbose print...", these are replaced with `self.logger.log()`. As such, you will need to **remove** the following sections:

```{.python}
# Print arrival time
if self.param.verbose:
    print(f"Patient {patient.patient_id} arrives " +
            f"at time: {patient.arrival_time:.3f}")
```

```{.python}
if self.param.verbose:
    print(f"Patient {patient.patient_id} starts consultation " +
            f"at: {self.env.now:.3f}")
```

```{.python}
# Record end time
if self.param.verbose:
    print(f"Patient {patient.patient_id} " +
            f"leaves at: {self.env.now:.3f}")
```

These are then replaced as below.

*Bold lines show where the code has changed since the previous version. The highlight may correspond to a change anywhere on that line, not necessarily to the entire line.*

```{python}
class Model:
    """
    Simulation model.

    Attributes
    ----------
    param : Parameters
        Simulation parameters.
    run_number : int
        Run number for random seed generation.
    env : simpy.Environment
        The SimPy environment for the simulation.
    doctor : simpy.Resource
        SimPy resource representing doctors.
    patients : list
        List of Patient objects.
    arrival_dist : Exponential
        Distribution used to generate random patient inter-arrival times.
    consult_dist : Exponential
        Distribution used to generate length of a doctor's consultation.
    logger : SimLogger#<<
        The logging instance used for logging messages.#<<
    """
    def __init__(self, param, run_number):
        """
        Create a new Model instance.

        Parameters
        ----------
        param : Parameters
            Simulation parameters.
        run_number : int
            Run number for random seed generation.
        """
        self.param = param
        self.run_number = run_number

        # Create SimPy environment
        self.env = simpy.Environment()

        # Create resource
        self.doctor = simpy.Resource(
            self.env, capacity=self.param.number_of_doctors
        )

        # Create a random seed sequence based on the run number
        ss = np.random.SeedSequence(self.run_number)
        seeds = ss.spawn(2)

        # Set up attributes to store results
        self.patients = []

        # Initialise distributions
        self.arrival_dist = Exponential(mean=self.param.interarrival_time,
                                        random_seed=seeds[0])
        self.consult_dist = Exponential(mean=self.param.consultation_time,
                                        random_seed=seeds[1])

        # Initialise logger#<<
        self.logger = SimLogger(verbose=self.param.verbose)#<<

    def generate_arrivals(self):
        """
        Process that generates patient arrivals.
        """
        while True:
            # Sample and pass time to next arrival
            sampled_iat = self.arrival_dist.sample()
            yield self.env.timeout(sampled_iat)

            # Create a new patient
            patient = Patient(patient_id=len(self.patients)+1,
                              arrival_time=self.env.now)
            self.patients.append(patient)

            # Log arrival time#<<
            self.logger.log(msg=f"Patient {patient.patient_id} arrives.",#<<
                            sim_time=self.env.now)#<<

            # Start process of consultation
            self.env.process(self.consultation(patient))

    def consultation(self, patient):
        """
        Process that simulates a consultation.

        Parameters
        ----------
        patient :
            Instance of the Patient() class representing a single patient.
        """
        # Patient requests access to a doctor (resource)
        with self.doctor.request() as req:
            yield req

            # Log consultation start time#<<
            self.logger.log(#<<
                msg=f"Patient {patient.patient_id} starts consultation.",#<<
                sim_time=self.env.now#<<
            )#<<

            # Sample consultation duration and pass time spent with doctor
            time_with_doctor = self.consult_dist.sample()
            yield self.env.timeout(time_with_doctor)

            # Log end time#<<
            self.logger.log(msg=f"Patient {patient.patient_id} leaves.",#<<
                            sim_time=self.env.now)#<<

    def run(self):
        """
        Run the simulation.
        """
        # Schedule arrival generator
        self.env.process(self.generate_arrivals())

        # Run the simulation
        self.env.run(until=self.param.run_length)
```

### Run the model

```{python}
param = Parameters()
model = Model(param=param, run_number=0)
model.run()
```

<br>

If we run the model with `verbose=False` in parameters, no logs are printed:

```{python}
param = Parameters(verbose=False)
model = Model(param=param, run_number=0)
model.run()
```

:::

::: {.r-content}

### Parameter function

No changes required.

::: {.callout title="▶&nbsp;&nbsp;Parameter function" collapse="true" style="minimal"}

```{r}
#' Generate parameter list.
#'
#' @param interarrival_time Numeric. Time between arrivals (minutes).
#' @param consultation_time Numeric. Mean length of doctor's
#'   consultation (minutes).
#' @param number_of_doctors Numeric. Number of doctors.
#' @param run_length Numeric. Total duration of simulation (minutes).
#' @param verbose Boolean. Whether to print messages as simulation runs.
#'
#' @return A named list of parameters.
#' @export

create_params <- function(
  interarrival_time = 5L,
  consultation_time = 10L,
  number_of_doctors = 3L,
  run_length = 50L,
  verbose = TRUE
) {
  list(
    interarrival_time = interarrival_time,
    consultation_time = consultation_time,
    number_of_doctors = number_of_doctors,
    run_length = run_length,
    verbose = verbose
  )
}
```

:::

### Model function

The required changes are explained line-by-line below.

*Bold lines show where the code has changed since the previous version. The highlight may correspond to a change anywhere on that line, not necessarily to the entire line.*

```{r}
#' Run simulation.
#'
#' @param param List. Model parameters.
#' @param run_number Numeric. Run number for random seed generation.
#'
#' @importFrom simmer add_generator add_resource get_attribute log_ release run
#' @importFrom simmer seize set_attribute simmer timeout trajectory
#'
#' @export

model <- function(param, run_number) {

  # Set random seed based on run number
  set.seed(run_number)

  # Create simmer environment
  env <- simmer("simulation", verbose = FALSE, #<<
                log_level = if (param[["verbose"]]) 1L else 0L)#<<

  # Define the patient trajectory
  patient <- trajectory("consultation") |>
    log_("🚶 Arrives.", level = 1L) |> #<<
    seize("doctor", 1L) |>
    # Record sampled length of consultation #<<
    set_attribute("doctor_serve_length", function() { #<<
      rexp(n = 1L, rate = 1L / param[["consultation_time"]])#<<
    }) |> #<<
    log_(function() {#<<
      paste0("🩺 Doctor consultation begins (length: ", #<<
             round(get_attribute(env, "doctor_serve_length"), 5L), ")")#<<
    }, level = 1L) |> #<<
    timeout(function() get_attribute(env, "doctor_serve_length")) |> #<<
    release("doctor", 1L) |>
    log_("🚪 Leaves.", level = 1L)#<<

  env <- env |>
    # Add doctor resource
    add_resource("doctor", param[["number_of_doctors"]]) |>
    # Add patient generator
    add_generator("patient", patient, function() {
      rexp(n = 1L, rate = 1L / param[["interarrival_time"]])
    }) |>
    # Run the simulation
    simmer::run(until = param[["run_length"]])

}
```

#### Explaining changes to `model()`

```{.r}
env <- simmer("simulation", verbose = FALSE,
              log_level = if (param[["verbose"]]) 1L else 0L)
```

Simmer's default logging is disabled by setting `verbose = FALSE` in `simmer::simmer()`.

The display of our custom log messages is then managed using the `log_level` argument in `simmer::simmer()`.

<br>

```{.r}
log_("🚶 Arrives.", level = 1L) |>
```

Custom log messages are created using simmer's [`log_()`](https://r-simmer.org/reference/log_) function, which allows us to insert our own messages into the simulation trajectory.

Each message can be assigned a `level`. We've set `level = 1`, meaning it will only be displayed if the simulation's `log_level` is greater than or equal to 1. 

<br>

```{.r}
# Record sampled length of consultation 
set_attribute("doctor_serve_length", function() {
  rexp(n = 1L, rate = 1L / param[["consultation_time"]])
}) |>
log_(function() {
  paste0("🩺 Doctor consultation begins (length: ",
         round(get_attribute(env, "doctor_serve_length"), 5L), ")")
}, level = 1L) |>
timeout(function() get_attribute(env, "doctor_serve_length")) |>
```

We need to sample the length of the consultation within the trajectory and use this value in two places: within `log_()` for reporting and in `timeout()` to control the duration.

To achieve this, we use simmer's `get_attribute()` to generate and store the value as an attribute. We then access this value later using `get_attribute()`.

<br>

```{.r}
log_("🚪 Leaves.", level = 1L)
```

Finally, we create a log message to record when the patient leaves.

### Run the model

```{r}
param <- create_params()
model(param = param, run_number = 1L)
```

<br>

If we run the model with `verbose = FALSE` in parameters, no logs are printed:

```{r}
param <- create_params(verbose = FALSE)
model(param = param, run_number = 1L)
```

:::

## Adding parameters to the logs

As mentioned on the [Parameters from script](../inputs/parameters_script.qmd) page, one of the benefits of storing parameters in a single object is that we can easily print a copy of all parameters used for a scenario by fetching the object.

::: {.python-content}

### Parameter class

No changes required.

### Patient class

No changes required.

### Logging class

We add a new method `sanitise_object()` to `SimLogger` which tidies up the provided parameter object so it is easily readable and doesn't contain memory addresses.

In Python, when you print objects directly, the log output often shows their memory address (e.g., `<simpy.core.Environment object at 0x752feb454d50>`), which is a reference to where the object is stored in the computer's memory and not relevant for us to include in our log.

The `sanitise_object()` method is applied in `log()` for when the provided `msg` is a dictionary.

```{python}
class SimLogger:
    """
    Provides log of events as the simulation runs.

    Attributes
    ----------
    verbose : bool
        Whether to print messages as simulation runs.
    """
    def __init__(self, verbose):
        """
        Initialise logging class.

        Parameters
        ----------
        verbose : bool
            Whether to print messages as simulation runs.
        """
        self.verbose = verbose
        if self.verbose:
            self.logger = logging.getLogger(__name__)
            self._configure_logging()

    def _configure_logging(self):
        """
        Configure the logger.
        """
        # Ensure any existing handlers are removed to avoid duplication
        for handler in self.logger.handlers[:]:
            self.logger.removeHandler(handler)

        # Configure RichHandler without INFO/ERROR labels, times or paths
        # to log message. Set up console with jupyter-specific behaviour
        # disabled to prevent large gaps between each log message on ipynb.
        console = Console()
        console.is_jupyter = False
        rich_handler = RichHandler(
            console=console, show_time=False,
            show_level=False, show_path=False
        )

        # Add handler to the logger
        self.logger.addHandler(rich_handler)

        # Set logging level and format. If don't set level info, it would
        # only show log messages which are warning, error or critical.
        self.logger.setLevel(logging.INFO)
        formatter = logging.Formatter("%(message)s")
        rich_handler.setFormatter(formatter)

    def sanitise_object(self, obj):#<<
        """#<<
        Sanitise object references to avoid memory addresses in logs.#<<

        Parameters#<<
        ----------#<<
        obj : object#<<
            Object to sanitise.#<<

        Returns#<<
        -------#<<
        str#<<
            Sanitised version of the object. For basic types (int, float,#<<
            bool, str, list, dict, tuple, set), returns the string#<<
            representation. For other objects, returns the class name in#<<
            angle brackets.#<<
        """#<<
        if isinstance(obj, object) and not isinstance(#<<
            obj, (int, float, bool, str, list, dict, tuple, set)#<<
        ):#<<
            # Return the class name instead of the memory address#<<
            return f"<{obj.__class__.__module__}.{obj.__class__.__name__}>"#<<
        return obj#<<

    def log(self, msg, sim_time=None):
        """
        Log a message if logging is enabled.

        Parameters
        ----------
        msg : str
            Message to log.
        sim_time : float or None, optional
            Current simulation time. If provided, prints before message.
        """
        # Format dictionaries so easier to read#<<
        if isinstance(msg, dict):#<<
            msg = {key: self.sanitise_object(value)#<<
                   for key, value in msg.items()}#<<
            msg = pformat(msg, indent=4)#<<

        if self.verbose:
            if sim_time is not None:
                self.logger.info("%0.3f: %s", sim_time, msg)
            else:
                self.logger.info(msg)
```

### Model class

Within `Model.__init__()`, we now provide the attributes of `Model` and our parameter object to the log.

```{python}
class Model:
    """
    Simulation model.

    Attributes
    ----------
    param : Parameters
        Simulation parameters.
    run_number : int
        Run number for random seed generation.
    env : simpy.Environment
        The SimPy environment for the simulation.
    doctor : simpy.Resource
        SimPy resource representing doctors.
    patients : list
        List of Patient objects.
    arrival_dist : Exponential
        Distribution used to generate random patient inter-arrival times.
    consult_dist : Exponential
        Distribution used to generate length of a doctor's consultation.
    logger : SimLogger
        The logging instance used for logging messages.
    """
    def __init__(self, param, run_number):
        """
        Create a new Model instance.

        Parameters
        ----------
        param : Parameters
            Simulation parameters.
        run_number : int
            Run number for random seed generation.
        """
        self.param = param
        self.run_number = run_number

        # Create SimPy environment
        self.env = simpy.Environment()

        # Create resource
        self.doctor = simpy.Resource(
            self.env, capacity=self.param.number_of_doctors
        )

        # Create a random seed sequence based on the run number
        ss = np.random.SeedSequence(self.run_number)
        seeds = ss.spawn(2)

        # Set up attributes to store results
        self.patients = []

        # Initialise distributions
        self.arrival_dist = Exponential(mean=self.param.interarrival_time,
                                        random_seed=seeds[0])
        self.consult_dist = Exponential(mean=self.param.consultation_time,
                                        random_seed=seeds[1])

        # Initialise logger
        self.logger = SimLogger(verbose=self.param.verbose)
        self.logger.log(sim_time=self.env.now, msg="Initialise model:")#<<
        self.logger.log(vars(self))#<<
        self.logger.log(sim_time=self.env.now, msg="Parameters:")#<<
        self.logger.log(vars(self.param))#<<

    def generate_arrivals(self):
        """
        Process that generates patient arrivals.
        """
        while True:
            # Sample and pass time to next arrival
            sampled_iat = self.arrival_dist.sample()
            yield self.env.timeout(sampled_iat)

            # Create a new patient
            patient = Patient(patient_id=len(self.patients)+1,
                              arrival_time=self.env.now)
            self.patients.append(patient)

            # Log arrival time
            self.logger.log(msg=f"Patient {patient.patient_id} arrives.",
                            sim_time=self.env.now)

            # Start process of consultation
            self.env.process(self.consultation(patient))

    def consultation(self, patient):
        """
        Process that simulates a consultation.

        Parameters
        ----------
        patient :
            Instance of the Patient() class representing a single patient.
        """
        # Patient requests access to a doctor (resource)
        with self.doctor.request() as req:
            yield req

            # Log consultation start time
            self.logger.log(
                msg=f"Patient {patient.patient_id} starts consultation.",
                sim_time=self.env.now
            )

            # Sample consultation duration and pass time spent with doctor
            time_with_doctor = self.consult_dist.sample()
            yield self.env.timeout(time_with_doctor)

            # Log end time
            self.logger.log(msg=f"Patient {patient.patient_id} leaves.",
                            sim_time=self.env.now)

    def run(self):
        """
        Run the simulation.
        """
        # Schedule arrival generator
        self.env.process(self.generate_arrivals())

        # Run the simulation
        self.env.run(until=self.param.run_length)
```

### Run the model

```{python}
param = Parameters()
model = Model(param=param, run_number=0)
model.run()
```

:::

::: {.r-content}

### Parameter function

No changes required.

### Model function

The `log_()` function is designed to be used within simmer trajectories. To record the model's parameter set as part of the simmer log, we create a simple trajectory `param_traj` containing a single `log_()` action. Here, the parameters are converted into string format and logged.

This is add to the simulation environment using `add_generator()` with `at(0)` ensuring that it just triggers once at the very start of the simulation.

```{r}
#' Run simulation.
#'
#' @param param List. Model parameters.
#' @param run_number Numeric. Run number for random seed generation.
#'
#' @importFrom simmer add_generator add_resource get_attribute log_ release run
#' @importFrom simmer seize set_attribute simmer timeout trajectory
#'
#' @export

model <- function(param, run_number) {

  # Set random seed based on run number
  set.seed(run_number)

  # Create simmer environment
  env <- simmer("simulation", verbose = FALSE,
                log_level = if (param[["verbose"]]) 1L else 0L)

  # Add blank trajectory used to add parameters to the log#<<
  param_string <- paste(names(param), param, sep = "=", collapse = ";\n ")#<<
  param_traj <- trajectory() |> #<<
    log_(param_string, level = 1L)#<<

  # Define the patient trajectory
  patient <- trajectory("consultation") |>
    log_("🚶 Arrives.", level = 1L) |>
    seize("doctor", 1L) |>
    # Record sampled length of consultation
    set_attribute("doctor_serve_length", function() {
      rexp(n = 1L, rate = 1L / param[["consultation_time"]])
    }) |>
    log_(function() {
      paste0("🩺 Doctor consultation begins (length: ",
             round(get_attribute(env, "doctor_serve_length"), 5L), ")")
    }, level = 1L) |>
    timeout(function() get_attribute(env, "doctor_serve_length")) |>
    release("doctor", 1L) |>
    log_("🚪 Leaves.", level = 1L)

  env <- env |>
    # Add parameter log#<<
    add_generator("Parameters", param_traj, at(0L)) |> #<<
    # Add doctor resource
    add_resource("doctor", param[["number_of_doctors"]]) |>
    # Add patient generator
    add_generator("patient", patient, function() {
      rexp(n = 1L, rate = 1L / param[["interarrival_time"]])
    }) |>
    # Run the simulation
    simmer::run(until = param[["run_length"]])

}
```

### Run the model

```{r}
param <- create_params()
model(param = param, run_number = 1L)
```

:::

## Saving logs to a file

Saving simulation logs to a file is very handy for reproducibility because it creates a permanent, detailed record of exactly what happened during the simulation run. This can be useful for debugging, peer review and similar.

::: {.python-content}

### Parameter class

Our argument `verbose` has been replace by three new arguments: `log_to_console`, `log_to_file` and `file_path`...

```{python}
class Parameters:
    """
    Parameter class.

    Attributes
    ----------
    interarrival_time : float
        Mean time between arrivals (minutes).
    consultation_time : float
        Mean length of doctor's consultation (minutes).
    number_of_doctors : int
        Number of doctors.
    run_length : int
        Total duration of simulation (minutes).
    log_to_console : bool#<<
        Whether to print activity log to console.#<<
    log_to_file : bool#<<
        Whether to save activity log to file.#<<
    file_path : str#<<
        Path to save log to file.#<<
    """
    def __init__(
        self, interarrival_time=5, consultation_time=10,
        number_of_doctors=3, run_length=50,
        log_to_console=False, log_to_file=False, file_path=None#<<
    ):
        """
        Initialise Parameters instance.

        Parameters
        ----------
        interarrival_time : float
            Time between arrivals (minutes).
        consultation_time : float
            Length of consultation (minutes).
        number_of_doctors : int
            Number of doctors.
        run_length : int
            Total duration of simulation (minutes).
        log_to_console : bool#<<
            Whether to print activity log to console.#<<
        log_to_file : bool#<<
            Whether to save activity log to file.#<<
        file_path : str#<<
            Path to save log to file.#<<
        """
        self.interarrival_time = interarrival_time
        self.consultation_time = consultation_time
        self.number_of_doctors = number_of_doctors
        self.run_length = run_length
        self.log_to_console = log_to_console#<<
        self.log_to_file = log_to_file#<<
        self.file_path = file_path#<<
```

### Patient class

No changes required.

### Logging class

The new parameter arguments `log_to_console`, `log_to_file` and `file_path` are passed to the `SimLogger` class. Where previously the class checked if verbose was true, it now checks whether log to console or file are true.

The class has a new `_validate_log_path()` which checks that the provided path is suitable (i.e., it exists and the filename ends with `.log`).

Within `_configure_logging()`, it either adds:

* The `RichHandler` (as before) if `log_to_console` is True.
* A `FileHandler` if `log_to_file` is True.

These are both add to `logger` and their logging level and format are set.

```{python}
class SimLogger:
    """
    Provides log of events as the simulation runs.

    Attributes
    ----------
    log_to_console : bool#<<
        Whether to print activity log to console.#<<
    log_to_file : bool#<<
        Whether to save activity log to file.#<<
    file_path : str#<<
        Path to save log to file.#<<
    """
    def __init__(self, log_to_console, log_to_file, file_path):#<<
        """
        Initialise logging class.

        Parameters
        ----------
        log_to_console : bool#<<
            Whether to print activity log to console.#<<
        log_to_file : bool#<<
            Whether to save activity log to file.#<<
        file_path : str#<<
            Path to save log to file.#<<
        """
        self.log_to_console = log_to_console#<<
        self.log_to_file = log_to_file#<<
        self.file_path = file_path#<<

        # If saving to file, check path is valid#<<
        if self.log_to_file:#<<
            self._validate_log_path()#<<

        if self.log_to_console or self.log_to_file:#<<
            self.logger = logging.getLogger(__name__)
            self._configure_logging()

    def _validate_log_path(self):#<<
        """#<<
        Validate the log file path.#<<

        Raises#<<
        ------#<<
        ValueError#<<
            If log path is invalid.#<<
        """
        directory = os.path.dirname(self.file_path)#<<
        if directory and not os.path.exists(directory):#<<
            raise ValueError(#<<
                f"The directory '{directory}' for the log file does not exist."#<<
            )#<<
        if not self.file_path.endswith(".log"):#<<
            raise ValueError(#<<
                f"The log file path '{self.file_path}' must end with '.log'."#<<
            )#<<

    def _configure_logging(self):
        """
        Configure the logger.
        """
        # Ensure any existing handlers are removed to avoid duplication
        for handler in self.logger.handlers[:]:
            self.logger.removeHandler(handler)

        # Configure RichHandler without INFO/ERROR labels, times or paths
        # to log message. Set up console with jupyter-specific behaviour
        # disabled to prevent large gaps between each log message on ipynb.
        console = Console()
        console.is_jupyter = False
        rich_handler = RichHandler(
            console=console, show_time=False,
            show_level=False, show_path=False
        )

        # Add handlers to the logger#<<
        handlers = []#<<
        if self.log_to_console:#<<
            handlers.append(rich_handler)#<<
        if self.log_to_file:#<<
            handlers.append(logging.FileHandler(self.file_path, mode="w"))#<<
        for handler in handlers:#<<
            self.logger.addHandler(handler)#<<

        # Set logging level and format. If don't set level info, it would
        # only show log messages which are warning, error or critical.
        self.logger.setLevel(logging.INFO)
        formatter = logging.Formatter("%(message)s")
        for handler in handlers:#<<
            handler.setFormatter(formatter)#<<

    def sanitise_object(self, obj):
        """
        Sanitise object references to avoid memory addresses in logs.

        Parameters
        ----------
        obj : object
            Object to sanitise.

        Returns
        -------
        str
            Sanitised version of the object. For basic types (int, float,
            bool, str, list, dict, tuple, set), returns the string
            representation. For other objects, returns the class name in
            angle brackets.
        """
        if isinstance(obj, object) and not isinstance(
            obj, (int, float, bool, str, list, dict, tuple, set)
        ):
            # Return the class name instead of the memory address
            return f"<{obj.__class__.__module__}.{obj.__class__.__name__}>"
        return obj

    def log(self, msg, sim_time=None):
        """
        Log a message if logging is enabled.

        Parameters
        ----------
        msg : str
            Message to log.
        sim_time : float or None, optional
            Current simulation time. If provided, prints before message.
        """
        # Format dictionaries so easier to read
        if isinstance(msg, dict):
            msg = {key: self.sanitise_object(value)
                   for key, value in msg.items()}
            msg = pformat(msg, indent=4)

        if self.log_to_console or self.log_to_file:#<<
            if sim_time is not None:
                self.logger.info("%0.3f: %s", sim_time, msg)
            else:
                self.logger.info(msg)
```

### Model class

The new Parameter attributes (`log_to_console`, `log_to_file`, `file_path`) are passed to the logger.

```{python}
class Model:
    """
    Simulation model.

    Attributes
    ----------
    param : Parameters
        Simulation parameters.
    run_number : int
        Run number for random seed generation.
    env : simpy.Environment
        The SimPy environment for the simulation.
    doctor : simpy.Resource
        SimPy resource representing doctors.
    patients : list
        List of Patient objects.
    arrival_dist : Exponential
        Distribution used to generate random patient inter-arrival times.
    consult_dist : Exponential
        Distribution used to generate length of a doctor's consultation.
    logger : SimLogger
        The logging instance used for logging messages.
    """
    def __init__(self, param, run_number):
        """
        Create a new Model instance.

        Parameters
        ----------
        param : Parameters
            Simulation parameters.
        run_number : int
            Run number for random seed generation.
        """
        self.param = param
        self.run_number = run_number

        # Create SimPy environment
        self.env = simpy.Environment()

        # Create resource
        self.doctor = simpy.Resource(
            self.env, capacity=self.param.number_of_doctors
        )

        # Create a random seed sequence based on the run number
        ss = np.random.SeedSequence(self.run_number)
        seeds = ss.spawn(2)

        # Set up attributes to store results
        self.patients = []

        # Initialise distributions
        self.arrival_dist = Exponential(mean=self.param.interarrival_time,
                                        random_seed=seeds[0])
        self.consult_dist = Exponential(mean=self.param.consultation_time,
                                        random_seed=seeds[1])

        # Initialise logger
        self.logger = SimLogger(log_to_console=self.param.log_to_console,#<<
                                log_to_file=self.param.log_to_file,#<<
                                file_path=self.param.file_path)#<<
        self.logger.log(sim_time=self.env.now, msg="Initialise model:")
        self.logger.log(vars(self))
        self.logger.log(sim_time=self.env.now, msg="Parameters:")
        self.logger.log(vars(self.param))

    def generate_arrivals(self):
        """
        Process that generates patient arrivals.
        """
        while True:
            # Sample and pass time to next arrival
            sampled_iat = self.arrival_dist.sample()
            yield self.env.timeout(sampled_iat)

            # Create a new patient
            patient = Patient(patient_id=len(self.patients)+1,
                              arrival_time=self.env.now)
            self.patients.append(patient)

            # Log arrival time
            self.logger.log(msg=f"Patient {patient.patient_id} arrives.",
                            sim_time=self.env.now)

            # Start process of consultation
            self.env.process(self.consultation(patient))

    def consultation(self, patient):
        """
        Process that simulates a consultation.

        Parameters
        ----------
        patient :
            Instance of the Patient() class representing a single patient.
        """
        # Patient requests access to a doctor (resource)
        with self.doctor.request() as req:
            yield req

            # Log consultation start time
            self.logger.log(
                msg=f"Patient {patient.patient_id} starts consultation.",
                sim_time=self.env.now
            )

            # Sample consultation duration and pass time spent with doctor
            time_with_doctor = self.consult_dist.sample()
            yield self.env.timeout(time_with_doctor)

            # Log end time
            self.logger.log(msg=f"Patient {patient.patient_id} leaves.",
                            sim_time=self.env.now)

    def run(self):
        """
        Run the simulation.
        """
        # Schedule arrival generator
        self.env.process(self.generate_arrivals())

        # Run the simulation
        self.env.run(until=self.param.run_length)
```

### Run the model

With `log_to_console` and `log_to_file` both set to `False`, no logs are generated.

```{python}
param = Parameters()
model = Model(param=param, run_number=0)
model.run()
```

<br>

If we set `log_to_console=True`, the log is printed as before.

```{python}
param = Parameters(log_to_console=True)
model = Model(param=param, run_number=0)
model.run()
```

<br>

If we set `log_to_file=True` and provide a `file_path`, a file is generated containing the log output.

```{python}
log_file = os.path.join("logs_resources", "python_log_example.log")
param = Parameters(log_to_file=True, file_path=log_file)
model = Model(param=param, run_number=0)
model.run()
```

If we import the log file, we'll see it contains the same output.

```{python}
# Read all lines from the file
with open(log_file, "r", encoding="utf-8") as file:
    log_contents = file.readlines()

# Print each line
for line in log_contents:
    print(line, end="")
```

:::

::: {.r-content}

### Parameter function

Our argument `verbose` has been replace by three new arguments: `log_to_console`, `log_to_file` and `file_path`...

```{r}
#' Generate parameter list.
#'
#' @param interarrival_time Numeric. Time between arrivals (minutes).
#' @param consultation_time Numeric. Mean length of doctor's
#'   consultation (minutes).
#' @param number_of_doctors Numeric. Number of doctors.
#' @param run_length Numeric. Total duration of simulation (minutes).
#' @param log_to_console Whether to print activity log to console.#<<
#' @param log_to_file Whether to save activity log to file.#<<
#' @param file_path Path to save log to file.#<<
#'
#' @return A named list of parameters.
#' @export

create_params <- function(
  interarrival_time = 5L,
  consultation_time = 10L,
  number_of_doctors = 3L,
  run_length = 50L,
  log_to_console = FALSE, #<<
  log_to_file = FALSE, #<<
  file_path = NULL#<<
) {
  list(
    interarrival_time = interarrival_time,
    consultation_time = consultation_time,
    number_of_doctors = number_of_doctors,
    run_length = run_length,
    log_to_console = log_to_console, #<<
    log_to_file = log_to_file, #<<
    file_path = file_path#<<
  )
}
```

### Model function

The `model()` function now determines whether `verbose` is True based on `log_to_console` and `log_to_file` (checking if either are True).

So that we are able to save the log messages, we use `capture.output` to store the logging outputs when our simulation is run into an object called `sim_log`. This can then be printed to the console or saved to a file.

```{r}
#' Run simulation.
#'
#' @param param List. Model parameters.
#' @param run_number Numeric. Run number for random seed generation.
#'
#' @importFrom simmer add_generator add_resource get_attribute log_ release run
#' @importFrom simmer seize set_attribute simmer timeout trajectory
#'
#' @export

model <- function(param, run_number) {

  # Set random seed based on run number
  set.seed(run_number)

  # Determine whether to get verbose activity logs#<<
  verbose <- any(c(param[["log_to_console"]], param[["log_to_file"]]))#<<

  # Create simmer environment
  env <- simmer("simulation", verbose = FALSE,
                log_level = if (verbose) 1L else 0L)#<<

  # Add blank trajectory used to add parameters to the log
  param_string <- paste(names(param), param, sep = "=", collapse = ";\n ")
  param_traj <- trajectory() |>
    log_(param_string, level = 1L)

  # Define the patient trajectory
  patient <- trajectory("consultation") |>
    log_("🚶 Arrives.", level = 1L) |>
    seize("doctor", 1L) |>
    # Record sampled length of consultation
    set_attribute("doctor_serve_length", function() {
      rexp(n = 1L, rate = 1L / param[["consultation_time"]])
    }) |>
    log_(function() {
      paste0("🩺 Doctor consultation begins (length: ",
             round(get_attribute(env, "doctor_serve_length"), 5L), ")")
    }, level = 1L) |>
    timeout(function() get_attribute(env, "doctor_serve_length")) |>
    release("doctor", 1L) |>
    log_("🚪 Leaves.", level = 1L)

  sim_log <- capture.output(#<<
    env <- env |> # nolint
      # Add parameter log
      add_generator("Parameters", param_traj, at(0L)) |>
      # Add doctor resource
      add_resource("doctor", param[["number_of_doctors"]]) |>
      # Add patient generator
      add_generator("patient", patient, function() {
        rexp(n = 1L, rate = 1L / param[["interarrival_time"]])
      }) |>
      # Run the simulation
      simmer::run(until = param[["run_length"]])
  )#<<

  # Print log to console#<<
  if (isTRUE(param[["log_to_console"]])) {#<<
    print(sim_log)#<<
  }#<<
  # Save log to file#<<
  if (isTRUE(param[["log_to_file"]])) {#<<
    writeLines(sim_log, param[["file_path"]])#<<
  }#<<
}
```

### Run the model

With `log_to_console` and `log_to_file` both set to `FALSE`, no logs are generated.

```{r}
param <- create_params()
model(param = param, run_number = 1L)
```

<br>

If we set `log_to_console = TRUE`, the log is printed as before.

```{r}
param <- create_params(log_to_console = TRUE)
model(param = param, run_number = 1L)
```

<br>

If we set `log_to_file = TRUE` and provide a `file_path`, a file is generated containing the log output.

```{r}
log_file <- file.path("logs_resources", "r_log_example.log")
param <- create_params(log_to_file = TRUE, file_path = log_file)
model(param = param, run_number = 1L)
```

If we import the log file, we'll see it contains the same output:

```{r}
log_contents <- readLines(log_file)
print(log_contents, sep = "\n")
```

:::

## Explore the example models

<div class="h3-tight"></div>

### Nurse visit simulation

:::: {.python-content}

{{< include /html/pydesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `notebooks/logs.ipynb`<br>`simulation/simlogger.py`<br>`simulation/param.py`<br>`simulation/model.py` |
| **What to look for?** | Using `SimLogger` with a similar set-up to above. |
: {tbl-colwidths="[20,80]"}
:::

::::

:::: {.r-content}

{{< include /html/rdesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `rmarkdown/logs.md` |
| **What to look for?** | Creates custom logs (not with parameters or saved to file - kept simple). |
| **Why it matters?** | In this example, we've kept the model simpler and stuck with the default simmer log message. Instead, custom logs are demonstrated in an rmarkdown file. |
: {tbl-colwidths="[20,80]"}
:::

::::

### Stroke pathway simulation

:::: {.python-content}

{{< include /html/pydesrapstroke.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `notebooks/logs.ipynb`<br>`simulation/logging.py`<br>`simulation/parameters.py`<br>`simulation/model.py` |
| **What to look for?** | Using `SimLogger` with a similar set-up to above. |
: {tbl-colwidths="[20,80]"}
:::

::::

:::: {.r-content}

{{< include /html/rdesrapstroke.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `rmarkdown/logs.md`<br>`R/model.R`<br>`R/parameters.R`<br>`R/create_asu_trajectory.R`<br>`R/create_rehab_trajectory.R` |
| **What to look for?** | Custom logs are implemented like above, but with parameters appended to `sim_log` rather than being a trajectory and part of `log_()`. |
| **Why it matters?** | Custom logs are part of the model for this example. They're set up a little differently - indeed, there are many ways you could do this! |
: {tbl-colwidths="[20,80]"}
:::

::::

## Test yourself

Try **adding logs to your model**. Experiment with different log messages and parameter options, then run your code and observe how the logs reflect your changes at each step.

If you extend your model (for example, with different patient types, destinations, or additional resources), continue making use of logs. Logging is especially valuable during **model development** - it helps you check that each part of your logic works as intended and makes it easier to troubleshoot issues as your simulation grows in complexity.

<br><br>
