---
title: Mathematical proof of correctness
filters:
  - line-highlight
date: "2025-10-23T16:24:23+01:00"
---

```{python}
#| echo: false
# pylint: disable=unused-import,undefined-variable,wrong-import-position
```

:::: {.pale-blue}

**Learning objectives:**

* Understand why and how to verify a model using mathematical proof of correctness.

**Pre-reading:**

This page implements mathematical proof of correctness implemented on the [verification and validation page](verification_validation.qmd). It does so within a test ([tests](tests.qmd) page).

The test is run on the model from the [parallel processing](../output_analysis/parallel.qmd) page (likewise, also used on the [scenario and sensitivity analysis](../experiments/scenarios.qmd) page and [tests](tests.qmd) pages).

> *Entity generation → Entity processing → Initialisation bias → Performance measures → Replications → Parallel processing → Mathematical proof of correctness*

::::

:::: {.very-pale-blue}

**Required packages:**

The following imports are required. These should be available from environment setup on the [Structuring as a package](/pages/guide/setup/package.qmd) page.

Since [Parallel processing](/pages/guide/output_analysis/parallel.qmd), we have one new imports: `math`.

::: {.python-content}

```{python}
# For display purposes:
from IPython.display import HTML
from itables import to_html_datatable

# For the simulation modelling and analysis:
import math#<<
import time
from joblib import Parallel, delayed, cpu_count
import numpy as np
import pandas as pd
import plotly.express as px
import scipy.stats as st
import simpy
from sim_tools.distributions import Exponential
```

```{python}
#| echo: false
# To ensure renders correctly in quarto
import plotly.io as pio  # pylint: disable=ungrouped-imports
pio.renderers.default = "plotly_mimetype"
```

:::

::: {.r-content}

```{r}
#| output: false
library(data.table)
library(dplyr)
library(future)
library(future.apply)
library(ggplot2)
library(knitr)
library(plotly)
library(queueing)
library(simmer)
library(tidyr)
```

:::

::::

## What is mathematical proof of correctness?

One of the suggested methods for [model verification](verification_validation.qmd) is **mathematical proof of correctness**. This involves showing that the simulation results align with those determined by a suitable **mathematical model**.

A mathematical model is a set of equations or formulas that describe how systems behave. Over time, researchers have studied simple systems (like people waiting in a queue) and discovered that, when rules are straightforward, their behaviour can be described exactly using math.

Real-world systems are often too complex for neat equations (e.g., variable arrivals, different service rules or other complex features). In these cases, discrete-event simulation allows us to better mimic the system. However, if your simulation is simple enough - or if you start with a basic model to verify your code (and later build in complexity) - you can use mathematical proof of correctness.

The small example we've been building throughout this book is an instance of an **M/M/s queue model** (as is the nurse visit simulation example). For a reminder on what an M/M/s queue model is, see the section "*Find out more about M/M/s models*" on the [example conceptual models](/pages/example_models/example_models.qmd) page. In the following sections, we'll show how to verify our small example by comparing it to a mathematical model.

## Simulation code

The model used is the same as that on the [parallel processing](../output_analysis/parallel.qmd) page - as also used on the [scenario and sensitivity analysis](../experiments/scenarios.qmd) and [tests](tests.qmd) pages.

```{python}
#| file: "tests_resources/simulation.py"
#| echo: false
```

```{r}
#| file: "tests_resources/simulation.R"
#| echo: false
```


## Mathematical M/M/s queue model

::: {.python-content}

The class implementing a mathematical M/M/s queue model is explained line-by-line below.

```{python}
class MMSQueue:
    """
    Analytical M/M/S queue formulas.

    Parameters
    ----------
    arrival_rate : float
        Customer arrival rate (λ).
    service_rate : float
        Service rate per server (μ).
    num_servers : int
        Number of servers (s).

    Attributes
    ----------
    rho : float
        Utilisation (λ / (sμ)).
    lambda_over_mu : float
        Arrival/service rate ratio (λ / μ).
    metrics : dict
        Calculated performance metrics.
    """

    def __init__(self, arrival_rate, service_rate, num_servers):
        """
        Initialise the M/M/S queue.

        Raises
        ------
        ValueError
            If parameters are invalid or system is unstable.
        """
        if arrival_rate <= 0:
            raise ValueError("Arrival rate must be positive")
        if service_rate <= 0:
            raise ValueError("Service rate must be positive")
        if num_servers < 1:
            raise ValueError("Number of servers must be at least 1")

        self.arrival_rate = arrival_rate
        self.service_rate = service_rate
        self.num_servers = num_servers

        # Calculate utilisation
        self.rho = self.get_traffic_intensity()

        # Check system stability
        if self.rho >= 1:
            raise ValueError(
                f"System is unstable: ρ = {self.rho:.4f} >= 1. "
                f"Need λ < s*μ ({arrival_rate} < {num_servers * service_rate})"
            )

        # Calculate λ/μ (average customers in service if infinite servers)
        self.lambda_over_mu = self.arrival_rate / self.service_rate

        # Calculate performance metrics using Little's Law
        self.metrics = self.calculate_metrics()

    def get_traffic_intensity(self):
        """
        Calculate the traffic intensity (server utilisation).

        Returns
        -------
        float
            Traffic intensity ρ = λ/(s*μ).
        """
        return self.arrival_rate / (self.num_servers * self.service_rate)

    def calculate_metrics(self):
        """
        Calculate all performance metrics for the queue.

        Returns
        -------
        dict[str, float]
            Dictionary containing performance metrics.
        """
        metrics = {}
        metrics["rho"] = self.rho
        metrics["L_q"] = self.get_mean_queue_length()
        metrics["L_s"] = metrics["L_q"] + (
            self.arrival_rate / self.service_rate
        )
        metrics["W_s"] = metrics["L_s"] / self.arrival_rate
        metrics["W_q"] = metrics["W_s"] - (1 / self.service_rate)
        return metrics

    def get_mean_queue_length(self):
        """
        Calculate the expected number of customers waiting in queue (L_q).

        Uses the formula:
        L_q = P₀ * (λ/μ)^s * ρ / (s! * (1-ρ)²)

        Returns
        -------
        float
            Expected queue length.
        """
        p0 = self.prob_system_empty()

        lq = (p0 * (self.lambda_over_mu**self.num_servers) * self.rho) / (
            math.factorial(self.num_servers) * (1 - self.rho) ** 2
        )

        return lq

    def prob_system_empty(self):
        """
        Calculate the probability that the system is empty (P₀).

        Uses the formula:
        P₀ = [Σ(n=0 to s-1) (λ/μ)^n/n! + (λ/μ)^s/(s!(1-ρ))]^(-1)

        Returns
        -------
        float
            Probability that system is empty.
        """
        # Sum for n = 0 to s-1
        sum_part = sum(
            (self.lambda_over_mu**n) / math.factorial(n)
            for n in range(self.num_servers)
        )

        # Term for n >= s
        server_term = (self.lambda_over_mu**self.num_servers) / (
            math.factorial(self.num_servers) * (1 - self.rho)
        )

        return 1 / (sum_part + server_term)
```

### Explaining `MMSQueue`

```{.python}
class MMSQueue:
    """
    Analytical M/M/S queue formulas.

    Parameters
    ----------
    arrival_rate : float
        Customer arrival rate (λ).
    service_rate : float
        Service rate per server (μ).
    num_servers : int
        Number of servers (s).

    Attributes
    ----------
    rho : float
        Utilisation (λ / (sμ)).
    lambda_over_mu : float
        Arrival/service rate ratio (λ / μ).
    metrics : dict
        Calculated performance metrics.
    """

    def __init__(self, arrival_rate, service_rate, num_servers):
        """
        Initialise the M/M/S queue.

        Raises
        ------
        ValueError
            If parameters are invalid or system is unstable.
        """
        if arrival_rate <= 0:
            raise ValueError("Arrival rate must be positive")
        if service_rate <= 0:
            raise ValueError("Service rate must be positive")
        if num_servers < 1:
            raise ValueError("Number of servers must be at least 1")

        self.arrival_rate = arrival_rate
        self.service_rate = service_rate
        self.num_servers = num_servers
```

`MMSQueue` requires three inputs:

* `arrival_rate` (λ) - average number of customers (patients) per time unit (minute).
* `service_rate` (μ) - average number of customers (patients) a single server (doctor) can serve per time unit (minute).
* `num_servers` (s) - number of servers (doctors) in the system.

It checks the inputs are valid (no negative or zero rates, at least one server) and stores them as attributes.

<br>

```{.python}
        # Calculate utilisation
        self.rho = self.get_traffic_intensity()

        # Check system stability
        if self.rho >= 1:
            raise ValueError(
                f"System is unstable: ρ = {self.rho:.4f} >= 1. "
                f"Need λ < s*μ ({arrival_rate} < {num_servers * service_rate})"
            )
```

It then calculates **utilisation (ρ)** by calling the `get_traffic_intensity()` method. Utilisation is a measure of how busy the servers are - and if it is close to 1, it means the servers are almost always busy.

If the calculation utilisation is **greater than 1**, it means that customers arrive faster than the servers can serve them. In this situation, the **queue would keep growing indefinitely**. There is no steady state, and the system is unstable. When this happens, the model stops and raises an error instead of producing meaningless results.

> **Note:** This behaviour is different from a discrete-event simulation model, where utilisation is capped at 1 (since servers cannot be more than fully busy). However, you can still have a queue that grows indefinitely - you would just notice it by looking at other metrics like queue length or waiting time, and see them continuing to increase over time.

<br>

```{.python}
        # Calculate λ/μ (average customers in service if infinite servers)
        self.lambda_over_mu = self.arrival_rate / self.service_rate

        # Calculate performance metrics using Little's Law
        self.metrics = self.calculate_metrics()
```

Next, we calculate `lambda_over_mu`, which represents the average number of customers being served if there were infinite servers. It's computed here to avoid code repetition, as it is used repeatedly in later formulas.

After that, the model calls `calculate_metrics()`.

<br>

```{.python}
    def get_traffic_intensity(self):
        """
        Calculate the traffic intensity (server utilisation).

        Returns
        -------
        float
            Traffic intensity ρ = λ/(s*μ).
        """
        return self.arrival_rate / (self.num_servers * self.service_rate)
```

The method `get_traffic_intensity()` (as called in `__init__()` computes utilisation by finding the fraction of total capacity being used.

<br>

```{.python}
    def calculate_metrics(self):
        """
        Calculate all performance metrics for the queue.

        Returns
        -------
        dict[str, float]
            Dictionary containing performance metrics.
        """
        metrics = {}
        metrics["rho"] = self.rho
        metrics["L_q"] = self.get_mean_queue_length()
        metrics["L_s"] = metrics["L_q"] + (
            self.arrival_rate / self.service_rate
        )
        metrics["W_s"] = metrics["L_s"] / self.arrival_rate
        metrics["W_q"] = metrics["W_s"] - (1 / self.service_rate)
        return metrics
```

The `calculate_metrics()` method uses several analytical formulas to compute the main queue performance measures.

| Symbol | Meaning | Formula |
| - | - | - |
| ρ	| Utilisation | λ / (sμ) |
| Lq | Average queue length | Computed using `get_mean_queue_length()` |
| Ls | Average number in system | Lq + (λ/μ) |
| Ws | Average time in system | Ls / λ |
| Wq | Average waiting time in queue | Ws − (1/μ) |

All results are returned in a dictionary.

> **Fun fact:** The formula for average time in system makes use of **Little's Law**, which Ls = λ * Ws. This is rearranged to calculate time in system: Ws = Ls / λ.

<br>

```{.python}
    def get_mean_queue_length(self):
        """
        Calculate the expected number of customers waiting in queue (L_q).

        Uses the formula:
        L_q = P₀ * (λ/μ)^s * ρ / (s! * (1-ρ)²)

        Returns
        -------
        float
            Expected queue length.
        """
        p0 = self.prob_system_empty()

        lq = (p0 * (self.lambda_over_mu**self.num_servers) * self.rho) / (
            math.factorial(self.num_servers) * (1 - self.rho) ** 2
        )

        return lq
```

As the calculation for mean queue length is a little more complicated, we have a separate method to calculate it.

First, we compute P0 which is the probability that the system is completely empty (using `prob_system_empty()`). We then plug this into the formula to calculate Lq (the mean queue length).

<br>

```{.python}
    def prob_system_empty(self):
        """
        Calculate the probability that the system is empty (P₀).

        Uses the formula:
        P₀ = [Σ(n=0 to s-1) (λ/μ)^n/n! + (λ/μ)^s/(s!(1-ρ))]^(-1)

        Returns
        -------
        float
            Probability that system is empty.
        """
        # Sum for n = 0 to s-1
        sum_part = sum(
            (self.lambda_over_mu**n) / math.factorial(n)
            for n in range(self.num_servers)
        )

        # Term for n >= s
        server_term = (self.lambda_over_mu**self.num_servers) / (
            math.factorial(self.num_servers) * (1 - self.rho)
        )

        return 1 / (sum_part + server_term)
```

This method calculates P0, which is the probability that the system is completely empty - i.e., no customers are waiting and all servers are idle.

:::

::: {.r-content}

In R, there is a package [queueing](https://cran.r-project.org/web/packages/queueing/index.html) which makes this task super easy!

To create a theoretical M/M/s queue model, we first need to calculate lambda and mu based on our inter-arrival time and service times:

```{r}
# Example parameters
interarrival_time <- 5L
service_length <- 5L
servers <- 2L

# Calculate lambda and mu
lambda <- 1L / interarrival_time
mu <- 1L / service_length
```

We can then use `queueing::NewInput.MMC()` to set-up an M/M/s (M/M/c) model. This function takes the arrival rate (`lambda`), service rate (`mu`) and the number of servers (`c`). There are two other parameters `n` and `method`, but we just specify the defaults here (0) (see [documentation](https://cran.r-project.org/web/packages/queueing/queueing.pdf) for more details).

```{r}
math_model <- queueing::NewInput.MMC(
  lambda = lambda, mu = mu, c = servers, n = 0L, method = 0L
)
math_model
```

We then provide this model to `queueing::QueueingModel()`, which builds the model and returns the relevant calculations.

```{r}
math_results <- queueing::QueueingModel(math_model)
math_results
```

:::

## Function to run our simulation

This function simply runs our simulation model and returns the relevant metrics, renamed to use queueing theory notation.

We could have included this code as part of the test function below - but have just made it in a separate function to keep that test function a little simpler.

Our short illustrative parameters (30 minute warm-up, 40 minute data collection, 5 replications) are very unstable, so we have increased them to get more reliable results for the comparison in this test.

::: {.python-content}

```{python}
def run_simulation_model(
    interarrival_time,
    consultation_time,
    number_of_doctors
):
    """
    Run simulation and return key performance indicators using standard
    queueing theory notation.

    The warm-up period should be sufficiently long to allow the system
    to reach steady-state before data collection begins.

    Parameters
    ----------
    interarrival_time : float
        Mean time between arrivals (minutes).
    consultation_time : float
        Mean length of doctor's consultation (minutes).
    number_of_doctors : int
        Number of doctors.
    """
    param = Parameters(
        interarrival_time=interarrival_time,
        consultation_time=consultation_time,
        number_of_doctors=number_of_doctors,
        warm_up_period=2000,
        data_collection_period=4000,
        number_of_runs=20,
        cores=1,
        verbose=False
    )
    runner = Runner(param)
    results = runner.run_reps()

    # Rename the columns using queuing theory notation
    mapping = {
        "mean_wait_time": "W_q",
        "mean_time_in_system": "W_s",
        "mean_utilisation_tw": "rho",
        "mean_queue_length": "L_q",
    }
    df = results["overall"].rename(columns=mapping)

    # Return relevant columns
    return df[mapping.values()].T["mean"]
```

For example:

```{python}
run_simulation_model(interarrival_time=5,
                     consultation_time=10,
                     number_of_doctors=3)
```

:::

::: {.r-content}

```{r}
#' Run simulation and return key performance indicators using standard queueing
#' theory notation.
#'
#' The warm-up period should be sufficiently long to allow the system to reach
#' steady-state before data collection begins.
#'
#' @param interarrival_time Mean time between arrivals (minutes).
#' @param consultation_time Mean length of doctor's consultation (minutes).
#' @param number_of_doctors Number of doctors.

run_simulation_model <- function(
  interarrival_time, consultation_time, number_of_doctors
) {
  # Run simulation
  param <- create_params(
    interarrival_time = interarrival_time,
    consultation_time = consultation_time,
    number_of_doctors = number_of_doctors,
    warm_up_period = 2000L,
    data_collection_period = 4000L,
    number_of_runs = 20L,
    verbose = FALSE,
    cores = 1L
  )
  run_results <- runner(param)[["run_results"]]

  # Get overall results, using queueing theory notation in column names
  results <- run_results |>
    summarise(
      RO = mean(.data[["utilisation_doctor"]]),
      Lq = mean(.data[["mean_queue_length_doctor"]]),
      W = mean(.data[["mean_time_in_system"]]),
      Wq = mean(.data[["mean_wait_time_doctor"]])
    )
  results
}
```

For example:

```{r}
run_simulation_model(interarrival_time = 5L,
                     consultation_time = 10L,
                     number_of_doctors = 3L)
```

:::

## Test comparing simulation with mathematical model results

::: {.python-content}

Our test function `test_simulation_against_theory()` uses `@pytest.mark.parametrize` to run several different test cases.

For each of these, it compares the results from the simulation model (from `run_simulation_model()`) with the results from the mathematical model (`MMSQueue`) and checks they are reasonably similar.

:::

::: {.r-content}

Our test function uses `patrick` to run several different test cases.

For each of these, it compares the results from the simulation model (from `run_simulation_model()`) with the results from the mathematical model from `queueing`, and checks they are reasonably similar.

:::

::: {.callout-note title="Tolerance"}

We use a 15% threshold for the relative difference between simulation and analytical results. This allows for expected variation, while still flagging results that differ by more than would be reasonably explained by randomness alone.

:::

::: {.python-content}

```{.python}
@pytest.mark.parametrize(
    "interarrival_time,consultation_time,number_of_doctors",
    [
        # Test case 1: Low utilisation (ρ ≈ 0.3)
        (10, 3, 2),
        # Test case 2: Medium utilisation (ρ ≈ 0.67)
        (6, 4, 2),
        # Test case 3: M/M/1 (ρ = 0.75)
        (4, 3, 1),
        # Test case 4: Multiple servers, high utilisation (ρ ≈ 0.91)
        (5.5, 5, 3),
        # Test case 5: Balanced system (ρ = 0.5)
        (8, 4, 1),
        # Test case 6: Many servers, low individual utilisation (ρ ≈ 0.63)
        (4, 10, 4),
        # Test case 7: Very low utilisation (ρ ≈ 0.167)
        (60, 10, 15),
    ],
)
def test_simulation_against_theory(
    interarrival_time,
    consultation_time,
    number_of_doctors
):
    """Test simulation results against theoretical M/M/S queue calculations."""

    # Create theoretical M/M/S queue model and get metrics
    lam = 1 / interarrival_time
    mu = 1 / consultation_time
    theory = MMSQueue(lam, mu, number_of_doctors).metrics

    # Run simulation
    sim = run_simulation_model(
        interarrival_time=interarrival_time,
        consultation_time=consultation_time,
        number_of_doctors=number_of_doctors
    )

    # Compare results with appropriate tolerance (round to 3dp + 15% tolerance)
    metrics = [
        ("rho", "Utilisation"),
        ("L_q", "Queue length"),
        ("W_q", "Wait time"),
        ("W_s", "System time")
    ]
    for key, label in metrics:
        sim_val = round(sim[key], 3)
        theory_val = round(theory[key], 3)
        assert sim_val == pytest.approx(theory_val, rel=0.15), (
            f"{label} mismatch: sim={sim_val}, theory={theory_val}"
        )
```

```{python}
#| echo: false
import pytest
pytest.main(["tests_resources/test_mms.py"])
```

:::

::: {.r-content}

```{r}
#| file: "tests_resources/test_mms.R"
#| eval: false
```

```{r}
#| echo: false
#| output: false
library(testthat)
```

```{r}
#| echo: false
testthat::test_file(file.path("tests_resources", "test_mms.R"))
```

:::

🎉 All tests have passed!

## Explore the example models

<div class="h3-tight"></div>

### Nurse visit simulation

:::: {.python-content}

{{< include /html/pydesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `tests/test_mms.py` |
: {tbl-colwidths="[20,80]"}
:::

::::

:::: {.r-content}

{{< include /html/rdesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `tests/testthat/test-mms.R` |
: {tbl-colwidths="[20,80]"}
:::

::::

### Stroke pathway simulation

Not relevant - model too complex.

## Test yourself

Try adding this test to the model set-up using this book.

<br><br>
