---
title: Tests
date: "2025-11-05T11:21:31+00:00"
---

```{python}
#| echo: false
# pylint: disable=undefined-variable
```

::: {.pale-blue}

**Learning objectives:**

* Understand how to write and run tests.
* Explore examples of the types of test you can run for your simulation model.

**Relevant reproducibility guidelines:**

* NHS Levels of RAP (🥈): Pipeline includes a testing framework (unit tests, back tests).

**Pre-reading:**

This page will run tests on the model from the [parallel processing](../output_analysis/parallel.qmd) page (likewise, also used on the [scenario and sensitivity analysis](../experiments/scenarios.qmd) page).

> *Entity generation → Entity processing → Initialisation bias → Performance measures → Replications → Parallel processing → Tests*

:::

<br>

Testing is the process of evaluating a model to ensure it works as expected, gives reliable results, and can handle different conditions. By **systematically checking for errors, inconsistencies, or unexpected behaviors**, testing helps improve the quality of a model, catch errors and prevent future issues. They can also be used to run test cases that check your results are consistent (i.e., that it is **reproducible**).

When you create a model, you will naturally carry out tests, with simple manual checks where you observe outputs and ensure they look right. These checks can be formalised and **automated** so that you can run them after any changes, and catch any issues that arise.

Writing tests is fundamental for [model verification](verification_validation.qmd). These tests should be run regularly and after code changes, to check that these changes haven't introduced new issues ("**regression testing**").

## Introduction to writing and running tests

<div class="h3-tight"></div>

### Writing a basic test

::: {.python-content}

A popular framework for testing in python is **pytest**.

Each test in pytest is a function that contains an assertion statement to check a condition (e.g., `number > 0`). If the condition fails, pytest will return an error message (e.g. "The number should be positive").

Test are typically stored in a folder called `tests/`, and filenames start with the prefix `test_`. This naming convention allows pytest to automatically discover and run all the tests in the folder.

Here's an example of a simple test using pytest:

```{python}
import pytest


def test_positive():
    """
    Confirm that the number is positive.
    """
    number = 5
    assert number > 0, "The number should be positive"
```

:::

::: {.r-content}

A popular framework for testing in R is **testthat**.

Each test uses the `test_that()` functions and is structured around expectations that check specific conditions (e.g., `expect_true()`, `expect_false()`, `expect_equal()`, `expect_error()` - see [package index](https://testthat.r-lib.org/reference/index.html) for more).

Tests are typically stored in a folder called `tests/testthat`, and filenames start with the prefix `test_`. This naming convention allows `testthat` to automatically discover and run all tests in the folder. You can create this structure with `usethis`, as explained below.

First, here's an example of a simple test using testthat:

```{r}
#| output: false
library(testthat)
```

```{r}
test_that("Confirm that the number is positive", {
  number <- 5L
  expect_gt(number, 0L)
})
```

### Setting up a testing structure with testthat

If you model is [structured as a package](/pages/guide/setup/package.qmd), the easiest way to get started with tests is to let `usethis` create the standard `testthat` layout for you:

```{.r}
usethis::use_testthat()
```

This creates the `tests/testthat/` directory, adds the necessary configuration files, and records `testthat` as a suggested dependency. You can then create individual test files with:

```{.r}
usethis::use_test("simulation")
```

...which will open a new file such as `tests/testthat/test-simulation.R` ready for you to add `test_that()` blocks.

:::

### Running tests

:::: {.python-content}

Tests are typically run from the terminal. Commands include:

* `pytest` - runs all tests.
* `pytest tests/filename.py` - runs tests from a specific file.

When you run a test, you'll see an output like this in the terminal:

::: {.callout-note icon=false title="Test output:"}

```{python}
#| echo: false
pytest.main(["tests_resources/test_example_simple.py"])
```

:::

::::

::: {.r-content}

Tests are typically run from the R console. Commands include:

* `testthat::test_dir()` - runs all tests in current directory (and sub-directories).
* `testthat::test_file("tests/testthat/test-filename.R")` - run tests from a specific file.
* `devtools::test()` - runs all tests in a package.

When you run a test, you'll see output in the console indicating whether tests passed or failed.

:::

### Parametrised tests

:::: {.python-content}

We can execute the same test with different parameters using `pytest.mark.parametrize`.

Here's an example:

```{python}
#| file: "tests_resources/test_example_param.py"
```

In this example, we're testing the same logic with four different values: `1`, `2`, `3`, and `-1`. The last value, `-1`, will cause the test to fail. The error message includes the failed value for easy debugging.

:::{.callout-note icon=false title="Test output:"}

```{python}
#| echo: false
pytest.main(["tests_resources/test_example_param.py"])
```

:::

::::

::: {.r-content}

We can execute the same test with different parameters using the [patrick](https://github.com/google/patrick) package.

Here's an example:

```{r}
#| file: "tests_resources/test_example_param.R"
#| eval: false
```

In this example, we're testing the same logic with four different values: `1`, `2`, `3`, and `-1`. The last value, `-1`, will cause the test to fail.

```{r}
#| echo: false
testthat::test_file(file.path("tests_resources", "test_example_param.R"))
```

:::

## Testing the model

There are many different ways of categorising tests. We will focus on three types:

* **Functional testing**
* **Unit testing**
* **Back testing**

As mentioned above, we are running these tests on the model from the [parallel processing](../output_analysis/parallel.qmd) page (which is also the one used on the [scenario and sensitivity analysis](../experiments/scenarios.qmd) page).

:::: {.python-content}

::: {.callout title="▶&nbsp;&nbsp; View simulaltion code" collapse="true" style="minimal"}

```{python}
#| file: "tests_resources/simulation.py"
```

:::

::::

:::: {.r-content}

::: {.callout title="▶&nbsp;&nbsp; View simulaltion code" collapse="true" style="minimal"}

```{r}
#| file: "tests_resources/simulation.R"
#| output: false
```

:::

:::

### Functional tests

Functional tests verify that the system or components perform their intended functionality.

For example, we expect that the number of arrivals should decrease if:

* The patient inter-arrival time increases.
* The length of the data collection period decreases.

:::: {.python-content}

```{python}
{{< include tests_resources/test_functional.py >}}
```

:::{.callout-note icon=false title="Test output"}

```{python}
#| echo: false
pytest.main(["tests_resources/test_functional.py"])
```

:::

::::

::: {.r-content}

```{r}
library(testthat)
library(patrick)

with_parameters_test_that("Test that adjusting parameters reduces arrivals", {
  # Run model with initial value
  init_param <- create_params(number_of_runs = 1L)
  init_param[[param_name]] <- init_value
  init_results <- runner(init_param)[["run_results"]]

  # Run model with adjusted value
  adj_param <- create_params(number_of_runs = 1L)
  adj_param[[param_name]] <- adj_value
  adj_results <- runner(adj_param)[["run_results"]]

  # Check that arrivals from adjusted model are less
  expect_lt(adj_results[["arrivals"]], init_results[["arrivals"]])
},
cases(
  list(param_name = "interarrival_time",
       init_value = 2L, adj_value = 15L),
  list(param_name = "data_collection_period",
       init_value = 2000L, adj_value = 500L)
))
```

:::

As another example, we expect that our model should fail if the number of doctors or the patient inter-arrival time were set to 0. This is tested using `test_zero_inputs`:

:::: {.python-content}

```{python}
#| file: "tests_resources/test_unit.py"
```

:::{.callout-note icon=false title="Test output"}

```{python}
#| echo: false
pytest.main(["tests_resources/test_unit.py"])
```

:::

::::

::: {.r-content}

```{r}
#| file: "tests_resources/test_unit.R"
#| eval: false
```

These tests fail as we do not have error handling for these values. In fact, with inter-arrival time set to 0, the model will run infinitely and crash your computer! Hence the use of `withTimeout` in the test.

To address this, we could add error handling which raises an error for users if they try to input a value of 0 (see the [parameter validation](../inputs/parameters_validation.qmd) page).

```{r}
#| echo: false
testthat::test_file(file.path("tests_resources", "test_unit.R"))
```

:::

<br>

We can identify other examples of functional tests from our [model verification](verification_validation.qmd) page. These include:

* **Assertion checking** - tests check that assertions hold true.
* **Stress testing** - tests simulate worst-case load and ensure model is robust under heavy demand.
* **Boundary testing** - tests check that the behaviour at, just inside, and just outside each boundary (if you have input variables with explicit limits).
* **Idle system** - tests run model with little or no activity/waits/service

### Unit tests

Unit tests are a type of functional testing that focuses on individual components (e.g. methods, classes) and tests them in isolation to ensure they work as intended. They can form part of "bottom-up" testing for model verification (see [verification and validation](verification_validation.qmd#bottom-up-testing) page).

::: {.python-content}

However, since we're using a simulation package (SimPy), much of the core functionality is already well-tested. As a result, it's often more relevant to write functional tests focused on your simulation logic, involving interactions between multiple classes or functions (e.g. parameters and model).

:::

::: {.r-content}

However, since we're using a simulation package (simmer), much of the core functionality is already well-tested. As a result, it's often more relevant to write functional tests focused on your simulation logic, involving interactions between multiple classes or functions (e.g. parameters and model).

:::

You may still find some uses for unit tests though, such as for custom validation methods or small utility functions, before running bigger integration checks.

### Back tests

Back tests check that the model code produces results consistent with those generated historically/from prior code.

First, we'll generate a set of expected results, with a specific set of parameters. Although this may seem unnecessary in this case, as they match our default parameters, these are still specified to ensure that we are testing on the same parameters, even if defaults change.

::: {.python-content}

```{python}
param = Parameters(
    interarrival_time=5,
    consultation_time=10,
    number_of_doctors=3,
    warm_up_period=30,
    data_collection_period=40,
    number_of_runs=5,
    verbose=False
)
```

:::

::: {.r-content}

```{r}
param <- create_params(
  interarrival_time = 5L,
  consultation_time = 10L,
  number_of_doctors = 3L,
  warm_up_period = 30L,
  data_collection_period = 40L,
  number_of_runs = 5L, verbose = FALSE
)
```

:::

We'll then run the model and save the results to `.csv` files.

::: {.python-content}

```{python}
runner = Runner(param=param)
results = runner.run_reps()
results["patient"].to_csv("tests_resources/python_patient.csv", index=False)
results["run"].to_csv("tests_resources/python_run.csv", index=False)
results["overall"].to_csv("tests_resources/python_overall.csv", index=False)
```

:::

::: {.r-content}

```{r}
results <- runner(param)
write.csv(arrange(results[["arrivals"]], replication, start_time),
          file.path("tests_resources", "r_arrivals.csv"),
          row.names = FALSE)
write.csv(results[["resources"]],
          file.path("tests_resources", "r_resources.csv"),
          row.names = FALSE)
write.csv(results[["run_results"]],
          file.path("tests_resources", "r_run_results.csv"),
          row.names = FALSE)
```

:::

In the test, we'll run the same model parameters, then import and compare against the saved `.csv` file to check for any differences.

:::: {.python-content}

```{python}
#| file: "tests_resources/test_back.py"
```

:::{.callout-note icon=false title="Test output"}

```{python}
#| echo: false
pytest.main(["tests_resources/test_back.py"])
```

:::

::::

::: {.r-content}

```{r}
testthat::test_file(file.path("tests_resources", "test_back.R"))
```

:::

::: {.python-content}

We generate the expected results for our backtest in a **seperate Python file or Jupyter notebook**, rather than within the test itself.

:::

::: {.r-content}

We generate the expected results for our backtest in a **seperate R file or R markdown file**, rather than within the test itself.

:::

We then would generally run tests using the same pre-generated `.csv` files, without regenerating them. However, the test will **fail if the model logic is intentionally changed**, leading to different results from the same parameters.

::: {.python-content}

In that case, if we are **certain** that these changes are the reason for differing results, we should re-run the Python file or notebook to regenerate the `.csv`.

:::

::: {.r-content}

In that case, if we are **certain** that these changes are the reason for differing results, we should re-run the R file or R markdown file to regenerate the `.csv`.

:::

It is crucial to exercise caution when doing this, to avoid unintentionally overwriting correct expected results.

## Test coverage

**Coverage** refers to the percentage of your code that is executed when you run your tests. It can help you spot parts of your code that are not included in any tests.

::: {.python-content}

The [pytest-cov](https://github.com/pytest-dev/pytest-cov) package can be used to run coverage calculations alongside pytest. After installing it, simply run tests with the `--cov` flag:

```{.bash}
pytest --cov
```

See the [GitHub actions](../style_docs/github_actions.qmd#example-test-coverage) page for guidance on generating a coverage badge for your README within CI/CD.

:::

::: {.r-content}

<br>

If your research is [structured as a package](../setup/package.qmd), then you can use `devtools` to calculate coverage:

```{.r}
devtools::test_coverage()
```

<br>

If not structured as a package, you can use [covr](https://github.com/r-lib/covr)'s `file_coverage()` function - for example:

```{.r}
file_coverage(source_files = c("param.R", "model.R"),
              test_files = c("test-unit.R", "test-back.R"))
```

:::

## Explore the example models

::: {.python-content}

{{< include /html/pydesrapmms.html >}}

{{< include /html/pydesrapstroke.html >}}

A wide variety of tests are available in the `tests/` folder.

Data for back tests are generated within `notebooks/generate_exp_results.ipynb`.

Tests are run and coverage are calculated on pushes to main via GitHub actions (`.github/workflows/tests.yaml`).

:::

::: {.r-content}

{{< include /html/rdesrapmms.html >}}

{{< include /html/rdesrapstroke.html >}}

A wide variety of tests are available in the `tests/` folder.

Data for back tests are generated within `notebooks/generate_exp_results.Rmd`.

Tests are run via GitHub actions (`.github/workflows/R-CMD-check.yaml`) and a coverage badge is updated (`.github/workflows/test-coverage.yaml`).

:::

## Test yourself

**Write tests**! Look at **example models** for inspiration on what and how to test.

Start writing tests **early**, and run them **often** to catch issues as you develop.

Use **coverage** calculates to help you spot parts of you code not run in any tests.

<br><br>
