---
title: ❌ Scenario analysis
bibliography: scenarios_resources/references.bib
---

{{< include ../../scripts/_reticulate-setup.md >}}

:::: {.pale-blue}

**Learning objectives:**

* Understand the importance of sharing your scenario code.
* Learn how to run scenarios from a script or csv.

**Relevant reproducibility guidelines:**

* STARS Reproducibility Recommendations (⭐): Provide code for all scenarios and sensitivity analyses.

**Pre-reading:**

This page continues on from: [Replications](../output_analysis/replications.qmd).

> *Entity generation → Entity processing → Initialisation bias → Performance measures → Replications → Scenario analysis*

**Required packages:**

These should be available from environment setup in the "Test yourself" section of [Environments](../setup/environment.qmd). 

::: {.python-content}

```{python}
from IPython.display import HTML
from itables import to_html_datatable
import itertools
import numpy as np
import pandas as pd
import plotly.express as px
import plotly.graph_objects as go
import scipy.stats as st
import simpy
from sim_tools.distributions import Exponential
```

```{python}
#| echo: false
# To ensure renders correctly in quarto
import plotly.io as pio  # pylint: disable=ungrouped-imports
pio.renderers.default = "plotly_mimetype"
```

:::

::: {.r-content}

```{r}
#| output: false
library(dplyr)
library(kableExtra)
library(knitr)
library(simmer)
library(tidyr)
```

:::

::::

## Running and sharing scenarios

When creating a DES, you won't just run one model - you run scenarios, which are different parameter set-ups. These are used to test how outcomes change under varying conditions.

When running scenarios, there are two important things to remember:

::: {.cream}

**1. Run scenarios programmatically.**

:::

Don't duplicate whole scripts just to test different set-ups. The idea is to build the model with functions and classes so you only need to change the parameters and re-run. That's why we guide you through building the model this way in this book.

::: {.cream}

**2. Share your scenario code.**

:::

You should share code and data for every scenario - not just the base case.

In @Heather2025, six out of seven studies with scenarios lacked scenario code. This made it challenging and time-consuming to reproduce their results, as descriptions of the scenarios in the articles were often ambiguous and key parameters unclear or missing.

<br>

This page demonstrates how to run scenarios with the example model from the [Replications](../output_analysis/replications.qmd) page - both directly in code and via CSV files.

```{python}
#| echo: false
{{< include ../output_analysis/replications_resources/parameters.py >}}
{{< include ../output_analysis/replications_resources/patient.py >}}
{{< include ../output_analysis/replications_resources/model.py >}}
{{< include ../output_analysis/replications_resources/summary_stats.py >}}
{{< include ../output_analysis/replications_resources/runner.py >}}
```

```{r}
#| echo: false
#| file: ../output_analysis/replications_resources/create_params.R
```

```{r}
#| echo: false
#| file: ../output_analysis/replications_resources/filter_warmup.R
```

```{r}
#| echo: false
#| file: ../output_analysis/replications_resources/calc_arrivals.R
```

```{r}
#| echo: false
#| file: ../output_analysis/replications_resources/calc_mean_wait.R
```

```{r}
#| echo: false
#| file: ../output_analysis/replications_resources/get_run_results.R
```

```{r}
#| echo: false
#| file: ../output_analysis/replications_resources/model.R
```

```{r}
#| echo: false
#| file: ../output_analysis/replications_resources/runner.R
```

## Scenarios from a script

::: {.python-content}

<div class="h3-tight"></div>

### Scenarios function

If you are running several scenarios, it can be useful to define a helper function like the one shown below. This example is adapted from:

> "[HSMA - little book of DES](https://des.hsma.co.uk/)" from Sammi Rosser, Dan Chalk and Amy Heather 2025 (MIT Licence).

We walk through it step-by-step below.

```{python}
def run_scenarios(scenarios, param_factory=None):
    """
    Execute a set of scenarios and return the results from each run.

    Parameters
    ----------
    scenarios : dict
        Dictionary where key is name of parameter and value is a list with
        different values to run in scenarios.
    param_factory : callable or None, optional
        A callable that returns a new Parameters object for each scenario run.
        This can be a class (e.g., `Parameters`) or a factory function/lambda
        with preset arguments (e.g., `lambda: Parameters(number_of_runs=4)`).
        If not provided, defaults to using `Parameters()` with no arguments.

    Returns
    -------
    pandas.DataFrame
        DataFrame with results from each run of each scenario.

    Notes
    -----
    Function adapted from Rosser, Chalk and Heather 2025.
    """
    # If none provided, use Parameters
    if param_factory is None:
        param_factory = Parameters

    # Find every possible permutation of the scenarios
    all_scenarios_tuples = list(itertools.product(*scenarios.values()))

    # Convert back into dictionaries
    all_scenarios_dicts = [
        dict(zip(scenarios.keys(), p)) for p in all_scenarios_tuples
    ]

    # Preview some of the scenarios
    print(f"There are {len(all_scenarios_dicts)} scenarios. Running:")

    # Run the scenarios...
    results = []
    for index, scenario_to_run in enumerate(all_scenarios_dicts):
        print(scenario_to_run)

        # Create fresh instance of parameter class for each scenario
        param = param_factory()

        # Update parameter list with the scenario parameters
        param.scenario_name = index
        for key in scenario_to_run:
            setattr(param, key, scenario_to_run[key])

        # Perform replications
        scenario_exp = Runner(param)
        scenario_res = scenario_exp.run_reps()["run"]

        # Add scenario number and values to the results dataframe
        scenario_res["scenario"] = index
        for key in scenario_to_run:
            scenario_res[key] = scenario_to_run[key]

        # Add results from scenario to list
        results.append(scenario_res)
    return pd.concat(results)
```

### Explaining the `run_scenarios` function

```{.python}
def run_scenarios(scenarios, param_factory=None):
    """
    Execute a set of scenarios and return the results from each run.

    Parameters
    ----------
    scenarios : dict
        Dictionary where key is name of parameter and value is a list with
        different values to run in scenarios.
    param_factory : callable or None, optional
        A callable that returns a new Parameters object for each scenario run.
        This can be a class (e.g., `Parameters`) or a factory function/lambda
        with preset arguments (e.g., `lambda: Parameters(number_of_runs=4)`).
        If not provided, defaults to using `Parameters()` with no arguments.

    Returns
    -------
    pandas.DataFrame
        DataFrame with results from each run of each scenario.

    Notes
    -----
    Function adapted from Rosser, Chalk and Heather 2025.
    """
    # If none provided, use Parameters
    if param_factory is None:
        param_factory = Parameters
```

The first argument, `scenarios`, is a dictionary. The keys should be parameter names, and the values should be lists of values to run. For example:

```
{"interarrival_time": [4, 5, 6],
 "number_of_doctors": [2, 3, 4]}
```

The second argument, `param_factory`, is a callable that returns a fresh Parameters instance for each scenario run. This ensures every scenario gets an independent set of parameters and avoids carrying state from one scenario to another.

* For a default base case, supply `param_factory = Parameters`.

* To use a specific preset (e.g., altering the number of runs for all scenarios), supply a lambda: `param_factory = lambda: Parameters(number_of_runs=4)`

<br>

```{.python}
# Find every possible permutation of the scenarios
all_scenarios_tuples = list(itertools.product(*scenarios.values()))
```

This line generates all possible combinations of values for each parameter.

<br>

```{.python}
# Convert back into dictionaries
all_scenarios_dicts = [
    dict(zip(scenarios.keys(), p)) for p in all_scenarios_tuples
]

# Preview some of the scenarios
print(f"There are {len(all_scenarios_dicts)} scenarios. Running:")
```

Each scenario is converted from a tuple of values into a dictionary, making it easier to update parameters by name when running scenarios.

<br>

```{.python}
# Run the scenarios...
results = []
for index, scenario_to_run in enumerate(all_scenarios_dicts):
    print(scenario_to_run)
```

We loop over each scenario dictionary and run it. Printing the scenario shows which parameter values are being used for each run.

<br>

```{.python}
# Create fresh instance of parameter class for each scenario
param = param_factory()

# Update parameter list with the scenario parameters
param.scenario_name = index
for key in scenario_to_run:
    setattr(param, key, scenario_to_run[key])     
```

For each scenario, we create a new (independent) parameter object, then update its attributes to match the scenario values. This ensures each run starts with a fresh configuration and avoids parameter sharing between scenarios.

<br>

```{.python}
# Perform replications
scenario_exp = Runner(param)
scenario_res = scenario_exp.run_reps()["run"]

# Add scenario number and values to the results dataframe
scenario_res["scenario"] = index
for key in scenario_to_run:
    scenario_res[key] = scenario_to_run[key]

# Add results from scenario to list
results.append(scenario_res)
```

Run the model for each scenario and collect results, adding the scenario numbers and settings to the output.

<br>

```{.python}
return pd.concat(results)
```

Combine results from all scenarios into a single dataframe.

### Run scenarios

We can now use `run_scenarios()` to execute every possible combination of the specified parameters.

```{python}
# Run scenarios
scenario_results = run_scenarios(
    scenarios = {"interarrival_time": [4, 5, 6, 7, 8],
                 "number_of_doctors": [3, 4, 5]},
    param_factory = lambda: Parameters(number_of_runs = 100)
)

# View scenario results
HTML(to_html_datatable(scenario_results))
```

We can find the average results for each scenario, and plot these, like below:

```{python}
def plot_scenario(results, x_var, result_var, colour_var=None, name_mappings=None):
    """
    Plot results from different model scenarios.

    Arguments:
        results : pd.DataFrame
            Contains results to plot.
        x_var : str
            Name of variable to plot on X axis.
        result_var : str
            Name of variable with results, to plot on Y axis.
        colour_var : str, optional
            Name of variable to colour lines with.
        name_mappings : dict, optional
            Dictionary mapping column names to labels. If not provided,
            function will default to variable names.
    """
    # Set up groupings
    if colour_var is not None:
        group_vars = [x_var, colour_var]
    else:
        group_vars = [x_var]

    # Calculate summary statistics for each scenario: mean, std, CI
    df = (
        results.groupby(group_vars)[result_var]
        .apply(summary_stats)
        .apply(pd.Series)
        .reset_index()
    )
    df.columns = (list(df.columns[:-4]) +
                  ["mean", "std_dev", "ci_lower", "ci_upper"])

    # Initialise figure
    fig = go.Figure()

    # Handling color mappings (optional)
    if colour_var:
        unique_groups = df[colour_var].unique()
        colors = px.colors.qualitative.Plotly
        color_map = {group: colors[i % len(colors)] for i, group in enumerate(unique_groups)}
    else:
        color_map = {None: "blue"}

    # Plot each scenario line and its shaded confidence interval
    groups = df.groupby(colour_var) if colour_var else [(None, df)]
    for group_name, group_data in groups:
        x = group_data[x_var].values
        mean = group_data["mean"].values
        ci_upper = group_data["ci_upper"].values
        ci_lower = group_data["ci_lower"].values
        color = color_map[group_name]

        # Filled confidence interval region
        fig.add_trace(go.Scatter(
            x = np.concatenate([x, x[::-1]]),
            y = np.concatenate([ci_upper, ci_lower[::-1]]),
            fill = "toself",
            fillcolor = "rgba(0,100,80,0.2)",
            line = dict(color="rgba(255,255,255,0)"),
            showlegend=False,
            name = f"95% CI {group_name}" if group_name is not None else "95% CI"
        ))

        # Mean line
        fig.add_trace(go.Scatter(
            x = x,
            y = mean,
            mode = "lines",
            line = dict(color = color, width=2),
            name = str(group_name) if group_name is not None else "Mean"
        ))

    # Set labels and layout
    fig.update_layout(
        xaxis_title=name_mappings.get(x_var, x_var) if name_mappings else x_var,
        yaxis_title=name_mappings.get(result_var, result_var) if name_mappings else result_var,
        legend_title_text=name_mappings.get(colour_var, colour_var) if colour_var and name_mappings else (colour_var if colour_var else None),
        template="plotly_white"
    )

    return df, fig
```

```{python}
result, plot = plot_scenario(
    results=scenario_results,
    x_var="interarrival_time",
    result_var="mean_wait_time",
    colour_var="number_of_doctors",
    name_mappings={"interarrival_time": "Patient inter-arrival time", "mean_wait_time": "Mean wait time for the doctor", "number_of_doctors": "Number of doctors"})
HTML(to_html_datatable(result))
plot.show()
```

:::

::: {.r-content}

```{r}
```

:::

## Scenarios in a CSV

TODO

::: {.python-content}

```{python}
```

:::

::: {.r-content}

```{r}
```

:::

## References

::: {#refs}
:::

<br><br>
