---
title: ❌ Scenario and sensitivity analysis
bibliography: scenarios_resources/references.bib
---

{{< include ../../scripts/_reticulate-setup.md >}}

:::: {.pale-blue}

**Learning objectives:**

* Understand the importance of sharing your scenario code.
* Learn how to run a combination of scenarios.

**Relevant reproducibility guidelines:**

* STARS Reproducibility Recommendations (⭐): Provide code for all scenarios and sensitivity analyses.

**Pre-reading:**

This page continues on from: [Replications](../output_analysis/replications.qmd).

> *Entity generation → Entity processing → Initialisation bias → Performance measures → Replications → Scenario analysis*

**Required packages:**

These should be available from environment setup in the "Test yourself" section of [Environments](../setup/environment.qmd). 

::: {.python-content}

```{python}
from IPython.display import HTML
from itables import to_html_datatable
import itertools
import matplotlib.colors as mcolors
import numpy as np
import pandas as pd
import plotly.express as px
import plotly.graph_objects as go
import scipy.stats as st
import simpy
from sim_tools.distributions import Exponential
```

```{python}
#| echo: false
# To ensure renders correctly in quarto
import plotly.io as pio  # pylint: disable=ungrouped-imports
pio.renderers.default = "plotly_mimetype"
```

:::

::: {.r-content}

```{r}
#| output: false
library(dplyr)
library(kableExtra)
library(knitr)
library(simmer)
library(tidyr)
```

:::

::::

## Running and sharing scenarios

When creating a DES, you won't just run one model - you run scenarios, which are different parameter set-ups. These are used to test how outcomes change under varying conditions.

When running scenarios, there are two important things to remember:

::: {.cream}

**1. Run scenarios programmatically.**

:::

Don't duplicate whole scripts just to test different set-ups. The idea is to build the model with functions and classes so you only need to change the parameters and re-run. That's why we guide you through building the model this way in this book.

::: {.cream}

**2. Share your scenario code.**

:::

You should share code and data for every scenario - not just the base case.

In @Heather2025, six out of seven studies with scenarios lacked scenario code. This made it challenging and time-consuming to reproduce their results, as descriptions of the scenarios in the articles were often ambiguous and key parameters unclear or missing.

```{python}
#| echo: false
{{< include ../output_analysis/replications_resources/parameters.py >}}
{{< include ../output_analysis/replications_resources/patient.py >}}
{{< include ../output_analysis/replications_resources/model.py >}}
{{< include ../output_analysis/replications_resources/summary_stats.py >}}
{{< include ../output_analysis/replications_resources/runner.py >}}
```

```{r}
#| echo: false
#| file: ../output_analysis/replications_resources/create_params.R
```

```{r}
#| echo: false
#| file: ../output_analysis/replications_resources/filter_warmup.R
```

```{r}
#| echo: false
#| file: ../output_analysis/replications_resources/calc_arrivals.R
```

```{r}
#| echo: false
#| file: ../output_analysis/replications_resources/calc_mean_wait.R
```

```{r}
#| echo: false
#| file: ../output_analysis/replications_resources/get_run_results.R
```

```{r}
#| echo: false
#| file: ../output_analysis/replications_resources/model.R
```

```{r}
#| echo: false
#| file: ../output_analysis/replications_resources/runner.R
```

## Simple scenarios

This code runs the simulation for each value of `number_of_doctors` from 3 to 6.

The returned dataframe contains the results from each of the five runs from the four scenarios.

::: {.python-content}

```{python}
results = []
# Loop through doctor counts 3-6, running the simulation
for i in range(3, 7):
    param = Parameters(number_of_doctors=i)
    runner = Runner(param=param)
    result = runner.run_reps()["run"]
    # Add scenario information to the results, then save to list
    result["number_of_doctors"] = i
    results.append(result)

# Combine results into one DataFrame
scenario_results = pd.concat(results, ignore_index = True)

# Show as interactive table
HTML(to_html_datatable(scenario_results))
```

:::

::: {.r-content}

```{r}
```

:::

## Running several scenario combinations

::: {.python-content}

If you are running several scenarios, it can be useful to define a helper function like the one shown below. This example is adapted from:

> "[HSMA - little book of DES](https://des.hsma.co.uk/)" from Sammi Rosser, Dan Chalk and Amy Heather 2025 (MIT Licence).

We walk through it step-by-step below.

```{python}
def run_scenarios(scenarios, param_factory=None):
    """
    Execute a set of scenarios and return the results from each run.

    Parameters
    ----------
    scenarios : dict
        Dictionary where key is name of parameter and value is a list with
        different values to run in scenarios.
    param_factory : callable or None, optional
        A callable that returns a new Parameters object for each scenario run.
        This can be a class (e.g., `Parameters`) or a factory function/lambda
        with preset arguments (e.g., `lambda: Parameters(number_of_runs=4)`).
        If not provided, defaults to using `Parameters()` with no arguments.

    Returns
    -------
    pandas.DataFrame
        DataFrame with results from each run of each scenario.

    Notes
    -----
    Function adapted from Rosser, Chalk and Heather 2025.
    """
    # If none provided, use Parameters
    if param_factory is None:
        param_factory = Parameters

    # Find every possible permutation of the scenarios
    all_scenarios_tuples = list(itertools.product(*scenarios.values()))

    # Convert back into dictionaries
    all_scenarios_dicts = [
        dict(zip(scenarios.keys(), p)) for p in all_scenarios_tuples
    ]

    # Preview some of the scenarios
    print(f"There are {len(all_scenarios_dicts)} scenarios. Running:")

    # Run the scenarios...
    results = []
    for index, scenario_to_run in enumerate(all_scenarios_dicts):
        print(scenario_to_run)

        # Create fresh instance of parameter class for each scenario
        param = param_factory()

        # Update parameter list with the scenario parameters
        param.scenario_name = index
        for key in scenario_to_run:
            setattr(param, key, scenario_to_run[key])

        # Perform replications
        scenario_exp = Runner(param)
        scenario_res = scenario_exp.run_reps()["run"]

        # Add scenario number and values to the results dataframe
        scenario_res["scenario"] = index
        for key in scenario_to_run:
            scenario_res[key] = scenario_to_run[key]

        # Add results from scenario to list
        results.append(scenario_res)
    return pd.concat(results)
```

### Explaining the `run_scenarios` function

```{.python}
def run_scenarios(scenarios, param_factory=None):
    """
    Execute a set of scenarios and return the results from each run.

    Parameters
    ----------
    scenarios : dict
        Dictionary where key is name of parameter and value is a list with
        different values to run in scenarios.
    param_factory : callable or None, optional
        A callable that returns a new Parameters object for each scenario run.
        This can be a class (e.g., `Parameters`) or a factory function/lambda
        with preset arguments (e.g., `lambda: Parameters(number_of_runs=4)`).
        If not provided, defaults to using `Parameters()` with no arguments.

    Returns
    -------
    pandas.DataFrame
        DataFrame with results from each run of each scenario.

    Notes
    -----
    Function adapted from Rosser, Chalk and Heather 2025.
    """
    # If none provided, use Parameters
    if param_factory is None:
        param_factory = Parameters
```

The first argument, `scenarios`, is a dictionary. The keys should be parameter names, and the values should be lists of values to run. For example:

```
{"interarrival_time": [4, 5, 6],
 "number_of_doctors": [2, 3, 4]}
```

The second argument, `param_factory`, is a callable that returns a fresh Parameters instance for each scenario run. This ensures every scenario gets an independent set of parameters and avoids carrying state from one scenario to another.

* For a default base case, supply `param_factory = Parameters`.

* To use a specific preset (e.g., altering the number of runs for all scenarios), supply a lambda: `param_factory = lambda: Parameters(number_of_runs=4)`

<br>

```{.python}
# Find every possible permutation of the scenarios
all_scenarios_tuples = list(itertools.product(*scenarios.values()))
```

This line generates all possible combinations of values for each parameter.

<br>

```{.python}
# Convert back into dictionaries
all_scenarios_dicts = [
    dict(zip(scenarios.keys(), p)) for p in all_scenarios_tuples
]

# Preview some of the scenarios
print(f"There are {len(all_scenarios_dicts)} scenarios. Running:")
```

Each scenario is converted from a tuple of values into a dictionary, making it easier to update parameters by name when running scenarios.

<br>

```{.python}
# Run the scenarios...
results = []
for index, scenario_to_run in enumerate(all_scenarios_dicts):
    print(scenario_to_run)
```

We loop over each scenario dictionary and run it. Printing the scenario shows which parameter values are being used for each run.

<br>

```{.python}
# Create fresh instance of parameter class for each scenario
param = param_factory()

# Update parameter list with the scenario parameters
param.scenario_name = index
for key in scenario_to_run:
    setattr(param, key, scenario_to_run[key])     
```

For each scenario, we create a new (independent) parameter object, then update its attributes to match the scenario values. This ensures each run starts with a fresh configuration and avoids parameter sharing between scenarios.

<br>

```{.python}
# Perform replications
scenario_exp = Runner(param)
scenario_res = scenario_exp.run_reps()["run"]

# Add scenario number and values to the results dataframe
scenario_res["scenario"] = index
for key in scenario_to_run:
    scenario_res[key] = scenario_to_run[key]

# Add results from scenario to list
results.append(scenario_res)
```

Run the model for each scenario and collect results, adding the scenario numbers and settings to the output.

<br>

```{.python}
return pd.concat(results)
```

Combine results from all scenarios into a single dataframe.

### Run scenarios

We can now use `run_scenarios()` to execute every possible combination of the specified parameters.

```{python}
# Run scenarios
scenario_results = run_scenarios(
    scenarios = {"interarrival_time": [4, 5, 6, 7, 8],
                 "number_of_doctors": [3, 4, 5]},
    param_factory = lambda: Parameters(number_of_runs = 100)
)

# Save to CSV
scenario_results.to_csv("scenarios_resources/scenario_results.csv", index=False)

# View scenario results
HTML(to_html_datatable(scenario_results.head(200)))
```

:::

::: {.r-content}

:::

## Sensitivity analysis

### How does sensitivity analysis differ from scenario analysis?

Scenario analysis focuses on a set of pre-defined situations which are plausible and relevant to the problem being studied. It can often involve varying multiple parameters simultaneously. The purpose is to understand how the system operates under different hypothetical scenarios.

Sensitivity analysis varies one parameter (or a small group of parameters) and assess the impact of small changes in that parameter on outcomes. The purpose is to understand **how uncertainty in the inputs affects the model**, and how **robust** the model is to variation in those inputs.

### Running sensitivity analyses

Sensitivity analyses should be run programmatically, just like scenario analyses, and the code should be shared for reproducibility.

They can be run using similar code. Below is an example using the `run_scenarios` function to vary the `interarrival_time` parameter.

```{python}
# Run scenarios
sensitivity_results = run_scenarios(
    scenarios = {"interarrival_time": [4, 4.5, 5, 5.5, 6, 6.5, 7]},
    param_factory = lambda: Parameters(number_of_runs = 100)
)

# Save to CSV
sensitivity_results.to_csv("scenarios_resources/sensitivity_results.csv",
                           index=False)

# View scenario results
HTML(to_html_datatable(sensitivity_results.head(200)))
```

## Explore the example models

<div class="h3-tight"></div>

### Nurse visit simulation

:::: {.python-content}

{{< include ../../html/pydesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | |
| **What to look for?** | |
| **Why it matters?** | |
: {tbl-colwidths="[20,80]"}
:::

::::

:::: {.r-content}

{{< include ../../html/rdesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | |
| **What to look for?** | |
| **Why it matters?** | |
: {tbl-colwidths="[20,80]"}
:::

::::

### Stroke pathway simulation

:::: {.python-content}

{{< include ../../html/pydesrapstroke.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | |
| **What to look for?** | |
| **Why it matters?** | |
: {tbl-colwidths="[20,80]"}
:::

::::

:::: {.r-content}

{{< include ../../html/rdesrapstroke.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | |
| **What to look for?** | |
| **Why it matters?** | |
: {tbl-colwidths="[20,80]"}
:::

::::

## Test yourself

<!--TODO-->

## References

::: {#refs}
:::

<br><br>
