---
title:  Tables and figures
bibliography: scenarios_resources/references.bib
---

{{< include ../../scripts/_reticulate-setup.md >}}

```{python}
#| echo: false
# pylint: disable=too-many-locals,undefined-variable,unused-import
```

:::: {.pale-blue}

**Learning objectives:**

* Recognise importance of sharing code for creation of tables and figures.
* View some example tables and figures.

**Relevant reproducibility guidelines:**

* STARS Reproducibility Recommendations (⭐): Include code to generate the tables, figures, and other reported results. 
* STARS Reproducibility Recommendations: Save outputs to a file.

**Pre-reading:**

This page will use the results saved on the [Scenario and sensitivity analysis](scenarios.qmd) page.

**Required packages:**


These should be available from environment setup in the "Test yourself" section of [Environments](../setup/environment.qmd). 

::: {.python-content}

```{python}
import os

from IPython.display import HTML
from itables import to_html_datatable
import matplotlib.colors as mcolors
import numpy as np
import pandas as pd
import plotly.express as px
import plotly.graph_objects as go
import scipy.stats as st
```

```{python}
#| echo: false
# To ensure renders correctly in quarto
import plotly.io as pio  # pylint: disable=ungrouped-imports
pio.renderers.default = "plotly_mimetype"
```

:::

::: {.r-content}

```{r}
#| output: false
library(dplyr)
library(ggplot2)
library(kableExtra)
library(knitr)
```

:::

::::

<br>

When reporting results from a simulation, you will usually include tables and figures. To make your work reproducible, you should:

::: {.cream}

**1. Make your tables and figures using code.**

**2. Share that code alongside your results.**

:::

This is just as important as sharing model and scenario code. In @Heather2025, only one study included the code needed to generate all its reported results. Without this code for the others, recreating outputs was time-consuming and challenging. It required guesswork about which results to extract, how to clean them, and what processing steps were taken when producing the published tables and figures.

This page will show some examples of creating tables and figures, but there are many different ways you could describe and visualise your simulation results. The key takeaway from the page is simple: **write code and share it!**

## Simulation results

This page will use the results saved on the [Scenario and sensitivity analysis](scenarios.qmd) page.

::: {.python-content}

### Scenario analysis

{{< downloadthis scenarios_resources/python_scenario_results.csv dname="python_scenario_results" label="Download the scenario analysis results" type="primary" >}}

```{python}
scenario = pd.read_csv("scenarios_resources/python_scenario_results.csv")
HTML(to_html_datatable(scenario.head(200)))
```

### Sensitivity analysis

{{< downloadthis scenarios_resources/python_sensitivity_results.csv dname="python_sensitivity_results" label="Download the sensitivity analysis results" type="primary" >}}

```{python}
sensitivity = pd.read_csv("scenarios_resources/python_sensitivity_results.csv")
HTML(to_html_datatable(sensitivity.head(200)))
```

:::

::: {.r-content}

### Scenario analysis

{{< downloadthis scenarios_resources/r_scenario_results.csv dname="r_scenario_results" label="Download the scenario analysis results" type="primary" >}}

```{r}
scenario <- read.csv(file.path("scenarios_resources", "r_scenario_results.csv"))
kable(scenario) |> scroll_box(height = "400px")
```

### Sensitivity analysis

{{< downloadthis scenarios_resources/r_sensitivity_results.csv dname="r_sensitivity_results" label="Download the sensitivity analysis results" type="primary" >}}

```{r}
sensitivity <- read.csv(
  file.path("scenarios_resources", "r_sensitivity_results.csv")
)
kable(sensitivity) |> scroll_box(height = "400px")
```

:::

## Example tables

This function finds the average results from each experiment.

In these examples, we've run it on the mean wait time for the doctor.

:::: {.python-content}

It uses the `summary_stats()` function introduced on the [Replications](../output_analysis/replications.qmd) page.

::: {.callout-note title="View `summary_stats()`" collapse="true"}

```{python}
{{< include ../output_analysis/replications_resources/summary_stats.py >}}
```

:::

```{python}
def summarise_scenarios(results, groups, result_var, path=None):
    """
    Find the average results from each scenario.

    Parameters
    ----------
    results : pd.DataFrame
        Run-level results from every scenario.
    groups : list
        List of columns to group by (provided as strings).
    result_var : str
        Performance measure to get results on.
    path : str
        Path to save table to.

    Returns
    -------
    summary_table : pd.DataFrame
        Average results from each scenario.
    """
    summary_table = (
        results.groupby(groups)[result_var]
        .apply(summary_stats)
        .apply(pd.Series)
        .reset_index()
    )
    summary_table.columns = (list(summary_table.columns[:-4]) +
                             ["mean", "std_dev", "ci_lower", "ci_upper"])
    if path:
        summary_table.to_csv(path)
    return summary_table
```

### Scenario analysis

```{python}
scenario_table = summarise_scenarios(
    scenario, groups=["scenario", "interarrival_time", "number_of_doctors"],
    result_var="mean_wait_time",
    path=os.path.join("tables_figures_resources", "python_scenario_table.csv")
)
HTML(to_html_datatable(scenario_table))
```

### Sensitivity analysis

```{python}
sensitivity_table = summarise_scenarios(
    sensitivity, groups=["scenario", "interarrival_time"],
    result_var="mean_wait_time",
    path=os.path.join("tables_figures_resources",
                      "python_sensitivity_table.csv")
)
HTML(to_html_datatable(sensitivity_table))
```

::::

::: {.r-content}

```{r}
#' Find the average results from each scenario
#'
#' @param results Dataframe with results from each replication of scenarios.
#' @param groups List of columns to group by (provided as strings).
#' @param result_var Name of performance measure to get results on.
#' @param path Path to save table to.
#'
#' @return Dataframe with the average results from each scenario.

summarise_scenario <- function(results, groups, result_var, path) {
  summary_table <- results |>
    group_by_at(groups) |>
    summarise(mean = mean(.data[[result_var]]),
              std_dev = sd(.data[[result_var]]),
              ci_lower = t.test(.data[[result_var]])[["conf.int"]][[1L]],
              ci_upper = t.test(.data[[result_var]])[["conf.int"]][[2L]])
  write.csv(summary_table, path)
  summary_table
}
```

### Scenario analysis

```{r}
scenario_table <- summarise_scenario(
  scenario, groups = c("scenario", "interarrival_time", "number_of_doctors"),
  result_var = "mean_wait_time_doctor",
  path = file.path("tables_figures_resources", "r_scenario_table.csv")
)
kable(scenario_table)
```

### Sensitivity analysis

```{r}
sensitivity_table <- summarise_scenario(
  sensitivity, groups = c("scenario", "interarrival_time"),
  result_var = "mean_wait_time_doctor",
  path = file.path("tables_figures_resources", "r_sensitivity_table.csv")
)
kable(sensitivity_table)
```

:::

## Example figures

::: {.python-content}

These functions are used to plot the results from the scenarios and sensitivity analysis.

```{python}
def pale_colour(colour, alpha=0.2):
    """
    Make pale version of a colour.

    Parameters
    ----------
    colour : str or tuple
        The colour to pale, as a Plotly-compatible name ('blue', 'red'), hex
        string (e.g., "#1f77b4"), or RGB tuple (e.g., (0.1, 0.3, 0.8)).
        Accepts any format readable by matplotlib.colors.to_rgb.
    alpha : float, optional
        Opacity for the pale colour (default = 0.2). Should be between 0 and 1.

    Returns
    -------
    str
        Colour in 'rgba(r, g, b, a)' string format
        (e.g., "rgba(31,119,180,0.2)").

    Acknowledgements
    ----------------
    This function was generated by Perplexity.
    """
    # Convert any accepted colour format to an RGB tuple
    rgb = mcolors.to_rgb(colour)
    # Scale values to 0-255 for plotly compatability
    r, g, b = [int(x * 255) for x in rgb]
    # Return as RGB tuple with added alpha parameter (which makes it paler)
    return f"rgba({r},{g},{b},{alpha})"
```

```{python}
def plot_scenario(
    summary_table, x_var, y_var, colour_var=None, name_mappings=None, path=None
):
    """
    Plot results from different model scenarios.

    Parameters
    ----------
    summary_table : pd.DataFrame
        Average results from each scenario.
    x_var : str
        Name of variable to plot on X axis.
    y_var : str
        Name of variable with results, to plot on Y axis.
    colour_var : str, optional
        Name of variable to colour lines with.
    name_mappings : dict, optional
        Dictionary mapping column names to labels. If not provided,
        function will default to variable names.
    path : str
        Path to save figure to.
    """
    # Initialise figure
    fig = go.Figure()

    # Handling color mappings (optional)
    if colour_var:
        unique_groups = summary_table[colour_var].unique()
        colours = px.colors.qualitative.Plotly
        colour_map = {
            group: colours[i % len(colours)]
            for i, group in enumerate(unique_groups)
        }
    else:
        colour_map = {None: "blue"}

    # Plot each scenario line and its shaded confidence interval
    if colour_var:
        groups = summary_table.groupby(colour_var)
    else:
        groups = [(None, summary_table)]
    for group_name, group_data in groups:
        x = group_data[x_var].values
        mean = group_data["mean"].values
        ci_upper = group_data["ci_upper"].values
        ci_lower = group_data["ci_lower"].values
        main_colour = colour_map[group_name]
        shade_colour = pale_colour(main_colour, alpha=0.2)

        # Filled confidence interval region
        fig.add_trace(go.Scatter(
            x=np.concatenate([x, x[::-1]]),
            y=np.concatenate([ci_upper, ci_lower[::-1]]),
            fill="toself",
            fillcolor=shade_colour,
            line={"color": "rgba(255,255,255,0)"},
            showlegend=False,
            name="95% CI",
            hoverinfo="name+y"
        ))

        # Mean line
        fig.add_trace(go.Scatter(
            x=x,
            y=mean,
            mode="lines",
            line={"color": main_colour, "width": 2},
            name=group_name,
            hoverinfo="y"
        ))

    # Set labels and layout
    fig.update_layout(
        xaxis_title=(name_mappings.get(x_var, x_var)
                     if name_mappings else x_var),
        yaxis_title=(name_mappings.get(y_var, y_var)
                     if name_mappings else y_var),
        legend_title_text=(name_mappings.get(colour_var, colour_var)
                           if colour_var and name_mappings
                           else (colour_var if colour_var else None)),
        template="plotly_white"
    )

    if not colour_var:
        fig.update_layout(showlegend=False)

    # Save figure if path is provided
    if path:
        fig.write_image(path)

    return fig
```

### Scenario analysis

```{python}
scenario_plot = plot_scenario(
    summary_table=scenario_table,
    x_var="interarrival_time",
    y_var="mean",
    colour_var="number_of_doctors",
    name_mappings={
        "interarrival_time": "Patient inter-arrival time",
        "mean": "Mean wait time for the doctor",
        "number_of_doctors": "Number of doctors"
    },
    path=os.path.join("tables_figures_resources", "python_scenario_plot.png")
)
scenario_plot.show()
```

### Sensitivity analysis

```{python}
sensitivity_plot = plot_scenario(
    summary_table=sensitivity_table,
    x_var="interarrival_time",
    y_var="mean",
    name_mappings={
        "interarrival_time": "Patient inter-arrival time",
        "mean": "Mean wait time for the doctor"
    },
    path=os.path.join("tables_figures_resources",
                      "python_sensitivity_plot.png")
)
sensitivity_plot.show()
```

:::

::: {.r-content}

This function is used to plot the results from the scenarios and sensitivity analysis.

```{r}
#' Plot results from different model scenarios.
#'
#' @param summary_table Dataframe with average results from each scenario.
#' @param x_var Name of variable to plot on x axis.
#' @param y_var Name of variable with results, to plot on y axis.
#' @param colour_var Name of variable to colour lines with.
#' @param xaxis_title Title for x axis.
#' @param yaxis_title Title for y axis.
#' @param legend_title Title for figure legend.
#' @param path Path to save figure to.

plot_scenario <- function(
  summary_table, x_var, y_var, colour_var,
  xaxis_title, yaxis_title, legend_title, path
) {
  if (!is.null(colour_var)) {
    # Convert colour variable to a factor so its treated as categorical
    summary_table[[colour_var]] <- as.factor(summary_table[[colour_var]])
    # Plot with colour groups
    p <- ggplot(summary_table, aes(x = .data[[x_var]], y = .data[[y_var]],
                                   group = .data[[colour_var]])) +
      geom_line(aes(color = .data[[colour_var]])) +
      geom_ribbon(aes(ymin = .data[["ci_lower"]], ymax = .data[["ci_upper"]],
                      fill = .data[[colour_var]]), alpha = 0.1)
  } else {
    # Plot without colour groups
    p <- ggplot(summary_table, aes(x = .data[[x_var]], y = .data[[y_var]])) +
      geom_line() +
      geom_ribbon(aes(ymin = .data[["ci_lower"]], ymax = .data[["ci_upper"]],
                      alpha = 0.1), show.legend = FALSE)
  }
  # Modify labels and style
  p <- p +
    labs(x = xaxis_title, y = yaxis_title, color = legend_title,
         fill = legend_title) +
    theme_minimal()

  # Save plot
  ggsave(filename = path, width = 6.5, height = 4L, bg = "white")

  p
}
```

### Scenario analysis

```{r}
plot_scenario(
  summary_table = scenario_table,
  x_var = "interarrival_time",
  y_var = "mean",
  colour_var = "number_of_doctors",
  xaxis_title = "Patient inter-arrival time",
  yaxis_title = "Mean wait time for the doctor",
  legend_title = "Number of doctors",
  path = file.path("tables_figures_resources", "r_scenario_plot.png")
)
```

### Sensitivity analysis

```{r}
plot_scenario(
  summary_table = sensitivity_table,
  x_var = "interarrival_time",
  y_var = "mean",
  colour_var = NULL,
  xaxis_title = "Patient inter-arrival time",
  yaxis_title = "Mean wait time for the doctor",
  legend_title = NULL,
  path = file.path("tables_figures_resources", "r_sensitivity_plot.png")
)
```

:::

## Explore the example models

<div class="h3-tight"></div>

### Nurse visit simulation

:::: {.python-content}

{{< include ../../html/pydesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `notebooks/analysis.ipynb` |
: {tbl-colwidths="[20,80]"}
:::

::::

:::: {.r-content}

{{< include ../../html/rdesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `rmarkdown/analysis.md` |
: {tbl-colwidths="[20,80]"}
:::

::::

### Stroke pathway simulation

:::: {.python-content}

{{< include ../../html/pydesrapstroke.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `notebooks/analysis.ipynb` |
: {tbl-colwidths="[20,80]"}
:::

::::

:::: {.r-content}

{{< include ../../html/rdesrapstroke.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `rmarkdown/analysis.md` |
: {tbl-colwidths="[20,80]"}
:::

::::

## Test yourself

Create at least one table and figure from your simulation results.

If you don't have any to hand, feel free to download ours:

::: {.python-content}

{{< downloadthis scenarios_resources/python_scenario_results.csv dname="python_scenario_results" label="Download the scenario analysis results" type="primary" >}}

{{< downloadthis scenarios_resources/python_sensitivity_results.csv dname="python_sensitivity_results" label="Download the sensitivity analysis results" type="primary" >}}

:::

::: {.r-content}

{{< downloadthis scenarios_resources/r_scenario_results.csv dname="r_scenario_results" label="Download the scenario analysis results" type="primary" >}}

{{< downloadthis scenarios_resources/r_sensitivity_results.csv dname="r_sensitivity_results" label="Download the sensitivity analysis results" type="primary" >}}

:::

Make sure to write and save code that generates your outputs, and that you save the outputs to files (e.g., `csv` for tables, `png` for figures).

::: {.python-content}

For example, create a Jupyter notebook for the analysis, and then push that to your GitHub repository.

:::

::: {.r-content}

For example, create an Rmarkdown file for the analysis, and then push that to your GitHub repository.

:::

## References

::: {#refs}
:::

<br><br>
