---
title: ❌ Tests
---

{{< include ../../scripts/_reticulate-setup.md >}}

::: {.pale-blue}

**Learning objectives:**

* TODO

**Relevant reproducibility guidelines:**

* NHS Levels of RAP (🥈): Pipeline includes a testing framework (unit tests, back tests).

**Pre-reading:**

This page will run tests on the model from [Replications](../output_analysis/replications.qmd) (which has two performance measures: total arrivals and mean wait time).

> *Entity generation → Entity processing → Initialisation bias → Performance measures → Replications → Tests*

:::

```{python}
#| file: "tests_resources/simulation.py"
#| echo: false
```

<br>

Testing is the process of evaluating a model to ensure it works as expected, gives reliable results, and can handle different conditions. By **systematically checking for errors, inconsistencies, or unexpected behaviors**, testing helps improve the quality of a model, catch errors and prevent future issues. They can also be used to run test cases that check your results are consistent (i.e., that it is **reproducible**).

When you create a model, you will naturally carry out tests, with simple manual checks where you observe outputs and ensure they look right. These checks can be formalised and **automated** so that you can run them after any changes, and catch any issues that arise.

## Introduction to writing and running tests

<div class="h3-tight"></div>

:::: {.python-content}

### Writing a basic test

A popular framework for testing in python is **pytest**.

Each test in pytest is a function that contains an assertion statement to check a condition (e.g., `number > 0`). If the condition fails, pytest will return an error message (e.g. "The number should be positive").

Test are typically stored in a folder called `tests/`, and filenames start with the prefix `test_`. This naming convention allows pytest to automatically discover and run all the tests in the folder.

Here's an example of a simple test using pytest:

```{python}
import pytest


def test_positive():
    """
    Confirm that the number is positive.
    """
    number = 5
    assert number > 0, "The number should be positive"
```

### Running tests

Tests are typically run from the terminal. Commands include:

* `pytest` - runs all tests.
* `pytest tests/filename.py` - runs tests from a specific file.

When you run a test, you'll see an output like this in the terminal:

::: {.callout-note icon=false title="Test output:"}

```{python}
#| echo: false
pytest.main(["tests_resources/test_example_simple.py"])
```

:::

### Parametrise

We can execute the same test on different parameters using `pytest.mark.parametrize`.

Here's an example:

```{python}
@pytest.mark.parametrize("number", [1, 2, 3, -1])
def test_positive_param(number):
    """
    Confirm that the number is positive.

    Arguments:
        number (float):
            Number to check.
    """
    assert number > 0, f"The number {number} is not positive."
```

In this example, we're testing the same logic with four different values: `1`, `2`, `3`, and `-1`. The last value, `-1`, will cause the test to fail. The error message includes the failed value for easy debugging.

:::{.callout-note icon=false title="Test output:"}

```{python}
#| echo: false
pytest.main(["tests_resources/test_example_param.py"])
```

:::

::::

:::: {.r-content}

### Writing a basic test

A popular framework for testing in R is **testthat**.

Each test in testthat is a...

::::

## Testing the model

There are many different ways of categorising tests. We will focus on three types:

* **Functional testing**
* **Unit testing**
* **Back testing**

### Functional tests

Functional tests verify that the system or components perform their intended functionality.

For example, we expect that the number of arrivals should decrease if:

* The patient inter-arrival time increases.
* The length of the data collection period decreases.

:::: {.python-content}

```{python}
{{< include tests_resources/test_functional.py >}}
```

:::{.callout-note icon=false title="Test output"}

```{python}
#| echo: false
pytest.main(["tests_resources/test_functional.py"])
```

:::

::::

::: {.r-content}

:::

### Unit tests

Unit tests are a type of functional testing that focuses on individual components (e.g. methods, classes) and tests them in isolation to ensure they work as intended.

For example, we expect that our model should fail if the number of doctors or the patient inter-arrival time were set to 0. This is tested using `test_zero_inputs`:

:::: {.python-content}

```{python}
#| file: "tests_resources/test_unit.py"
```

:::{.callout-note icon=false title="Test output"}

```{python}
#| echo: false
pytest.main(["tests_resources/test_unit.py"])
```

:::

::::

::: {.r-content}

```{r}
```

:::

### Back tests

Back tests check that the model code produces results consistent with those generated historically/from prior code.

First, we'll generate a set of expected results, with a specific set of parameters. Although this may seem unnecessary in this case, as they match our default parameters, these are still specified to ensure that we are testing on the same parameters, even if defaults change.

::: {.python-content}

```{python}
param = Parameters(
    interarrival_time=5,
    consultation_time=10,
    number_of_doctors=3,
    warm_up_period=30,
    data_collection_period=40,
    number_of_runs=5,
    verbose=False
)
```

:::

::: {.r-content}

```{r}
```

:::

We'll then run the model and save the results to `.csv` files.

::: {.python-content}

```{python}
runner = Runner(param=param)
results = runner.run_reps()
results["patient"].to_csv("tests_resources/python_patient.csv", index=False)
results["run"].to_csv("tests_resources/python_run.csv", index=False)
results["overall"].to_csv("tests_resources/python_overall.csv", index=False)
```

:::

::: {.r-content}

```{r}
```

:::

In the test, we'll run the same model parameters, then import and compare against the saved `.csv` file to check for any differences.

:::: {.python-content}

```{python}
#| file: "tests_resources/test_back.py"
```

:::{.callout-note icon=false title="Test output"}

```{python}
#| echo: false
pytest.main(["tests_resources/test_back.py"])
```

:::

::::

::: {.r-content}

```{r}
```

:::

::: {.python-content}

We generate the expected results for our backtest in a **seperate Python file or Jupyter notebook**, rather than within the test itself.

:::

::: {.r-content}

We generate the expected results for our backtest in a **seperate R file or R markdown file**, rather than within the test itself.

:::

We then would generally run tests using the same pre-generated `.csv` files, without regenerating them. However, the test will **fail if the model logic is intentionally changed**, leading to different results from the same parameters.

::: {.python-content}

In that case, if we are **certain** that these changes are the reason for differing results, we should re-run the Python file or notebook to regenerate the `.csv`.

:::

::: {.r-content}

In that case, if we are **certain** that these changes are the reason for differing results, we should re-run the R file or R markdown file to regenerate the `.csv`.

:::

It is crucial to exercise caution when doing this, to avoid unintentionally overwriting correct expected results.

## Test coverage

<!-- R testthat covr CI
1. usethis::use_github_action("test-coverage") (adds action, and badge to README)
2. login to codecov with github. if repo is in organisation then need to click the button to add to apps, and authorise access to organisation. in codecov, can then switch to organisation from top left
3. github repo > settings > secrets and variables > actions > new token and copy in the info
4. run workflow -->

## Explore the example models

<div class="h3-tight"></div>

### Nurse visit simulation

:::: {.python-content}

{{< include ../../html/pydesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | |
| **What to look for?** | |
| **Why it matters?** | |
: {tbl-colwidths="[20,80]"}
:::

::::

:::: {.r-content}

{{< include ../../html/rdesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | |
| **What to look for?** | |
| **Why it matters?** | |
: {tbl-colwidths="[20,80]"}
:::

::::

### Stroke pathway simulation

:::: {.python-content}

{{< include ../../html/pydesrapstroke.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | |
| **What to look for?** | |
| **Why it matters?** | |
: {tbl-colwidths="[20,80]"}
:::

::::

:::: {.r-content}

{{< include ../../html/rdesrapstroke.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | |
| **What to look for?** | |
| **Why it matters?** | |
: {tbl-colwidths="[20,80]"}
:::

::::

## Test yourself

<!--TODO-->

<br><br>
