---
title: ❌ Number of replications
---

{{< include ../../scripts/_reticulate-setup.md >}}

:::: {.pale-blue}

**Learning objectives:**

* TODO

**Pre-reading:**

This page continues on from: [Replications](outputs.qmd).

> *Entity generation → Entity processing → Initialisation bias → Performance measures → Replications → Number of replications*

**Required packages:**

These should be available from environment setup in the "🧪 Test yourself" section of [Environments](../setup/environment.qmd). 

::: {.python-content}

```{python}
from IPython.display import HTML
from itables import to_html_datatable
import numpy as np
import pandas as pd
import scipy.stats as st
import simpy
from sim_tools.distributions import Exponential
```

:::

::: {.r-content}

```{r}
#| output: false
library(dplyr)
library(kableExtra)
library(knitr)
library(simmer)
library(tidyr)
```

:::

::::

## Confidence interval method

The **confidence interval method** can be used to select the number of replications to run. The more replications you run, the narrower your confidence interval becomes, leading to a more precise estimate of the model's mean performance.

First, you select a **desired confidence interval** - for example, 95%. Then, run the model with an **increasing number of replications**, and identify the number required to achieve that precision in the estimate of a given metric - and also, to maintain that precision (as the intervals may converge or expand again later on).

When selecting the number of replications you should **repeat the analysis for all performance measures** and select the highest value as your number of replications.

We'll show two ways of implementing this: **manually** and **automated**.

The model being used is that from the [Replications](replications.qmd) page.

```{python}
#| echo: false
{{< include replications_resources/parameters.py >}}
{{< include replications_resources/patient.py >}}
{{< include replications_resources/model.py >}}
{{< include replications_resources/summary_stats.py >}}
{{< include replications_resources/runner.py >}}
```

```{r}
#| echo: false
{{< include replications_resources/create_params.R >}}
{{< include replications_resources/filter_warmup.R >}}
{{< include replications_resources/calc_arrivals.R >}}
{{< include replications_resources/calc_mean_wait.R >}}
{{< include replications_resources/get_run_results.R >}}
{{< include replications_resources/model.R >}}
{{< include replications_resources/runner.R >}}
```

## Manual inspection

<!--TODO: Import from sim-tools...waiting for Tom's update...-->

## Algorithm

## 🔍 Explore the example models

<div class="h3-tight"></div>

### 🩺 Nurse visit simulation

<!--TODO: Mention sensitivity analysis-->

:::: {.python-content}

{{< include ../../html/pydesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | |
| **What to look for?** | |
: {tbl-colwidths="[20,80]"}
:::

::::

:::: {.r-content}

{{< include ../../html/rdesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | |
| **What to look for?** | |
: {tbl-colwidths="[20,80]"}
:::

::::

### 🧠 Stroke pathway simulation

:::: {.python-content}

{{< include ../../html/pydesrapstroke.html >}}

Not relevant - replicating an existing model described in a paper, so just used the number of replications stated by the authors.

::::

:::: {.r-content}

{{< include ../../html/rdesrapstroke.html >}}

Not relevant - replicating an existing model described in a paper, so just used the number of replications stated by the authors.

::::

## 🧪 Test yourself

TODO

<br><br>
