---
title: Replications
filters:
  - line-highlight
---

{{< include ../../scripts/_reticulate-setup.md >}}

:::: {.pale-blue}

**Learning objectives:**

* **Adapt** our model to run multiple replications.

**Pre-reading:**

This page continues on from: [Performance measures](outputs.qmd).

> *Entity generation → Entity processing → Initialisation bias → Performance measures → Replications*

**Required packages:**

These should be available from environment setup in the "🧪 Test yourself" section of [Environments](../setup/environment.qmd). 

::: {.python-content}

```{python}
# Uses HTML(to_html_datatable(df)) instead of show() due to reticulate
from IPython.display import HTML
from itables import to_html_datatable
import numpy as np
import pandas as pd
import scipy.stats as st
import simpy
from sim_tools.distributions import Exponential
```

:::

::: {.r-content}

```{r}
#| output: false
library(dplyr)
library(kableExtra)
library(knitr)
library(simmer)
library(tidyr)
```

```{r}
#| echo: false
#| output: false
# For illustrative plot only, code hidden
library(ggplot2)
```

:::

::::

<br>

Replications in discrete event simulation are necessary because each run includes randomness, leading to different results every time.

Running multiple replications reveals the **range and variability** of possible system outcomes, allowing the calculation of **averages and confidence intervals** for results.

```{r}
#| echo: false
set.seed(9L)

df <- data.frame(
  Category = rep(c("Performance measure 1", "Performance measure 2"),
                 each = 7L),
  Value = runif(14L, min = 0L, max = 1L),
  Run = rep(paste("Run", 1L:7L), 2L)
)

ggplot(df, aes(x = Category, y = Value)) +
  geom_point(size = 3L) +
  geom_text(aes(label = Run), vjust = -1, size = 3L) +
  labs(x = NULL, y = "Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 12L))
```

## 🔁 How to run replications

We will continue working with the model developed on the previous pages. To keep things simple, this version of the model does not include all the metrics described on the [Performance measures](outputs.qmd) page. Instead, it focuses on two examples: **total arrivals** and **mean wait time**.

:::: {.python-content}

### Parameter class

Let's add a new attribute `number_of_runs`. We've arbitrarily set this to five, but the [Number of replications](n_reps.qmd) page shows how to choose a suitable number to run.

```{python}
class Parameters:
    """
    Parameter class.

    Attributes
    ----------
    interarrival_time : float
        Mean time between arrivals (minutes).
    consultation_time : float
        Mean length of doctor's consultation (minutes).
    number_of_doctors : int
        Number of doctors.
    warm_up_period : int
        Duration of the warm-up period (minutes).
    data_collection_period : int
        Duration of the data collection period (minutes).
    number_of_runs : int#<<
        The number of runs (i.e., replications).#<<
    verbose : bool
        Whether to print messages as simulation runs.
    """
    def __init__(
        self, interarrival_time=5, consultation_time=10,
        number_of_doctors=3, warm_up_period=30, data_collection_period=40,
        number_of_runs=5, verbose=False#<<
    ):
        """
        Initialise Parameters instance.

        Parameters
        ----------
        interarrival_time : float
            Time between arrivals (minutes).
        consultation_time : float
            Length of consultation (minutes).
        number_of_doctors : int
            Number of doctors.
        warm_up_period : int
            Duration of the warm-up period (minutes).
        data_collection_period : int
            Duration of the data collection period (minutes).
        number_of_runs : int#<<
            The number of runs (i.e., replications).#<<
        verbose : bool
            Whether to print messages as simulation runs.
        """
        self.interarrival_time = interarrival_time
        self.consultation_time = consultation_time
        self.number_of_doctors = number_of_doctors
        self.warm_up_period = warm_up_period
        self.data_collection_period = data_collection_period
        self.number_of_runs = number_of_runs#<<
        self.verbose = verbose
```

### Patient class

No changes required.

::: {.callout title="▶&nbsp;&nbsp; Patient class" collapse="true" style="minimal"}

```{python}
class Patient:
    """
    Represents a patient.

    Attributes
    ----------
    patient_id : int
        Unique patient identifier.
    period : str
        Arrival period (warm up or data collection) with emoji.
    wait_time : float
        Time spent waiting for the doctor (minutes).
    """
    def __init__(self, patient_id, period):
        """
        Initialises a new patient.

        Parameters
        ----------
        patient_id : int
            Unique patient identifier.
        period : str
            Arrival period (warm up or data collection) with emoji.
        """
        self.patient_id = patient_id
        self.period = period
        self.wait_time = np.nan
```

:::

### Model class

No changes required.

::: {.callout title="▶&nbsp;&nbsp;Model class" collapse="true" style="minimal"}

```{python}
class Model:
    """
    Simulation model.

    Attributes
    ----------
    param : Parameters
        Simulation parameters.
    run_number : int
        Run number for random seed generation.
    env : simpy.Environment
        The SimPy environment for the simulation.
    doctor : simpy.Resource
        SimPy resource representing doctors.
    patients : list
        List of Patient objects.
    results_list : list
        List of dictionaries with the attributes of each patient.
    arrival_dist : Exponential
        Distribution used to generate random patient inter-arrival times.
    consult_dist : Exponential
        Distribution used to generate length of a doctor's consultation.
    """
    def __init__(self, param, run_number):
        """
        Create a new Model instance.

        Parameters
        ----------
        param : Parameters
            Simulation parameters.
        run_number : int
            Run number for random seed generation.
        """
        self.param = param
        self.run_number = run_number

        # Create SimPy environment
        self.env = simpy.Environment()

        # Create resource
        self.doctor = simpy.Resource(
            self.env, capacity=self.param.number_of_doctors
        )

        # Create a random seed sequence based on the run number
        ss = np.random.SeedSequence(self.run_number)
        seeds = ss.spawn(2)

        # Set up attributes to store results
        self.patients = []
        self.results_list = []

        # Initialise distributions
        self.arrival_dist = Exponential(mean=self.param.interarrival_time,
                                        random_seed=seeds[0])
        self.consult_dist = Exponential(mean=self.param.consultation_time,
                                        random_seed=seeds[1])

    def generate_arrivals(self):
        """
        Process that generates patient arrivals.
        """
        while True:
            # Sample and pass time to next arrival
            sampled_iat = self.arrival_dist.sample()
            yield self.env.timeout(sampled_iat)

            # Check whether arrived during warm-up or data collection
            if self.env.now < self.param.warm_up_period:
                period = "\U0001F538 WU"
            else:
                period = "\U0001F539 DC"

            # Create a new patient
            patient = Patient(patient_id=len(self.patients)+1,
                              period=period)
            self.patients.append(patient)

            # Print arrival time
            if self.param.verbose:
                print(f"{patient.period} Patient {patient.patient_id} " +
                      f"arrives at time: {self.env.now:.3f}")

            # Start process of consultation
            self.env.process(self.consultation(patient))

    def consultation(self, patient):
        """
        Process that simulates a consultation.

        Parameters
        ----------
        patient :
            Instance of the Patient() class representing a single patient.
        """
        start_wait = self.env.now
        # Patient requests access to a doctor (resource)
        with self.doctor.request() as req:
            yield req

            # Record how long patient waited before consultation
            patient.wait_time = self.env.now - start_wait

            if self.param.verbose:
                print(f"{patient.period} Patient {patient.patient_id} " +
                      f"starts consultation at: {self.env.now:.3f}")

            # Sample consultation duration and pass time spent with doctor
            time_with_doctor = self.consult_dist.sample()
            yield self.env.timeout(time_with_doctor)

    def reset_results(self):
        """
        Reset results.
        """
        self.patients = []

    def warmup(self):
        """
        Reset result collection after the warm-up period.
        """
        if self.param.warm_up_period > 0:
            # Delay process until warm-up period has completed
            yield self.env.timeout(self.param.warm_up_period)
            # Reset results variables
            self.reset_results()
            if self.param.verbose:
                print(f"Warm up period ended at time: {self.env.now}")

    def run(self):
        """
        Run the simulation.
        """
        # Schedule arrival generator and warm-up
        self.env.process(self.generate_arrivals())
        self.env.process(self.warmup())

        # Run the simulation
        self.env.run(until=(self.param.warm_up_period +
                            self.param.data_collection_period))

        # Create list of dictionaries containing each patient's attributes
        self.results_list = [x.__dict__ for x in self.patients]
```

:::

### Summary statistics function

This function computes descriptive statistics for a dataset, returning the mean, standard deviation, and 95% confidence interval.

* If the input has **no valid values**, all outputs are `NaN`.

* If the input has **fewer than 3 values**, it returns the mean only, with the rest as `NaN`.

* If the input has **3 or more values**, it calculates the mean, standard deviation, and a 95% confidence interval using the t-distribution (`scipy.stats`).

* If all values are identical (zero variance), the confidence interval equals the mean.

```{python}
def summary_stats(data):
    """
    Calculate mean, standard deviation and 95% confidence interval (CI).

    Parameters
    ----------
    data : pd.Series
        Data to use in calculation.

    Returns
    -------
    tuple
        (mean, standard deviation, CI lower, CI upper).
    """
    # Remove any NaN from the series
    data = data.dropna()

    # Find number of observations
    count = len(data)

    # If there are no observations, then set all to NaN
    if count == 0:
        mean, std_dev, ci_lower, ci_upper = np.nan, np.nan, np.nan, np.nan
    # If there is only one or two observations, can do mean but not others
    elif count < 3:
        mean = data.mean()
        std_dev, ci_lower, ci_upper = np.nan, np.nan, np.nan
    # With more than one observation, can calculate all...
    else:
        mean = data.mean()
        std_dev = data.std()
        # Special case for CI if variance is 0
        if np.var(data) == 0:
            ci_lower, ci_upper = mean, mean
        else:
            # Calculation of CI uses t-distribution, which is suitable for
            # smaller sample sizes (n<30)
            ci_lower, ci_upper = st.t.interval(
                confidence=0.95,
                df=count-1,
                loc=mean,
                scale=st.sem(data))
    return mean, std_dev, ci_lower, ci_upper
```

### Runner class

We run replications with a new method `run_reps()`, explained line-by-line below.

```{python}
class Runner:
    """
    Run the simulation.

    Attributes
    ----------
    param : Parameters
        Simulation parameters.
    """
    def __init__(self, param):
        """
        Initialise a new instance of the Runner class.

        Parameters
        ----------
        param : Parameters
            Simulation parameters.
        """
        self.param = param

    def run_single(self, run):
        """
        Runs the simulation once and performs results calculations.

        Parameters
        ----------
        run : int
            Run number for the simulation.

        Returns
        -------
        dict
            Contains patient-level results and results from each run.
        """
        model = Model(param=self.param, run_number=run)
        model.run()

        # Patient results
        patient_results = pd.DataFrame(model.results_list)
        patient_results["run"] = run

        # Run results
        run_results = {
            "run_number": run,
            "arrivals": len(patient_results),
            "mean_wait_time": patient_results["wait_time"].mean()
        }

        return {
            "patient": patient_results,
            "run": run_results
        }

    def run_reps(self):#<<
        """#<<
        Execute a single model configuration for multiple runs.#<<
        """#<<
        # Run replications#<<
        all_results = [self.run_single(run)#<<
                       for run in range(self.param.number_of_runs)]#<<

        # Separate results from each run into appropriate lists#<<
        patient_results_list = [result["patient"] for result in all_results]#<<
        run_results_list = [result["run"] for result in all_results]#<<

        # Convert lists into dataframes#<<
        patient_results_df = pd.concat(#<<
            patient_results_list, ignore_index=True#<<
        )#<<
        run_results_df = pd.DataFrame(run_results_list)#<<

        # Calculate average results and uncertainty from across all runs#<<
        uncertainty_metrics = {}#<<
        run_col = run_results_df.columns#<<

        # Loop through the run performance measure columns#<<
        # Calculate mean, standard deviation and 95% confidence interval#<<
        for col in run_col[~run_col.isin(["run_number", "scenario"])]:#<<
            uncertainty_metrics[col] = dict(zip(#<<
                ["mean", "std_dev", "lower_95_ci", "upper_95_ci"],#<<
                summary_stats(run_results_df[col])#<<
            ))
        # Convert to dataframe#<<
        overall_results_df = pd.DataFrame(uncertainty_metrics)#<<

        return {#<<
            "patient": patient_results_df,#<<
            "run": run_results_df,#<<
            "overall": overall_results_df#<<
        }#<<
```

#### Explaining `Runner.run_reps()`

```{.python}
def run_reps(self):
    """
    Execute a single model configuration for multiple runs.
    """
    # Run replications
    all_results = [self.run_single(run)
                    for run in range(self.param.number_of_runs)]
```

It starts by calling `run_single()` repeatedly for the requested number of simulation runs.

<br>

```{.python}
# Separate results from each run into appropriate lists
patient_results_list = [result["patient"] for result in all_results]
run_results_list = [result["run"] for result in all_results]

# Convert lists into dataframes
patient_results_df = pd.concat(
    patient_results_list, ignore_index=True
)
run_results_df = pd.DataFrame(run_results_list)
```

The results then get stacked on top of each other, so you end up with one big dataframe for all patients, and another for all runs. The `replication` column shows which run each row came from.

<br>

```{.python}
# Calculate average results and uncertainty from across all runs
uncertainty_metrics = {}
run_col = run_results_df.columns

# Loop through the run performance measure columns
# Calculate mean, standard deviation and 95% confidence interval
for col in run_col[~run_col.isin(["run_number", "scenario"])]:
    uncertainty_metrics[col] = dict(zip(
        ["mean", "std_dev", "lower_95_ci", "upper_95_ci"],
        summary_stats(run_results_df[col])
    ))
# Convert to dataframe
overall_results_df = pd.DataFrame(uncertainty_metrics)
```

We then calculate summary statistics (mean, standard deviation, and 95% confidence interval) for each run in `run_results_df` using `summary_stats()`.

<br>

```{.python}
return {
    "patient": patient_results_df,
    "run": run_results_df,
    "overall": overall_results_df
}
```

At the end, the function returns a dictionary with all patient results (`patient`), all run results (`run`), and a summary table with statistics calculated across all runs (`overall`).

### Run the model

```{python}
runner = Runner(param=Parameters())
result = runner.run_reps()
HTML(to_html_datatable(result["patient"]))
HTML(to_html_datatable(result["run"]))
HTML(to_html_datatable(result["overall"]))
```

::::

:::: {.r-content}

### Parameter function

Let's add a new attribute `number_of_runs`. We've arbitrarily set this to five, but the [Number of replications](n_reps.qmd) page shows how to choose a suitable number to run.

```{r}
#' Generate parameter list.
#'
#' @param interarrival_time Numeric. Time between arrivals (minutes).
#' @param consultation_time Numeric. Mean length of doctor's
#'   consultation (minutes).
#' @param number_of_doctors Numeric. Number of doctors.
#' @param warm_up_period Numeric. Duration of the warm-up period (minutes).
#' @param data_collection_period Numeric. Duration of the data collection
#'   period (minutes).
#' @param number_of_runs Numeric. The number of runs (i.e., replications).#<<
#' @param verbose Boolean. Whether to print messages as simulation runs.
#'
#' @return A named list of parameters.

create_params <- function(
  interarrival_time = 5L,
  consultation_time = 10L,
  number_of_doctors = 3L,
  warm_up_period = 30L,
  data_collection_period = 40L,
  number_of_runs = 5L,#<<
  verbose = FALSE
) {
  list(
    interarrival_time = interarrival_time,
    consultation_time = consultation_time,
    number_of_doctors = number_of_doctors,
    warm_up_period = warm_up_period,
    data_collection_period = data_collection_period,
    number_of_runs = number_of_runs,#<<
    verbose = verbose
  )
}
```

### Warm-up function

No changes required.

::: {.callout title="▶&nbsp;&nbsp;Warm-up function" collapse="true" style="minimal"}

```{r}
#' Filters arrivals and resources to remove warm-up results.
#'
#' @param result Named list with two tables: monitored arrivals & resources.
#' @param warm_up_period Length of warm-up period.
#'
#' @importFrom dplyr ungroup arrange slice n filter
#'
#' @return The name list `result`, but with the tables (`arrivals` and
#' `resources`) filtered to remove warm-up results.
#' @export

filter_warmup <- function(result, warm_up_period) {
  # Skip filtering if the warm-up period is zero
  if (warm_up_period == 0L) return(result)

  # Arrivals: Keep only patients who came during the data collection period
  result[["arrivals"]] <- result[["arrivals"]] |>
    group_by(.data[["name"]]) |>
    filter(all(.data[["start_time"]] >= warm_up_period)) |>
    ungroup()

  # Resources: Filter to resource events in the data collection period
  dc_resources <- filter(result[["resources"]],
                         .data[["time"]] >= warm_up_period)

  if (nrow(dc_resources) > 0L) {

    # For each resource, get the last even during warm-up, and replace time
    # for that event with the start time of the data collection period
    last_usage <- result[["resources"]] |>
      filter(.data[["time"]] < warm_up_period) |>
      arrange(.data[["time"]]) |>
      group_by(.data[["resource"]]) |>
      slice(n()) |>
      mutate(time = warm_up_period) |>
      ungroup()

    # Set that last event of the resource/s as the first row of the dataframe
    # else calculations would assume all resources were idle at the start of
    # the data collection period
    result[["resources"]] <- rbind(last_usage, dc_resources)

  } else {
    # No events after warm-up; use filtered resource table as is
    result[["resources"]] <- dc_resources
  }

  result
}
```

:::

### Arrivals function

No changes required.

::: {.callout title="▶&nbsp;&nbsp;Arrivals function" collapse="true" style="minimal"}

```{r}
#' Calculate the number of arrivals
#'
#' @param arrivals Dataframe with times for each patient with each resource.
#'
#' @return Tibble with column containing total number of arrivals.
#' @export

calc_arrivals <- function(arrivals) {
  arrivals |>
    summarise(arrivals = n_distinct(.data[["name"]])) |>
    ungroup()
}
```

:::

### Wait time function

No changes required.

::: {.callout title="▶&nbsp;&nbsp;Wait time function" collapse="true" style="minimal"}

```{r}
#' Calculate the mean wait time for each resource
#'
#' @param arrivals Dataframe with times for each patient with each resource.
#'
#' @importFrom tidyr drop_na
#' 
#' @return Tibble with columns containing result for each resource.
#' @export

calc_mean_wait <- function(arrivals) {

  # Create subset of data that removes patients who were still waiting
  complete_arrivals <- drop_na(arrivals, any_of("wait_time"))

  # Calculate mean wait time for each resource
  complete_arrivals |>
    group_by(resource) |>
    summarise(mean_wait_time = mean(wait_time)) |>
    pivot_wider(names_from = "resource",
                values_from = "mean_wait_time",
                names_glue = "mean_wait_time_{resource}") |>
    ungroup()
}
```

:::

### Run results function

No changes required.

::: {.callout title="▶&nbsp;&nbsp;Run results function" collapse="true" style="minimal"}

```{r}
#' Get average results for the provided single run.
#'
#' @param results Named list with `arrivals` from `get_mon_arrivals()` and
#'   `resources` from `get_mon_resources()`
#'   (`per_resource = TRUE` and `ongoing = TRUE`).
#' @param run_number Integer index of current simulation run.
#'
#' @importFrom dplyr bind_cols
#' @importFrom tibble tibble
#'
#' @return Tibble with processed results from replication.
#' @export

get_run_results <- function(results, run_number) {
  metrics <- list(
    calc_arrivals(results[["arrivals"]]),
    calc_mean_wait(results[["arrivals"]])
  )
  dplyr::bind_cols(tibble(replication = run_number), metrics)
}
```

:::

### Model function

No changes required.

::: {.callout title="▶&nbsp;&nbsp;Model function" collapse="true" style="minimal"}

```{r}
#' Run simulation.
#'
#' @param param List. Model parameters.
#' @param run_number Numeric. Run number for random seed generation.
#'
#' @importFrom dplyr mutate
#' @importFrom simmer add_generator get_mon_arrivals get_mon_resources run
#' @importFrom simmer set_attribute simmer timeout trajectory

model <- function(param, run_number) {

  # Set random seed based on run number
  set.seed(run_number)

  # Create simmer environment
  env <- simmer("simulation", verbose = param[["verbose"]])

  # Define the patient trajectory
  patient <- trajectory("consultation") |>
    seize("doctor", 1L) |>
    # Record time resource is obtained
    set_attribute("doctor_serve_start", function() now(env)) |>
    timeout(function() {
      rexp(n = 1L, rate = 1L / param[["consultation_time"]])
    }) |>
    release("doctor", 1L)

  env <- env |>
    # Add doctor resource
    add_resource("doctor", param[["number_of_doctors"]]) |>
    # Add patient generator (mon=2 so can get manual attributes)
    add_generator("patient", patient, function() {
      rexp(n = 1L, rate = 1L / param[["interarrival_time"]])
    }, mon = 2L) |>
    # Run the simulation
    simmer::run(until = (param[["warm_up_period"]] +
                           param[["data_collection_period"]]))

  # Extract information on arrivals and resources from simmer environment
  result <- list(
    arrivals = get_mon_arrivals(env, per_resource = TRUE, ongoing = TRUE), 
    resources = get_mon_resources(env)
  )

  # Get the extra arrivals attributes
  extra_attributes <- get_mon_attributes(env) |>
    dplyr::select("name", "key", "value") |>
    # Add column with resource name, and remove that from key
    mutate(resource = gsub("_.+", "", .data[["key"]]),
           key = gsub("^[^_]+_", "", .data[["key"]])) |>
    pivot_wider(names_from = "key", values_from = "value")

  # Merge extra attributes with the arrival data
  result[["arrivals"]] <- left_join(
    result[["arrivals"]], extra_attributes, by = c("name", "resource")
  )

  # Filter to remove results from the warm-up period
  result <- filter_warmup(result, param[["warm_up_period"]])

  # Replace "replication" value with appropriate run number
  result[["arrivals"]] <- mutate(result[["arrivals"]],
                                 replication = run_number)
  result[["resources"]] <- mutate(result[["resources"]],
                                  replication = run_number)

  # Calculate wait time for each patient
  result[["arrivals"]] <- result[["arrivals"]] |>
    mutate(wait_time = .data[["serve_start"]] - .data[["start_time"]])

  # Calculate the average results for that run
  result[["run_results"]] <- get_run_results(result, run_number)

  result
}
```

:::

### Runner function

This new function `runner()` runs the simulation model multiple times based on the specified `param[["number_of_runs"]]`. It combines the results into unified output tables for arrivals, resources and run-level results. If only one replication is requested, it simply returns that single result without combining.

```{r}
#' Run simulation for multiple replications.
#' 
#' @param param Named list of model parameters.
#' 
#' @return List with three tables: arrivals, resources and run_results.
#' @export

runner <- function(param) {
  # Run simulation for multiple replications
  results <- lapply(
    seq_len(param[["number_of_runs"]]),
    function(i) {
      model(run_number = i, param = param)
    }
  )

  # If only one replication, return without changes
  if (param[["number_of_runs"]] == 1L) {
    return(results[[1L]])
  }

  # Combine results from all replications into unified tables
  all_arrivals <- do.call(
    rbind, lapply(results, function(x) x[["arrivals"]])
  )
  all_resources <- do.call(
    rbind, lapply(results, function(x) x[["resources"]])
  )

  # run_results may have missing data - use bind_rows to fill NAs automatically
  all_run_results <- dplyr::bind_rows(
    lapply(results, function(x) x[["run_results"]])
  )

  # Return a list containing the combined tables
  return(
    list(
      arrivals = all_arrivals,
      resources = all_resources,
      run_results = all_run_results
    )
  )
}
```

**Note:** If you are recording patients in service, just add:

```
all_patients_in_service <- dplyr::bind_rows(
  lapply(results, function(x) x[["patients_in_service"]])
)
```

### Run the model

```{r}
param <- create_params()
result <- runner(param = param)
result[["arrivals"]] |>
  arrange(replication, start_time) |>
  kable() |>
  scroll_box(height = "400px")
kable(result[["resources"]]) |> scroll_box(height = "400px")
kable(result[["run_results"]])
```

::::

## 🔍 Explore the example models

<div class="h3-tight"></div>

### 🩺 Nurse visit simulation

:::: {.python-content}

{{< include ../../html/pydesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `simulation/param.py`<br>`simulation/summary_stats.py`<br>`simulation/runner.py` |
| **What to look for?** | This is set up to run replications in parallel by default, as introduced on the [Parallel processing](parallel.qmd) page. |
: {tbl-colwidths="[20,80]"}
:::

::::

:::: {.r-content}

{{< include ../../html/rdesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `R/parameters.R`<br>`R/runner.R` |
| **What to look for?** | This is set up to run replications sequentially by default, but has functionality for running in parallel as introduced on the [Parallel processing](parallel.qmd) page. This means some differences in set up (with sequential also using `future_lapply` in that case - see parallel page for more details). |
: {tbl-colwidths="[20,80]"}
:::

::::

### 🧠 Stroke pathway simulation

:::: {.python-content}

{{< include ../../html/pydesrapstroke.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `simulation/parameters.py`<br>`simulation/runner.py` |
| **What to look for?** | Like nurse visit simulation, this is also set up to support parallel processing. |
: {tbl-colwidths="[20,80]"}
:::

::::

:::: {.r-content}

{{< include ../../html/rdesrapstroke.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `R/parameters.R`<br>`R/runner.R` |
| **What to look for?** | Like nurse visit simulation, this is also set up to support parallel processing. |
: {tbl-colwidths="[20,80]"}
:::

::::

## 🧪 Test yourself

Extend your model to support multiple replications, as shown above.

<br><br>
