---
title: Replications
filters:
  - line-highlight
---

{{< include ../../scripts/_reticulate-setup.md >}}

```{python}
#| echo: false
# pylint: disable=undefined-variable, unused-import
```

:::: {.pale-blue}

**Learning objectives:**

* **Adapt** our model to run multiple replications.

**Pre-reading:**

This page continues on from: [Performance measures](outputs.qmd).

> *Entity generation → Entity processing → Initialisation bias → Performance measures → Replications*

**Required packages:**

These should be available from environment setup in the "Test yourself" section of [Environments](../setup/environment.qmd). 

::: {.python-content}

```{python}
from IPython.display import HTML
from itables import to_html_datatable
import numpy as np
import pandas as pd
import scipy.stats as st
import simpy
from sim_tools.distributions import Exponential
```

:::

::: {.r-content}

```{r}
#| output: false
library(dplyr)
library(kableExtra)
library(knitr)
library(simmer)
library(tidyr)
```

```{r}
#| echo: false
#| output: false
# For illustrative plot only, code hidden
library(ggplot2)
```

:::

::::

<br>

Replications in discrete event simulation are necessary because each run includes randomness, leading to different results every time.

Running multiple replications reveals the **range and variability** of possible system outcomes, allowing the calculation of **averages and confidence intervals** for results.

```{r}
#| echo: false
set.seed(9L)

df <- data.frame(
  Category = rep(c("Performance measure 1", "Performance measure 2"),
                 each = 7L),
  Value = runif(14L, min = 0L, max = 1L),
  Run = rep(paste("Run", 1L:7L), 2L),
  stringsAsFactors = FALSE
)

ggplot(df, aes(x = Category, y = Value)) +
  geom_point(size = 3L) +
  geom_text(aes(label = Run), vjust = -1L, size = 3L) +
  labs(x = NULL, y = "Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 12L))
```

## How to run replications

We will continue working with the model developed on the previous pages. This model contains all the metrics from the [Performance measures](outputs.qmd) page (with utilisation implemented via the time-weighted approach).

:::: {.python-content}

### Parameter class

Let's add a new attribute `number_of_runs`. We've arbitrarily set this to five, but the [Number of replications](n_reps.qmd) page shows how to choose a suitable number to run.

```{python}
{{< include replications_resources/parameters.py >}}
```

### Patient class

No changes required.

::: {.callout title="▶&nbsp;&nbsp; Patient class" collapse="true" style="minimal"}

```{python}
{{< include replications_resources/patient.py >}}
```

:::

### Monitored resource class

No changes required.

::: {.callout title="▶&nbsp;&nbsp; Monitored resource class" collapse="true" style="minimal"}

```{python}
{{< include replications_resources/monitored_resource.py >}}
```

:::

### Model class

No changes required.

::: {.callout title="▶&nbsp;&nbsp;Model class" collapse="true" style="minimal"}

```{python}
{{< include replications_resources/model.py >}}
```

:::

### Summary statistics function

This function computes descriptive statistics for a dataset, returning the mean, standard deviation, and 95% confidence interval.

* If the input has **no valid values**, all outputs are `NaN`.

* If the input has **fewer than 3 values**, it returns the mean only, with the rest as `NaN`.

* If the input has **3 or more values**, it calculates the mean, standard deviation, and a 95% confidence interval using the t-distribution (`scipy.stats`).

* If all values are identical (zero variance), the confidence interval equals the mean.

```{python}
{{< include replications_resources/summary_stats.py >}}
```

### Runner class

We run replications with a new method `run_reps()`, explained line-by-line below.

```{python}
{{< include replications_resources/runner.py >}}
```

#### Explaining `Runner.run_reps()`

```{.python}
def run_reps(self):
    """
    Execute a single model configuration for multiple runs.

    Returns
    -------
    dict
        Contains patient-level results, results from each run and
        overall results.
    """
    # Run replications
    all_results = [self.run_single(run)
                    for run in range(self.param.number_of_runs)]
```

It starts by calling `run_single()` repeatedly for the requested number of simulation runs, passing the run number to set the random seed (ensuring each run uses a different seed!).

<br>

```{.python}
# Separate results from each run into appropriate lists
patient_results_list = [result["patient"] for result in all_results]
run_results_list = [result["run"] for result in all_results]

# Convert lists into dataframes
patient_results_df = pd.concat(
    patient_results_list, ignore_index=True
)
run_results_df = pd.DataFrame(run_results_list)
```

The results then get stacked on top of each other, so you end up with one big dataframe for all patients, and another for all runs. The `replication` column shows which run each row came from.

<br>

```{.python}
# Calculate average results and uncertainty from across all runs
uncertainty_metrics = {}
run_col = run_results_df.columns

# Loop through the run performance measure columns
# Calculate mean, standard deviation and 95% confidence interval
for col in run_col[~run_col.isin(["run_number", "scenario"])]:
    uncertainty_metrics[col] = dict(zip(
        ["mean", "std_dev", "lower_95_ci", "upper_95_ci"],
        summary_stats(run_results_df[col])
    ))
# Convert to dataframe
overall_results_df = pd.DataFrame(uncertainty_metrics)
```

We then calculate summary statistics (mean, standard deviation, and 95% confidence interval) for each run in `run_results_df` using `summary_stats()`.

<br>

```{.python}
return {
    "patient": patient_results_df,
    "run": run_results_df,
    "overall": overall_results_df
}
```

At the end, the function returns a dictionary with all patient results (`patient`), all run results (`run`), and a summary table with statistics calculated across all runs (`overall`).

### Run the model

```{python}
runner = Runner(param=Parameters())
result = runner.run_reps()
HTML(to_html_datatable(result["patient"]))
HTML(to_html_datatable(result["run"]))
HTML(to_html_datatable(result["overall"]))
```

::::

:::: {.r-content}

### Parameter function

Let's add a new attribute `number_of_runs`. We've arbitrarily set this to five, but the [Number of replications](n_reps.qmd) page shows how to choose a suitable number to run.

```{r}
#| file: replications_resources/create_params.R
```

### Warm-up function

No changes required.

::: {.callout title="▶&nbsp;&nbsp;Warm-up function" collapse="true" style="minimal"}

```{r}
#| file: replications_resources/filter_warmup.R
```

:::

### Arrivals function

No changes required.

::: {.callout title="▶&nbsp;&nbsp;Arrivals function" collapse="true" style="minimal"}

```{r}
#| file: replications_resources/calc_arrivals.R
```

:::

### Wait time function

No changes required.

::: {.callout title="▶&nbsp;&nbsp;Wait time function" collapse="true" style="minimal"}

```{r}
#| file: replications_resources/calc_mean_wait.R
```

:::

### Run results function

No changes required.

::: {.callout title="▶&nbsp;&nbsp;Run results function" collapse="true" style="minimal"}

```{r}
#| file: replications_resources/get_run_results.R
```

:::

### Model function

No changes required.

::: {.callout title="▶&nbsp;&nbsp;Model function" collapse="true" style="minimal"}

```{r}
#| file: replications_resources/model.R
```

:::

### Runner function

This new function `runner()` runs the simulation model multiple times based on the specified `param[["number_of_runs"]]`. It passes the run number to set the random seed (ensuring each run uses a different seed!).

It combines the results into unified output tables for arrivals, resources and run-level results. If only one replication is requested, it simply returns that single result without combining.

```{r}
#| file: replications_resources/runner.R
```

**Note:** If you are recording patients in service, just add:

```
all_patients_in_service <- dplyr::bind_rows(
  lapply(results, function(x) x[["patients_in_service"]])
)
```

### Run the model

```{r}
param <- create_params()
result <- runner(param = param)
result[["arrivals"]] |>
  arrange(replication, start_time) |>
  kable() |>
  scroll_box(height = "400px")
kable(result[["resources"]]) |> scroll_box(height = "400px")
kable(result[["run_results"]])
```

::::

## Explore the example models

<div class="h3-tight"></div>

### Nurse visit simulation

:::: {.python-content}

{{< include ../../html/pydesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `simulation/param.py`<br>`simulation/summary_stats.py`<br>`simulation/runner.py` |
| **What to look for?** | This is set up to run replications in parallel by default, as introduced on the [Parallel processing](parallel.qmd) page. |
: {tbl-colwidths="[20,80]"}
:::

::::

:::: {.r-content}

{{< include ../../html/rdesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `R/parameters.R`<br>`R/runner.R` |
| **What to look for?** | This is set up to run replications sequentially by default, but has functionality for running in parallel as introduced on the [Parallel processing](parallel.qmd) page. This means some differences in set up (with sequential also using `future_lapply` in that case - see parallel page for more details). |
: {tbl-colwidths="[20,80]"}
:::

::::

### Stroke pathway simulation

:::: {.python-content}

{{< include ../../html/pydesrapstroke.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `simulation/parameters.py`<br>`simulation/runner.py` |
| **What to look for?** | Like nurse visit simulation, this is also set up to support parallel processing. |
: {tbl-colwidths="[20,80]"}
:::

::::

:::: {.r-content}

{{< include ../../html/rdesrapstroke.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `R/parameters.R`<br>`R/runner.R` |
| **What to look for?** | Like nurse visit simulation, this is also set up to support parallel processing. |
: {tbl-colwidths="[20,80]"}
:::

::::

## Test yourself

Extend your model to support multiple replications, as shown above.

<br><br>
