---
title: ✅ Parameters from file
---

{{< include ../../scripts/_reticulate-setup.md >}}

:::{.guidelines}
* Heather et al. 2025: Avoid hard-coded parameters.
* NHS Levels of RAP (🥈): Data is handled and output in a Tidy data format.
:::

## Introduction

Discrete-event simulations (DES) require many parameters - like arrival rates, resource times, and probabilities - which often need to be changed for different scenarios and analyses. **Managing these parameters well makes your simulations easier to update, track, and reuse**.

This page focuses on the **storage of parameters within a file**. If you want to see how to store parameters within a script, see the [parameters from script](parameters_script.qmd) page.

<br>

## Why use external parameter files?

External parameter files can offer some advantages over storing parameters directly in scripts:

* Easier for non-programmers to modify parameters.
* Can easily share your simulation with different parameters versions (e.g. real parameters for internal use, synthetic parameters for external sharing, as discussed in [input data management](input_data.qmd)) without editing the code itself.
* Less hard coding as parameters are completely separate.

Though storing parameters within a script is commonplace - particularly for simpler models - due to its simplicity.

<br>

## Create parameter file

There are several file types you could consider for storing parameters. CSV files are often a good choice - particularly better than Excel - because they are:

* Open and non-proprietary format.
* Compatible with version-control systems (being text-based).
* Readable in wide range of spreadsheet software (Excel, LibreOffice), text editors, and programming environments
* Easy to import and parse in major programming languages.

The parameter file should be structured following tidy data principles. This means each row represents a single observation and each column represents a variable. Avoid using merged cells or stacked headers as these make the data harder to process programmatically.

:::{.python-content}

```{python}
# Import required packages
import pandas as pd

# Import and preview parameter file
pd.read_csv("parameters_file_resources/example_parameters.csv")
```

:::

:::{.r-content}

```{r}
# Import and preview parameter file
read.csv(file.path("parameters_file_resources", "example_parameters.csv"))
```

:::

<br>

## Create data dictionary

As discussed on the [input data management page](input_data.qmd), you should include a data dictionary or documentation describing each parameter. This documentation should explain the parameter's meaning, units, and any abbreviations or codes used.

For example:

{{< include parameters_file_resources/example_data_dictionary.md >}}

<br>

## Import parameters into function or class

The [parameters from script page](parameters_script.qmd) provides detailed guidance on how parameters should be structured within your simulation. The key principles are:

1. **Group parameters into a dedicated object**.
2. **Pass these objects explicitly to your model**.

These same principles apply when importing parameters from files.  We recommend using **structured parameter objects** rather than passing raw CSV data directly to your simulation functions because:

* Your files might contain baseline parameters plus scenario-specific overrides, but your simulation needs **one complete object** with all parameters merged together. Having all required parameters for a run in one object also makes it clear exactly what's being used.
* Classes allow you to incorporate parameter **validation** (e.g. ranges, types) before running the simulation.
* With an object, you can keep **consistent parameter names** in your code, while allowing names or file formats to change, as helper functions can translate between the file and parameter object.

However, there is one major difference when working with parameters from files: **you should not include default values in your parameter classes**. All values should come explicitly from your data sources. This ensures your parameter choices are intentional and missing parameters fail clearly instead of silently using unexpected defaults.

<br>

### Define parameter function or class

Parameters can be stored using one or multiple functions or classes (see the [parameters from script page](parameters_script.qmd) for detailed discussion).

:::{.python-content}

Here, we show a single class approach:

```{python}
class Parameters:  # pylint: disable=too-many-instance-attributes, too-many-arguments, too-many-positional-arguments, too-few-public-methods
    """
    Container for simulation parameters across patient demographics.
    """
    def __init__(
        self,
        adult_interarrival, adult_consultation, adult_transfer,
        child_interarrival, child_consultation, child_transfer,
        elderly_interarrival, elderly_consultation, elderly_transfer
    ):
        """
        Initialise Parameters instance.

        Parameters
        ----------
        adult_interarrival : float
            Time between adult patient arrivals (e.g., in minutes or hours).
        adult_consultation : float  
            Duration of consultation time for adult patients.
        adult_transfer : float
            Time required to transfer adult patients between stages.
        child_interarrival : float
            Time between child patient arrivals (e.g., in minutes or hours).
        child_consultation : float
            Duration of consultation time for child patients.
        child_transfer : float
            Time required to transfer child patients between stages.
        elderly_interarrival : float
            Time between elderly patient arrivals (e.g., in minutes or hours).
        elderly_consultation : float
            Duration of consultation time for elderly patients.
        elderly_transfer : float
            Time required to transfer elderly patients between stages.
        """
        # Adult parameters
        self.adult_interarrival = adult_interarrival
        self.adult_consultation = adult_consultation
        self.adult_transfer = adult_transfer
        # Child parameters
        self.child_interarrival = child_interarrival
        self.child_consultation = child_consultation
        self.child_transfer = child_transfer
        # Elderly parameters
        self.elderly_interarrival = elderly_interarrival
        self.elderly_consultation = elderly_consultation
        self.elderly_transfer = elderly_transfer
```

:::

:::{.r-content}

Here, we show a single function approach:

```{r}
#' Create parameter list for simulation
#'
#' @param adult_interarrival Numeric. Time between adult patient arrivals
#'   (e.g. in minutes or hours).
#' @param adult_consultation Numeric. Duration of consultation time for
#'   adult patients.
#' @param adult_transfer Numeric. Time required to transfer adult patients
#'   between stages.
#' @param child_interarrival Numeric. Time between child patient arrivals
#'   (e.g. in minutes or hours).
#' @param child_consultation Numeric. Duration of consultation time for
#'   child patients.
#' @param child_transfer Numeric. Time required to transfer child patients
#'   between stages.
#' @param elderly_interarrival Numeric. Time between elderly patient arrivals
#'   (e.g. in minutes or hours).
#' @param elderly_consultation Numeric. Duration of consultation time for
#'   elderly patients.
#' @param elderly_transfer Numeric. Time required to transfer elderly patients
#'   between stages.
#'
#' @return A named list containing all simulation parameters organised by
#'   patient demographic groups (adult, child, elderly). Each group contains
#'   interarrival, consultation, and transfer time parameters.

create_params <- function(
  adult_interarrival, adult_consultation, adult_transfer,
  child_interarrival, child_consultation, child_transfer,
  elderly_interarrival, elderly_consultation, elderly_transfer
) {
  list(
    # Adult parameters
    adult_interarrival = adult_interarrival,
    adult_consultation = adult_consultation,
    adult_transfer = adult_transfer,
    # Child parameters
    child_interarrival = child_interarrival,
    child_consultation = child_consultation,
    child_transfer = child_transfer,
    # Elderly parameters
    elderly_interarrival = elderly_interarrival,
    elderly_consultation = elderly_consultation,
    elderly_transfer = elderly_transfer
  )
}
```

:::

<br>

### Write helper function to import parameters

Create helper functions to load parameters from your CSV file into your functions/classes. How you write this function will depend on your parameter class structure and how your parameter files are organised.

For example:

:::{.python-content}

```{python}
def setup_param_from_csv(csv_path):
    """
    Create a Parameters instance using parameter values loaded from a CSV file.

    Parameters
    ----------
    csv_path : str
        Path to CSV file containing the parameters. Should have columns
        "patient", "metric", and "value".

    Returns
    -------
    Parameters
        Instance of Parameters initialised with parameters from the CSV file.
    """

    # Import the parameters
    param_file = pd.read_csv(csv_path)

    # Create name-value mappings
    values = {
      f"{row.patient}_{row.metric}": row.value
      for row in param_file.itertuples()
    }

    # Pass these values to the Parameters class
    return Parameters(**values)
```

:::

:::{.r-content}

```{r}
#' Create parameter list from CSV file
#'
#' @param csv_path Character. Path to CSV file containing the parameters.
#'   The CSV file should have columns "patient", "metric", and "value".
#'   The "patient" column should contain demographic groups (e.g., "adult",
#'   "child", "elderly"), the "metric" column should contain parameter types
#'   (e.g., "interarrival", "consultation", "transfer"), and the "value"
#'   column should contain the numeric parameter values.
#'
#' @return A named list containing all simulation parameters organised by
#'   patient demographic groups, identical to the output of
#'   \code{create_params()}. The list contains interarrival, consultation, and
#'   transfer time parameters for each patient group.

setup_param_from_csv <- function(csv_path) {

  # Import the parameters
  param_file <- read.csv(csv_path)

  # Create named vector where names match function arguments
  values <- setNames(
    param_file$value, paste(param_file$patient, param_file$metric, sep = "_")
  )

  # Pass the named list to create_params
  do.call(create_params, as.list(values))
}
```

:::

<br>

### Load and use parameters

:::{.python-content}

```{python}
# Import parameters
params = setup_param_from_csv(
    "parameters_file_resources/example_parameters.csv")

# View object
print(params.__dict__)
```

:::

:::{.r-content}

```{r}
# Import parameters
params <- setup_param_from_csv(
  file.path("parameters_file_resources", "example_parameters.csv")
)

# View objects
params
```

:::

## Examples

This section contains full code examples for our [example conceptual models](../intro/examples.qmd).

---

<details>
<summary>Show/Hide example 1: 🩺 Nurse visit simulation</summary>
<br>

We didn't create a file-based example for the nurse M/M/s model due to its simplicity with only a few parameters.

<br>
</details>

---

<details>
<summary>Show/Hide example 2: 🧠 Stroke pathway simulation</summary>
<br>

:::{.python-content}
This example is from `notebooks/parameters_csv.ipynb` in [pydesrap_stroke](https://github.com/pythonhealthdatascience/pydesrap_stroke).
:::
:::{.r-content}
This example is from `rmarkdown/parameters_csv.Rmd` in [rdesrap_stroke](https://github.com/pythonhealthdatascience/rdesrap_stroke).
:::

**If you're building the stroke model:** Use the code from [parameters from script](parameters_script.qmd) - you don't need this file-based example. This is because we have actually used script-based parameters - this code was just created to demonstrate how file-based parameters could work instead.

<br>

:::{.python-content}

```{.python}
from IPython.display import display
import pandas as pd

from simulation.parameters import (
    ASUArrivals, RehabArrivals, ASULOS, RehabLOS,
    ASURouting, RehabRouting, Param
)

def init_param_class(df, unit, parameter, param_class):
    """
    Instantiate a parameter class using values from a DataFrame.

    Parameters
    ----------
    df : pd.DataFrame
        Dataframe with columns "unit", "parameter", "type", "mean" and "sd".
    unit : str
        Unit name to filter by ("asu" or "rehab").
    parameter : str
        Parameter name to filter by ("iat", "los" or "routing").
    param_class: class
        Class to instantiate.

    Returns
    -------
    object
        An instance of param_class initialised with parameters from the
        DataFrame.
    """
    # Filter data to the specified unit and parameter
    df_subset = df[(df["unit"] == unit) & (df["parameter"] == parameter)]

    # If all SD values are missing, create a dict: {type: mean}
    if df_subset["sd"].isnull().all():
        param_dict = df_subset.set_index("type")["mean"].to_dict()
    # Otherwise, create a nested dict with mean and SD for each type
    else:
        param_dict = {}
        for _, row in df_subset.iterrows():
            param_dict[f"{row["type"]}_mean"] = row["mean"]
            param_dict[f"{row["type"]}_sd"] = row["sd"]

    # Instantiate parameter class using dict
    return param_class(**param_dict)



def setup_param_from_csv(csv_path):
    """
    Create a Param instance using parameter values loaded from a CSV file.

    Parameters
    ----------
    csv_path : str
        Path to csv file containing the parameters. Should have columns "unit",
        "parameter", "type", "mean" and "sd". Missing values should be marked
        as "NA".

    Returns
    -------
    Param
        An instance of Param initialised with the parameters from the CSV file.
    """
    # Load parameter data from CSV, treating "NA" as missing values
    df = pd.read_csv(csv_path, na_values=["NA"])

    # Specify mapping of Param() arguments to their corresponding units,
    # parameter types, and parameter classes
    param_specs = [
        ("asu_arrivals", "asu", "iat", ASUArrivals),
        ("rehab_arrivals", "rehab", "iat", RehabArrivals),
        ("asu_los", "asu", "los", ASULOS),
        ("rehab_los", "rehab", "los", RehabLOS),
        ("asu_routing", "asu", "routing", ASURouting),
        ("rehab_routing", "rehab", "routing", RehabRouting),
    ]

    # Instantiate each parameter class and store in a dictionary
    param_kwargs = {
        name: init_param_class(
            df=df, unit=unit, parameter=parameter, param_class=param_class)
        for name, unit, parameter, param_class in param_specs
    }

    # Return a Param instance initialised with all parameter classes
    return Param(**param_kwargs)


display(setup_param_from_csv(csv_path="../inputs/parameters.csv").__dict__)
```

:::

:::{.r-content}

```{.r}
#' Use parameter function to create parameter list using values from dataframe.
#'
#' @param data Dataframe with columns "unit", "parameter", "type", "mean" and
#' "sd".
#' @param unit Unit name to filter by ("asu" or "rehab").
#' @param parameter Parameter name to filter by ("iat", "los" or "routing").
#' @param param_function Function to run
#'
#' @return Named list of parameters

init_param_class <- function(data, unit, parameter, param_function) {

  # Filter data to specified unit and parameter
  df_subset <- data[data[["unit"]] == unit &
                      data[["parameter"]] == parameter, ]

  # Create named list of parameter values.
  # If all SD values are missing, use only means and name by 'type'.
  # Otherwise, include both mean and sd for each type, with names like
  # 'type_mean' and 'type_sd'.
  if (all(is.na(df_subset$sd))) {
    param_list <- as.list(setNames(df_subset$mean, df_subset$type))
  } else {
    param_list <- list()
    for (i in seq_len(nrow(df_subset))) {
      df_row <- df_subset[i, ]
      param_list[[paste0(df_row$type, "_mean")]] <- df_row$mean
      param_list[[paste0(df_row$type, "_sd")]] <- df_row$sd
    }
  }

  # Run parameter function using list
  do.call(param_function, param_list)
}


#' Generate named_list with create_parameters() using values loaded from a CSV
#' file.
#'
#' @param csv_path Path to csv file containing the parameters. Should have
#' columns "unit", "parameter", "type", "mean" and "sd". Missing values should
#' be marked as "NA".
#'
#' @return Named list generated by create_parameters()

setup_param_from_csv <- function(csv_path) {
  # Load parameter data from CSV, treating "NA" as missing values
  param_data <- read.csv(csv_path, na.strings = "NA")

  # Specify mappings of create_parameters() arguments to their corresponding
  # units, parameter types and parameter classes
  param_specs <- list(
    list(name = "asu_arrivals",
         unit = "asu",
         parameter = "iat",
         param_function = create_asu_arrivals),
    list(name = "rehab_arrivals",
         unit = "rehab",
         parameter = "iat",
         param_function = create_rehab_arrivals),
    list(name = "asu_los",
         unit = "asu",
         parameter = "los",
         param_function = create_asu_los),
    list(name = "rehab_los",
         unit = "rehab",
         parameter = "los",
         param_function = create_rehab_los),
    list(name = "asu_routing",
         unit = "asu",
         parameter = "routing",
         param_function = create_asu_routing),
    list(name = "rehab_routing",
         unit = "rehab",
         parameter = "routing",
         param_function = create_rehab_routing)
  )
  param_kwargs <- list()
  for (spec in param_specs) {
    param_kwargs[[spec$name]] <- init_param_class(
      data = param_data, unit = spec$unit, parameter = spec$parameter,
      param_function = spec$param_function
    )
  }

  do.call(create_parameters, param_kwargs)
}


setup_param_from_csv(file.path("..", "inputs", "parameters.csv"))
```

:::

<br>
</details>

---

<br>

## Further information

:::{.python-content}
The `sim-tools` package provides a class for managing probability distributions stored within JSON files - check it out: <https://tommonks.github.io/sim-tools/01_sampling/03_distributions_registry.html>. 
:::

:::{.r-content}
N/A
:::
