---
title: Parameter validation
format: html
filters:
  - line-highlight
---

::: {.pale-blue}

🎯 **Objectives**

**Parameter validation** refers to checking that inputs provided to functions or classes are correct and as expected. This page focuses on two examples of validation we can perform.

* **🚨 Accidental creation of new parameters:** A dangerous but common mistake in discrete-event simulation (DES), where new parameters are created instead of modifying existing ones.
* **🛡️ Validating parameter values (e.g. range):** to ensure they fall within the expected values.
* **🧾 Examples:** Full code examples from our example simulation models.

[🔗](../intro/guidelines.qmd) **Reproducibility guidelines**

Whilst this is not necessary to reproducibility or a reproducible analytical pipeline, it is good practice when working, and may help avoid mistakes.

:::

:::{.python-content}
```{python}
# Import required packages
from collections import UserDict
import inspect
```
:::
:::{.r-content}
```{r}
# Import required packages
library(R6) # nolint: undesirable_function_linter
```
:::

## 🚨 Accidental creation of new parameters

<div class="h3-tight"></div>

### The problem

When defining your parameters in a function, if you mistype a parameter name, an error won't be raised. Instead, a **new, unused parameter** is created. This can **silently** invalidate your results, as you may not realise that your parameter is unchanged.

:::{.r-content}
**Note:** This parameter typo issue doesn't occur when using R6 classes directly (as on the right), as they prevent adding new fields by default. However, if you extract the parameter list from an R6 class and modify that list separately, you'll encounter the same silent error problem. For this reason, it's better to modify parameters through the class interface rather than extracting and modifying the underlying list.
:::

::::: {.columns}

:::: {.column width="47.5%"}

Function-based example:

:::{.python-content}
```{python}
#| file: parameters_validation_resources/param_function.py
```

```{python}
# pylint: disable=undefined-variable
# Use function to create params dict
params = param_function()

# Mistype transfer_prob
params["transfer_probs"] = 0.4
print(params)
```

:::

:::{.r-content}

```{r}
#| file: parameters_validation_resources/param_function.R
```

```{r}
# Use function to create params list
params <- param_function()

# Mistype transfer_prob
params$transfer_probs <- 0.4
params
```

:::

::::

:::: {.column width="5%"}
::::

:::: {.column width="47.5%"}

Class-based example:

:::{.python-content}

```{python}
#| file: parameters_validation_resources/ParamClass.py
```

```{python}
# pylint: disable=used-before-assignment
# Create instance of ParamClass
params = ParamClass()

# Mistype transfer_prob
params.transfer_probs = 0.4
print(params.__dict__)
```

:::

:::{.r-content}
```{r}
#| file: parameters_validation_resources/ParamClass.R
``` 

```{r}
# Create instance of ParamClass
params <- ParamClass$new()

# Mistype transfer_prob
try({
  params$transfer_probs <- 0.4
})
```
:::

::::

:::::

### The solution

::::{.python-content}
There are three approaches to prevent these silent failures:

:::{.pale-blue}
* **When using functions:** Implement parameter validation within your model functions.
* **When using classes:** Build validation directly into the class structure.
* **When using dictionaries:** Use a custom wrapper that locks the keys. *This is relevant when using dictionaries independently, or as part of a nested structure in a function or class.*
:::

Class-based validation is often preferable to function-based as it catches errors at the point when parameters are defined - but both approaches are effective! Also, if either included nested structures of parameters, a wrapper for dictionaries may also be necessary.

::::

::::{.r-content}
There are two main approaches to prevent these silent failures **when using functions**:

:::{.pale-blue}
* Implement parameter validation within your model functions.
* Switch to using R6 classes.
:::

Class-based validation is often preferable as it catches errors at the point when parameters are defined - but both approaches are effective!

Functions are more commonly used in R than classes, so you may prefer to stick with the function-based approach for consistency with R conventions.
:::

### Parameter validation within the model function

:::{.python-content}
Our parameter function returns a dictionary. While you could validate arguments within the parameter function, this would not prevent accidental modification after the dictionary is returned.

When calling the function, you are restricted to the defined arguments - so extra entries cannot be added at that stage. However, once the dictionary is generated, it can be altered by adding or removing entries. For this reason, we incorporate validation into the model function (rather than the parameter function).
:::
:::{.r-content}
Our parameter function returns a list. While you could validate arguments within the parameter function, this would not prevent accidental modification after the list is returned.

When calling the function, you are restricted to the defined arguments - so extra entries cannot be added at that stage. However, once the list is generated, it can be altered by adding or removing entries. For this reason, we incorporate validation into the model function (rather than the parameter function).
:::

As a reminder, this is our parameter function:

:::{.python-content}

```{python}
#| file: "parameters_validation_resources/param_function.py"
```

:::

:::{.r-content}

```{r}
#| file: "parameters_validation_resources/param_function.R"
```

:::

We can write a validation function which checks that **all the required parameters are present**, and that **no extra parameters** are provided:

:::{.python-content}
```{python}
def check_param_names(param_dict, param_function):
    """
    Validate parameter names.

    Ensure that all required parameters are present, and no extra parameters
    are provided.

    Parameters
    ----------
    param_dict : dict
        Dictionary containing parameters for the simulation.
    param_function : function
        Function used to generate the parameter dictionary.
    """
    # Get the set of valid parameter names from the function signature
    valid_params = set(inspect.signature(param_function).parameters)

    # Get the set of input parameter names from the provided dictionary
    input_params = set(param_dict)

    # Identify missing and extra parameters
    missing = valid_params - input_params
    extra = input_params - valid_params

    # If there are any missing or extra parameters, raise an error message
    if missing or extra:
        raise ValueError("; ".join([
            f"Missing keys: {', '.join(missing)}" if missing else "",
            f"Extra keys: {', '.join(extra)}" if extra else ""
        ]).strip("; "))
```

:::

:::{.r-content}

```{r}
#' Validate parameter names.
#'
#' Ensure that all required parameters are present, and no extra parameters are
#' provided.
#'
#' @param param List containing parameters for the simulation.
#' @param param_function Function used to generate parameter list.
#'
#' @return Throws an error if there are missing or extra parameters.

check_param_names <- function(param, param_function) {

  # Get valid argument names from the function
  valid_names <- names(formals(param_function))

  # Get names from input parameter list
  input_names <- names(param)

  # Find missing keys (i.e. are there things in valid_names not in input)
  # and extra keys (i.e. are there things in input not in valid_names)
  missing_keys <- setdiff(valid_names, input_names)
  extra_keys <- setdiff(input_names, valid_names)

  # If there are any missing or extra keys, throw an error
  if (length(missing_keys) > 0L || length(extra_keys) > 0L) {
    error_message <- ""
    if (length(missing_keys) > 0L) {
      error_message <- paste0(
        error_message, "Missing keys: ", toString(missing_keys), ". "
      )
    }
    if (length(extra_keys) > 0L) {
      error_message <- paste0(
        error_message, "Extra keys: ", toString(extra_keys), ". "
      )
    }
    stop(error_message, call. = FALSE)
  }
}
```

:::

Then, in our **model function**, we call the validation function to check all inputs before proceeding with the simulation:

:::{.python-content}
```{python}
def model(param_dict, param_function):
    """
    Run simulation after validating parameter names.

    Parameters
    ----------
    param_dict : dict
        Dictionary of parameters.
    param_function : function
        Function used to generate the parameter dictionary.
    """
    # Check all inputs are valid
    check_param_names(param_dict=param_dict, param_function=param_function)

    # Simulation code...


# Example usage
# No extra or missing parameters - model runs without issue
params = param_function()
model(params, param_function)

# Mistype transfer_prob - model returns an error
params["transfer_probs"] = 0.4
try:
    model(params, param_function)
except ValueError as e:
    print(e)
```
:::

:::{.r-content}
```{r}
#' Run simulation after validating parameter names.
#'
#' @param param Named list of model parameters.
#' @param param_function Function used to generate parameter list.
model <- function(param, param_function) {

  # Check all inputs are valid
  check_param_names(param = param, param_function = param_function)

  # Simulation code...
}


# Example usage
# No extra or missing parameters - model runs without issue
params <- param_function()
model(params, param_function)

# Mistype transfer_prob - model returns an error
params$transfer_probs <- 0.4
try(model(params, param_function))
```
:::

### Parameter validation within the class

:::{.python-content}

We can add logic to prevent the addition of new attributes to our Python classes. This can be implemented either:

* Directly within the class, or-
* Using class inheritance.

#### Direct implementation within the class

This approach implements validation logic directly within the class using a custom `__setattr__` method.

```{python}
# pylint: disable=too-few-public-methods
class Param:
    """
    Parameter class with validation to prevent the addition of new attributes.
    """
    def __init__(self, param1="test", param2=42):
        """
        Initialise Param instance.
        """
        # Disable restriction during initialisation
        object.__setattr__(self, "_initialising", True)

        # Set the attributes
        self.param1 = param1
        self.param2 = param2

        # Re-enable attribute checks after initialisation
        object.__setattr__(self, "_initialising", False)

    def __setattr__(self, name, value):
        """
        Prevent addition of new attributes.

        This method overrides the default `__setattr__` behavior to restrict
        the addition of new attributes to the instance. It allows modification
        of existing attributes but raises an `AttributeError` if an attempt is
        made to create a new attribute. This ensures that accidental typos in
        attribute names do not silently create new attributes.

        Parameters
        ----------
        name : str
          The name of the attribute to set.
        value : Any
          The value to assign to the attribute.

        Raises
        -------
        AttributeError:
            If `name` is not an existing attribute and an attempt is made
            to add it to the instance.
        """
        # Skip validation if still initialising
        # pylint: disable=maybe-no-member
        if hasattr(self, "_initialising") and self._initialising:
            super().__setattr__(name, value)
        else:
            # Check if attribute already exists
            if name in self.__dict__:
                super().__setattr__(name, value)
            else:
                raise AttributeError(
                    f"Cannot add new attribute '{name}' - only possible to "
                    f"modify existing attributes: {self.__dict__.keys()}"
                )


# Example usage...

# Create an instance of the class
params = Param()

# Successfully modify an existing attribute
params.param1 = "newtest"

# Attempts to add new attributes should raise an error
try:
    params.new_attribute = 3  # pylint: disable=attribute-defined-outside-init
except AttributeError as e:
    print(f"Error: {e}")
```

#### Using class inheritance

**Class inheritance** allows a class to serve as a blueprint for another, passing down attributes and methods. This approach is useful for sharing logic - such as restricting new attribute creation—across multiple classes - **reducing code duplication and making each class simpler**.

Here, we have three classes, with each subsequent class inheriting from those above:

1. `RestrictAttributesMeta` (metaclass).
2. `RestrictAttributes` (parent/base class).
3. `Param` (child/derived class).

`RestrictAttributesMeta` (metaclass): A metaclass controls how classes and instances are created. In this case, it adds an `_initialised` flag after `__init__` completes.

```{python}
class RestrictAttributesMeta(type):
    """
    Metaclass for attribute restriction.

    A metaclass modifies class construction. It intercepts instance creation
    via __call__, adding the _initialised flag after __init__ completes. This
    is later used by RestrictAttributes to enforce attribute restrictions.
    """
    def __call__(cls, *args, **kwargs):
        # Create instance using the standard method
        instance = super().__call__(*args, **kwargs)
        # Set the "_initialised" flag to True, marking end of initialisation
        instance.__dict__["_initialised"] = True
        return instance
```

`RestrictAttributes` (parent/base class):  A parent or base class controls the behaviour (but doesn't impact initialisation). Here, the class inherits from `RestrictAttributesMeta`, and adds a new method which prevents the addition of new attributes after initialisation.

```{python}
# pylint: disable=too-few-public-methods
class RestrictAttributes(metaclass=RestrictAttributesMeta):
    """
    Base class that prevents the addition of new attributes after
    initialisation.

    This class uses RestrictAttributesMeta as its metaclass to implement
    attribute restriction. It allows for safe initialisation of attributes
    during the __init__ method, but prevents the addition of new attributes
    afterwards.

    The restriction is enforced through the custom __setattr__ method, which
    checks if the attribute already exists before allowing assignment.
    """
    def __setattr__(self, name, value):
        """
        Prevent addition of new attributes.

        Parameters
        ----------
        name: str
            The name of the attribute to set.
        value: any
            The value to assign to the attribute.

        Raises
        ------
        AttributeError
            If `name` is not an existing attribute and an attempt is made
            to add it to the class instance.
        """
        # Check if the instance is initialised and the attribute doesn"t exist
        if hasattr(self, "_initialised") and not hasattr(self, name):
            # Get a list of existing attributes for the error message
            existing = ", ".join(self.__dict__.keys())
            raise AttributeError(
                f"Cannot add new attribute '{name}' - only possible to " +
                f"modify existing attributes: {existing}."
            )
        # If checks pass, set the attribute using the standard method
        object.__setattr__(self, name, value)
```

As a child or derived class, `Param` inherits behavior from `RestrictAttributes` (and, by extension, its metaclass `RestrictAttributesMeta`). This means `Param` automatically gains the validation logic defined in its parent classes.

```{python}
# pylint: disable=too-few-public-methods,function-redefined
class Param(RestrictAttributes):
    """
    Parameter class with validation to prevent the addition of new attributes.
    """
    def __init__(self, param1="test", param2=42):
        """
        Initialise Param instance.
        """
        self.param1 = param1
        self.param2 = param2


# Example usage...

# Create an instance of the class
params = Param()

# Successfully modify an existing attribute
params.param1 = "newtest"

# Attempts to add new attributes should raise an error
try:
    params.new_attribute = 3
except AttributeError as e:
    print(f"Error: {e}")
```

### Custom wrapper for dictionaries

`LockedDict` is a dictionary wrapper that locks its initial set of keys after creation. It ensures that existing keys can still be updated, but adding or deleting top-level keys is not possible, preventing bugs caused by typos or unintended modifications.

```{python}
class LockedDict(UserDict):
    """
    Wrapper that prevents adding or deleting top-level keys in a dictionary
    after creation.

    This prevents accidental addition or deletion of keys after construction,
    helping to avoid subtle bugs (e.g., typos introducing new keys).

    Attributes
    ----------
    _locked_keys : set
        The set of top-level keys locked after initialisation.
    _locked_keys_initialised : bool
        Indicates whether locked key enforcement is active (True after
        __init__ completes). This flag is crucial for compatibility with
        joblib and multiprocessing, as object reconstruction during
        deserialisation may call __setitem__ before additional attributes are
        restored.
    """
    def __init__(self, *args, **kwargs):
        """
        Initialise the LockedDict.

        Parameters
        ----------
        *args, **kwargs : any
            Arguments passed to dict for initialisation. Top-level keys will
            be locked.

        Notes
        -----
        """
        self._locked_keys_initialised = False
        super().__init__(*args, **kwargs)
        self._locked_keys = set(self.data)
        self._locked_keys_initialised = True

    def __setattr__(self, name, value):
        """
        Block silent attribute assignment by only allowing internal, private
        and known base-class-needed attributes.

        This avoids silent failure when users try to set dict.new_attribute,
        where it does not change the dictionary as it set an attribute.
        """
        if name.startswith("_") or name == "data":
            super().__setattr__(name, value)
        else:
            raise AttributeError(
                f"Cannot set attribute '{name}'. "
                f"Use item syntax: obj['{name}'] = value"
            )

    def __setitem__(self, key, value):
        """
        Restrict assignment to existing top-level keys.

        Parameters
        ----------
        key : str
            Top-level dictionary key.
        value : any
            Value to assign.

        Raises
        ------
        KeyError
            If key is not an original top-level key.

        Notes
        -----
        Key restriction is enforced only after initialisation to prevent errors
        when used with joblib or multiprocessing, which may call __setitem__
        before all attributes are available.
        """
        if getattr(self, '_locked_keys_initialised', False):
            if key not in self._locked_keys:
                raise KeyError(
                    f"Attempted to add or update key '{key}', which is not "
                    f"one of the original locked keys. This is likely due to "
                    f"a typo or unintended new parameter. Allowed top-level "
                    f"keys are: [{self._locked_keys}]"
                )
        super().__setitem__(key, value)

    def __delitem__(self, key):
        """
        Prevent deletion of top-level keys.

        Parameters
        ----------
        key : str
            Top-level dictionary key.

        Raises
        ------
        KeyError
            Always, to disallow top-level key deletion.
        """
        raise KeyError(
            f"Deletion of key '{key}' is not allowed. The set of top-level "
            f"keys is locked to prevent accidental removal of expected "
            f"parameters. Allowed top-level keys are: [{self._locked_keys}]"
        )
```

This can be used when parameters are simply stored in a dictionary:

```{python}
param = LockedDict({
  "param1": "test",
  "param2": 42
})

# Successfuly modify an existing attribute
param["param1"] = "newtest"

# Attempts to add new attributes should raise an error
try:
  param["newattribute"] = 3
except KeyError as e:
  print(f"Error: {e}")

print(param)
```

It is also useful for nested structures - for example, when a nested dictionary of parameters is a class attribute.

```{python}
class Param(RestrictAttributes):
    """
    Parameter class with validation to prevent the addition of new attributes.
    """
    def __init__(self, param1="test", param2=42):
        """
        Initialise Param instance.
        """
        self.param_dict = LockedDict({
          "param1": "test",
          "param2": 42
        })


param_instance = Param()

# Successfully modify an existing attribute
param_instance.param_dict["param1"] = "newtest"

# Attempts to add new attributes should raise an error
try:
    param_instance.param_dict["new_attribute"] = 3
except KeyError as e:
    print(f"Error: {e}")

print(param_instance.param_dict)
```

:::

:::::{.r-content}

By default, **R6 classes prevent the addition of new fields**. However, there are ways you can override this behavior or set up your classes differently that would not have this protection for your parameters. These include:

* Setting `lock_objects = FALSE`.
* Setting parameters within a list.

#### Setting `lock_objects = FALSE`

The prevention of new fields is thanks to the default `lock_objects = TRUE` setting. If we override this and set `lock_objects = FALSE`, it will not raise an error when new fields are added. Therefore, it's important not to override this default behavior.

:::: {.columns}

::: {.column width="47.5%"}
Original class with `lock_objects = TRUE`:

```{r}
#| file = "parameters_validation_resources/ParamClass.R"
``` 

```{r}
# Create instance of ParamClass
params <- ParamClass$new()

# Mistype transfer_prob
try({
  params$transfer_probs <- 0.4
})
```
:::

::: {.column width="5%"}
:::

::: {.column width="47.5%"}
Same class with `lock_objects = TRUE`:

```{r}
#' @title Returns transfer_prob for validation example.
#'
#' @field transfer_prob Numeric. Transfer probability (0-1).

ParamClass <- R6Class( # nolint: object_name_linter
  lock_objects = FALSE, #<<
  public = list(
    transfer_prob = NULL,

    #' @description
    #' Initialises the R6 object.

    initialize = function(transfer_prob = 0.3) {
      self$transfer_prob <- transfer_prob
    }
  )
)
```

```{r}
# Create instance of ParamClass
params <- ParamClass$new()

# Mistype transfer_prob
try({
  params$transfer_probs <- 0.4
})
params #<<
```
:::

::::

#### Setting parameters within a list

If you set up your R6 class with each parameter as a class field, then by default, it will have validation to prevent the addition of new fields.

You may choose to store parameters in a list instead, to make it easier to access them all at once - but, if you do this, the validation won't apply.

:::: {.columns}

::: {.column width="47.5%"}
Original class with individual fields:

```{r}
#| file: parameters_validation_resources/ParamClass.R
``` 

```{r}
# Create instance of ParamClass
params <- ParamClass$new()

# Mistype transfer_prob
try({
  params$transfer_probs <- 0.4
})
```
:::

::: {.column width="5%"}
:::

::: {.column width="47.5%"}
Class with parameters in a list:
```{r}
#' @title Returns transfer_prob for validation example.
#'
#' @field transfer_prob Numeric. Transfer probability (0-1).

ParamClass <- R6Class( # nolint: object_name_linter
  public = list(
    parameters = NULL, #<<

    #' @description
    #' Initialises the R6 object.

    initialize = function(transfer_prob = 0.3) {
      self$parameters <- list( #<<
        transfer_prob = transfer_prob #<<
      ) #<<
    }
  )
)
```

```{r}
# Create instance of ParamClass
params <- ParamClass$new()

# Mistype transfer_prob
try({
  params$parameters$transfer_probs <- 0.4
})
params$parameters #<<
```
:::

::::

<br>

However, there is a clean solution that allows you to access parameters easily while maintaining individual fields with built-in validation. By adding a `get_params()` method, you can extract parameters without storing them in a list.

:::: {.columns}

::: {.column width="47.5%"}
Original class with individual fields:

```{r}
#| file: parameters_validation_resources/ParamClass.R
``` 

```{r}
# Create instance of ParamClass
params <- ParamClass$new()

# Mistype transfer_prob
try({
  params$transfer_probs <- 0.4
})
```
:::

::: {.column width="5%"}
:::

::: {.column width="47.5%"}
Same class with added `get_params()` method:
```{r}
#' @title Returns transfer_prob for validation example.
#'
#' @field transfer_prob Numeric. Transfer probability (0-1).

ParamClass <- R6Class( # nolint: object_name_linter
  public = list(
    transfer_prob = NULL,

    #' @description
    #' Initialises the R6 object.

    initialize = function(transfer_prob = 0.3) {
      self$transfer_prob <- transfer_prob
    },

    #' @description #<<
    #' Returns parameters as a named list. #<<

    get_params = function() { #<<
      # Get all non-function fields #<<
      all_names <- ls(self) #<<
      is_not_function <- vapply( #<<
        all_names, #<<
        function(x) !is.function(self[[x]]), #<<
        FUN.VALUE = logical(1L) #<<
      ) #<<
      param_names <- all_names[is_not_function] #<<
      mget(param_names, envir = self) #<<
    } #<<
  )
)
```

```{r}
# Create instance of ParamClass
params <- ParamClass$new()

# Mistype transfer_prob
try({
  params$transfer_probs <- 0.4
})

# Get all parameters #<<
params$get_params() #<<
```
:::

::::

:::::

## 🛡️ Validating parameter values (e.g. range)

You can check that the provided inputs are valid - expected format, range, etc. This can either be:

:::{.pale-blue}
* **When using functions:** Implement parameter validation within your model functions.
* **When using classes:** Build validation directly into the class structure.
:::

### Parameter validation within the model functions

As a reminder, this is our parameter function:

:::{.python-content}
```{python}
#| file: parameters_validation_resources/param_function.py
```

:::

:::{.r-content}
```{r}
#| file: parameters_validation_resources/param_function.R
```

:::

We can write a validation function which checks that the provided transfer probability is between 0 and 1.

:::{.python-content}
```{python}
def validate_param(parameters):
    """
    Check that the transfer probability is between 0 and 1.

    Parameters
    ----------
    parameters : dict
      Dictionary of parameters.
    """
    transfer_prob = parameters["transfer_prob"]
    if transfer_prob < 0 or transfer_prob > 1:
        raise ValueError(
          f"transfer_prob must be between 0 and 1, but is: {transfer_prob}"
        )
```
:::

:::{.r-content}
```{r}
#' @title Check that the transfer probability is between 0 and 1.
#'
#' @param parameters Named list of model parameters.

validate_param <- function(parameters) {
  transfer_prob <- parameters$transfer_prob
  if (transfer_prob < 0L || transfer_prob > 1L) {
    stop(
      "transfer_prob must be between 0 and 1, but is: ", transfer_prob,
      call. = FALSE
    )
  }
}
```
:::

Then, in our **model function**, we call the validation function to check all inputs before proceeding with the simulation:

:::{.python-content}
```{python}
def model(param_dict):
    """
    Run simulation.

    Parameters
    ----------
    param_dict : dict
        Dictionary of parameters.
    """
    # Check all inputs are valid
    validate_param(parameters=param_dict)

    # Simulation code...


# Example usage
# Run model() invalid transfer_prob - raises an error
param = param_function(transfer_prob=1.4)
try:
    model(param)
except ValueError as e:
    print(e)
```
:::

:::{.r-content}
```{r}
#' Run simulation.
#'
#' @param param Named list of model parameters.

model <- function(param) {

  # Check all inputs are valid
  validate_param(parameters = param)

  # Simulation code...
}


# Example usage
# Run model() invalid transfer_prob - raises an error
params <- param_function({
  transfer_prob <- 1.4
})
try(model(params))
```
:::

### Parameter validation within the class

The new `validate_param()` method checks whether `transfer_prob` is between 0 and 1.

Although this is defined within the class, it could also be called from within the model function, so that all parameters are checked before the simulation runs.

::::: {.columns}

:::: {.column width="47.5%"}

Original `ParamClass`:

:::{.python-content}

```{python}
#| file: parameters_validation_resources/ParamClass.py
```

```{python}
# Create instance with invalid transfer_prob - no error raised
param = ParamClass(transfer_prob=1.4)
```

:::

:::{.r-content}

```{r}
#| file: parameters_validation_resources/ParamClass.R
```

```{r}
# Create instance with invalid transfer_prob - no error raised
param <- ParamClass$new(transfer_prob = 1.4)
```

:::

::::

:::: {.column width="5%"}
::::

:::: {.column width="47.5%"}

Same class with added `validate_param()` method:

:::{.python-content}
```{python}
class ParamClass:
    """
    Returns transfer_prob for validation example.
    """
    def __init__(self, transfer_prob=0.3):
        """
        Initialise ParamClass instance.

        Parameters
        ----------
        transfer_prob : float
            Transfer probability (0-1).
        """
        self.transfer_prob = transfer_prob

    def validate_param(self):  # <<
        """  # <<
        Check that transfer_prob is between 0 and 1.  # <<
        """  # <<
        if self.transfer_prob < 0 or self.transfer_prob > 1:  # <<
            raise ValueError("transfer_prob must be between 0 and 1" +  # <<
                             f", but is: {self.transfer_prob}")  # <<
```

```{python}
# Create instance of ParamClass with invalid transfer_prob and run method
param = ParamClass(transfer_prob=1.4)
try:  # <<
    param.validate_param()  # <<
except ValueError as e:  # <<
    print(f"Error: {e}")  # <<
```
:::

:::{.r-content}
```{r}
#' @title Returns transfer_prob for validation example.
#'
#' @field transfer_prob Numeric. Transfer probability (0-1).

ParamClass <- R6Class( # nolint: object_name_linter
  public = list(
    transfer_prob = NULL,

    #' @description
    #' Initialises the R6 object.

    initialize = function(transfer_prob = 0.3) {
      self$transfer_prob <- transfer_prob
    },

    #' @description #<<
    #' Check that transfer_prob is between 0 and 1. #<<
    #' @return No return value; throws an error if invalid. #<<

    validate_param = function() { #<<
      if (self$transfer_prob < 0L || self$transfer_prob > 1L) { #<<
        stop( #<<
          "transfer_prob must be between 0 and 1, but is: ", #<<
          self$transfer_prob, call. = FALSE #<<
        ) #<<
      } #<<
    } #<<
  )
)
```

```{r}
# Create instance with invalid transfer_prob and run method
param <- ParamClass$new(transfer_prob = 1.4)
try(param$validate_param()) #<<
```
:::

::::

:::::

## 🧾 Examples

This section contains full code examples for our [example conceptual models](../intro/examples.qmd).

---

<details>
<summary>Show/Hide example 1: 🩺 Nurse visit simulation</summary>
<br>

:::{.python-content}
This example is from `simulation/param.py` in [pydesrap_mms](https://github.com/pythonhealthdatascience/pydesrap_mms).

```{.r}
{{< include ../../examples/pydesrap_mms/simulation/param.py >}}
```
:::

:::{.r-content}
This example is from `R/validate_model_inputs.R` in [rdesrap_mms](https://github.com/pythonhealthdatascience/rdesrap_mms).

```{.r}
{{< include ../../examples/rdesrap_mms/R/validate_model_inputs.R >}}
```
:::

<br>
</details>

---

<details>
<summary>Show/Hide example 2: 🧠 Stroke pathway simulation</summary>
<br>

:::{.python-content}

This example is from `simulation/parameters.py` in [pydesrap_stroke](https://github.com/pythonhealthdatascience/pydesrap_stroke).

```{.python}
{{< include ../../examples/pydesrap_stroke/simulation/parameters.py >}}
```

:::

:::{.r-content}
This example is from `R/validate_model_inputs.R` in [rdesrap_stroke](https://github.com/pythonhealthdatascience/rdesrap_stroke).

```{.r}
{{< include ../../examples/rdesrap_stroke/R/validate_model_inputs.R >}}
```
:::

<br>
</details>

---
