---
title: Parameters from file
---

{{< include ../../scripts/_reticulate-setup.md >}}

::: {.pale-blue}

🎯 **Objectives**

Discrete-event simulations (DES) require many parameters - like arrival rates, resource times, and probabilities - which often need to be changed for different scenarios and analyses. Managing these parameters well makes your simulations easier to update, track, and reuse.

This page focuses on the storage of parameters within a file.

* **❓ Why use external parameter files?**
* **📝 Create parameter file**
* **📖 Create data dictionary**
* **📥 Methods for importing parameters**
* **🧾 Examples:** Full code examples from our example simulation models.

If you want to see how to store parameters within a script, see the [parameters from script](parameters_script.qmd) page.

[🔗](../intro/guidelines.qmd) **Reproducibility guidelines**

This page helps you meet reproducibility criteria from:

* Heather et al. 2025: Avoid hard-coded parameters.
* NHS Levels of RAP (🥈): Data is handled and output in a Tidy data format.

:::

:::: {.python-content}

::: {.callout-note title="Utility function" collapse="true"}

The `print_dict` function is just a small helper used on this page to make dictionary outputs easier to read.

```{python}
import json
import itertools


def print_dict(dictionary, max_items_per_level):
    """
    Pretty print the first n items of a dictionary in JSON format.

    Parameters
    ----------
    dictionary : dict
        Dictionary to print.
    max_items_per_level: int or list of int.
        If int, applies limit to all levels. If list, limit for
        [level1, level2, ...]. If the nesting is deeper, the last value is
        used for deeper levels.
    """
    # Ensure max_items_per_level is a list of limits per level
    if isinstance(max_items_per_level, int):
        levels = [max_items_per_level]
    else:
        levels = list(max_items_per_level)

    def _truncate(obj, level=0):
        """
        Recursively copy the dictionary, trimming to the allowed number
        of items at each level.

        Parameters
        ----------
        obj : dict or other
            The object (dict or non-dict) to be truncated.
        level : int
            Current depth in nested structure (0 = top/root).
        """
        # Choose the current level's limit, or last limit if deeper
        max_items = levels[min(level, len(levels)-1)]

        # If obj is a dict, slice its items to the limit
        if isinstance(obj, dict):
            # Use islice for efficiency; only take max_items items
            return {
                k: _truncate(v, level + 1)
                for k, v in itertools.islice(obj.items(), max_items)
            }
        # If obj is not a dict, return as is
        return obj

    # Create a truncated copy and print as pretty JSON
    trimmed_dict = _truncate(dictionary)
    print(json.dumps(trimmed_dict, indent=4, sort_keys=True))
```

:::

::: {.callout-note title="Linting" collapse="true"}

The following `# pylint: disable` lines simply turn off some linting warnings that would otherwise appear in this page's examples. If you want to learn more about linting, see the [linting](../style_docs/linting.qmd) page.

```{python}
# pylint: disable=wrong-import-position,reimported,wrong-import-order
# pylint: disable=ungrouped-imports,too-many-instance-attributes
# pylint: disable=too-many-arguments, too-many-positional-arguments
# pylint: disable=too-few-public-methods,redefined-outer-name
# pylint: disable=function-redefined
```

:::

```{python}
import pandas as pd
import json
from collections import defaultdict
```

::::

::: {.r-content}

```{r}
# nolint start: undesirable_function_linter
library(jsonlite)
library(R6)
# nolint end
```

:::

## ❓ Why use external parameter files?

External parameter files can offer some advantages over storing parameters directly in scripts:

* **Easier for non-programmers** to modify parameters.
* Can easily share your simulation with **different parameters versions** (e.g. real parameters for internal use, synthetic parameters for external sharing, as discussed in [input data management](input_data.qmd)) **without editing the code itself**.
* **Less hard coding** as parameters are completely separate.

In many cases though, storing parameters within a script is commonplace - particularly for simpler models - due to its simplicity (everything in one place, fewer files).

## 📝 Create parameter file

There are several file types you can use to store parameters, with **CSV** and **JSON** being two good options. Both have advantages over Excel files:

* **Open and non-proprietary** format.
* Work well with **version control** (as they are plain text).
* **Supported** by many tools: spreadsheets, text editors, programming languages.
* **Easier to import** and parse programmatically than Excel.

### CSV example

For CSV files, use a simple tabular structure and follow tidy data principles - each row is an observation, each column a variable. Avoid merged cells or stacked headers, as these make data processing difficult.

:::{.python-content}

```{python}
# Import and preview parameter file
csv_content = pd.read_csv("parameters_file_resources/example_parameters.csv")
csv_content.head()
```

:::

:::{.r-content}

```{r}
# Import and view the parameter table
read.csv(file.path("parameters_file_resources", "example_parameters.csv"))
```

:::

### JSON example

For JSON files, data are stored in hierachical / nested structures. Keep the structure clear and consistent.

::: {.python-content}

```{python}
# Import parameters
# pylint:disable=invalid-name
json_file = "parameters_file_resources/example_parameters.json"
with open(json_file, "r", encoding="utf-8") as f:
    json_content = json.load(f)

# Preview the dictionary
print_dict(json_content, [1, 3])
```

:::

::: {.r-content}

```{r}
json_file <- fromJSON(
  file.path("parameters_file_resources", "example_parameters.json")
)
str(json_file)
```

:::

## 📖 Create data dictionary

As discussed on the [input data management](input_data.qmd) page, you should include a data dictionary or documentation describing each parameter. This documentation should explain the parameter's meaning, units, and any abbreviations or codes used.

```{python}
#| echo: false
#| output: false

import subprocess

subprocess.run([
    "pandoc",
    "parameters_file_resources/_csv_data_dictionary.md",
    "-o", "parameters_file_resources/_csv_data_dictionary.pdf"
], check=False)
subprocess.run([
    "pandoc",
    "parameters_file_resources/_json_data_dictionary.md",
    "-o", "parameters_file_resources/_json_data_dictionary.pdf"
], check=False)
```

### CSV data dictionary example

<iframe src="parameters_file_resources/_csv_data_dictionary.pdf" width="100%" height="500px" style="border:1px solid #dee2e6"></iframe>

### JSON data dictionary example

<iframe src="parameters_file_resources/_json_data_dictionary.pdf" width="100%" height="500px" style="border:1px solid #dee2e6"></iframe>

## 📥 Methods for importing parameters

As explained in the [parameters from script](parameters_script.qmd) page, there are two key principles for managing parameters effectively in simulation models:

1. **Group parameters into a dedicated object**.
2. **Pass this object explicitly to your model**.

If you haven't already, see the [parameters from script](parameters_script.qmd) page for more detail on why this explicit grouping and passing is important.

::: {.python-content}

On that page, we discussed three implementation options: **dictionary**, **function**, and **class**. When defining parameters within scripts, we recommended using a function or class. But what about when you're importing parameters from a file?

:::

::: {.r-content}

On that page, we discussed three implementation options: **list**, **function**, and **R6 class**. When defining parameters within scripts, we recommended using a function or class. But what about when you're importing parameters from a file?

:::


### Implementation options for file-based parameters

::: {.python-content}
You still have two main choices: a **function** or a **class**. 

**Direct dictionary import** (loading the file straight into a dictionary) is rarely practical. It only works if the file structure exactly matches your model and no extra parameters are needed.

**Function (returns a dictionary)**
:::

::: {.r-content}
You still have two main choices: a **function** or a **R6 class**. 

**Direct list import** (loading the file straight into a named list) is rarely practical. It only works if the file structure exactly matches your model and no extra parameters are needed.

**Function (returns a named list)**
:::

* ✅ Handles any processing needed to go from file to parameter structure.

* ✅ Lets you merge file-based and script-based parameters into one object.

* ✅ Keeps all import/processing logic together for easy reuse and debugging.

::: {.python-content}
* ❌ Dictionaries can be modified after creation, so validation must be performed in the model functions/classes (ensuring the final set of parameters is being checked). This means that errors are only caught later, rather than immediately at the point of import/change.

**Class**
:::

::: {.r-content}
* ❌ Named list can be modified after creation, so validation must be performed in the model functions/classes (ensuring the final set of parameters is being checked). This means that errors are only caught later, rather than immediately at the point of import/change.

**R6 class**
:::

* ✅ Includes all function benefits: processing, merging, in one place.

* ✅ Supports validation at creation or update, catching errors earlier.

For more context on validation, see the [parameter validation](parameters_validation.qmd) page.

### Using a function

<div class="h4-tight"></div>

::: {.python-content}

#### Nested or flat dictionary

A function loads parameters and returns them as a dictionary. You can choose between a **nested** or **flat** structure, depending on complexity.

A **nested dictionary** is good for grouped parameters, like distributions with multiple values (`mean`, `sd`, etc.), or if you want to organise by category like patient type.

```{python}
params = {
    "adult": {"interarrival": 5, "service_time": 10},
    "child": {"interarrival": 3, "service_time": 7},
    "distribution": {"mean": 0, "sd": 1}
}

print(params["adult"]["interarrival"])
```

A **flat dictionary** can be simpler to use if everything fits in one layer.

```{python}
params = {
    "adult_interarrival": 5,
    "adult_service_time": 10,
    "child_interarrival": 3,
    "child_service_time": 7
}

print(params["adult_interarrival"])
```

:::

::: {.r-content}

#### Nested or flat list

A function loads parameters and returns them as a named list. You can choose between a **nested** or **flat** structure, depending on complexity.

A **nested list** is good for grouped parameters, like distributions with multiple values (`mean`, `sd`, etc.), or if you want to organise by category like patient type.

```{r}
params <- list(
  adult = list(interarrival = 5L, service_time = 10L),
  child = list(interarrival = 3L, service_time = 7L),
  distribution = list(mean = 0L, sd = 1L)
)

params$adult$interarrival
```

A **flat list** can be simpler to use if everything fits in one layer.

```{r}
params <- list(
  adult_interarrival = 5L,
  adult_service_time = 10L,
  child_interarrival = 3L,
  child_service_time = 7L
)

params$adult_interarrival
```

:::

#### JSON example

This uses a nested format, as the JSON was already in that format:

::: {.python-content}
```{python}
def create_parameters(json_file, number_of_runs=10):
    """
    Create parameter dictionary by combining parameters from JSON with those
    defined in this function.

    Parameters
    ----------
    json_file : str
        Path to JSON file containing simulation parameters.
    number_of_runs : int, optional
        Number of simulation runs (default is 10).

    Returns
    -------
    param_dict : dict
        Nested dictionary of simulation parameters.
    """
    with open(json_file, "r", encoding="utf-8") as f:
        json_content = json.load(f)
    param_dict = json_content["simulation_parameters"]
    param_dict["number_of_runs"] = number_of_runs
    return param_dict


param = create_parameters("parameters_file_resources/example_parameters.json")
print_dict(param, 4)
```
:::

::: {.r-content}
```{r}
#' Create parameter list by combining parameters from JSON with those
#' defined in this function.
#'
#' @param json_file Character, path to JSON file containing simulation
#' parameters.
#' @param number_of_runs Integer, number of simulation runs (default is 10).
#'
#' @return Named list of simulation parameters (nested).
#' @export
create_parameters <- function(json_file, number_of_runs = 10L) {
  json_content <- jsonlite::fromJSON(json_file)
  param_list <- json_content$simulation_parameters
  param_list$number_of_runs <- number_of_runs
  return(param_list)
}

# Example usage:
param <- create_parameters(
  file.path("parameters_file_resources", "example_parameters.json")
)
print(param[1L:4L])
```
:::

#### CSV example (flat)

::: {.python-content}
```{python}
def create_parameters(csv_file, number_of_runs=10):
    """
    Create flat parameter dictionary by combining parameters from CSV with
    those defined in this function.

    Parameters
    ----------
    csv_file : str
        Path to CSV file containing simulation parameters.
    number_of_runs : int, optional
        Number of simulation runs (default is 10).

    Returns
    -------
    param_dict : dict
        Flat dictionary of simulation parameters.
    """
    csv_content = pd.read_csv(csv_file)
    param_dict = {
        f"{row.patient}_{row.metric}": float(row.value)
        for row in csv_content.itertuples(index=False)
    }
    param_dict["number_of_runs"] = number_of_runs
    return param_dict


param = create_parameters("parameters_file_resources/example_parameters.csv")
print_dict(param, 4)
```
:::

::: {.r-content}
```{r}
#' Create flat parameter list by combining parameters from CSV with those
#' defined in this function.
#'
#' @param csv_file Character, path to CSV file containing simulation
#' parameters.
#' @param number_of_runs Integer, number of simulation runs (default is 10).
#'
#' @return Named list of simulation parameters (flat).
#' @export
create_parameters <- function(csv_file, number_of_runs = 10L) {
  csv_content <- read.csv(csv_file)
  lst <- as.list(csv_content$value)
  names(lst) <- paste(csv_content$patient, csv_content$metric, sep = "_")
  lst$number_of_runs <- number_of_runs
  return(lst)
}

# Example usage:
param <- create_parameters(
  file.path("parameters_file_resources", "example_parameters.csv")
)
print(param[1L:4L])
```
:::

#### CSV example (nested)

::: {.python-content}
```{python}
def create_parameters(csv_file, number_of_runs=10):
    """
    Create nested parameter dictionary by combining parameters from CSV with
    those defined in this function.

    Parameters
    ----------
    csv_file : str
        Path to CSV file containing simulation parameters.
    number_of_runs : int, optional
        Number of simulation runs (default is 10).

    Returns
    -------
    dict
        Nested dictionary of simulation parameters.
    """
    csv_content = pd.read_csv(csv_file)
    param_dict = defaultdict(dict)
    for row in csv_content.itertuples(index=False):
        param_dict[row.patient][row.metric] = float(row.value)
    param_dict["number_of_runs"] = number_of_runs
    return dict(param_dict)


param = create_parameters("parameters_file_resources/example_parameters.csv")
print_dict(param, 4)
```
:::

::: {.r-content}
```{r}
#' Create nested parameter list by combining parameters from CSV with those
#' defined in this function.
#'
#' @param csv_file Character, path to CSV file containing simulation
#' parameters.
#' @param number_of_runs Integer, number of simulation runs (default is 10).
#'
#' @return Named list of simulation parameters (nested).
#' @export
create_parameters <- function(csv_file, number_of_runs = 10L) {
  csv_content <- read.csv(csv_file)
  lst <- lapply(split(csv_content, csv_content$patient), function(subset) {
    setNames(as.list(subset$value), subset$metric)
  })
  lst$number_of_runs <- number_of_runs
  return(lst)
}

# Example usage:
param <- create_parameters(
  file.path("parameters_file_resources", "example_parameters.csv")
)
print(param[1L:4L])
```
:::

### Using a class

In the class, parameters can again be nested or flat, but this time they are **attributes** of the class instance. Those attributes can themselves them be simple values or dictionaries.

As described above, classes also allow you to "lock down" the allowed parameters and add validation checks during initialisation and any changes. For examples of how to add validation when using a class, see the [parameter validation](parameters_validation.qmd) page.

::: {.python-content}

#### JSON example

```{python}
class CreateParameters:
    """
    Store simulation parameters, combining those loaded from a JSON file with
    values defined in the class.
    """
    def __init__(self, json_file, number_of_runs=10):
        """
        Parameters
        ----------
        json_file : str
            Path to JSON file containing simulation parameters.
        number_of_runs : int, optional
            Number of simulation runs (default is 10).
        """
        # Import parameters from JSON file
        with open(json_file, "r", encoding="utf-8") as f:
            json_content = json.load(f)
        params = json_content["simulation_parameters"]

        # Assign file-based parameters as attributes
        for key, value in params.items():
            setattr(self, key, value)

        # Add class-defined parameters
        self.number_of_runs = number_of_runs


param = CreateParameters("parameters_file_resources/example_parameters.json")
print_dict(param.__dict__, 4)
```

#### CSV example (flat)

```{python}
class CreateParameters:
    """
    Store simulation parameters, combining those loaded from a CSV file with
    values defined in the class.
    """
    def __init__(self, csv_file, number_of_runs=10):
        """
        Parameters
        ----------
        csv_file : str
            Path to CSV file containing simulation parameters.
        number_of_runs : int, optional
            Number of simulation runs (default is 10).
        """
        # Import parameters from CSV file
        csv_content = pd.read_csv(csv_file)

        # Assign CSV parameters as attributes (flat style)
        for row in csv_content.itertuples(index=False):
            param_key = f"{row.patient}_{row.metric}"
            setattr(self, param_key, float(row.value))

        # Add class-defined parameters
        self.number_of_runs = number_of_runs


param = CreateParameters("parameters_file_resources/example_parameters.csv")
print_dict(param.__dict__, 4)
```

#### CSV example (nested)

```{python}
class CreateParameters:
    """
    Store simulation parameters, combining those loaded from a CSV file with
    values defined in the class.
    """
    def __init__(self, csv_file, number_of_runs=10):
        """
        Parameters
        ----------
        csv_file : str
            Path to CSV file containing simulation parameters.
        number_of_runs : int, optional
            Number of simulation runs (default is 10).
        """
        # Import parameters from CSV file and structure as nested dictionary
        csv_content = pd.read_csv(csv_file)
        param_dict = defaultdict(dict)
        for row in csv_content.itertuples(index=False):
            param_dict[row.patient][row.metric] = float(row.value)
        param_dict["number_of_runs"] = number_of_runs

        # Assign file-based parameters as attributes
        for key, value in param_dict.items():
            setattr(self, key, value)

        # Add class-defined parameters
        self.number_of_runs = number_of_runs


param = CreateParameters("parameters_file_resources/example_parameters.csv")
print_dict(param.__dict__, 4)
```

#### Alternative set-up: class with explicit attributes

The previous examples dynamically set attributes based on whatever is present in your input file. Here, we show another approach - explicitly defining each attribute. This provides clarity and structure, although is more verbose.

You'll notice that this class is similar to those in [parameters from script](parameters_script.qmd) - except that it has no default values.

```{python}
class Parameters:
    """
    Container for simulation parameters across patient demographics.
    """
    def __init__(
        self,
        adult_interarrival, adult_consultation, adult_transfer,
        child_interarrival, child_consultation, child_transfer,
        elderly_interarrival, elderly_consultation, elderly_transfer
    ):
        """
        Initialise Parameters instance.

        Parameters
        ----------
        adult_interarrival : float
            Time between adult patient arrivals (e.g., in minutes or hours).
        adult_consultation : float
            Duration of consultation time for adult patients.
        adult_transfer : float
            Time required to transfer adult patients between stages.
        child_interarrival : float
            Time between child patient arrivals (e.g., in minutes or hours).
        child_consultation : float
            Duration of consultation time for child patients.
        child_transfer : float
            Time required to transfer child patients between stages.
        elderly_interarrival : float
            Time between elderly patient arrivals (e.g., in minutes or hours).
        elderly_consultation : float
            Duration of consultation time for elderly patients.
        elderly_transfer : float
            Time required to transfer elderly patients between stages.
        """
        # Adult parameters
        self.adult_interarrival = adult_interarrival
        self.adult_consultation = adult_consultation
        self.adult_transfer = adult_transfer
        # Child parameters
        self.child_interarrival = child_interarrival
        self.child_consultation = child_consultation
        self.child_transfer = child_transfer
        # Elderly parameters
        self.elderly_interarrival = elderly_interarrival
        self.elderly_consultation = elderly_consultation
        self.elderly_transfer = elderly_transfer
```

Use a helper function that loads the data and matches it to the class attributes. For example, from a CSV file:

```{python}
def setup_param_from_csv(csv_path):
    """
    Create a Parameters instance using parameter values loaded from a CSV file.

    Parameters
    ----------
    csv_path : str
        Path to CSV file containing the parameters. Should have columns
        "patient", "metric", and "value".

    Returns
    -------
    Parameters
        Instance of Parameters initialised with parameters from the CSV file.
    """

    # Import the parameters
    param_file = pd.read_csv(csv_path)

    # Create name-value mappings
    values = {
      f"{row.patient}_{row.metric}": row.value
      for row in param_file.itertuples()
    }

    # Pass these values to the Parameters class
    return Parameters(**values)


# Import parameters
params = setup_param_from_csv(
    "parameters_file_resources/example_parameters.csv"
)

# View object
print_dict(params.__dict__, 4)
```

:::

::: {.r-content}
This requires each parameter to be a field, rather than storing all parameters in a list within the class (as explained on the [parameter validation](parameters_validation.qmd) page).

In R, we cannot dynamically add these fields when initialising - i.e. we have to explicitly specify each attribute.

If you didn't want to do that (e.g. for large numbers of parameters), it would be possible to store the parameters as a named list within the class, but the list items themselves would still be modifiable, so this has no advantages over just using a function - for either, validation of the parameters present would need to be done within the model function/class.

You'll notice that this class is similar to those in [parameters from script](parameters_script.qmd) - except that it has no default values.

```{r}
#' @title Parameter class.
#'
#' @field adult_interarrival Numeric. Time between adult patient arrivals
#' (minutes).
#' @field adult_consultation Numeric. Length of adult patient consultation
#' (minutes).
#' @field adult_transfer Numeric. Adult patient transfer probability (0-1).
#' @field child_interarrival Numeric. Time between child patient arrivals
#' (minutes).
#' @field child_consultation Numeric. Length of child patient consultation
#' (minutes).
#' @field child_transfer Numeric. Child patient transfer probability (0-1).
#' @field elderly_interarrival Numeric. Time between elderly patient arrivals
#' (minutes).
#' @field elderly_consultation Numeric. Length of elderly patient consultation
#' (minutes).
#' @field elderly_transfer Numeric. Elderly patient transfer probability (0-1).

Parameters <- R6Class( # nolint: object_name_linter
  classname = "Parameters",
  public = list(
    # Adult parameters
    adult_interarrival = NULL,
    adult_consultation = NULL,
    adult_transfer = NULL,

    # Child parameters
    child_interarrival = NULL,
    child_consultation = NULL,
    child_transfer = NULL,

    # Elderly parameters
    elderly_interarrival = NULL,
    elderly_consultation = NULL,
    elderly_transfer = NULL,

    #' @description
    #' Initialise Parameters instance.

    initialize = function(
      adult_interarrival, adult_consultation, adult_transfer,
      child_interarrival, child_consultation, child_transfer,
      elderly_interarrival, elderly_consultation, elderly_transfer
    ) {
      # Adult parameters
      self$adult_interarrival <- adult_interarrival
      self$adult_consultation <- adult_consultation
      self$adult_transfer <- adult_transfer

      # Child parameters
      self$child_interarrival <- child_interarrival
      self$child_consultation <- child_consultation
      self$child_transfer <- child_transfer

      # Elderly parameters
      self$elderly_interarrival <- elderly_interarrival
      self$elderly_consultation <- elderly_consultation
      self$elderly_transfer <- elderly_transfer
    },

    #' @description
    #' Returns parameters as a named list.
    #'
    #' @return A named list containing all parameter values.

    get_params = function() {
      # Get all non-function fields
      all_names <- ls(self)
      is_not_function <- vapply(
        all_names,
        function(x) !is.function(self[[x]]),
        FUN.VALUE = logical(1L)
      )
      param_names <- all_names[is_not_function]
      mget(param_names, envir = self)
    }
  )
)
```

Use a helper function that loads the data and matches it to the class attributes. For example, from a CSV file:

```{r}
#' Create parameter list from CSV file
#'
#' @param csv_path Character.  Path to CSV file with columns:
#'   "patient", "metric", "value".
#'
#' @return An instance of `Parameters` R6 class.

setup_param_from_csv <- function(csv_path) {

  # Import the parameters
  param_file <- read.csv(csv_path, stringsAsFactors = FALSE)

  # Create named vector (e.g. "adult_interarrival" = 5)
  values <- setNames(
    param_file$value,
    paste(param_file$patient, param_file$metric, sep = "_")
  )

  # Instantiate Parameters with values
  do.call(Parameters$new, as.list(values))
}


# Import parameters
params <- setup_param_from_csv(
  file.path("parameters_file_resources", "example_parameters.csv")
)

# View object
params$get_params()
```

Or, from a JSON file:

```{r}
#' Create parameter list from JSON file
#'
#' @param json_path Character. Path to JSON file containing the parameters.
#'
#' @return A named list containing all simulation parameters.

setup_param_from_json <- function(json_path) {

  # Import the JSON parameters
  json_file <- jsonlite::fromJSON(json_path)

  # Extract the simulation_parameters list
  sim_params <- json_file$simulation_parameters

  # Pull out the mean values from the nested structure
  values <- lapply(sim_params, function(x) x$params$mean)

  # Pass them into Parameters (arguments matched by name)
  do.call(Parameters$new, values)
}


# Import parameters
params <- setup_param_from_json(
  file.path("parameters_file_resources", "example_parameters.json")
)

# View object
params$get_params()
```

:::

## 🧾 Examples

This section contains full code examples for our [example conceptual models](../intro/examples.qmd).

---

<details>
<summary>Show/Hide example 1: 🩺 Nurse visit simulation</summary>
<br>

We didn't create a file-based example for the nurse M/M/s model due to its simplicity with only a few parameters.

<br>
</details>

---

<details>
<summary>Show/Hide example 2: 🧠 Stroke pathway simulation</summary>
<br>

:::{.python-content}
This example is from `notebooks/parameters_csv.ipynb` in [pydesrap_stroke](https://github.com/pythonhealthdatascience/pydesrap_stroke).
:::
:::{.r-content}
This example is from `rmarkdown/parameters_csv.Rmd` in [rdesrap_stroke](https://github.com/pythonhealthdatascience/rdesrap_stroke).
:::

**If you're building the stroke model:** Use the code from [parameters from script](parameters_script.qmd) - you don't need this file-based example. This is because we have actually used script-based parameters - this code was just created to demonstrate how file-based parameters could work instead.

<br>

:::{.python-content}

```{.python}
from IPython.display import display
import pandas as pd

from simulation.parameters import (
    ASUArrivals, RehabArrivals, ASULOS, RehabLOS,
    ASURouting, RehabRouting, Param
)

def init_param_class(df, unit, parameter, param_class):
    """
    Instantiate a parameter class using values from a DataFrame.

    Parameters
    ----------
    df : pd.DataFrame
        Dataframe with columns "unit", "parameter", "type", "mean" and "sd".
    unit : str
        Unit name to filter by ("asu" or "rehab").
    parameter : str
        Parameter name to filter by ("iat", "los" or "routing").
    param_class: class
        Class to instantiate.

    Returns
    -------
    object
        An instance of param_class initialised with parameters from the
        DataFrame.
    """
    # Filter data to the specified unit and parameter
    df_subset = df[(df["unit"] == unit) & (df["parameter"] == parameter)]

    # If all SD values are missing, create a dict: {type: mean}
    if df_subset["sd"].isnull().all():
        param_dict = df_subset.set_index("type")["mean"].to_dict()
    # Otherwise, create a nested dict with mean and SD for each type
    else:
        param_dict = {}
        for _, row in df_subset.iterrows():
            param_dict[f"{row["type"]}_mean"] = row["mean"]
            param_dict[f"{row["type"]}_sd"] = row["sd"]

    # Instantiate parameter class using dict
    return param_class(**param_dict)



def setup_param_from_csv(csv_path):
    """
    Create a Param instance using parameter values loaded from a CSV file.

    Parameters
    ----------
    csv_path : str
        Path to csv file containing the parameters. Should have columns "unit",
        "parameter", "type", "mean" and "sd". Missing values should be marked
        as "NA".

    Returns
    -------
    Param
        An instance of Param initialised with the parameters from the CSV file.
    """
    # Load parameter data from CSV, treating "NA" as missing values
    df = pd.read_csv(csv_path, na_values=["NA"])

    # Specify mapping of Param() arguments to their corresponding units,
    # parameter types, and parameter classes
    param_specs = [
        ("asu_arrivals", "asu", "iat", ASUArrivals),
        ("rehab_arrivals", "rehab", "iat", RehabArrivals),
        ("asu_los", "asu", "los", ASULOS),
        ("rehab_los", "rehab", "los", RehabLOS),
        ("asu_routing", "asu", "routing", ASURouting),
        ("rehab_routing", "rehab", "routing", RehabRouting),
    ]

    # Instantiate each parameter class and store in a dictionary
    param_kwargs = {
        name: init_param_class(
            df=df, unit=unit, parameter=parameter, param_class=param_class)
        for name, unit, parameter, param_class in param_specs
    }

    # Return a Param instance initialised with all parameter classes
    return Param(**param_kwargs)


display(setup_param_from_csv(csv_path="../inputs/parameters.csv").__dict__)
```

:::

:::{.r-content}

```{.r}
#' Use parameter function to create parameter list using values from dataframe.
#'
#' @param data Dataframe with columns "unit", "parameter", "type", "mean" and
#' "sd".
#' @param unit Unit name to filter by ("asu" or "rehab").
#' @param parameter Parameter name to filter by ("iat", "los" or "routing").
#' @param param_function Function to run
#'
#' @return Named list of parameters

init_param_class <- function(data, unit, parameter, param_function) {

  # Filter data to specified unit and parameter
  df_subset <- data[data[["unit"]] == unit &
                      data[["parameter"]] == parameter, ]

  # Create named list of parameter values.
  # If all SD values are missing, use only means and name by 'type'.
  # Otherwise, include both mean and sd for each type, with names like
  # 'type_mean' and 'type_sd'.
  if (all(is.na(df_subset$sd))) {
    param_list <- as.list(setNames(df_subset$mean, df_subset$type))
  } else {
    param_list <- list()
    for (i in seq_len(nrow(df_subset))) {
      df_row <- df_subset[i, ]
      param_list[[paste0(df_row$type, "_mean")]] <- df_row$mean
      param_list[[paste0(df_row$type, "_sd")]] <- df_row$sd
    }
  }

  # Run parameter function using list
  do.call(param_function, param_list)
}


#' Generate named_list with create_parameters() using values loaded from a CSV
#' file.
#'
#' @param csv_path Path to csv file containing the parameters. Should have
#' columns "unit", "parameter", "type", "mean" and "sd". Missing values should
#' be marked as "NA".
#'
#' @return Named list generated by create_parameters()

setup_param_from_csv <- function(csv_path) {
  # Load parameter data from CSV, treating "NA" as missing values
  param_data <- read.csv(csv_path, na.strings = "NA")

  # Specify mappings of create_parameters() arguments to their corresponding
  # units, parameter types and parameter classes
  param_specs <- list(
    list(name = "asu_arrivals",
         unit = "asu",
         parameter = "iat",
         param_function = create_asu_arrivals),
    list(name = "rehab_arrivals",
         unit = "rehab",
         parameter = "iat",
         param_function = create_rehab_arrivals),
    list(name = "asu_los",
         unit = "asu",
         parameter = "los",
         param_function = create_asu_los),
    list(name = "rehab_los",
         unit = "rehab",
         parameter = "los",
         param_function = create_rehab_los),
    list(name = "asu_routing",
         unit = "asu",
         parameter = "routing",
         param_function = create_asu_routing),
    list(name = "rehab_routing",
         unit = "rehab",
         parameter = "routing",
         param_function = create_rehab_routing)
  )
  param_kwargs <- list()
  for (spec in param_specs) {
    param_kwargs[[spec$name]] <- init_param_class(
      data = param_data, unit = spec$unit, parameter = spec$parameter,
      param_function = spec$param_function
    )
  }

  do.call(create_parameters, param_kwargs)
}


setup_param_from_csv(file.path("..", "inputs", "parameters.csv"))
```

:::

<br>
</details>

---

## 💬 Comments
