---
title: Parameter validation
format: html
filters:
  - line-highlight
---

> This is not necessary to reproducibility or a reproducible analytical pipeline, but is good practice when working, and may help avoid mistakes.

<br>

## Introduction

**Parameter validation** refers to checking that inputs provided to functions or classes are correct and as expected.

This page focuses on preventing a dangerous but common mistake in discrete-event simulation (DES): accidentally creating new parameters through typos instead of modifying existing ones.

:::{.r-content}
```{r}
# Import required packages
library(R6)
```
:::

<br>

### The problem

When defining your parameters in a function, if you mistype a parameter name, an error won't be raised. Instead, a **new, unused parameter** is created. This can **silently** invalidate your results, as you may not realise that your parameter is unchanged.

:::{.r-content}
**Note:** This parameter typo issue doesn't occur when using R6 classes directly (as on the right), as they prevent adding new fields by default. However, if you extract the parameter list from an R6 class and modify that list separately, you'll encounter the same silent error problem. For this reason, it's better to modify parameters through the class interface rather than extracting and modifying the underlying list.
:::

::::: {.columns}

:::: {.column width="47.5%"}

:::{.python-content}
```{python}
# Function-based example

def param_function(transfer_prob=0.3):
    """
    Returns transfer_prob for validation example.

    Parameters
    ----------
    transfer_prob : float
        Transfer probability (0-1).

    Returns
    -------
    Dictionary containing the transfer_prob parameter.
    """
    return {"transfer_prob": transfer_prob}

# Use function to create params dict
params = param_function()

# Mistype transfer_prob
params["transfer_probs"] = 0.4
params
```
:::

:::{.r-content}
```{r}
# Function-based example

#' Returns transfer_prob for validation example.
#' 
#' @param transfer_prob Numeric. Transfer probability (0-1).
#' 
#' @return A named list containing the transfer_prob parameter.

param_function <- function(transfer_prob = 0.3) {
    list(transfer_prob = transfer_prob)
}

# Use function to create params list
params <- param_function()

# Mistype transfer_prob
params$transfer_probs <- 0.4
params
```
:::

::::

:::: {.column width="5%"}
::::

:::: {.column width="47.5%"}

:::{.python-content}
```{python}
# Class-based example

class ParamClass:
    """
    Returns transfer_prob for validation example.
    """
    def __init__(self, transfer_prob=0.3):
        """
        Initialise ParamClass instance.

        Parameters
        ----------
        transfer_prob : float
            Transfer probability (0-1).
        """
        self.transfer_prob = transfer_prob

# Create instance of ParamClass
params = ParamClass()

# Mistype transfer_prob
params.transfer_probs = 0.4
params.__dict__
```
:::

:::{.r-content}
```{r}
# Class-based example

{{< include parameters_validation_resources/ParamClass.R >}}
```
:::

::::

:::::

<br>

### The solution

::::{.python-content}
There are two main approaches to prevent these silent failures:

:::{.pale-blue}
* **When using functions:** Implement parameter validation within your model functions.
* **When using classes:** Build validation directly into the class structure.
:::
::::

::::{.r-content}
There are two main approaches to prevent these silent failures **when using functions**:

:::{.pale-blue}
* Implement parameter validation within your model functions.
* Switch to using R6 classes.
:::
::::

Class-based validation is often preferable as it catches errors at the point when parameters are defined - but both approaches are effective!

:::{.r-content}
Functions are more commonly used in R than classes, so you may prefer to stick with the function-based approach for consistency with R conventions.
:::

<br>

## Parameter validation within the simulation functions

for functions, this is done in model code, as if you did it in function, changes could be made to dict between making it with function and using it

<br>

## Parameter validation within the class

:::{.python-content}

Prevent typos and unauthorised changes by locking down classes after initialisation.

for classes, can do it direct in class, anytime try to update an attribute, which is preferable approach
you then have two options:
- directly in class

```{.python}
class Param:
    def __init__(self, ...):
        # Disable restriction on attribute modification during initialisation
        object.__setattr__(self, '_initialising', True)

        ...

        # Re-enable attribute checks after initialisation
        object.__setattr__(self, '_initialising', False)

    def __setattr__(self, name, value):
        # Skip the check if the object is still initialising
        # pylint: disable=maybe-no-member
        if hasattr(self, '_initialising') and self._initialising:
            super().__setattr__(name, value)
        else:
            # Check if attribute of that name is already present
            if name in self.__dict__:
                super().__setattr__(name, value)
            else:
                raise AttributeError(
                    f'Cannot add new attribute "{name}" - only possible to ' +
                    f'modify existing attributes: {self.__dict__.keys()}')
```

- super class

```{.python}
class RestrictAttributesMeta(type):
    """
    Metaclass for attribute restriction.

    A metaclass modifies class construction. It intercepts instance creation
    via __call__, adding the _initialised flag after __init__ completes. This
    is later used by RestrictAttributes to enforce attribute restrictions.
    """
    def __call__(cls, *args, **kwargs):
        # Create instance using the standard method
        instance = super().__call__(*args, **kwargs)
        # Set the "_initialised" flag to True, marking end of initialisation
        instance.__dict__["_initialised"] = True
        return instance


class RestrictAttributes(metaclass=RestrictAttributesMeta):
    """
    Base class that prevents the addition of new attributes after
    initialisation.

    This class uses RestrictAttributesMeta as its metaclass to implement
    attribute restriction. It allows for safe initialisation of attributes
    during the __init__ method, but prevents the addition of new attributes
    afterwards.

    The restriction is enforced through the custom __setattr__ method, which
    checks if the attribute already exists before allowing assignment.
    """
    def __setattr__(self, name, value):
        """
        Prevent addition of new attributes.

        Parameters
        ----------
        name: str
            The name of the attribute to set.
        value: any
            The value to assign to the attribute.

        Raises
        ------
        AttributeError
            If `name` is not an existing attribute and an attempt is made
            to add it to the class instance.
        """
        # Check if the instance is initialised and the attribute doesn"t exist
        if hasattr(self, "_initialised") and not hasattr(self, name):
            # Get a list of existing attributes for the error message
            existing = ", ".join(self.__dict__.keys())
            raise AttributeError(
                f"Cannot add new attribute '{name}' - only possible to " +
                f"modify existing attributes: {existing}."
            )
        # If checks pass, set the attribute using the standard method
        object.__setattr__(self, name, value)
```

:::

:::::{.r-content}

By default, **R6 classes prevent the addition of new fields**. However, there are ways you can override this behavior or set up your classes differently that would not have this protection for your parameters. These include:

* Setting `lock_objects = FALSE`.
* Setting parameters within a list.

<br>

**Setting `lock_objects = FALSE`**

The prevention of new fields is thanks to the default `lock_objects = TRUE` setting. If we override this and set `lock_objects = FALSE`, it will not raise an error when new fields are added. Therefore, it's important not to override this default behavior.

:::: {.columns}

::: {.column width="47.5%"}
Original class with `lock_objects = TRUE`:
```{r}
{{< include parameters_validation_resources/ParamClass.R >}}
```
:::

::: {.column width="5%"}
:::

::: {.column width="47.5%"}
Same class with `lock_objects = TRUE`:
```{r}
#' @title Returns transfer_prob for validation example.
#'
#' @field transfer_prob Numeric. Transfer probability (0-1).

ParamClass <- R6Class(
  lock_objects = FALSE, #<<
  public = list(
    transfer_prob = NULL,

    #' @description
    #' Initialises the R6 object.

    initialize = function(
        transfer_prob = 0.3
    ) {
      self$transfer_prob <- transfer_prob
    }
  )
)

# Create instance of ParamClass
params <- ParamClass$new()

# Mistype transfer_prob
try(params$transfer_probs <- 0.4)
params #<<
```
:::

::::

<br>

**Setting parameters within a list**

If you set up your R6 class with each parameter as a class field, then by default, it will have validation to prevent the addition of new fields.

You may choose to store parameters in a list instead, to make it easier to access them all at once - but, if you do this, the validation won't apply.

:::: {.columns}

::: {.column width="47.5%"}
Original class with individual fields:
```{r}
{{< include parameters_validation_resources/ParamClass.R >}}
```
:::

::: {.column width="5%"}
:::

::: {.column width="47.5%"}
Class with parameters in a list:
```{r}
#' @title Returns transfer_prob for validation example.
#'
#' @field transfer_prob Numeric. Transfer probability (0-1).

ParamClass <- R6Class(
  public = list(
    parameters = NULL, #<<

    #' @description
    #' Initialises the R6 object.

    initialize = function(
        transfer_prob = 0.3
    ) {
      self$parameters <- list( #<<
        transfer_prob = transfer_prob #<<
      ) #<<
    }
  )
)

# Create instance of ParamClass
params <- ParamClass$new()

# Mistype transfer_prob
try(params$parameters$transfer_probs <- 0.4)
params$parameters #<<
```
:::

::::

<br>

However, there is a clean solution that allows you to access parameters easily while maintaining individual fields with built-in validation. By adding a `get_params()` method, you can extract parameters without storing them in a list.

:::: {.columns}

::: {.column width="47.5%"}
Original class with individual fields:
```{r}
{{< include parameters_validation_resources/ParamClass.R >}}
```
:::

::: {.column width="5%"}
:::

::: {.column width="47.5%"}
Same class with added `get_params()` method:
```{r}
#' @title Returns transfer_prob for validation example.
#'
#' @field transfer_prob Numeric. Transfer probability (0-1).

ParamClass <- R6Class(
  public = list(
    transfer_prob = NULL,

    #' @description
    #' Initialises the R6 object.

    initialize = function(
        transfer_prob = 0.3
    ) {
      self$transfer_prob <- transfer_prob
    },

    #' @description #<<
    #' Returns parameters as a named list. #<<

    get_params = function() { #<<
      # Get all non-function fields #<<
      all_names <- ls(self) #<<
      param_names <- all_names[ #<<
        !sapply(all_names, function(x) is.function(self[[x]])) #<<
      ] #<<
      mget(param_names, envir = self) #<<
    } #<<
  )
)

# Create instance of ParamClass
params <- ParamClass$new()

# Mistype transfer_prob
try(params$transfer_probs <- 0.4)

# Get all parameters #<<
params$get_params() #<<
```
:::

::::

:::::

<br>
