---
title: Scenario and sensitivity analysis
filters:
  - line-highlight
bibliography: scenarios_resources/references.bib
date: "2025-11-06T10:44:04+00:00"
---

```{python}
#| echo: false
# pylint: disable=undefined-variable,unused-import
```

:::: {.pale-blue}

**Learning objectives:**

* Recognise the importance of sharing your scenario and sensitivity analysis code.
* Learn how to run a combination of scenarios.
* Understand differences between scenario and sensitivity analysis.

**Relevant reproducibility guidelines:**

* STARS Reproducibility Recommendations (⭐): Provide code for all scenarios and sensitivity analyses.
* STARS Reproducibility Recommendations: Save outputs to a file.
* STARS Reproducibility Recommendations: Avoid excessive output files.
* STARS Reproducibility Recommendations: Address large file sizes.

**Pre-reading:**

This page continues on from: [Parallel processing](../output_analysis/parallel.qmd).

> *Entity generation → Entity processing → Initialisation bias → Performance measures → Replications → Parallel processing → Scenario and sensitivity analysis*

::::

:::: {.very-pale-blue}

**Required packages:**

The following imports are required. These should be available from environment setup on the [Structuring as a package](/pages/guide/setup/package.qmd) page.

We have one new package: `itertools`.

::: {.python-content}

```{python}
# For display purposes:
from IPython.display import HTML
from itables import to_html_datatable

# For the simulation modelling and analysis:
import itertools#<<
import time
from joblib import Parallel, delayed, cpu_count
import numpy as np
import pandas as pd
import plotly.express as px
import scipy.stats as st
import simpy
from sim_tools.distributions import Exponential
```

```{python}
#| echo: false
# To ensure renders correctly in quarto
import plotly.io as pio  # pylint: disable=ungrouped-imports
pio.renderers.default = "plotly_mimetype"
```

:::

::: {.r-content}

```{r}
#| output: false
library(dplyr)
library(kableExtra)
library(knitr)
library(simmer)
library(tidyr)
```

:::

::::

## Running and sharing scenarios

When creating a DES, you won't just run one model - you run scenarios, which can involve different parameter set‑ups and/or changes to the process flow or simulation logic (e.g., alternative configurations of resources, routing, or priority rules). These are used to test how outcomes change under varying conditions.

When running scenarios, there are two important things to remember:

::: {.cream}

**1. Run scenarios programmatically.**

:::

**Don't duplicate whole scripts** just to test different set-ups.

Using functions and classes is important here, as:

* If your scenarios only change **parameters**, you just run the same model with different inputs.
* If your scenarios change the **configuration** of the system (e.g., routing or scheduling), you can create alternative methods, functions, or configuration objects.

On this page, we only show examples where scenarios differ by parameters.

::: {.cream}

**2. Share your scenario code.**

:::

You should share code and data for every scenario - not just the base case.

In @Heather2025, six out of seven studies with scenarios lacked scenario code. This made it challenging and time-consuming to reproduce their results, as descriptions of the scenarios in the articles were often ambiguous and key parameters unclear or missing.

```{python}
#| echo: false
#| file: ../output_analysis/parallel_resources/parameters.py
```

```{python}
#| echo: false
#| file: ../output_analysis/replications_resources/patient.py
```

```{python}
#| echo: false
#| file: ../output_analysis/replications_resources/monitored_resource.py
```

```{python}
#| echo: false
#| file: ../output_analysis/replications_resources/model.py
```

```{python}
#| echo: false
#| file: ../output_analysis/replications_resources/summary_stats.py
```

```{python}
#| echo: false
#| file: ../output_analysis/parallel_resources/runner.py
```

```{r}
#| echo: false
#| file: ../output_analysis/replications_resources/create_params.R
```

```{r}
#| echo: false
#| file: ../output_analysis/replications_resources/filter_warmup.R
```

```{r}
#| echo: false
#| file: ../output_analysis/length_warmup_resources/metrics.R
```

```{r}
#| echo: false
#| file: ../output_analysis/replications_resources/get_run_results.R
```

```{r}
#| echo: false
#| file: ../output_analysis/replications_resources/model.R
```

```{r}
#| echo: false
#| file: ../output_analysis/replications_resources/runner.R
```

## Simple scenarios

This code runs the simulation for each value of `number_of_doctors` from 3 to 6.

The returned dataframe contains the results from each of the five runs from the four scenarios.

::: {.python-content}

```{python}
results = []
# Loop through doctor counts 3-6, running the simulation
for i in range(3, 7):
    param = Parameters(number_of_doctors=i)
    runner = Runner(param=param)
    result = runner.run_reps()["run"]
    # Add scenario information to the results, then save to list
    result["number_of_doctors"] = i
    results.append(result)

# Combine results into a single dataframe
scenario_results = pd.concat(results, ignore_index=True)

# Show as interactive table
HTML(to_html_datatable(scenario_results))
```

:::

::: {.r-content}

```{r}
results <- list()
# Loop through doctor counts 3-6, running the simulation
for (i in 3L:6L) {
  param <- create_params(number_of_doctors = i)
  result <- runner(param = param)[["run_results"]]
  # Add scenario information to the results, then save to list
  result["number_of_doctors"] <- i
  results[[length(results) + 1L]] <- result
}

# Combine results into a single dataframe
scenario_results <- do.call(rbind, results)

# Show as interactive table
kable(scenario_results) |> scroll_box(height = "400px")
```

:::

## Running several scenario combinations

::: {.python-content}

If you are running several scenarios, it can be useful to define a helper function like the one shown below. This example is adapted from:

> "[HSMA - little book of DES](https://des.hsma.co.uk/)" from Sammi Rosser, Dan Chalk and Amy Heather 2025 (MIT Licence).

```{python}
def run_scenarios(scenarios, param_factory=None):
    """
    Execute a set of scenarios and return the results from each run.

    Parameters
    ----------
    scenarios : dict
        Dictionary where key is name of parameter and value is a list with
        different values to run in scenarios.
    param_factory : callable or None, optional
        A callable that returns a new Parameters object for each scenario run.
        This can be a class (e.g., `Parameters`) or a factory function/lambda
        with preset arguments (e.g., `lambda: Parameters(number_of_runs=4)`).
        If not provided, defaults to using `Parameters()` with no arguments.

    Returns
    -------
    pandas.DataFrame
        DataFrame with results from each run of each scenario.

    Notes
    -----
    Function adapted from Rosser, Chalk and Heather 2025.
    """
    # If none provided, use Parameters
    if param_factory is None:
        param_factory = Parameters

    # Find every possible permutation of the scenarios
    all_scenarios_tuples = list(itertools.product(*scenarios.values()))

    # Convert back into dictionaries
    all_scenarios_dicts = [
        dict(zip(scenarios.keys(), p)) for p in all_scenarios_tuples
    ]

    # Preview some of the scenarios
    print(f"There are {len(all_scenarios_dicts)} scenarios. Running:")

    # Run the scenarios...
    results = []
    for index, scenario_to_run in enumerate(all_scenarios_dicts):
        print(scenario_to_run)

        # Create fresh instance of parameter class for each scenario
        param = param_factory()

        # Update parameter list with the scenario parameters
        param.scenario_name = index
        for key in scenario_to_run:
            setattr(param, key, scenario_to_run[key])

        # Perform replications
        scenario_exp = Runner(param)
        scenario_res = scenario_exp.run_reps()["run"]

        # Add scenario number and values to the results dataframe
        scenario_res["scenario"] = index
        for key in scenario_to_run:
            scenario_res[key] = scenario_to_run[key]

        # Add results from scenario to list
        results.append(scenario_res)
    return pd.concat(results)
```

:::

::: {.r-content}

If you are running several scenarios, it can be useful to define a helper function like the one shown below.

```{r}
#' Run a set of scenarios
#'
#' @param scenarios List where key is name of parameter and value is a list of
#' different values to run in scenarios
#' @param base_list List of parameters to use as base for scenarios, which can
#' be partial (as will input to create_params() function).
#' @param verbose Boolean, whether to print messages about scenarios as run.
#'
#' @return Tibble with results from each replication for each scenario.
#' @export

run_scenarios <- function(scenarios, base_list, verbose = TRUE) {
  # Generate all permutations of the scenarios
  all_scenarios <- expand.grid(scenarios)

  # Preview the number of scenarios
  if (isTRUE(verbose)) {
    message(sprintf("There are %d scenarios.", nrow(all_scenarios)))
    message("Base parameters:")
    print(base_list)
  }

  results <- list()

  # Iterate through each scenario
  for (index in seq_len(nrow(all_scenarios))) {

    # Filter to one of the scenarios
    scenario_to_run <- all_scenarios[index, , drop = FALSE]

    # Print the scenario parameters
    formatted_scenario <- toString(
      paste0(names(scenario_to_run), " = ", scenario_to_run)
    )

    # Print the scenario currently running
    if (isTRUE(verbose)) {
      message("Scenario: ", formatted_scenario)
    }

    # Create parameter list with scenario-specific values
    s_args <- c(scenario_to_run, list(scenario_name = index))

    # Create instance of parameter class with specified base parameters
    s_param <- do.call(create_params, base_list)

    # Update parameter list with the scenario parameters
    for (name in names(s_args)) {
      s_param[[name]] <- s_args[[name]]
    }

    # Run replications for the current scenario and get processed results
    scenario_result <- runner(s_param)[["run_results"]]

    # Append scenario parameters to the results
    scenario_result[["scenario"]] <- index
    for (key in names(scenario_to_run)) {
      scenario_result[[key]] <- scenario_to_run[[key]]
    }

    # Append to results list
    results[[index]] <- scenario_result
  }
  do.call(rbind, results)
}
```

:::

The function generates all possible scenario combinations from the values supplied in `scenarios`.

It then iterates through these, creating a fresh set of parameters for each scenario run. This ensures every scenario gets an independent set of parameters and avoids carrying state from one scenario to another.

The replication results from each scenario are saved in a list and finally combined into a single table.

Let's run it!

::: {.python-content}

```{python}
# Run scenarios
scenario_results = run_scenarios(
    scenarios={"interarrival_time": [4, 5, 6, 7, 8],
               "number_of_doctors": [3, 4, 5]},
    param_factory=Parameters
)
```

We'll save results to CSV - you can change the file path listed, or otherwise, you'll just need to create the `scenarios_resources` directory first.

```{python}
# Save to CSV
scenario_results.to_csv(
    "scenarios_resources/python_scenario_results.csv", index=False
)

# View scenario results
HTML(to_html_datatable(scenario_results.head(200)))
```

:::

:::: {.r-content}

```{r}
# Run scenarios
scenarios <- list(
  interarrival_time = c(4L, 5L, 6L, 7L, 8L),
  number_of_doctors = c(3L, 4L, 5L)
)
scenario_results <- run_scenarios(
  scenarios = scenarios, base_list = create_params()
)

# Save to CSV
write.csv(scenario_results,
          file.path("scenarios_resources", "r_scenario_results.csv"))

# View scenario results
kable(scenario_results) |> scroll_box(height = "400px")
```

<br>

::: {.callout-note title="Selective re-running"}

For simulations with very long run times, you may wish to explore the [{targets}](https://books.ropensci.org/targets/) package. 

With `{targets}`, you can set up your pipeline so that input changes (like updates to parameter files or scenario definitions) trigger the package to automatically re-run only the affected simulations and analyses, while all other results are reused from cache. This targeted re-running helps save computation time when there are many slow scenarios to run.

If you change the code inside your model function, `{targets}` will detect those updates and make sure all impacted analysis steps are re-run.

*Note: This package hasn't been personally tried in this project, so we can't vouch for whether it definitely works reliably with simmer simulation models - it's just a cool package we've heard about! For guides on using this, see the official [{targets} manual](https://books.ropensci.org/targets/)*.

:::

::::

## Sensitivity analysis

<div class="h3-tight"></div>

### How does sensitivity analysis differ from scenario analysis?

Scenario analysis focuses on a set of pre-defined situations which are plausible and relevant to the problem being studied. It can often involve varying multiple parameters simultaneously. The purpose is to understand how the system operates under different hypothetical scenarios.

Sensitivity analysis varies one parameter (or a small group of parameters) and assess the impact of small changes in that parameter on outcomes. The purpose is to understand **how uncertainty in the inputs affects the model**, and how **robust** the model is to variation in those inputs.

Sensitivity analysis is an important technique for model validation (and forms part of "[experimentation validation](../verification_validation/verification_validation.qmd#experimentation-validation)").

### Running sensitivity analyses

Sensitivity analyses should be run programmatically, just like scenario analyses, and the code should be shared for reproducibility.

They can be run using similar code. Below is an example using the `run_scenarios` function to vary the `interarrival_time` parameter.

::: {.python-content}

```{python}
# Run sensitivity analysis
sensitivity_results = run_scenarios(
    scenarios={"interarrival_time": [4, 4.5, 5, 5.5, 6, 6.5, 7]},
    param_factory=Parameters
)
```

We'll save results to CSV - you can change the file path listed, or otherwise, you'll just need to ensure you have already made the `scenarios_resources` directory first.

```{python}
# Save to CSV
sensitivity_results.to_csv(
    "scenarios_resources/python_sensitivity_results.csv", index=False
)

# View sensitivity results
HTML(to_html_datatable(sensitivity_results.head(200)))
```

:::

::: {.r-content}

```{r}
# Run sensitivity analysis
sensitivity_results <- run_scenarios(
  scenarios = list(interarrival_time = c(4L, 4.5, 5L, 5.5, 6L, 6.5, 7L)),
  base_list = create_params()
)

# Save to CSV
write.csv(sensitivity_results,
          file.path("scenarios_resources", "r_sensitivity_results.csv"))

# View sensitivity results
kable(sensitivity_results) |> scroll_box(height = "400px")
```

:::

::: {.callout-note title="Sensitivity analysis across scenarios"}

In practice, sensitivity analysis is often performed within each scenario - for example, you might define several scenarios (different configurations of arrivals, capacity, or routing) and then vary a key parameter (such as interarrival time) around each of them. In this book, for simplicity, we show sensitivity analysis for a single scenario at a time, and you can repeat the same approach for any other plausible scenarios you define.

:::

### Two-way and factorial sensitivity analysis

The previous example varied one parameter at a time. However, model outputs can depend on several uncertain parameters at once. Varying two or more parameters together can show:

* The **main effect** of each parameter: its average effect on the outcome across the values of other parameters.
* An **interaction effect**: whether the effect of one parameter changes depending on the value of another parameter.

One way to investigate this is to use a 2k factorial design, as described in @Robinson2007. For each parameter (called a "factor"), you choose a plausible low (`-`) and high (`+`) value, then run the model for every possible combination of these values. For example:

| Factor 1 | Factor 2 | Factor 3 | Response |
| - | - | - | - |
| - | - | - | $R_1$ |
| + | - | - | $R_2$ |
| - | + | - | $R_3$ |
| + | + | - | $R_4$ |
| - | - | + | $R_5$ |
| + | - | + | $R_6$ |
| - | + | + | $R_7$ |
| + | + | + | $R_8$ |

The main effect of Factor 1 is the average change in the response when Factor 1 changes from its low to high value, across all combinations of Factors 2 and 3:

$$
\frac{
  (R_2 - R_1) + (R_4 - R_3) + (R_6 - R_5) + (R_8 - R_7)
}{4}
$$

For example, a main effect of `+2` minutes for mean waiting time would indicate that changing Factor 1 from its low to high value increases mean waiting time by 2 minutes *on average* across the tested combinations of the other factors.

An interaction occurs when the effect of one factor is different at different values of another factor. For example, increasing the number of doctors may reduce waiting time substantially when demand is high, but have little effect when demand is low. In that case, the effect of the number of doctors depends on the arrival rate: the two factors interact. See @Robinson2007 for the formula for calculating interaction effects.

For an applied healthcare example, please see Monks, T., Pearson, M., Pitt, M., Stein, K., & James, M. A. (2015). *Evaluating the impact of a simulation study in emergency stroke care*. Operations research for health care, 6, 40-49. <https://doi.org/10.1016/j.orhc.2015.09.002>. They used a 2k factorial design to explore uncertainty in an emergency stroke care simulation:

> "Here we focus on the uncertainty in the model results. To illustrate the results of the model we include a 23 factorial design including two of the early referral parameters and the proportion of exclusions between midnight and 11 am (due to our concern about underestimating the proportion of 'wake-up' strokes where the onset time is unknown). Full results, main and interaction effects are provided in the online supplementary material (see Appendix A)."

## Saving results

Saving your simulation results to file is important for reproducility, as it allows others to verify your findings and generate consistent (or new) figures and analyses, even if they can't re-run your simulation.
This practice is transparent, providing a clear record of what you found, and it is valuable for you as well, ensuring you always know exactly what results you obtained, and can regenerate your own tables and figures from the results.

However, there are two key things to keep in mind:

**1. Number of files.** Running many scenarios or replications can easily lead to an explosion of output files. Do not save each scenario or run as a separate file unless there is a specific need. Instead, combine all results into a **single file with columns marking scenario and replication IDs**. 

**2. Avoid large file sizes.** Be strategic about what you save. For short tests and debugging, saving detailed ("patient-level") results makes sense. But for full-scale runs with many replications, those files can become unmanageably large. Generally, **save summary outputs** to file for analysis (e.g., means from each run), not massive raw datasets. If you absolutely need to save or share large files, use compressed formats (e.g., `csv.gz`). Also, keep in mind practical size limits for version control: for example, GitHub's individual file size limit is 100 MB.

## Explore the example models

<div class="h3-tight"></div>

### Nurse visit simulation

:::: {.python-content}

{{< include /html/pydesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `simulation/run_scenarios.py`<br>`notebooks/analysis.ipynb` |
| **What to look for?** | Similar implementation to above. |
: {tbl-colwidths="[20,80]"}
:::

::::

:::: {.r-content}

{{< include /html/rdesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `R/scenarios.R`<br>`rmarkdown/analysis.md` |
| **What to look for?** | Similar implementation to above. |
: {tbl-colwidths="[20,80]"}
:::

::::

### Stroke pathway simulation

:::: {.python-content}

{{< include /html/pydesrapstroke.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `notebooks/analysis.ipynb` |
| **What to look for?** | These scenarios are run one-by-one in the Jupyter notebook. They involve altering arrivals before running the simulation (5% increase in admissions), or changes in calculations (pooling beds). |
: {tbl-colwidths="[20,80]"}
:::

::::

:::: {.r-content}

{{< include /html/rdesrapstroke.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `rmarkdown/analysis.md` |
| **What to look for?** | These scenarios are run one-by-one in the Rmarkdown. They involve altering arrivals before running the simulation (5% increase in admissions), or changes in calculations (pooling beds). |
: {tbl-colwidths="[20,80]"}
:::

::::

## Test yourself

<div class="h3-tight"></div>

### Quiz

```{r}
#| echo: false
library(webexercises) # nolint: library_call_linter
```

::: {.callout-note}

## Why should code be shared for every scenario, not just the base case?

```{r}
#| output: asis
#| echo: false
cat(longmcq(c(
  answer = paste0(
    "Because published descriptions are often ambiguous, and without code ",
    "for each scenario it can be hard or impossible to reproduce the results."
  ),
  "Because reviewers only care about the scenario code and not the base case.",
  "Because scenario code runs faster than base‑case code."
)))
```

:::

::: {.callout-note}

## Conceptually, how does scenario analysis differ from sensitivity analysis?

```{r}
#| output: asis
#| echo: false
cat(longmcq(c(
  answer = paste0(
    "Scenario analysis studies predefined, plausible configurations (often ",
    "changing several parameters), while sensitivity analysis varies one or ",
    "a small set of parameters to see how input uncertainty affects outputs."
  ),
  "Scenario analysis only changes one parameter; sensitivity changes many.",
  "Scenario analysis is exploratory; sensitivity is only about performance."
)))
```

:::

### Activity

Try running your own scenario and sensitivity analyses.

You could build your own simple loop to test different values - just make sure you use a fresh set of parameters for each run.

Alternatively, use the `run_scenarios` function.

## References

::: {#refs}
:::

<br><br>
