---
title:  Tables and figures
bibliography: scenarios_resources/references.bib
date: "2025-10-24T10:55:25+01:00"
---

```{python}
#| echo: false
# pylint: disable=too-many-locals,undefined-variable,unused-import
# pylint: disable=pointless-statement
```

:::: {.pale-blue}

**Learning objectives:**

* Recognise importance of sharing code for creation of tables and figures.
* View some example tables and figures.

**Relevant reproducibility guidelines:**

* STARS Reproducibility Recommendations (⭐): Include code to generate the tables, figures, and other reported results. 
* STARS Reproducibility Recommendations: Save outputs to a file.

**Pre-reading:**

This page uses the results saved on the [Scenario and sensitivity analysis](scenarios.qmd) page.

::::

:::: {.very-pale-blue}

**Required packages:**

The following imports are required. These should be available from environment setup on the [Structuring as a package](/pages/guide/setup/package.qmd) page. 

::: {.python-content}

On this page, we have three new imports: `os`, `matplotlib.colors` and `plotly.graph_objects`.

```{python}
# For display purposes:
from IPython.display import HTML
from itables import to_html_datatable

# For the simulation modelling and analysis:
import itertools
import os#<<
import time
from joblib import Parallel, delayed, cpu_count
import matplotlib.colors as mcolors#<<
import numpy as np
import pandas as pd
import plotly.express as px
import plotly.graph_objects as go#<<
import scipy.stats as st
import simpy
from sim_tools.distributions import Exponential
```

```{python}
#| echo: false
# To ensure renders correctly in quarto
import plotly.io as pio  # pylint: disable=ungrouped-imports
pio.renderers.default = "plotly_mimetype"
```

:::

::: {.r-content}

```{r}
#| output: false
library(dplyr)
library(ggplot2)
library(kableExtra)
library(knitr)
```

:::

::::

<br>

When reporting results from a simulation, you will usually include tables and figures. To make your work reproducible, you should:

::: {.cream}

**1. Make your tables and figures using code.**

**2. Share that code alongside your results.**

:::

This is just as important as sharing model and scenario code. In @Heather2025, only one study included the code needed to generate all its reported results. Without this code for the others, recreating outputs was time-consuming and challenging. It required guesswork about which results to extract, how to clean them, and what processing steps were taken when producing the published tables and figures.

This page will show some examples of creating tables and figures, but there are many different ways you could describe and visualise your simulation results. The key takeaway from the page is simple: **write code and share it!**

## Simulation results

This page will use the results saved on the [Scenario and sensitivity analysis](scenarios.qmd) page.

::: {.python-content}

### Scenario analysis

{{< downloadthis scenarios_resources/python_scenario_results.csv dname="python_scenario_results" label="Download the scenario analysis results" type="primary" >}}

```{python}
scenario = pd.read_csv("scenarios_resources/python_scenario_results.csv")
HTML(to_html_datatable(scenario.head(200)))
```

### Sensitivity analysis

{{< downloadthis scenarios_resources/python_sensitivity_results.csv dname="python_sensitivity_results" label="Download the sensitivity analysis results" type="primary" >}}

```{python}
sensitivity = pd.read_csv("scenarios_resources/python_sensitivity_results.csv")
HTML(to_html_datatable(sensitivity.head(200)))
```

:::

::: {.r-content}

### Scenario analysis

{{< downloadthis scenarios_resources/r_scenario_results.csv dname="r_scenario_results" label="Download the scenario analysis results" type="primary" >}}

```{r}
scenario <- read.csv(file.path("scenarios_resources", "r_scenario_results.csv"))
kable(scenario) |> scroll_box(height = "400px")
```

### Sensitivity analysis

{{< downloadthis scenarios_resources/r_sensitivity_results.csv dname="r_sensitivity_results" label="Download the sensitivity analysis results" type="primary" >}}

```{r}
sensitivity <- read.csv(
  file.path("scenarios_resources", "r_sensitivity_results.csv")
)
kable(sensitivity) |> scroll_box(height = "400px")
```

:::

## Example tables

This function finds the average results from each experiment.

:::: {.python-content}

It uses the `summary_stats()` function introduced on the [Replications](../output_analysis/replications.qmd) page.

::: {.callout-note title="View `summary_stats()`" collapse="true"}

```{python}
#| echo: false
#| file: ../output_analysis/replications_resources/summary_stats.py
```

:::

```{python}
def summarise_scenarios(results, groups, result_vars, path_prefix=None):
    """
    Find the average results from each scenario for multiple metrics.

    Parameters
    ----------
    results : pd.DataFrame
        Run-level results from every scenario.
    groups : list
        List of columns to group by (provided as strings).
    result_vars : list
        List of performance measures to get results on (provided as strings).
    path_prefix : str, optional
        Path prefix to save tables to. Each metric will be saved as
        {path_prefix}_{metric}.csv

    Returns
    -------
    summary_tables : dict
        Dictionary with metric names as keys and summary DataFrames as values.
    """
    summary_tables = {}

    for result_var in result_vars:
        summary_table = (
            results.groupby(groups)[result_var]
            .apply(summary_stats)
            .apply(pd.Series)
            .reset_index()
        )
        summary_table.columns = (
            list(summary_table.columns[:-4]) +
            ["mean", "std_dev", "ci_lower", "ci_upper"]
        )

        # Add column to identify which metric this is
        summary_table["metric"] = result_var

        summary_tables[result_var] = summary_table

        # Save if path provided
        if path_prefix:
            output_path = f"{path_prefix}_{result_var}.csv"
            summary_table.to_csv(output_path, index=False)

    return summary_tables
```

### Scenario analysis

```{python}
result_variables = [
    "mean_wait_time",
    "mean_utilisation_tw",
    "mean_queue_length",
    "mean_time_in_system",
    "mean_patients_in_system"
]
```

```{python}
scenario_tables = summarise_scenarios(
    results=scenario,
    groups=["scenario", "interarrival_time", "number_of_doctors"],
    result_vars=result_variables,
    path_prefix=os.path.join("tables_figures_resources", "python_scenario")
)
```

Click the name of each metric below to view the relevant dataframe:

::: {.panel-tabset}

## Mean wait time

```{python}
HTML(to_html_datatable(scenario_tables["mean_wait_time"]))
```

## Mean utilisation

```{python}
HTML(to_html_datatable(scenario_tables["mean_utilisation_tw"]))
```

## Mean queue length

```{python}
HTML(to_html_datatable(scenario_tables["mean_queue_length"]))
```

## Mean time in system

```{python}
HTML(to_html_datatable(scenario_tables["mean_time_in_system"]))
```

## Mean patients in system

```{python}
HTML(to_html_datatable(scenario_tables["mean_patients_in_system"]))
```

:::

### Sensitivity analysis

```{python}
sensitivity_tables = summarise_scenarios(
    results=sensitivity,
    groups=["scenario", "interarrival_time"],
    result_vars=result_variables,
    path_prefix=os.path.join("tables_figures_resources", "python_sensitivity")
)
```

Click the name of each metric below to view the relevant dataframe:

::: {.panel-tabset}

## Mean wait time

```{python}
HTML(to_html_datatable(sensitivity_tables["mean_wait_time"]))
```

## Mean utilisation

```{python}
HTML(to_html_datatable(sensitivity_tables["mean_utilisation_tw"]))
```

## Mean queue length

```{python}
HTML(to_html_datatable(sensitivity_tables["mean_queue_length"]))
```

## Mean time in system

```{python}
HTML(to_html_datatable(sensitivity_tables["mean_time_in_system"]))
```

## Mean patients in system

```{python}
HTML(to_html_datatable(sensitivity_tables["mean_patients_in_system"]))
```

:::

::::

:::: {.r-content}

```{r}
#' Summarise multiple performance measures at once
#'
#' @param results Dataframe with results from each replication of scenarios.
#' @param groups Vector of grouping columns (as strings).
#' @param result_vars Vector of performance measures to summarise (as strings).
#' @param path_prefix Optional path prefix to save each table to.
#'
#' @return A named list of summary data frames.

summarise_scenarios <- function(
  results, groups, result_vars, path_prefix = NULL
) {
  summary_tables <- list()

  for (result_var in result_vars) {
    summary_table <- results |>
      group_by(across(all_of(groups))) |>
      summarise(
        mean = mean(.data[[result_var]], na.rm = TRUE),
        std_dev = sd(.data[[result_var]], na.rm = TRUE),
        ci_lower = t.test(.data[[result_var]])$conf.int[1L],
        ci_upper = t.test(.data[[result_var]])$conf.int[2L],
        .groups = "drop"
      ) |>
      mutate(metric = result_var)

    # Save table if prefix specified
    if (!is.null(path_prefix)) {
      output_path <- paste0(path_prefix, "_", result_var, ".csv")
      write.csv(summary_table, output_path, row.names = FALSE)
    }

    # Store in list
    summary_tables[[result_var]] <- summary_table
  }

  return(summary_tables)
}
```

### Scenario analysis

```{r}
result_variables <- c(
  "mean_wait_time_doctor",
  "utilisation_doctor",
  "mean_queue_length_doctor",
  "mean_time_in_system",
  "mean_patients_in_system"
)
```

```{r}
scenario_tables <- summarise_scenarios(
  results = scenario,
  groups = c("scenario", "interarrival_time", "number_of_doctors"),
  result_vars = result_variables,
  path_prefix = file.path("tables_figures_resources", "r_scenario")
)
```

Click the name of each metric below to view the relevant table:

::: {.panel-tabset}

## Mean wait time

```{r}
kable(scenario_tables[["mean_wait_time_doctor"]])
```

## Mean utilisation

```{r}
kable(scenario_tables[["utilisation_doctor"]])
```

## Mean queue length

```{r}
kable(scenario_tables[["mean_queue_length_doctor"]])
```

## Mean time in system

```{r}
kable(scenario_tables[["mean_time_in_system"]])
```

## Mean patients in system

```{r}
kable(scenario_tables[["mean_patients_in_system"]])
```

:::

### Sensitivity analysis

```{r}
sensitivity_tables <- summarise_scenarios(
  results = sensitivity,
  groups = c("scenario", "interarrival_time"),
  result_vars = result_variables,
  path_prefix = file.path("tables_figures_resources", "r_sensitivity")
)
```

Click the name of each metric below to view the relevant table:

::: {.panel-tabset}

## Mean wait time

```{r}
kable(sensitivity_tables[["mean_wait_time_doctor"]])
```

## Mean utilisation

```{r}
kable(sensitivity_tables[["utilisation_doctor"]])
```

## Mean queue length

```{r}
kable(sensitivity_tables[["mean_queue_length_doctor"]])
```

## Mean time in system

```{r}
kable(sensitivity_tables[["mean_time_in_system"]])
```

## Mean patients in system

```{r}
kable(sensitivity_tables[["mean_patients_in_system"]])
```

:::

::::

## Example figures

:::: {.python-content}

These functions are used to plot the results from the scenarios and sensitivity analysis.

```{python}
def pale_colour(colour, alpha=0.2):
    """
    Make pale version of a colour.

    Parameters
    ----------
    colour : str or tuple
        The colour to pale, as a Plotly-compatible name ('blue', 'red'), hex
        string (e.g., "#1f77b4"), or RGB tuple (e.g., (0.1, 0.3, 0.8)).
        Accepts any format readable by matplotlib.colors.to_rgb.
    alpha : float, optional
        Opacity for the pale colour (default = 0.2). Should be between 0 and 1.

    Returns
    -------
    str
        Colour in 'rgba(r, g, b, a)' string format
        (e.g., "rgba(31,119,180,0.2)").

    Acknowledgements
    ----------------
    This function was generated by Perplexity.
    """
    # Convert any accepted colour format to an RGB tuple
    rgb = mcolors.to_rgb(colour)
    # Scale values to 0-255 for plotly compatability
    r, g, b = [int(x * 255) for x in rgb]
    # Return as RGB tuple with added alpha parameter (which makes it paler)
    return f"rgba({r},{g},{b},{alpha})"
```

```{python}
def plot_metrics(
    summary_tables, colour_var=None, name_mappings=None, path_prefix=None
):
    """
    Plot results from different model scenarios.

    Parameters
    ----------
    summary_tables : dict
        Dictionary with metric names as keys and summary DataFrames as values.
    colour_var : str, optional
        Name of variable to colour lines with.
    name_mappings : dict, optional
        Dictionary mapping column names to labels. If not provided,
        function will default to variable names.
    path : str
        Path to save figure to.

    Returns
    -------
    figs : dict
        Dictionary with metric names as keys and plotly figures as values.
    """
    figs = {}

    for metric_name, summary_table in summary_tables.items():

        # Initialise figure
        fig = go.Figure()

        # Handling color mappings (optional)
        if colour_var:
            unique_groups = summary_table[colour_var].unique()
            colours = px.colors.qualitative.Plotly
            colour_map = {
                group: colours[i % len(colours)]
                for i, group in enumerate(unique_groups)
            }
        else:
            colour_map = {None: "blue"}

        # Plot each scenario line and its shaded confidence interval
        if colour_var:
            groups = summary_table.groupby(colour_var)
        else:
            groups = [(None, summary_table)]
        for group_name, group_data in groups:
            x = group_data["interarrival_time"].values
            mean = group_data["mean"].values
            ci_upper = group_data["ci_upper"].values
            ci_lower = group_data["ci_lower"].values
            main_colour = colour_map[group_name]
            shade_colour = pale_colour(main_colour, alpha=0.2)

            # Filled confidence interval region
            fig.add_trace(go.Scatter(
                x=np.concatenate([x, x[::-1]]),
                y=np.concatenate([ci_upper, ci_lower[::-1]]),
                fill="toself",
                fillcolor=shade_colour,
                line={"color": "rgba(255,255,255,0)"},
                showlegend=False,
                name="95% CI",
                hoverinfo="name+y"
            ))

            # Mean line
            fig.add_trace(go.Scatter(
                x=x,
                y=mean,
                mode="lines",
                line={"color": main_colour, "width": 2},
                name=group_name,
                hoverinfo="y"
            ))

        # Set labels and layout
        fig.update_layout(
            xaxis_title=(
                name_mappings.get("interarrival_time", "interarrival_time")
                if name_mappings else "interarrival_time"
            ),
            yaxis_title=(
                name_mappings.get(metric_name, metric_name)
                if name_mappings else metric_name
            ),
            legend_title_text=(
                name_mappings.get(colour_var, colour_var)
                if colour_var and name_mappings
                else (colour_var if colour_var else None)
            ),
            template="plotly_white"
        )

        if not colour_var:
            fig.update_layout(showlegend=False)

        figs[metric_name] = fig

        # Save if path provided
        if path_prefix:
            output_path = f"{path_prefix}_{metric_name}.png"
            fig.write_image(output_path)

    return figs
```

### Scenario analysis

```{python}
name_mappings = {
    "interarrival_time": "Patient inter-arrival time",
    "number_of_doctors": "Number of doctors",
    "mean_wait_time": "Mean wait time for the doctor",
    "mean_utilisation_tw": "Mean utilisation",
    "mean_queue_length": "Mean queue length",
    "mean_time_in_system": "Mean patient time in the system",
    "mean_patients_in_system": "Mean number of patients in the system"
}
```

```{python}
scenario_plots = plot_metrics(
    summary_tables=scenario_tables,
    colour_var="number_of_doctors",
    name_mappings=name_mappings,
    path_prefix=os.path.join("tables_figures_resources", "python_scenario")
)
```

Click the name of each metric below to view the relevant plot:

::: {.panel-tabset}

## Mean wait time

```{python}
scenario_plots["mean_wait_time"]
```

## Mean utilisation

```{python}
scenario_plots["mean_utilisation_tw"]
```

## Mean queue length

```{python}
scenario_plots["mean_queue_length"]
```

## Mean time in system

```{python}
scenario_plots["mean_time_in_system"]
```

## Mean patients in system

```{python}
scenario_plots["mean_patients_in_system"]
```

:::

### Sensitivity analysis

```{python}
sensitivity_plots = plot_metrics(
    summary_tables=sensitivity_tables,
    name_mappings=name_mappings,
    path_prefix=os.path.join("tables_figures_resources", "python_sensitivity")
)
```

Click the name of each metric below to view the relevant plot:

::: {.panel-tabset}

## Mean wait time

```{python}
sensitivity_plots["mean_wait_time"]
```

## Mean utilisation

```{python}
sensitivity_plots["mean_utilisation_tw"]
```

## Mean queue length

```{python}
sensitivity_plots["mean_queue_length"]
```

## Mean time in system

```{python}
sensitivity_plots["mean_time_in_system"]
```

## Mean patients in system

```{python}
sensitivity_plots["mean_patients_in_system"]
```

:::

::::

::: {.r-content}

This function is used to plot the results from the scenarios and sensitivity analysis.

```{r}
#' Plot multiple performance measures at once
#'
#' @param summary_tables Named list of summary tables, one per metric
#' (like output from summarise_scenarios()).
#' @param x_var Name of variable to plot on x axis.
#' @param colour_var Name of variable to colour lines with (can be NULL).
#' @param name_mappings Optional named list for prettier axis/legend labels.
#' @param path_prefix Optional path prefix to save figures.
#' @return Named list of ggplot objects

plot_metrics <- function(
  summary_tables, x_var, colour_var = NULL, name_mappings = NULL,
  path_prefix = NULL
) {
  # List to store plots for each metric
  plot_list <- list()

  for (metric_name in names(summary_tables)) {
    # Extract relevant results table
    summary_table <- summary_tables[[metric_name]]

    # Helper to map a variable to display name if available in `name_mappings`.
    # Just uses variable name if no mapping is found.
    get_name <- function(var) {
      if (!is.null(var) && !is.null(name_mappings[[var]])) {
        name_mappings[[var]]
      } else {
        var
      }
    }
    xaxis_title <- get_name(x_var)
    yaxis_title <- get_name(metric_name)
    legend_title <- get_name(colour_var)

    # Create plot, with or without grouping colour variable
    if (!is.null(colour_var)) {
      summary_table[[colour_var]] <- as.factor(summary_table[[colour_var]])
      p <- ggplot(
        summary_table,
        aes_string(
          x = x_var,
          y = "mean",
          group = colour_var,
          color = colour_var,
          fill = colour_var
        )
      ) +
        geom_line() +
        geom_ribbon(
          aes_string(
            ymin = "ci_lower",
            ymax = "ci_upper"
          ),
          alpha = 0.1
        ) +
        labs(
          x = xaxis_title,
          y = yaxis_title,
          color = legend_title,
          fill = legend_title
        ) +
        theme_minimal()
    } else {
      p <- ggplot(
        summary_table,
        aes_string(
          x = x_var,
          y = "mean"
        )
      ) +
        geom_line() +
        geom_ribbon(
          aes_string(
            ymin = "ci_lower",
            ymax = "ci_upper"
          ),
          alpha = 0.1,
          show.legend = FALSE
        ) +
        labs(
          x = xaxis_title,
          y = yaxis_title
        ) +
        theme_minimal()
    }

    # Save plot if prefix supplied
    if (!is.null(path_prefix)) {
      output_path <- paste0(path_prefix, "_", metric_name, ".png")
      ggsave(filename = output_path, plot = p, width = 6.5, height = 4L,
             bg = "white")
    }

    plot_list[[metric_name]] <- p
  }

  return(plot_list)
}
```

### Scenario analysis

```{r}
name_mappings <- list(
  interarrival_time = "Patient inter-arrival time",
  number_of_doctors = "Number of doctors",
  mean_wait_time_doctor = "Mean wait time for the doctor",
  utilisation_doctor = "Mean utilisation",
  mean_queue_length_doctor = "Mean queue length",
  mean_time_in_system = "Mean patient time in the system",
  mean_patients_in_system = "Mean number of patients in the system"
)
```

```{r}
scenario_plots <- plot_metrics(
  summary_tables = scenario_tables,
  x_var = "interarrival_time",
  colour_var = "number_of_doctors",
  name_mappings = name_mappings,
  path_prefix = file.path("tables_figures_resources", "r_scenario")
)
```

Click the name of each metric below to view the relevant plot:

::: {.panel-tabset}

## Mean wait time

```{r}
scenario_plots[["mean_wait_time_doctor"]]
```

## Mean utilisation

```{r}
scenario_plots[["utilisation_doctor"]]
```

## Mean queue length

```{r}
scenario_plots[["mean_queue_length_doctor"]]
```

## Mean time in system

```{r}
scenario_plots[["mean_time_in_system"]]
```

## Mean patients in system

```{r}
scenario_plots[["mean_patients_in_system"]]
```

:::

These plots look a little funny simple due to the wide confidence intervals - with very short run times and few replications in our illustrative example, the confidence intervals are very wide!

### Sensitivity analysis

```{r}
sensitivity_plots <- plot_metrics(
  summary_tables = sensitivity_tables,
  x_var = "interarrival_time",
  name_mappings = name_mappings,
  path_prefix = file.path("tables_figures_resources", "r_sensitivity")
)
```

Click the name of each metric below to view the relevant plot:

::: {.panel-tabset}

## Mean wait time

```{r}
sensitivity_plots[["mean_wait_time_doctor"]]
```

## Mean utilisation

```{r}
sensitivity_plots[["utilisation_doctor"]]
```

## Mean queue length

```{r}
sensitivity_plots[["mean_queue_length_doctor"]]
```

## Mean time in system

```{r}
sensitivity_plots[["mean_time_in_system"]]
```

## Mean patients in system

```{r}
sensitivity_plots[["mean_patients_in_system"]]
```

:::

::::

## Explore the example models

<div class="h3-tight"></div>

### Nurse visit simulation

:::: {.python-content}

{{< include /html/pydesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `notebooks/analysis.ipynb` |
: {tbl-colwidths="[20,80]"}
:::

::::

:::: {.r-content}

{{< include /html/rdesrapmms.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `rmarkdown/analysis.md` |
: {tbl-colwidths="[20,80]"}
:::

::::

### Stroke pathway simulation

:::: {.python-content}

{{< include /html/pydesrapstroke.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `notebooks/analysis.ipynb` |
: {tbl-colwidths="[20,80]"}
:::

::::

:::: {.r-content}

{{< include /html/rdesrapstroke.html >}}

::: {.blue-table}
| | |
| - | - |
| **Key files** | `rmarkdown/analysis.md` |
: {tbl-colwidths="[20,80]"}
:::

::::

## Test yourself

Create at least one table and figure from your simulation results.

If you don't have any to hand, feel free to download ours:

::: {.python-content}

{{< downloadthis scenarios_resources/python_scenario_results.csv dname="python_scenario_results" label="Download the scenario analysis results" type="primary" >}}

{{< downloadthis scenarios_resources/python_sensitivity_results.csv dname="python_sensitivity_results" label="Download the sensitivity analysis results" type="primary" >}}

:::

::: {.r-content}

{{< downloadthis scenarios_resources/r_scenario_results.csv dname="r_scenario_results" label="Download the scenario analysis results" type="primary" >}}

{{< downloadthis scenarios_resources/r_sensitivity_results.csv dname="r_sensitivity_results" label="Download the sensitivity analysis results" type="primary" >}}

:::

Make sure to write and save code that generates your outputs, and that you save the outputs to files (e.g., `csv` for tables, `png` for figures).

::: {.python-content}

For example, create a Jupyter notebook for the analysis, and then push that to your GitHub repository.

:::

::: {.r-content}

For example, create an Rmarkdown file for the analysis, and then push that to your GitHub repository.

:::

## References

::: {#refs}
:::

<br><br>
